### 2024-04-03 paper/paper.tex
first update to paper text

%  LaTeX support: latex@mdpi.com 
%  For support, please attach all files needed for compiling as well as the log file, and specify your operating system, LaTeX version, and LaTeX editor.

%=================================================================
\documentclass[remotesensing,article,submit,pdftex,moreauthors]{Definitions/mdpi}

%=================================================================
% MDPI internal commands - do not modify
\firstpage{1} 
\makeatletter 
\setcounter{page}{\@firstpage} 
\makeatother
\pubvolume{1}
\issuenum{1}
\articlenumber{0}
\pubyear{2024}
\copyrightyear{2024}
%\externaleditor{Academic Editor: Firstname Lastname}
\datereceived{ } 
\daterevised{ } % Comment out if no revised date
\dateaccepted{ } 
\datepublished{ } 
%\datecorrected{} % For corrected papers: "Corrected: XXX" date in the original paper.
%\dateretracted{} % For corrected papers: "Retracted: XXX" date in the original paper.
\hreflink{https://doi.org/} % If needed use \linebreak
%\doinum{}
%\pdfoutput=1 % Uncommented for upload to arXiv.org
%\CorrStatement{yes}  % For updates

%=================================================================
% Add packages and commands here. The following packages are loaded in our class file: fontenc, inputenc, calc, indentfirst, fancyhdr, graphicx, epstopdf, lastpage, ifthen, float, amsmath, amssymb, lineno, setspace, enumitem, mathpazo, booktabs, titlesec, etoolbox, tabto, xcolor, colortbl, soul, multirow, microtype, tikz, totcount, changepage, attrib, upgreek, array, tabularx, pbox, ragged2e, tocloft, marginnote, marginfix, enotez, amsthm, natbib, hyperref, cleveref, scrextend, url, geometry, newfloat, caption, draftwatermark, seqsplit
% cleveref: load \crefname definitions after \begin{document}

%=================================================================
% Please use the following mathematics environments: Theorem, Lemma, Corollary, Proposition, Characterization, Property, Problem, Example, ExamplesandDefinitions, Hypothesis, Remark, Definition, Notation, Assumption
%% For proofs, please use the proof environment (the amsthm package is loaded by the MDPI class).

%=================================================================
% Full title of the paper (Capitalized)
\Title{Robot Team III: Scotty Strikes Back}

% MDPI internal command: Title for citation in the left column
\TitleCitation{Title}

% Author Orchid ID: enter ID or remove command
\newcommand{\orcidauthorA}{0000-0002-5910-0183} % John
\newcommand{\orcidauthorE}{0000-0003-4265-9543} % David

% Authors, for the paper (add full first names)
\Author{John Waczak$^{1,\dagger,\ddagger}$\orcidA{}, David J. Lary $^{2,\ddagger}$\orcidE{}*}

%\longauthorlist{yes}

% MDPI internal command: Authors, for metadata in PDF
\AuthorNames{John Waczak, and David Lary}

% MDPI internal command: Authors, for citation in the left column
\AuthorCitation{Waczak, J.; Lary, D.J.;}
% If this is a Chicago style journal: Lastname, Firstname, Firstname Lastname, and Firstname Lastname.

% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
\address{%
$^{1}$ \quad Affiliation 1; e-mail@e-mail.com\\
$^{2}$ \quad Affiliation 2; e-mail@e-mail.com}

% Contact information of the corresponding author
\corres{Correspondence: e-mail@e-mail.com; Tel.: (optional; include country code; if there are multiple corresponding authors, add author initials) +xx-xxxx-xxx-xxxx (F.L.)}

% Current address and/or shared authorship
\firstnote{Current address: Affiliation 3.} 
\secondnote{These authors contributed equally to this work.}
% The commands \thirdnote{} till \eighthnote{} are available for further notes

%\simplesumm{} % Simple summary

%\conference{} % An extended version of a conference paper

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{FIXME}

% Keywords
\keyword{FIXME}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}

1. Classification is a common remote sensing and hyperspectral imaging problem
    - Endmember Identification and Spectral Unmixing is sub-problem specific to hyperspectral data

2. HSI Size allows their incorporation onto drones

3. Provided with in situ measurements, HSI can be used to predict a variety of relevant water quality and water composition parameters. Here we can describe our previous work

4. However, we would like to be able to identify the presence of spectral signatures for which we do not already have an in situ sensor.

5. Unsupervised classification techniques are nice however standard algorithms like k-means clustering require prior knowledge of the number of constituent endmembers and do not, in general, provide any interpretable relationship between classes e.g. how is class 1 more similar to class 2 than to class 10.

6. "In this paper, we extend our previous work into the unsupervised classification domain..." Here we demonstrate the application of Generative Topographic mapping which addresses these issues by providing a probability distribution of resulting classes.

7. We apply the GTM to datasets collected by our Robot Team specifically showcasing the identification of spectral signatures for Rhodamine Dye, Algae, etc...

8. With the signatures identified, class abundance, and therefore area estimation, can be performed rapidly, in real time, by utilizing the learned signatures with the nromalized spectral similarity score (or other similar metrics).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Materials and Methods}

\subsection{Autonomous Robot Team}
\subsection{Hyperspectral Data Cubes}
\subsection{Generative Topographic Mapping}
\subsection{Image Segmentation with the Normalized Spectral Similarity Score}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}

\subsection{GTM Fit}
\subsection{Hyperparamter Optimization with BIC}
\subsection{Spectral Signatures}
\subsection{Water Class Map}
\subsection{Algae Identification}
\subsection{Rhodamine Dye Plume Identification}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}

\subsection{Summary of GTM applications in our paper}
\subsection{Comparison to other existing approaches in literature}
- Is there much done for classification in real time? 
\subsection{Other applications of GTM}
- Discuss use of GTM means/modes/responsibilities feature transformer for supervised models (e.g. better than PCA)

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\vspace{6pt}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% optional
%\supplementary{The following supporting information can be downloaded at:  \linksupplementary{s1}, Figure S1: title; Table S1: title; Video S1: title.}

% Only for journal Methods and Protocols:
% If you wish to submit a video article, please do so with any other supplementary material.
% \supplementary{The following supporting information can be downloaded at: \linksupplementary{s1}, Figure S1: title; Table S1: title; Video S1: title. A supporting video article is available at doi: link.}

% Only for journal Hardware:
% If you wish to submit a video article, please do so with any other supplementary material.
% \supplementary{The following supporting information can be downloaded at: \linksupplementary{s1}, Figure S1: title; Table S1: title; Video S1: title.\vspace{6pt}\\
%\begin{tabularx}{\textwidth}{lll}
%\toprule
%\textbf{Name} & \textbf{Type} & \textbf{Description} \\
%\midrule
%S1 & Python script (.py) & Script of python source code used in XX \\
%S2 & Text (.txt) & Script of modelling code used to make Figure X \\
%S3 & Text (.txt) & Raw data from experiment X \\
%S4 & Video (.mp4) & Video demonstrating the hardware in use \\
%... & ... & ... \\
%\bottomrule
%\end{tabularx}
%}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\authorcontributions{FIXME}

\funding{FIXME}

\institutionalreview{Not applicable}

\informedconsent{Not applicable}

\dataavailability{We encourage all authors of articles published in MDPI journals to share their research data. In this section, please provide details regarding where data supporting reported results can be found, including links to publicly archived datasets analyzed or generated during the study. Where no new data were created, or where data is unavailable due to privacy or ethical restrictions, a statement is still required. Suggested Data Availability Statements are available in section ``MDPI Research Data Policies'' at \url{https://www.mdpi.com/ethics}.}

\acknowledgments{In this section you can acknowledge any support given which is not covered by the author contribution or funding sections. This may include administrative and technical support, or donations in kind (e.g., materials used for experiments).}

\conflictsofinterest{Declare conflicts of interest or state ``The authors declare no conflicts of interest.'' Authors must identify and declare any personal circumstances or interest that may be perceived as inappropriately influencing the representation or interpretation of reported research results. Any role of the funders in the design of the study; in the collection, analyses or interpretation of data; in the writing of the manuscript; or in the decision to publish the results must be declared in this section. If there is no role, please state ``The funders had no role in the design of the study; in the collection, analyses, or interpretation of data; in the writing of the manuscript; or in the decision to publish the results''.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Optional

%% Only for journal Encyclopedia
%\entrylink{The Link to this entry published on the encyclopedia platform.}

\abbreviations{Abbreviations}{
The following abbreviations are used in this manuscript:\\

\noindent 
\begin{tabular}{@{}ll}
MDPI & Multidisciplinary Digital Publishing Institute\\
DOAJ & Directory of open access journals\\
TLA & Three letter acronym\\
LD & Linear dichroism
\end{tabular}
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Optional
\appendixtitles{no} % Leave argument "no" if all appendix headings stay EMPTY (then no dot is printed after "Appendix A"). If the appendix sections contain a heading then change the argument to "yes".
\appendixstart
\appendix
\section[\appendixname~\thesection]{}
\subsection[\appendixname~\thesubsection]{}
The appendix is an optional section that can contain details and data supplemental to the main text---for example, explanations of experimental details that would disrupt the flow of the main text but nonetheless remain crucial to understanding and reproducing the research shown; figures of replicates for experiments of which representative data are shown in the main text can be added here if brief, or as Supplementary Data. Mathematical proofs of results not central to the paper can be added as an appendix.

\begin{table}[H] 
\caption{This is a table caption.\label{tab5}}
\newcolumntype{C}{>{\centering\arraybackslash}X}
\begin{tabularx}{\textwidth}{CCC}
\toprule
\textbf{Title 1}	& \textbf{Title 2}	& \textbf{Title 3}\\
\midrule
Entry 1		& Data			& Data\\
Entry 2		& Data			& Data\\
\bottomrule
\end{tabularx}
\end{table}

\section[\appendixname~\thesection]{}
All appendix sections must be cited in the main text. In the appendices, Figures, Tables, etc. should be labeled, starting with ``A''---e.g., Figure A1, Figure A2, etc.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{adjustwidth}{-\extralength}{0cm}
%\printendnotes[custom] % Un-comment to print a list of endnotes

\reftitle{References}

% Please provide either the correct journal abbreviation (e.g. according to the “List of Title Word Abbreviations” http://www.issn.org/services/online-services/access-to-the-ltwa/) or the full name of the journal.
% Citations and References in Supplementary files are permitted provided that they also appear in the reference list here.

%=====================================
% References, variant A: external bibliography
%=====================================
%\bibliography{your_external_BibTeX_file}

%=====================================
% References, variant B: internal bibliography
%=====================================
\begin{thebibliography}{999}
% Reference 1
\bibitem[Author1(year)]{ref-journal}
Author~1, T. The title of the cited article. {\em Journal Abbreviation} {\bf 2008}, {\em 10}, 142--149.
% Reference 2
\bibitem[Author2(year)]{ref-book1}
Author~2, L. The title of the cited contribution. In {\em The Book Title}; Editor 1, F., Editor 2, A., Eds.; Publishing House: City, Country, 2007; pp. 32--58.
% Reference 3
\bibitem[Author3(year)]{ref-book2}
Author 1, A.; Author 2, B. \textit{Book Title}, 3rd ed.; Publisher: Publisher Location, Country, 2008; pp. 154--196.
% Reference 4
\bibitem[Author4(year)]{ref-unpublish}
Author 1, A.B.; Author 2, C. Title of Unpublished Work. \textit{Abbreviated Journal Name} year, \textit{phrase indicating stage of publication (submitted; accepted; in press)}.
% Reference 5
\bibitem[Author5(year)]{ref-communication}
Author 1, A.B. (University, City, State, Country); Author 2, C. (Institute, City, State, Country). Personal communication, 2012.
% Reference 6
\bibitem[Author6(year)]{ref-proceeding}
Author 1, A.B.; Author 2, C.D.; Author 3, E.F. Title of presentation. In Proceedings of the Name of the Conference, Location of Conference, Country, Date of Conference (Day Month Year); Abstract Number (optional), Pagination (optional).
% Reference 7
\bibitem[Author7(year)]{ref-thesis}
Author 1, A.B. Title of Thesis. Level of Thesis, Degree-Granting University, Location of University, Date of Completion.
% Reference 8
\bibitem[Author8(year)]{ref-url}
Title of Site. Available online: URL (accessed on Day Month Year).
\end{thebibliography}

% If authors have biography, please use the format below
%\section*{Short Biography of Authors}
%\bio
%{\raisebox{-0.35cm}{\includegraphics[width=3.5cm,height=5.3cm,clip,keepaspectratio]{Definitions/author1.pdf}}}
%{\textbf{Firstname Lastname} Biography of first author}
%
%\bio
%{\raisebox{-0.35cm}{\includegraphics[width=3.5cm,height=5.3cm,clip,keepaspectratio]{Definitions/author2.jpg}}}
%{\textbf{Firstname Lastname} Biography of second author}

% For the MDPI journals use author-date citation, please follow the formatting guidelines on http://www.mdpi.com/authors/references
% To cite two works by the same author: \citeauthor{ref-journal-1a} (\citeyear{ref-journal-1a}, \citeyear{ref-journal-1b}). This produces: Whittaker (1967, 1975)
% To cite two works by the same author with specific pages: \citeauthor{ref-journal-3a} (\citeyear{ref-journal-3a}, p. 328; \citeyear{ref-journal-3b}, p.475). This produces: Wong (1999, p. 328; 2000, p. 475)

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% for journal Sci
%\reviewreports{\\
%Reviewer 1 comments and authors’ response\\
%Reviewer 2 comments and authors’ response\\
%Reviewer 3 comments and authors’ response
%}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\PublishersNote{}
\end{adjustwidth}
\end{document}

### 2024-04-04 1-datacube-exemplar-viz.jl
updated figures

@@ -210,12 +210,12 @@ important_coords = Dict(
         "R" => Data[1:462, 400, 80] ./ maximum(Data[1:idx_900, 400, 80]),
         "λs" => λs,
     ),
-    "road" => Dict(
-        "x" => x_road,
-        "y" => y_road,
-        "R" => R_road,
-        "λs" => λs
-    )
+    # "road" => Dict(
+    #     "x" => x_road,
+    #     "y" => y_road,
+    #     "R" => R_road,
+    #     "λs" => λs
+    # )
 )
 
 
@@ -235,9 +235,10 @@ darktan = colorant"#ba8a50"
 
 
 fig = Figure();
-ax = CairoMakie.Axis(fig[2,1]);
+ax = CairoMakie.Axis(fig[2,1],
+                     );
 
-# hidedecorations!(ax)
+hidedecorations!(ax)
 # hidespines!(ax)
 
 img = heatmap!(ax, rgb_image);
@@ -265,8 +266,9 @@ la = lines!(ax, important_coords["algae"]["λs"][1:idx_900], important_coords["a
 lp = lines!(ax, important_coords["plume"]["λs"][1:idx_900], important_coords["plume"]["R"][1:idx_900], linewidth=2, label="Rhodamine", color=red)
 lw = lines!(ax, important_coords["water"]["λs"][1:idx_900], important_coords["water"]["R"][1:idx_900], linewidth=2, label="Water", color=blue)
 lg = lines!(ax, important_coords["grass"]["λs"][1:idx_900], important_coords["grass"]["R"][1:idx_900], linewidth=2, label="Grass", color=tan)
-lr = lines!(ax, important_coords["road"]["λs"][1:idx_900], important_coords["road"]["R"][1:idx_900], linewidth=2, label="Road", color=brown)
-fig[1,1] = Legend(fig, [la, lp, lw, lg, lr], ["Algae", "Rhodamine", "Water", "Grass", "Road"], framevisible=false, orientation=:horizontal, padding=(0,0,0,0), labelsize=13, height=-5)
+# lr = lines!(ax, important_coords["road"]["λs"][1:idx_900], important_coords["road"]["R"][1:idx_900], linewidth=2, label="Road", color=brown)
[…]

### 2024-04-04 3-gtm-robotteam.jl

@@ -266,6 +266,7 @@ end
 fig[1,1] = Legend(fig, ls, ["Water Spectrum", "GTM Node Signatures"], framevisible=false, orientation=:horizontal, padding=(0,0,0,0), labelsize=13, height=-5)
 xlims!(ax, wavelengths[1], 900)
 ylims!(ax, 0, 1)
+
 heatmap!(ax_inset, rgb_image)
 scatter!(ax_inset, [sample_coords["water"]["img_idxs"][1]], [sample_coords["water"]["img_idxs"][2]], markersize=7, color=:white, marker = :xcross)
 
@@ -378,6 +379,17 @@ save(joinpath(figures_path, "square-ndvi.pdf"), fig)
 
 
 # colored class map with identified pixels
+function map_color(ξ₁, ξ₂)
+    red = (ξ₁ + 1.0) / 2.0
+    blue = (ξ₂ + 1.0) / 2.0
+
+    red = (red < 0.0) ? 0.0 : red
+    blue = (blue < 0.0) ? 0.0 : blue
+
+    RGBA(red, 0.0, blue, 1.0)
+end
+
+
 fig = Figure();
 ax = CairoMakie.Axis(fig[2,1], xlabel="ξ₁", ylabel="ξ₂");
 mean_proj = DataFrame(MLJ.transform(mach, X))
@@ -473,19 +485,6 @@ f_list_2 = joinpath.(hsipath, "11-23", "Scotty_2", f_list_2)
 @assert all(ispath.(f_list_1))
 @assert all(ispath.(f_list_2))
 
-
-
-
-
-function map_color(ξ₁, ξ₂)
-    red = (ξ₁ + 1.0) / 2.0
-    blue = (ξ₂ + 1.0) / 2.0
-
-    red = (red < 0.0) ? 0.0 : red
-    blue = (blue < 0.0) ? 0.0 : blue
-
-    RGBA(red, 0.0, blue, 1.0)
-end
 
 
 

### 2024-04-04 3b-gtm-supervised.jl

@@ -1,0 +1,407 @@
+using CSV, DataFrames, DelimitedFiles, Tables
+using MLJ
+using GenerativeTopographicMapping
+using CairoMakie
+using Random
+using HDF5
+
+Random.seed!(42)
+
+using JSON
+using ArgParse
+using Random
+using ProgressMeter
+using LaTeXStrings
+using MultivariateStats
+using Distances
+using LinearAlgebra
+
+include("utils/makie-defaults.jl")
+include("utils/config.jl")
+include("utils/viz.jl")
+
+set_theme!(mints_theme)
+update_theme!(
+    figure_padding=30,
+    Axis=(
+        xticklabelsize=20,
+        yticklabelsize=20,
+        xlabelsize=22,
+        ylabelsize=22,
+        titlesize=25,
+    ),
+    Colorbar=(
+        ticklabelsize=20,
+        labelsize=22
+    )
+)
+
+
+
+# get map of exemplar points
+h5path= "/Users/johnwaczak/data/robot-team/processed/hsi/12-09/Dye_1/Dye_1-6.h5"
+rgb_image = get_h5_rgb(h5path)
+
+
+blue = colorant"#0366fc"
+darkblue = colorant"#013582"
+
+red = colorant"#d10f0f"
+darkred = colorant"#a10d0d"
+
+green = colorant"#28a10d"
+darkgreen = colorant"#157000"
+
+brown = colorant"#703e00"
+darkbrown = colorant"#543105"
+
+tan = colorant"#f0ba78"
+darktan = colorant"#ba8a50"
+
+
+
+datapath="./data/robot-team"
+@assert ispath(datapath)
+
+figures_path = joinpath("./figures", "robot-team-sup")
+if !ispath(figures_path)
+    mkpath(figures_path)
+end
+
+df_X = CSV.read(joinpath(datapath, "df_features_sup.csv"), DataFrame);
+df_Y = CSV.read(joinpath(datapath, "df_targets_sup.csv"), DataFrame);
+df_lat_lon = CSV.read(joinpath(datapath, "df_lat_lon_sup.csv"), DataFrame);
+
+
+idx_900 = findfirst(wavelengths .≥ 900)
+X = df_X[:, 1:idx_900]
+
[…]

### 2024-04-04 1-datacube-exemplar-viz.jl
added datacube figures with colorbar

@@ -111,9 +111,12 @@ R_road = Data[1:idx_900, 20, 760] ./ maximum(Data[1:idx_900, 20, 760])
 
 
 fig1, fig2 = vis_rectified_cube(h5path; azimuth=-π/3, elevation=5π/16)
+fig3, _ = vis_rectified_cube(h5path; azimuth=-π/3, elevation=5π/16, colorbar=true)
 
 GLMakie.activate!()
+colgap!(fig3.layout, 1, -200)
 save(joinpath(figs_path, "11-23_Scotty_1-2.png"), fig1)
+save(joinpath(figs_path, "11-23_Scotty_1-2_cb.png"), fig3)
 
 CairoMakie.activate!()
 save(joinpath(figs_path, "11-23_Scotty_1-2__colorbar.svg"), fig2)
@@ -122,9 +125,14 @@ save(joinpath(figs_path, "11-23_Scotty_1-2__colorbar.svg"), fig2)
 
 h5path = files_dict["11-23"]["no-dye"][2]
 fig1, fig2 = vis_rectified_cube(h5path; azimuth=-π/3, elevation=π/4)
+fig3, _ = vis_rectified_cube(h5path; azimuth=-π/3, elevation=5π/16, colorbar=true)
 
 GLMakie.activate!()
+colgap!(fig3.layout, 1, -200)
+fig3
+
 save(joinpath(figs_path, "11-23_Scotty_2-1.png"), fig1)
+save(joinpath(figs_path, "11-23_Scotty_2-1_cb.png"), fig3)
 
 CairoMakie.activate!()
 save(joinpath(figs_path, "11-23_Scotty_2-1__colorbar.svg"), fig2)
@@ -133,9 +141,14 @@ save(joinpath(figs_path, "11-23_Scotty_2-1__colorbar.svg"), fig2)
 
 h5path = files_dict["11-23"]["dye"][1]
 fig1, fig2 = vis_rectified_cube(h5path; azimuth=0)
+fig3, _ = vis_rectified_cube(h5path; azimuth=-π/3, elevation=5π/16, colorbar=true)
 
 GLMakie.activate!()
+colgap!(fig3.layout, 1, -200)
+fig3
+
 save(joinpath(figs_path, "11-23_Scotty_4-1.png"), fig1)
+save(joinpath(figs_path, "11-23_Scotty_4-1_cb.png"), fig3)
 
 CairoMakie.activate!()
[…]

### 2024-04-04 1-datacube-exemplar-viz.jl
updated with new figures

@@ -125,7 +125,7 @@ save(joinpath(figs_path, "11-23_Scotty_1-2__colorbar.svg"), fig2)
 
 h5path = files_dict["11-23"]["no-dye"][2]
 fig1, fig2 = vis_rectified_cube(h5path; azimuth=-π/3, elevation=π/4)
-fig3, _ = vis_rectified_cube(h5path; azimuth=-π/3, elevation=5π/16, colorbar=true)
+fig3, _ = vis_rectified_cube(h5path; azimuth=-π/3, elevation=π/4, colorbar=true)
 
 GLMakie.activate!()
 colgap!(fig3.layout, 1, -200)
@@ -141,7 +141,7 @@ save(joinpath(figs_path, "11-23_Scotty_2-1__colorbar.svg"), fig2)
 
 h5path = files_dict["11-23"]["dye"][1]
 fig1, fig2 = vis_rectified_cube(h5path; azimuth=0)
-fig3, _ = vis_rectified_cube(h5path; azimuth=-π/3, elevation=5π/16, colorbar=true)
+fig3, _ = vis_rectified_cube(h5path; azimuth=0, colorbar=true)
 
 GLMakie.activate!()
 colgap!(fig3.layout, 1, -200)
@@ -159,7 +159,7 @@ h5path = files_dict["11-23"]["dye"][2]
 
 GLMakie.activate!()
 fig1, fig2 = vis_rectified_cube(h5path; azimuth=-π/3, elevation=π/4)
-fig3, _ = vis_rectified_cube(h5path; azimuth=-π/3, elevation=5π/16, colorbar=true)
+fig3, _ = vis_rectified_cube(h5path; azimuth=-π/3, elevation=π/4, colorbar=true)
 colgap!(fig3.layout, 1, -200)
 
 save(joinpath(figs_path, "11-23_Scotty_5-1.png"), fig1)
@@ -174,7 +174,7 @@ h5path = files_dict["12-09"]["dye"][1]
 
 GLMakie.activate!()
 fig1, fig2 = vis_rectified_cube(h5path; azimuth=3π/4, elevation=3π/16, ibounds=(100, length(xs)))
-fig3, _ = vis_rectified_cube(h5path; azimuth=-π/3, elevation=5π/16, colorbar=true)
[…]

### 2024-04-04 paper/paper.tex
Updates from Overleaf

@@
 \newcommand{\orcidauthorA}{0000-0002-5910-0183} % John
+\newcommand{\orcidauthorB}{0000-0003-2688-648X} % Lakitha
+\newcommand{\orcidauthorC}{0000-0002-9841-6703} % Shawhin
+\newcommand{\orcidauthorD}{0000-0003-2657-3416} % Prabuddha
 \newcommand{\orcidauthorE}{0000-0003-4265-9543} % David
+\newcommand{\orcidauthorF}{0000-0003-0667-2345} % Ash
+\newcommand{\orcidauthorG}{0000-0002-0126-218X} % Adam
 
 % Authors, for the paper (add full first names)
-\Author{John Waczak$^{1,\dagger,\ddagger}$\orcidA{}, David J. Lary $^{2,\ddagger}$\orcidE{}*}
+% \Author{John Waczak \orcidA{}, Adam Aker \orcidG{}, Lakitha O.
+H.
+Wijeratne \orcidB{}, Shawhin Talebi \orcidC{}, Ashen Fernando \orcidF{}, Prabuddha M.
+H.
+Dewage \orcidD{}, Mazhar Iqbal, Matthew Lary, David Schaefer and David J.
+Lary *\orcidE{}}
+\Author{John Waczak \orcidA{} and David J. Lary *\orcidE{}}
 
@@
 % MDPI internal command: Authors, for metadata in PDF
-\AuthorNames{John Waczak, and David Lary}
+% \AuthorNames{John Waczak, Adam Aker, Lakitha O. H. Wijeratne, Shawhin Talebi, Bharana Fernando, Prabuddha Hathurusinghe, Mazhar Iqbal, Matthew Lary, David Schaefer and David J. Lary}
+\AuthorNames{John Waczak and David J. Lary}
 
 % MDPI internal command: Authors, for citation in the left column
-\AuthorCitation{Waczak, J.; Lary, D.J.;}
+% \AuthorCitation{Waczak, J.; Aker, A.; Wijeratne, L.O.H.; Talebi, S.; Fernando, B.; Hathurusinghe, P.; Iqbal, M.; Lary, M.; Schaefer, D.; Lary, D.J.}
+\AuthorCitation{Waczak, J.; Lary, D.J.}
 % If this is a Chicago style journal: Lastname, Firstname, Firstname Lastname, and Firstname Lastname.
@@
 % Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
-\address{%
-$^{1}$ \quad Affiliation 1; e-mail@e-mail.com\\
-$^{2}$ \quad Affiliation 2; e-mail@e-mail.com}
+% \address[1]{%
+% Hanson Center for Space Sciences, University of Texas at Dallas, Richardson, TX 75080, USA;  john.waczak@utdallas.edu (J.W.); adam.aker@utdallas.edu (A.A.); lhw150030@utdallas.edu (L.O.H.W.); shawhintalebi@gmail.com (S.T.); ashen.fernando@utdallas.edu  (B.F.); pxh180012@utdallas.edu (P.M.H.D.); mazhar.iqbal@utdallas.edu (M.I.); MDL210001@utdallas.edu (M.L.); captdaveschaefer@gmail.com (D.S.)}
+
+\address[1]{%
+Hanson Center for Space Sciences, University of Texas at Dallas, Richardson, TX 75080, USA;  john.waczak@utdallas.edu (J.W.);}
+\corres{\hangafter=1 \hangindent=1.05em \hspace{-0.82em}Correspondence: {David.Lary@utdallas.edu}}
 
 % Contact information of the corresponding author
-\corres{Correspondence: e-mail@e-mail.com; Tel.: (optional; include country code; if there are multiple corresponding authors, add author initials) +xx-xxxx-xxx-xxxx (F.L.)}
-
-% Current address and/or shared authorship
-\firstnote{Current address: Affiliation 3.} 
-\secondnote{These authors contributed equally to this work.}
-% The commands \thirdnote{} till \eighthnote{} are available for further notes
-
-%\simplesumm{} % Simple summary
-
-%\conference{} % An extended version of a conference paper
+\corres{Correspondence: david.lary@utdallas.edu} %; Tel.: (optional; include country code; if there are multiple corresponding authors, add author initials) +xx-xxxx-xxx-xxxx (F.L.)}
 
@@
 
+\begin{figure}[t]
+\begin{adjustwidth}{-\extralength}{0cm}
+\centering
+\includegraphics[width=15.5cm]{Definitions/logo-mdpi}
+\end{adjustwidth}
+\caption{This is a wide figure.\label{fig2}}
+\end{figure}
+
 \subsection{Autonomous Robot Team}
-\subsection{Hyperspectral Data Cubes}
+\subsection{Data Collection and Pre-Processing}
 \subsection{Generative Topographic Mapping}
@@
 - Discuss use of GTM means/modes/responsibilities feature transformer for supervised models (e.g. better than PCA) 
+- Use of GTM classes to aid autonomous data collection by identification of "interesting" regons
 
@@
 \subsection[\appendixname~\thesubsection]{}
-The appendix is an optional section that can contain details and data supplemental to the main text---for example, explanations of experimental details that would disrupt the flow of the main text but nonetheless remain crucial to understanding and reproducing the research shown; figures of replicates for experiments of which representative data are shown in the main text can be added here if brief, or as Supplementary Data.
-Mathematical proofs of results not central to the paper can be added as an appendix.
-
-\begin{table}[H] 
-\caption{This is a table caption.\label{tab5}}
-\newcolumntype{C}{>{\centering\arraybackslash}X}
-\begin{tabularx}{\textwidth}{CCC}
-\toprule
-\textbf{Title 1}	& \textbf{Title 2}	& \textbf{Title 3}\\
-\midrule
-Entry 1		& Data			& Data\\
-Entry 2		& Data			& Data\\
-\bottomrule
-\end{tabularx}
-\end{table}
-
-\section[\appendixname~\thesection]{}
-All appendix sections must be cited in the main text. In the appendices, Figures, Tables, etc. should be labeled, starting with ``A''---e.g., Figure A1, Figure A2, etc.
 
@@
 
-% Please provide either the correct journal abbreviation (e.g.
-according to the “List of Title Word Abbreviations” http://www.issn.org/services/online-services/access-to-the-ltwa/) or the full name of the journal.
-% Citations and References in Supplementary files are permitted provided that they also appear in the reference list here.
-
 %=====================================
@@
 %=====================================
-%\bibliography{your_external_BibTeX_file}
+\bibliography{paper/references.bib}
 
-%=====================================
-% References, variant B: internal bibliography
-%=====================================
-\begin{thebibliography}{999}
-% Reference 1
-\bibitem[Author1(year)]{ref-journal}
-Author~1, T. The title of the cited article. {\em Journal Abbreviation} {\bf 2008}, {\em 10}, 142--149.
-% Reference 2
-\bibitem[Author2(year)]{ref-book1}
-Author~2, L. The title of the cited contribution. In {\em The Book Title}; Editor 1, F., Editor 2, A., Eds.; Publishing House: City, Country, 2007; pp. 32--58.
-% Reference 3
-\bibitem[Author3(year)]{ref-book2}
-Author 1, A.; Author 2, B. \textit{Book Title}, 3rd ed.; Publisher: Publisher Location, Country, 2008; pp. 154--196.
-% Reference 4
-\bibitem[Author4(year)]{ref-unpublish}
-Author 1, A.B.; Author 2, C. Title of Unpublished Work. \textit{Abbreviated Journal Name} year, \textit{phrase indicating stage of publication (submitted; accepted; in press)}.
-% Reference 5
-\bibitem[Author5(year)]{ref-communication}
-Author 1, A.B. (University, City, State, Country); Author 2, C. (Institute, City, State, Country). Personal communication, 2012.
-% Reference 6
-\bibitem[Author6(year)]{ref-proceeding}
-Author 1, A.B.; Author 2, C.D.; Author 3, E.F.
-Title of presentation.
-In Proceedings of the Name of the Conference, Location of Conference, Country, Date of Conference (Day Month Year); Abstract Number (optional), Pagination (optional).
-% Reference 7
-\bibitem[Author7(year)]{ref-thesis}
-Author 1, A.B. Title of Thesis. Level of Thesis, Degree-Granting University, Location of University, Date of Completion.
-% Reference 8
-\bibitem[Author8(year)]{ref-url}
-Title of Site. Available online: URL (accessed on Day Month Year).
-\end{thebibliography}
-
-% If authors have biography, please use the format below
-%\section*{Short Biography of Authors}
-%\bio
-%{\raisebox{-0.35cm}{\includegraphics[width=3.5cm,height=5.3cm,clip,keepaspectratio]{Definitions/author1.pdf}}}
-%{\textbf{Firstname Lastname} Biography of first author}
-%
-%\bio
-%{\raisebox{-0.35cm}{\includegraphics[width=3.5cm,height=5.3cm,clip,keepaspectratio]{Definitions/author2.jpg}}}
-%{\textbf{Firstname Lastname} Biography of second author}
-
-% For the MDPI journals use author-date citation, please follow the formatting guidelines on http://www.mdpi.com/authors/references
-% To cite two works by the same author: \citeauthor{ref-journal-1a} (\citeyear{ref-journal-1a}, \citeyear{ref-journal-1b}). This produces: Whittaker (1967, 1975)
-% To cite two works by the same author with specific pages: \citeauthor{ref-journal-3a} (\citeyear{ref-journal-3a}, p.
-328; \citeyear{ref-journal-3b}, p.475).
-This produces: Wong (1999, p.
-328; 2000, p.
-475)
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% for journal Sci
-%\reviewreports{\\
-%Reviewer 1 comments and authors’ response\\
-%Reviewer 2 comments and authors’ response\\
-%Reviewer 3 comments and authors’ response
-%}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2024-04-04 5-gtm-comparison.jl
updated table for hyperparam search

@@ -232,7 +232,7 @@ function generate_tex_table(df)
     out = out * "  \\newcolumntype{C}{>{\\centering\\arraybackslash}X}\n"
     out = out * "  \\begin{tabularx}{\\fulllength}{CCCCCC}\n"
     out = out * "    \\toprule\n"
-    out = out * "    \\textbf{k} & \\textbf{\$\alpha\$} & \\textbf{s} & \\textbf{k} & \\textbf{BIC} & \\textbf{AIC} \\\\\n"
+    out = out * "    \\textbf{m} & \\textbf{\$\alpha\$} & \\textbf{s} & \\textbf{k} & \\textbf{BIC} & \\textbf{AIC} \\\\\n"
     out = out * "    \\midrule\n"
 
     for i ∈ 1:25
@@ -258,7 +258,9 @@ end
 
 tex_table = generate_tex_table(df)
 
-open("./models/hp-search-table.tex", "w") do f
+println(tex_table)
+
+open("./paper/hp-search-table.tex", "w") do f
     println(f, tex_table)
 end
 

### 2024-04-04 paper/paper.tex
Updates from Overleaf

@@
 \centering
-\includegraphics[width=15.5cm]{Definitions/logo-mdpi}
+\includegraphics[width=15.5cm]{paper/figures/methods/sample-spectra.pdf}
 \end{adjustwidth}
-\caption{This is a wide figure.\label{fig2}}
+\caption{(a) Sample hyperspectral data cube.
+Spectra are plotted using their geographic position with the log10-reflectance colored along the z axis and a pseudo-color image on top.
+(b) Points taken from a sample hyperspectral data cube corresponding to algae, rhodamine dye, water, and dry grass.
+(c) The associated spectra for the exemplar points scaled so that the peak value of each spectrum is 1.
+(d) The location of the pond in Montague, Texas where data were collected for this study.\label{fig:sample-spectra}}
 \end{figure}

### 2024-04-04 paper/paper.tex
Updates from Overleaf

@@
 \begin{figure}[t]
-\begin{adjustwidth}{-\extralength}{0cm}
+%\begin{adjustwidth}{-\extralength}{0cm}
 \centering
-\includegraphics[width=15.5cm]{paper/figures/methods/sample-spectra.pdf}
-\end{adjustwidth}
+%\includegraphics[width=15.5cm]{paper/figures/methods/sample-spectra.pdf}
+\includegraphics[width=\columnwidth]{paper/figures/methods/sample-spectra.pdf}
+%\end{adjustwidth}
 \caption{(a) Sample hyperspectral data cube.
@@
 \subsection{GTM Fit}
+
+\begin{figure}[t]
+\centering
+\includegraphics[width=0.8\columnwidth]{paper/figures/results/square-ndvi.pdf}
+\caption{GTM Map: We plot the mean position of each data point in the latent space with points colored by the their associated NDVI values.
+The locations of the exemplar points for algae, rhodamine dye, water, and dry grass are shown in the latent space illustrating that the GTM has clearly distinguished land, near-land, and water pixels.\label{fig:}}
+\end{figure}
+
 \subsection{Hyperparamter Optimization with BIC}
+
+\begin{figure}[t]
+%\begin{adjustwidth}{-\extralength}{0cm}
+\centering
+\includegraphics[width=\textwidth]{paper/figures/results/bic.pdf}
+%\end{adjustwidth}
+\caption{Results of the hyperparameter search.
+Left: Variation of BIC with $m$ and $s$ for fixed $\alpha=0.1$.
+Right: Variation of the BIC with with $m$ and $\alpha$ for fixed $s=1.0$.
+The white star in each plot indicates the parameters with the lowest BIC.\label{fig:hp-results}}
+\end{figure}
+
 \subsection{Spectral Signatures}
+
+\begin{figure}[t]
+\begin{adjustwidth}{-\extralength}{0cm}
+\centering
+\includegraphics[width=15.5cm]{paper/figures/results/spectral-signatures.png}
+\end{adjustwidth}
+\caption{GTM Node Signatures corresponding to the exemplar spectra for the rhodamine plume (top left), algae (top right), dry grass (bottom left), and open water (bottom right).
+\label{fig:spectral-signautes}}
+\end{figure}
+
 \subsection{Water Class Map}
+
+\begin{figure}[t]
+\begin{adjustwidth}{-\extralength}{0cm}
+\centering
+\includegraphics[width=18.0cm]{paper/figures/results/gtm-water.png}
+\end{adjustwidth}
+\caption{Classification map for GTM trained solely on water pixels (no land and no rhodamine plume).
+\textbf{(a)} GTM applied to all water pixels colored by their associated position in the latent space.
+\textbf{(b)} Representation of the poitns from the training set in the latent space.
+The points are colored with $\xi_1$ corresponding to red and $\xi_2$ corresponding to blue.
+\textbf{(c)} The associated GTM spectral signatures, $\Psi_i$, corresponding to the four corners and center of the GTM latent space.
+\label{fig:gtm-water}}
+\end{figure}
+
 \subsection{Algae Identification}
+
+\begin{figure}[t]
+\begin{adjustwidth}{-\extralength}{0cm}
+\centering
+\includegraphics[width=15.5cm]{paper/figures/results/algae.png}
+\end{adjustwidth}
+\caption{\label{fig:algae-map}}
+\end{figure}
+
 \subsection{Rhodamine Dye Plume Identification}
+
+\begin{figure}[t]
+\begin{adjustwidth}{-\extralength}{0cm}
+\centering
+\includegraphics[width=15.5cm]{paper/figures/results/rhodamine.png}
+\end{adjustwidth}
+\caption{\label{fig:rhodamine-map}}
+\end{figure}
 
@@
 %% Optional
-\appendixtitles{no} % Leave argument "no" if all appendix headings stay EMPTY (then no dot is printed after "Appendix A"). If the appendix sections contain a heading then change the argument to "yes".
+\appendixtitles{yes} % Leave argument "no" if all appendix headings stay EMPTY (then no dot is printed after "Appendix A").
+If the appendix sections contain a heading then change the argument to "yes".
 \appendixstart
 \appendix
-\section[\appendixname~\thesection]{}
-\subsection[\appendixname~\thesubsection]{}
+\section[\appendixname~\thesection]{Hyperparameter Search Results}
+
+\begin{table}[H]
+  \caption{The top 25 models from the hyperparameter search. A variety of GTM were trained to explore the the impact of varying m, $\alpha$, and s.\label{tab:hp-search}}
+  \begin{adjustwidth}{-\extralength}{0cm}
+  \newcolumntype{C}{>{\centering\arraybackslash}X}
+  \begin{tabularx}{\fulllength}{CCCCCC}
+    \toprule
+    \textbf{m} & \textbf{$\alpha$} & \textbf{s} & \textbf{k} & \textbf{BIC} & \textbf{AIC} \\
+    \midrule
+    14 & 0.1 & 1.0 & 32 & -1.918e8 & -1.926e8\\
+    13 & 0.01 & 1.0 & 32 & -1.917e8 & -1.923e8\\
+    16 & 0.01 & 1.5 & 32 & -1.917e8 & -1.926e8\\
+    14 & 10.0 & 1.0 & 32 & -1.917e8 & -1.924e8\\
+    16 & 0.001 & 1.5 & 32 & -1.917e8 & -1.926e8\\
+    13 & 1.0 & 1.0 & 32 & -1.917e8 & -1.923e8\\
+    13 & 10.0 & 1.0 & 32 & -1.917e8 & -1.923e8\\
+    14 & 0.001 & 1.5 & 32 & -1.916e8 & -1.924e8\\
+    13 & 0.1 & 1.0 & 32 & -1.916e8 & -1.923e8\\
+    14 & 0.01 & 1.0 & 32 & -1.916e8 & -1.924e8\\
+    15 & 0.01 & 1.5 & 32 & -1.916e8 & -1.925e8\\
+    14 & 0.01 & 1.5 & 32 & -1.916e8 & -1.923e8\\
+    15 & 1.0 & 1.0 & 32 & -1.916e8 & -1.924e8\\
+    18 & 0.01 & 1.5 & 32 & -1.916e8 & -1.928e8\\
+    12 & 0.01 & 1.0 & 32 & -1.916e8 & -1.921e8\\
+    15 & 0.01 & 0.5 & 32 & -1.915e8 & -1.924e8\\
+    17 & 1.0 & 1.0 & 32 & -1.915e8 & -1.926e8\\
+    16 & 0.1 & 1.0 & 32 & -1.915e8 & -1.925e8\\
+    18 & 0.001 & 1.5 & 32 & -1.915e8 & -1.928e8\\
+    13 & 0.001 & 1.0 & 32 & -1.915e8 & -1.922e8\\
+    12 & 1.0 & 1.0 & 32 & -1.915e8 & -1.921e8\\
+    17 & 0.001 & 1.5 & 32 & -1.915e8 & -1.926e8\\
+    15 & 0.001 & 1.5 & 32 & -1.915e8 & -1.923e8\\
+    15 & 10.0 & 1.0 & 32 & -1.915e8 & -1.923e8\\
+    12 & 0.1 & 1.5 & 32 & -1.915e8 & -1.92e8\\
+    \bottomrule
+  \end{tabularx}
+  \end{adjustwidth}
+\end{table}
 

### 2024-04-10 paper/paper.tex
more annotation to bibliography

@@
 \newcommand{\orcidauthorG}{0000-0002-0126-218X} % Adam
+% \newcommand{\orcidauthorH}{} % Gokul
 
 % Authors, for the paper (add full first names)
-% \Author{John Waczak \orcidA{}, Adam Aker \orcidG{}, Lakitha O.
+\Author{John Waczak \orcidA{}, Adam Aker \orcidG{}, Lakitha O.
 H.
@@
 H.
-Dewage \orcidD{}, Mazhar Iqbal, Matthew Lary, David Schaefer and David J.
+Dewage \orcidD{}, Mazhar Iqbal, Matthew Lary, David Schaefer, Gokul Balagopal, and David J.
 Lary *\orcidE{}}
-\Author{John Waczak \orcidA{} and David J. Lary *\orcidE{}}
 
@@
 % MDPI internal command: Authors, for metadata in PDF
-% \AuthorNames{John Waczak, Adam Aker, Lakitha O. H. Wijeratne, Shawhin Talebi, Bharana Fernando, Prabuddha Hathurusinghe, Mazhar Iqbal, Matthew Lary, David Schaefer and David J. Lary}
-\AuthorNames{John Waczak and David J. Lary}
+\AuthorNames{John Waczak, Adam Aker, Lakitha O. H. Wijeratne, Shawhin Talebi, Bharana Fernando, Prabuddha Hathurusinghe, Mazhar Iqbal, Matthew Lary, David Schaefer, Gokul Balagopal, and David J. Lary}
 
 % MDPI internal command: Authors, for citation in the left column
-% \AuthorCitation{Waczak, J.; Aker, A.; Wijeratne, L.O.H.; Talebi, S.; Fernando, B.; Hathurusinghe, P.; Iqbal, M.; Lary, M.; Schaefer, D.; Lary, D.J.}
-\AuthorCitation{Waczak, J.; Lary, D.J.}
+ \AuthorCitation{Waczak, J.; Aker, A.; Wijeratne, L.O.H.; Talebi, S.; Fernando, B.; Hathurusinghe, P.; Iqbal, M.; Lary, M.; Schaefer, Balagopal, G.; D.; Lary, D.J.}
 % If this is a Chicago style journal: Lastname, Firstname, Firstname Lastname, and Firstname Lastname.
@@
 % Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
-% \address[1]{%
-% Hanson Center for Space Sciences, University of Texas at Dallas, Richardson, TX 75080, USA;  john.waczak@utdallas.edu (J.W.); adam.aker@utdallas.edu (A.A.); lhw150030@utdallas.edu (L.O.H.W.); shawhintalebi@gmail.com (S.T.); ashen.fernando@utdallas.edu  (B.F.); pxh180012@utdallas.edu (P.M.H.D.); mazhar.iqbal@utdallas.edu (M.I.); MDL210001@utdallas.edu (M.L.); captdaveschaefer@gmail.com (D.S.)}
-
 \address[1]{%
-Hanson Center for Space Sciences, University of Texas at Dallas, Richardson, TX 75080, USA;  john.waczak@utdallas.edu (J.W.);}
-\corres{\hangafter=1 \hangindent=1.05em \hspace{-0.82em}Correspondence: {David.Lary@utdallas.edu}}
+Hanson Center for Space Sciences, University of Texas at Dallas, Richardson, TX 75080, USA;  john.waczak@utdallas.edu (J.W.); adam.aker@utdallas.edu (A.A.); lhw150030@utdallas.edu (L.O.H.W.); shawhintalebi@gmail.com (S.T.); ashen.fernando@utdallas.edu  (B.F.); pxh180012@utdallas.edu (P.M.H.D.); mazhar.iqbal@utdallas.edu (M.I.); MDL210001@utdallas.edu (M.L.); captdaveschaefer@gmail.com (D.S.); gokul.balagopal@utdallas.edu (G.B.);}
 
@@
 2. HSI Size allows their incorporation onto drones
+    - \cite{robot-team-1}
+    - \cite{robot-team-2}
 
@@
 how is class 1 more similar to class 2 than to class 10.
+The Self Organizing map introduced by Kohonen is a popular technique which enables an unsupervised classification of data where classes
+
+- We should make a note about Autoencoders and VAEs here.
+While they work well in many contexts, they do not provide a relationship between classes and there is no obvious way to choose the size of the encoding
 
@@
 With the signatures identified, class abundance, and therefore area estimation, can be performed rapidly, in real time, by utilizing the learned signatures with the nromalized spectral similarity score (or other similar metrics).
+
+IDEA: spectral unmixing of water is challenging due to nonlinear effects of scattering by solids and absorption by dissolved chemicals (dyes etc...) this motivates a nonlinear approach to endmember identification, etc...
 
@@
 \subsection{Generative Topographic Mapping}
+
+- GTM paper oricinal \cite{gtm-bishop-1}
+- GTM developments paper \cite{gtm-biship-2}
+
 \subsection{Image Segmentation with the Normalized Spectral Similarity Score}
@@
     \midrule
-    14 & 0.1 & 1.0 & 32 & -1.918e8 & -1.926e8\\
+    14 & 0.1 & 1.0 & 32 & \textbf{-1.918e8} & -1.926e8\\
     13 & 0.01 & 1.0 & 32 & -1.917e8 & -1.923e8\\

### 2024-04-13 paper/paper.tex
updated outline in introduction

@@
 % Keywords
-\keyword{FIXME}
+\keyword{Hyperspectral Imaging; Remote Sensing; Unsupervised Classification; Spectral Unmixing}
 
@@
 
+0. Specific applications of remote sensing and HSI include harmful algal bloom (HAB) identification and oil spill mapping. 
+    - Macine learning methods have been used for species identification of algal blooms from remote sensing imagery \cite{ghatkar2019classification}
+    - Another paper on HAB identification. Machine learning is identified as a useful tehcnique \cite{thenkabail2018hyperspectral}
+    - Remote sensing has been used for analysis of oil spills such as the 2010 Deepwater Horizon spill \cite{kokaly2013spectroscopic}
+    - Neural networks and other ML techniques used in conjunction with multispectral data from AVIRIS and MODIS for oil slick thickness and emulsion ratio identification for Deepwater Horizon spill \cite{leifer2012state}
+    - Sun glitter remains a challenge.
+Multi-spectral and hyperspectral imaging have been utilized for oil spill identification and to identify their impacts via vegetation stress and mortality \cite{fingas2014review, khan2018modern}
+    - Hyperspectral data can help distinguish oils by type, e.g. crude, diesel, gasoline, and palm oil \cite{yang2020characterization}
+
 1. Classification is a common remote sensing and hyperspectral imaging problem
-    - Endmember Identification and Spectral Unmixing is sub-problem specific to hyperspectral data
+    - Remote sensing applications for water quality include monitoring optically-active parameters like turbidity, temerpature, chlorophyll-a, blue-green algae, etc. \cite{ritchie2003remote}.
+    - Remote sensing useful for mapping natural hazzards such as earthquakes, volcanic activity, flooding, and wildfires \cite{joyce2009review}.
+    - Other applications of hyperspectral imaging include food quality, food safety, medical diagnosis, precision agriculture, and forensic document examination \cite{khan2018modern}
+    - Remote sensing imagery from satellite sources such as Landsat have been used for inversion of water quality parameters such as Chlorophyll-a and Secchi Disk Transparency \cite{bonansea2015using}.
+    - Analysis of inland water ways remains a challenge due to small-scale spatial variability and complicated spectral signatures. 
+    - Spectral indices together with machine learning and simulation remain key approaches \cite{sagan2020monitoring}.
+    - When high resolution ground truth data are not available, one approach is to classify water quality into a set of categories ranging from bad to good \cite{koponen2002lake}.
+    - When ground truth data are not available, we may still want to partition imagery into groups or clusters. 
+    - Many data-driven, unsupervised classification techniques have been employed for this task for remotely sensed multi-spectral and hyperspectral imagery including various matrix factorization methods, k-nearest neighbors, fuzzy c-means, density estimation, etc...
+\cite{zhang2019hyperspectral}
+    - Additionally, ICA, LDA, and similar tecniques can be used to perform nonlinear dimensionality reduction (as opposed to PCA) for visualization or as a preprocessing step for supervised regression and classification tasks \cite{liang2010hyperspectral, zhangSVM2014}
+
+1b.
+Endmember extraction and spectral unmixing are a related problem in whic the goal is to identify the spectral signatures of relevant sources within a scene.
+Once signatures have been identified, we then desire to model their interaction, typically through linear mixings, in order to determine the relative abundance of each source for each pixel of an HSI \cite{spectral-unmixing-orig, heylen2014review}.
+    - statistical approaches for finding endmembers using least squares \cite{berman2004ice}
+    - Popular approach is to treat pure endmembers as vertices of a simplex \cite{plaza2012endmember, nascimento2005vertex}
+    - Sparce PCA used for mineral endmember extraction and abundance mapping \cite{yousefi2016mineral}
+    - Self organizing map used for endmember extraction and neural network used for abundance mapping \cite{cantero2004analysis}
+    - spectral angle mapper is popular similarity function used in endmember extraction. The SAM is nice because it is more sensitive to shape than scale \cite{jiang2020classification}
+    - Convolutional neural networks and auto-encoders have also been used for endmember extraction and abundance mapping \cite{palsson2020convolutional}
+    - Non-negative autoencoders with a structure and loss function that enforces the sum-to-one and non-negativity constraints \cite{non-negative-autoencoders}
+    - Variational Autoencoders build on this by trying to learn the distribution of latent vectors \cite{su2019daen}
+    - Another autoencoder paper \cite{borsoi2019deep}
 
 2. HSI Size allows their incorporation onto drones
+    -  Nearth-earth HSI addresses spatial, spectral, temporal limitations of satellite and airborne platforms. UAV-based HSI enable fine-scale mapping applications \cite{banerjee2020uav}
+    - Drones can readily be equipped with HSI sensors and on-board compute enables rapid generation of spectral indices like the NDVI for applications to precision agriculture \cite{horstrand2019uav}
+    - UAV-based HSI can be geocrorrected accurately to the centimeter scale w/o ground control points \cite{arroyo2019implementation}
     - \cite{robot-team-1}
-    - \cite{robot-team-2}
+    - UAVs provide significant advantages over satellite approaches \cite{adao2017hyperspectral}
 
 3. Provided with in situ measurements, HSI can be used to predict a variety of relevant water quality and water composition parameters. Here we can describe our previous work
+    - Simple polynomial regression models can be used with HSI to estimate levels of relevant water quality parameters like chlorophyll-a, turbidity, etc... \cite{zhang2022selection}
+    - UAVs have also been used to rapidly estimate turbidity \cite{vogt2016near}
+    - ML models can be used to extract relevant optically active water quality parameters from HSI \cite{lu2021retrieval}
+    - Key challenge in supervised regression of wate rquality parameters is the collection of necessary in-situ data.
+    - Ross et al.
+created a comprehensive dataset with over 600,000 water quality records matching optically active water quality parameters with associated satellite imagery from Landsat 5,7,8.
+To achieve this quantity of data, they needed records spanning 1984 to 2019 \cite{ross2019aquasat}.
+    - Aurin et al. took a similar route combining 30 years of remote sensing imagery with in-situ data from over 500 field compaigns for CDOM, organic carbon, etc... \cite{aurin2018remote}.
+    - This can be extended with an autonomous robotic team. \cite{robot-team-2}
 
@@
 The Self Organizing map introduced by Kohonen is a popular technique which enables an unsupervised classification of data where classes
-
-- We should make a note about Autoencoders and VAEs here.
-While they work well in many contexts, they do not provide a relationship between classes and there is no obvious way to choose the size of the encoding
+    - Original SOM paper \cite{kohonen-som-1}
+    - SOM has been a popular technique for remote sensing as well. 
+    - SOM used for identifying synoptic-scale patterns in wind and sea surface temperature \cite{som-satellite}
+    - There are drawbacks to the SOM.
+However Heskes reinterpreted the algorithm so that the training procedure can bee seen as stochastic gradient descent on an energy surface, i.e.
+with a cost function \cite{som-cost-function}
+    - SOM have also been used with HSI for feature extraction \cite{som-hsi}
+    - SOM used for classification of remote sensing imagery \cite{msom-remote-sensing}
+    - SOM used for clustering and data compression of HSI data on-board a cube sat \cite{danielsen2021self}
+    - SOM used for analysis of land-use and landcover change \cite{penfound2021analysis}
 
@@
 "In this paper, we extend our previous work into the unsupervised classification domain..." Here we demonstrate the application of Generative Topographic mapping which addresses these issues by providing a probability distribution of resulting classes.
+    - Original GTM paper \cite{gtm-bishop-1}
+    - Extensions of the GTM for incremental learning and other approaches \cite{gtm-bishop-2}
 
@@
 With the signatures identified, class abundance, and therefore area estimation, can be performed rapidly, in real time, by utilizing the learned signatures with the nromalized spectral similarity score (or other similar metrics).
+    - Many different spectral distance functions exist for comparing spectra and abundance mapping such as spectral anagle mapper, euclidean distance, spectral correlation mapper, etc...
+These have trade-offs between the ability to discriminate changes in intensity and changes in hue (i.e.
+location of peaks) \cite{deborah2015comprehensive}
+    - The spectral correlation mapper was introduced as a statistically motivated alternative to the spectral angle mapper \cite{de2000spectral}
+    - We adopt the normalized spectral similarity score developed by nNidamanuri et al.
+which combines the mean-squared-error euclidean type distance with the spectral angle \cite{nidamanuri2010normalized}
 
@@
 \subsection{Data Collection and Pre-Processing}
+
+- georetification \cite{muller2002program, baumker2001new, mostafa2000multi}
+
+- Julia programming language used \cite{bezanson2012julia}
+
 \subsection{Generative Topographic Mapping}
@@
 - GTM developments paper \cite{gtm-biship-2}
+
+- implementation designed to work in MLJ ecosystem \cite{blaom2020mlj}
 
@@
 - Is there much done for classification in real time? 
+\subsection{Comparison to Spectral Unmixing}
+- Here we can re-interpret the GTM model as performing a type of Bayesian spectral unmixing and suggest how the method could be further extended to work in this domain.
+Critically, other traditional methods work by observing that pure endmembers form the vertices of a simplex.
+
 \subsection{Other applications of GTM}

### 2024-04-15 paper/paper.tex
updated text for intro

--- 2024-04-10 paper/paper.tex
@@
 
-1. Classification is a common remote sensing and hyperspectral imaging problem
-    - Endmember Identification and Spectral Unmixing is sub-problem specific to hyperspectral data
+1. Water quality classification
 
-2. HSI Size allows their incorporation onto drones
-    - \cite{robot-team-1}
-    - \cite{robot-team-2}
+Inland water bodies pose a unique challenge for classification by remote sensing imagery due to their complex spectral characteristics, which stem from a combination of chemical, biological, and environmental factors.
 
-3. Provided with in situ measurements, HSI can be used to predict a variety of relevant water quality and water composition parameters. Here we can describe our previous work
+The complex spectral characteristics of inland water bodies pose a unique challenge to their classification by remote sensing methods due to the combination of chemical, biological, and environmental sources.
+Remote sensing applications for water quality include monitoring optically-active parameters like turbidity, temperature, chlorophyll-a, blue-green algae, etc.
+\cite{ritchie2003remote}.
+Spectral indices together with machine learning and simulation remain key approaches \cite{sagan2020monitoring}.
+Remote sensing has been used for analysis of oil spills such as the 2010 Deepwater Horizon spill \cite{leifer2012state,kokaly2013spectroscopic}.
+Another application is for the identification and classification of harmful algal blooms (HABs) \cite{thenkabail2018hyperspectral,ghatkar2019classification}.
+The application of remote sensing to water quality monitoring is largely focused on the estimation of optically-active constituents such as chlorophyll-a, blue-green algae, temperature, and turbidity \cite{ritchie2003remote}.
 
-4. However, we would like to be able to identify the presence of spectral signatures for which we do not already have an in situ sensor.
+2. Limitations of traditional remote sensing: Limited spectral resolution and need for expensive in-situ reference data
+
+A key limitation is the need for sufficient quantities of in-situ reference data.
+Aurin et al.
+took a similar route combining 30 years of remote sensing imagery with in-situ data from over 500 field compaigns for CDOM, organic carbon, etc...
+\cite{aurin2018remote}.
+Ross et al.
+created a comprehensive dataset with over 600,000 water quality records matching optically active water quality parameters with associated satellite imagery from Landsat 5,7,8.
+To achieve this quantity of data, they needed records spanning 1984 to 2019 \cite{ross2019aquasat}.
+
+3. Data Driven methods for classification
+
+When high resolution ground truth data are not available, one approach is to classify water quality into a set of categories ranging from bad to good \cite{koponen2002lake}.
+Many data-driven, unsupervised classification techniques have been employed for this task for remotely sensed multi-spectral and hyperspectral imagery including various matrix factorization methods, k-nearest neighbors, fuzzy c-means, density estimation, etc...
+\cite{zhang2019hyperspectral}.
+Additionally, ICA, LDA, and similar techniques can be used to perform nonlinear dimensionality reduction (as opposed to PCA) for visualization or as a preprocessing step for supervised regression and classification tasks \cite{liang2010hyperspectral, zhangSVM2014}.
+
+4. Hyperspectral imaging addresses the first limitation by providing improved spectral resolution
+
+Hyperspectral data can help distinguish oils by type, e.g.
+crude, diesel, gasoline, and palm oil \cite{yang2020characterization}.
+Other applications of hyperspectral imaging include food quality, food safety, medical diagnosis, precision agriculture, and forensic document examination \cite{khan2018modern}
 
 5.
-Unsupervised classification techniques are nice however standard algorithms like k-means clustering require prior knowledge of the number of constituent endmembers and do not, in general, provide any interpretable relationship between classes e.g.
-how is class 1 more similar to class 2 than to class 10.
-The Self Organizing map introduced by Kohonen is a popular technique which enables an unsupervised classification of data where classes
 
-- We should make a note about Autoencoders and VAEs here.
-While they work well in many contexts, they do not provide a relationship between classes and there is no obvious way to choose the size of the encoding
+1.
+Hyperspectral imaging is a powerful tool.
+Specifically for water quality, the increased spectral resolution enables the identification and discernment of a variety of contaminants from harmful algal blooms (HABs) to oil spills which pose a significant challenge to traditional multi-spectral remote sensing systems...
+
+2.
+Traditionally, the applications of such imagers to water quality monitoring relies on ground-truth observations using reference instruments.
+For example...
+chorophyll-a, blue-green algae, crude oil, etc.
+However reference instruments are expensive to deploy and challenging to maintain often requiring complicated calibration
+
+3. Unsupervised classification of hyperspectral imagery. Discuss ML approaches including Self Organizing Map (SOM).
+
+4.
+A related involves the data-driven identification of unique spectral constituents within a scene known as endmember extraction.
+A variety of statistical and machine learning based approaches have been developed to extract unique spectral signatures corresponding to different sources.
+Once these libraries of spectral signatures are obtained, they can be combined to model the spectrum of each pixel in a HSI.
+This approach is known as spectral unmixing and has been used to great success for a variety of applications including
+
+4.
+While HSI offer superior spectral resolution, the limited spatial and temporal resolution provided by satellite-based platforms remains a challenge for the analysis of inland waters with small-scale spatial variability.
+Additionally, the infrequency of satellite overpasses makes it challenging to identify sources of contamination or spills in real time.
+With the continual development of hyperspectral imaging technology, the incorporation of hyperspectral imagers as the payload of autonomous aerial vehicles (UAVs) such as quadcopters, hexacopters, and other drones offers unparalleled capabilities with nearly centimeter accuracy.
+
+5. Robot team
 
 6.
-"In this paper, we extend our previous work into the unsupervised classification domain..." Here we demonstrate the application of Generative Topographic mapping which addresses these issues by providing a probability distribution of resulting classes.
-
-7. We apply the GTM to datasets collected by our Robot Team specifically showcasing the identification of spectral signatures for Rhodamine Dye, Algae, etc...
-
-8.
-With the signatures identified, class abundance, and therefore area estimation, can be performed rapidly, in real time, by utilizing the learned signatures with the nromalized spectral similarity score (or other similar metrics).
-
-IDEA: spectral unmixing of water is challenging due to nonlinear effects of scattering by solids and absorption by dissolved chemicals (dyes etc...) this motivates a nonlinear approach to endmember identification, etc...
+In this paper, we demonstrate how Generative Topographic Mapping (GTM), a Bayesian alternative to the Self Organizing Map, can be utilized to rapidly identify relevant spectral signatures within a scene.
+Using data collected at site in Montague, North Texas, we demonstrate the ability of this paradigm with two case studies.
+First, we demonstrate how the GTM can be used to provide a high-resolution unsupervised classification of a water body enabling the rapid identification of relevant regions which can then be used to direct further investigation by reference-sensor equipped boats.
+Second, we demonstrate the ability to perform endmember extraction and abundance mapping using imagery from a test-release of rhodamine dye.
+Together with the normalized spectral similarity score, these spectral signatures are identified and their associated area in the scene is determined.
 
@@
 \begin{tabular}{@{}ll}
-MDPI & Multidisciplinary Digital Publishing Institute\\
-DOAJ & Directory of open access journals\\
-TLA & Three letter acronym\\
-LD & Linear dichroism
+HAB & Harmful Algal Bloom \\ 
+HSI & Hyperspectral Image \\ 
+SOM & Self Organizing Map \\ 
+GTM & Generative Topographic Mapping 
 \end{tabular}

### 2024-04-16 paper/paper.tex
Updated intro text

--- 2024-04-10 paper/paper.tex
@@
 
-1. Classification is a common remote sensing and hyperspectral imaging problem
-    - Endmember Identification and Spectral Unmixing is sub-problem specific to hyperspectral data
+Inland water bodies pose a unique challenge for analysis by remote sensing imagery due to their complex spectral characteristics and small-scale spatial variability; nevertheless, significant effort has gone into the development of techniques for estimating optically-active water quality parameters such as chlorophyll-a, blue-green algae, turbidity, and temperature (citation).
+Machine learning techniques have proven to be a powerful tool in this domain, facilitating the development of data-driven models which map observed reflectance to parameters of interest.
+However, these methods rely on the availability of significant volumes of in-situ reference data needed to calibrate models and asses their uncertainty.
+For example Aurin et al...
+Similarly, Ross et al developed...
+The challenge posed by in-situ data collection together with long satellite revisit times limits the utility of remote sensing imagery for real-time applications.
 
-2. HSI Size allows their incorporation onto drones
-    - \cite{robot-team-1}
-    - \cite{robot-team-2}
+- supervised classification of water quality with XGBoost \cite{ghatkar2019classification}
 
-3. Provided with in situ measurements, HSI can be used to predict a variety of relevant water quality and water composition parameters. Here we can describe our previous work
+Unsupervised classification
 
-4. However, we would like to be able to identify the presence of spectral signatures for which we do not already have an in situ sensor.
+endmember extraction, spectral unmixing
 
-5.
-Unsupervised classification techniques are nice however standard algorithms like k-means clustering require prior knowledge of the number of constituent endmembers and do not, in general, provide any interpretable relationship between classes e.g.
-how is class 1 more similar to class 2 than to class 10.
-The Self Organizing map introduced by Kohonen is a popular technique which enables an unsupervised classification of data where classes
+Hyperspectral imaging has emerged as a pivotal technology in this space.
+With their increased spectral resolution, hyperspectral imagers (HSIs) capture the complex chemical, biological, environmental characteristics of reflectance spectra.
+(add references for specific applications of HSIs).
+Continued development of HSI technology has led to a dramatic reduction in their size enabling their utilize as the payload of small autonomous aerial vehicles (UAVs) (citation).
+By flying close to the ground UAV-born hyperspectral imagers can produce centimeter-scale imagery while simulatenously limiting the need for complicated atmospheric corrections required for satellite platforms (citation).
 
-- We should make a note about Autoencoders and VAEs here.
-While they work well in many contexts, they do not provide a relationship between classes and there is no obvious way to choose the size of the encoding
+In our previous work, we introduced a robot team designed to rapidly characterize water composition in novel environments \cite{robot-team-1}. 
+Joining UAV-borne HSI together with collection of in-situ reference data by an autonomous boat dramatically reduces the time required to generate .
 
-6.
-"In this paper, we extend our previous work into the unsupervised classification domain..." Here we demonstrate the application of Generative Topographic mapping which addresses these issues by providing a probability distribution of resulting classes.
+By carefully coordinating the simultanous collection of ground-truth water quality data and hyperspectral imagery, this team can rapidly estimate concentrations of optically-active components including colored dissolved organic matter (CDOM), blue-green algae, crude oil, temperature as well
 
-7. We apply the GTM to datasets collected by our Robot Team specifically showcasing the identification of spectral signatures for Rhodamine Dye, Algae, etc...
-
-8.
-With the signatures identified, class abundance, and therefore area estimation, can be performed rapidly, in real time, by utilizing the learned signatures with the nromalized spectral similarity score (or other similar metrics).
-
-IDEA: spectral unmixing of water is challenging due to nonlinear effects of scattering by solids and absorption by dissolved chemicals (dyes etc...) this motivates a nonlinear approach to endmember identification, etc...
+In this paper, we demonstrate how Generative Topographic Mapping (GTM), a Bayesian alternative to the Self Organizing Map, can be utilized to rapidly identify relevant spectral signatures within a scene.
+Using data collected at site in Montague, North Texas, we demonstrate the ability of this paradigm with two case studies.
+First, we demonstrate how the GTM can be used to provide a high-resolution unsupervised classification of a water body enabling the rapid identification of relevant regions which can then be used to direct further investigation by reference-sensor equipped boats.
+Second, we demonstrate the ability to perform endmember extraction and abundance mapping using imagery from a test-release of rhodamine dye.
+Together with the normalized spectral similarity score, these spectral signatures are identified and their associated area in the scene is determined.
 

### 2024-04-16 paper/notes/outline.qmd
updated with notes from outline

---
title: "Paper Outline"
format:
    html:
      self-contained: true
    pdf:
      keep-tex: true

bibliography: ../references.bib
---

# Introduction

1. Characterization of Inland Water Bodies via Remote Sensing is Challenging
   - complex spectral signatures
   - limited spatial and temporal resolution (long satellite overpasses)
   - nevertheless remote sensing has significant potential
   - Harmful Algal Bloom (HAB) classification and mapping
   - Oil Spill mapping

2. Hyperspectral Imaging addresses spectral limitations of traditional remote sensing platforms

3. Supervised methods
   - Inversion of optically-active parameters
   - Key limitation: availability of in-situ reference data
     - Cite Aurin et al. and Ross et al. here
   - When ground truth data are limited, classification is an alternative approach, e.g. "bad" to "good" water quality

4. Unsupervised methods
   - Unsupervised classification & ML Techniques
   - Endmember Extraction and Spectral Unmixing

5. Robot Teams and HSI equipped Drones

6. This paper
   - Extend capabilities of robot team by showcasing capabilities for unsupervised classification
   - Utilize GTM as Bayesian extension of SOM to enable unsupervised classification
   - GTM also enables extraction of spectral features associated with each node
   - Two case studies
     1. GTM trained on water-only HSI data for lake
        - Identify interesting regions for further examination by robot team
        - Investigate spectral variability of water
     2. GTM on dataset combining grass and water pixels with a simulated pollution event using Rhodamine tracer dye
        - Use extracted GTM "endmembers" together with NS3 to map abundance
        - Identify abundance of algae
        - Map evolution of Rhodamine dye plume

# Materials and Methods

1. Data Collection
   - Describe collection at site in Montague, North Texas
2. Autonomous Robot Team
   - Describe drone
     - Alta X Quadcopter
     - Resonon Pika XC2 HSI
     - Intel Nuc compute
   - describe georectification procedure
3. Pre-Processing
4. Generative Topographic Mapping
  - Original SOM paper [@kohonen-som-1]
  - Heskes reinterpretation of SOM algorithm as stochastic gradient descent on an energy suface i.e. justification with a cost function [@som-cost-function]
  - GTM paper [@gtm-bishop-1]
    - Describe algorithm
    - Describe parameters
  - Bayesian Information Criterion and Akaike Information Criterion for Hyperparameter Optimization
    - A full Bayesian treatment would involve marginalization over all possible parameter values but this is much more computational efficient.
5. Abundance Mapping with the Normalized Spectral Similarity Score
   - Describe different metrics for comparison of spectra
     - Trade-off between metrics that distinguish changes in overall intensity and changes in  peak location, i.e. hue

# Results

1. Two Case Studies
2. GTM Fit on Full Dataset
3. Hyperparameter Optimization with BIC
4. Spectral Signature Identification
5. Water Class Map (Water Only GTM)
6. Algae Identification
7. Dye Plume Identification

# Discussion

1. GTM Applications in our Paper
2. Comparison to existing approaches
   - Unsupervised Classification
   - Endmember Extraction & Spectral Unmixing
   - Applications for either of these with Drones?
3. Use of GTM and Unsupervised methods for Robot Team
   - As a feature transformer/preprocessor for supervised methods (as opposed to, say, PCA or k-NN)
   - identification of "interesting" regions for intelligent deployment of robotic boat
     - don't continue collecting data in region that is similar
     - provision boat to maximize data collection across distribution of GTM classes
 4. Extensions and future work
    - Incorporation of GTM extensions for efficient online-learning
      - Currently GTM classes can quickly be applied (in real time) once they are trained to generate abundance maps e.g. chlorophyll, algae, etc...
      - Key limitation of GTM is similar to SOM and k-NN: you need to compute distances between samples and latent nodes. This can scale poorly for very large datasets
      - Bishop suggests a way to augment GTM training algorithm to enable online training of GTM. This could be deployed on-board the UAV to perform classification and identify spectral signatures in near-real time.
      - Extensions of GTM [@gtm-bishop-2]
    - Utilize the GTM with in-situ data collection to identify spectral signatures of specific algal species

# Conclusions

# NOTES AND CITATIONS

1. Applications of remote sensing to water quality
   - machine learning (XGBoost) used with paired remote sensing imagery and in-situ data to classify imagery into 5 categories including 3 for harmful algal blooms [@ghatkar2019classification]
   - Spectral signatures of classes are complex and often overlapping [@thenkabail2018hyperspectral]
   - Remote sensing used for oil spill analysis: extent and thickness mapping [@kokaly2013spectroscopic; @leifer2012state]
   - Sun glitter remains a key challenge for sensing in the visible portion of the spectrum but multi and hyperspectral imagers have been used for oil spill identification and to identify their impacts on vegetation stress and mortality [@fingas2014review; @khan2018modern]

2. Hyperspectral Imagery
   - applications include food quality & safety, medical diagnoses, precision agriculture, and forensic document examination [@khan2018modern]
   - Hyperspectral data were used to distinguish oils by type, e.g. crude, diesel, gasoline, and palm [@yang2020characterization]

3. Supervised regression and classification for water quality
   - common approach is inversion of optically-active water quality parameters such as chlorophyll-a, blue-green algae, turbidity, and temperature [@ritchie2003remote]
   - Combining spectral indices such as the NDVI together with machine learning is a popular approach [@thenkabail2018hyperspectral; @sagan2020monitoring; @lu2021retrieval]
   - Polynomial regression models for chlorophyll-a, turbidity [@zhang2022selection]
   - Key limitation is collection of sufficient quantity of in-situ reference data
   - Ross et al. created a comprehensive dataset with over 600,000 water quality records matching optically active water quality parameters with associated satellite imagery from Landsat 5,7,8. To achieve this quantity of data, they needed records spanning 1984 to 2019 [@ross2019aquasat]
   - Aurin et al. took a similar route combining 30 years of remote sensing imagery with in-situ data from over 500 field compaigns for CDOM, organic carbon, etc... [@aurin2018remote].
   - When ground truth data aren't available in sufficient quantity, classification into water quality categories is another approach (ground truth can be easier to obtain by expert analysis of scene) [@koponen2002lake]

- When no ground-truth data are available, unsupervised classification can still help partition imagery into groups or clusters.
   - Many data-driven, ML methods have been employed for the task including various matrix factorizations, k-nearest neighbors, fuzzy c-means, density estimation methods, etc. [@zhang2019hyperspectral]
   - Additionally unsupervised approaches can be used to perform nonlinear dimensionality reduction & pre-processing for supervised approaches.
   - SOM used for remote sensing imagery classification [@msom-remote-sensing]
   - SOM used for clustering and data compression of HSI cube-sat [@danielsen2021self]
   - SOM used for land-use and land-cover change analysis [@penfound2021analysis]

4. Endmember Extraction and Spectral Unmixing
   - This is a related problem where the goal is to identify unique spectral signatures which combine (linearly or non-linearly) to produce the measured signal.
   - Having identified the set of endmembers, we then seek to determine their relative abundance in each pixel
   - Many statistical approaches for extracting endmembers [@berman2004ice]
   - Sparse PCA used for endmember extraction [@yousefi2016mineral]
   - Popular spectral unmixing approach is to treat "pure" endmembers as vertices of a simplex [@plaza2012endmember; @nascimento2005vertex]
   - Convolutional Neural Networks with Autoencoder architectures are a popular ML approach which identify endmembers and perform non-linear unmixing [@palsson2020convolutional; @non-negative-autoencoders; @su2019daen; @borsoi2019deep]
   - Self Organizing Map is another approach which has been used for endmember extraction together with neural networks for abundance mapping (unmixing) [@cantero2004analysis]
   - SOM has bee nused for identifying synoptic-scale patterns in wind and sea surface temperature data [@som-satellite]
   - SOM used for HSI feature extraction [@som-hsi]
   - drawback of autoencoder and other statistical approach is lack of a (topological) relationship between the classes/endmembers. Is endmember 1 closer to endmember 2 or endmember 10? The SOM addresses this. The GTM is even better...

5. Abundance Mapping
   - Many different spectral similarity functions exist for comparing spectra. Each have trade-offs between ability to distinguish differences in intensity versus differences in peak location (hue) [@deborah2015comprehensive]
   - spectral angle mapper is popular similarity function used in endmember extraction. The SAM is nice because it is more sensitive to shape than scale [@jiang2020classification]
   - spectral correlation mapper introduced to as a statistical alternative to SAM based on covariance instead of spectral angle [@de2000spectral]
   - Normalized spectral similarity score developed by Nidamanuri et al. combine the mean-squared error with spectral angle for a happy medium [@nidamanuri2010normalized]

6. Drone-based HSI
   - Near-earth HSI addresses spatial, spectral, and temporal limitations of satellite and airborne platforms. UAV-based HSI enable fine-scale mapping [@banerjee2020uav]
   - Drones can be equipped with HSI and compute to enable rapid generation of spectral indices like the NDVI for applications such as precision agriculture [@horstrand2019uav]
   - UAV-based HSI can be georectified to centimeter-scales without need for ground control points by using on-board GPS and IMU [@arroyo2019implementation]
   - UAV-based HSI for turbidity estimation [@vogt2016near]

### 2024-04-22 paper/template.tex
added them back

%  LaTeX support: latex@mdpi.com 
%  For support, please attach all files needed for compiling as well as the log file, and specify your operating system, LaTeX version, and LaTeX editor.

%=================================================================
\documentclass[journal,article,submit,pdftex,moreauthors]{Definitions/mdpi}

%--------------------
% Class Options:
%--------------------
%----------
% journal
%----------
% Choose between the following MDPI journals:
% acoustics, actuators, addictions, admsci, adolescents, aerobiology, aerospace, agriculture, agriengineering, agrochemicals, agronomy, ai, air, algorithms, allergies, alloys, analytica, analytics, anatomia, animals, antibiotics, antibodies, antioxidants, applbiosci, appliedchem, appliedmath, applmech, applmicrobiol, applnano, applsci, aquacj, architecture, arm, arthropoda, arts, asc, asi, astronomy, atmosphere, atoms, audiolres, automation, axioms, bacteria, batteries, bdcc, behavsci, beverages, biochem, bioengineering, biologics, biology, biomass, biomechanics, biomed, biomedicines, biomedinformatics, biomimetics, biomolecules, biophysica, biosensors, biotech, birds, bloods, blsf, brainsci, breath, buildings, businesses, cancers, carbon, cardiogenetics, catalysts, cells, ceramics, challenges, chemengineering, chemistry, chemosensors, chemproc, children, chips, cimb, civileng, cleantechnol, climate, clinpract, clockssleep, cmd, coasts, coatings, colloids, colorants, commodities, compounds, computation, computers, condensedmatter, conservation, constrmater, cosmetics, covid, crops, cryptography, crystals, csmf, ctn, curroncol, cyber, dairy, data, ddc, dentistry, dermato, dermatopathology, designs, devices, diabetology, diagnostics, dietetics, digital, disabilities, diseases, diversity, dna, drones, dynamics, earth, ebj, ecologies, econometrics, economies, education, ejihpe, electricity, electrochem, electronicmat, electronics, encyclopedia, endocrines, energies, eng, engproc, entomology, entropy, environments, environsciproc, epidemiologia, epigenomes, est, fermentation, fibers, fintech, fire, fishes, fluids, foods, forecasting, forensicsci, forests, foundations, fractalfract, fuels, future, futureinternet, futurepharmacol, futurephys, futuretransp, galaxies, games, gases, gastroent, gastrointestdisord, gels, genealogy, genes, geographies, geohazards, geomatics, geosciences, geotechnics, geriatrics, grasses, gucdd, hazardousmatters, healthcare, hearts, hemato, hematolrep, heritage, higheredu, highthroughput, histories, horticulturae, hospitals, humanities, humans, hydrobiology, hydrogen, hydrology, hygiene, idr, ijerph, ijfs, ijgi, ijms, ijns, ijpb, ijtm, ijtpp, ime, immuno, informatics, information, infrastructures, inorganics, insects, instruments, inventions, iot, j, jal, jcdd, jcm, jcp, jcs, jcto, jdb, jeta, jfb, jfmk, jimaging, jintelligence, jlpea, jmmp, jmp, jmse, jne, jnt, jof, joitmc, jor, journalmedia, jox, jpm, jrfm, jsan, jtaer, jvd, jzbg, kidneydial, kinasesphosphatases, knowledge, land, languages, laws, life, liquids, literature, livers, logics, logistics, lubricants, lymphatics, machines, macromol, magnetism, magnetochemistry, make, marinedrugs, materials, materproc, mathematics, mca, measurements, medicina, medicines, medsci, membranes, merits, metabolites, metals, meteorology, methane, metrology, micro, microarrays, microbiolres, micromachines, microorganisms, microplastics, minerals, mining, modelling, molbank, molecules, mps, msf, mti, muscles, nanoenergyadv, nanomanufacturing,\gdef\@continuouspages{yes}} nanomaterials, ncrna, ndt, network, neuroglia, neurolint, neurosci, nitrogen, notspecified, %%nri, nursrep, nutraceuticals, nutrients, obesities, oceans, ohbm, onco, %oncopathology, optics, oral, organics, organoids, osteology, oxygen, parasites, parasitologia, particles, pathogens, pathophysiology, pediatrrep, pharmaceuticals, pharmaceutics, pharmacoepidemiology,\gdef\@ISSN{2813-0618}\gdef\@continuous pharmacy, philosophies, photochem, photonics, phycology, physchem, physics, physiologia, plants, plasma, platforms, pollutants, polymers, polysaccharides, poultry, powders, preprints, proceedings, processes, prosthesis, proteomes, psf, psych, psychiatryint, psychoactives, publications, quantumrep, quaternary, qubs, radiation, reactions, receptors, recycling, regeneration, religions, remotesensing, reports, reprodmed, resources, rheumato, risks, robotics, ruminants, safety, sci, scipharm, sclerosis, seeds, sensors, separations, sexes, signals, sinusitis, skins, smartcities, sna, societies, socsci, software, soilsystems, solar, solids, spectroscj, sports, standards, stats, std, stresses, surfaces, surgeries, suschem, sustainability, symmetry, synbio, systems, targets, taxonomy, technologies, telecom, test, textiles, thalassrep, thermo, tomography, tourismhosp, toxics, toxins, transplantology, transportation, traumacare, traumas, tropicalmed, universe, urbansci, uro, vaccines, vehicles, venereology, vetsci, vibration, virtualworlds, viruses, vision, waste, water, wem, wevj, wind, women, world, youth, zoonoticdis 
% For posting an early version of this manuscript as a preprint, you may use "preprints" as the journal. Changing "submit" to "accept" before posting will remove line numbers.

%---------
% article
%---------
% The default type of manuscript is "article", but can be replaced by: 
% abstract, addendum, article, book, bookreview, briefreport, casereport, comment, commentary, communication, conferenceproceedings, correction, conferencereport, entry, expressionofconcern, extendedabstract, datadescriptor, editorial, essay, erratum, hypothesis, interestingimage, obituary, opinion, projectreport, reply, retraction, review, perspective, protocol, shortnote, studyprotocol, systematicreview, supfile, technicalnote, viewpoint, guidelines, registeredreport, tutorial
% supfile = supplementary materials

%----------
% submit
%----------
% The class option "submit" will be changed to "accept" by the Editorial Office when the paper is accepted. This will only make changes to the frontpage (e.g., the logo of the journal will get visible), the headings, and the copyright information. Also, line numbering will be removed. Journal info and pagination for accepted papers will also be assigned by the Editorial Office.

%------------------
% moreauthors
%------------------
% If there is only one author the class option oneauthor should be used. Otherwise use the class option moreauthors.

%---------
% pdftex
%---------
% The option pdftex is for use with pdfLaTeX. Remove "pdftex" for (1) compiling with LaTeX & dvi2pdf (if eps figures are used) or for (2) compiling with XeLaTeX.

%=================================================================
% MDPI internal commands - do not modify
\firstpage{1} 
\makeatletter 
\setcounter{page}{\@firstpage} 
\makeatother
\pubvolume{1}
\issuenum{1}
\articlenumber{0}
\pubyear{2024}
\copyrightyear{2024}
%\externaleditor{Academic Editor: Firstname Lastname}
\datereceived{ } 
\daterevised{ } % Comment out if no revised date
\dateaccepted{ } 
\datepublished{ } 
%\datecorrected{} % For corrected papers: "Corrected: XXX" date in the original paper.
%\dateretracted{} % For corrected papers: "Retracted: XXX" date in the original paper.
\hreflink{https://doi.org/} % If needed use \linebreak
%\doinum{}
%\pdfoutput=1 % Uncommented for upload to arXiv.org
%\CorrStatement{yes}  % For updates

%=================================================================
% Add packages and commands here. The following packages are loaded in our class file: fontenc, inputenc, calc, indentfirst, fancyhdr, graphicx, epstopdf, lastpage, ifthen, float, amsmath, amssymb, lineno, setspace, enumitem, mathpazo, booktabs, titlesec, etoolbox, tabto, xcolor, colortbl, soul, multirow, microtype, tikz, totcount, changepage, attrib, upgreek, array, tabularx, pbox, ragged2e, tocloft, marginnote, marginfix, enotez, amsthm, natbib, hyperref, cleveref, scrextend, url, geometry, newfloat, caption, draftwatermark, seqsplit
% cleveref: load \crefname definitions after \begin{document}

%=================================================================
% Please use the following mathematics environments: Theorem, Lemma, Corollary, Proposition, Characterization, Property, Problem, Example, ExamplesandDefinitions, Hypothesis, Remark, Definition, Notation, Assumption
%% For proofs, please use the proof environment (the amsthm package is loaded by the MDPI class).

%=================================================================
% Full title of the paper (Capitalized)
\Title{Title}

% MDPI internal command: Title for citation in the left column
\TitleCitation{Title}

% Author Orchid ID: enter ID or remove command
\newcommand{\orcidauthorA}{0000-0000-0000-000X} % Add \orcidA{} behind the author's name
%\newcommand{\orcidauthorB}{0000-0000-0000-000X} % Add \orcidB{} behind the author's name

% Authors, for the paper (add full first names)
\Author{Firstname Lastname $^{1,\dagger,\ddagger}$\orcidA{}, Firstname Lastname $^{2,\ddagger}$ and Firstname Lastname $^{2,}$*}

%\longauthorlist{yes}

% MDPI internal command: Authors, for metadata in PDF
\AuthorNames{Firstname Lastname, Firstname Lastname and Firstname Lastname}

% MDPI internal command: Authors, for citation in the left column
\AuthorCitation{Lastname, F.; Lastname, F.; Lastname, F.}
% If this is a Chicago style journal: Lastname, Firstname, Firstname Lastname, and Firstname Lastname.

% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
\address{%
$^{1}$ \quad Affiliation 1; e-mail@e-mail.com\\
$^{2}$ \quad Affiliation 2; e-mail@e-mail.com}

% Contact information of the corresponding author
\corres{Correspondence: e-mail@e-mail.com; Tel.: (optional; include country code; if there are multiple corresponding authors, add author initials) +xx-xxxx-xxx-xxxx (F.L.)}

% Current address and/or shared authorship
\firstnote{Current address: Affiliation 3.} 
\secondnote{These authors contributed equally to this work.}
% The commands \thirdnote{} till \eighthnote{} are available for further notes

%\simplesumm{} % Simple summary

%\conference{} % An extended version of a conference paper

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{A single paragraph of about 200 words maximum. For research articles, abstracts should give a pertinent overview of the work. We strongly encourage authors to use the following style of structured abstracts, but without headings: (1) Background: place the question addressed in a broad context and highlight the purpose of the study; (2) Methods: describe briefly the main methods or treatments applied; (3) Results: summarize the article's main findings; (4) Conclusions: indicate the main conclusions or interpretations. The abstract should be an objective representation of the article, it must not contain results which are not presented and substantiated in the main text and should not exaggerate the main conclusions.}

% Keywords
\keyword{keyword 1; keyword 2; keyword 3 (List three to ten pertinent keywords specific to the article; yet reasonably common within the subject discipline.)}

% The fields PACS, MSC, and JEL may be left empty or commented out if not applicable
%\PACS{J0101}
%\MSC{}
%\JEL{}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Only for the journal Diversity
%\LSID{\url{http://}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Only for the journal Applied Sciences
%\featuredapplication{Authors are encouraged to provide a concise description of the specific application or a potential application of the work. This section is not mandatory.}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Only for the journal Data
%\dataset{DOI number or link to the deposited data set if the data set is published separately. If the data set shall be published as a supplement to this paper, this field will be filled by the journal editors. In this case, please submit the data set as a supplement.}
%\datasetlicense{License under which the data set is made available (CC0, CC-BY, CC-BY-SA, CC-BY-NC, etc.)}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Only for the journal Toxins
%\keycontribution{The breakthroughs or highlights of the manuscript. Authors can write one or two sentences to describe the most important part of the paper.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Only for the journal Encyclopedia
%\encyclopediadef{For entry manuscripts only: please provide a brief overview of the entry title instead of an abstract.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Only for the journal Advances in Respiratory Medicine
%\addhighlights{yes}
%\renewcommand{\addhighlights}{%

%\noindent This is an obligatory section in “Advances in Respiratory Medicine”, whose goal is to increase the discoverability and readability of the article via search engines and other scholars. Highlights should not be a copy of the abstract, but a simple text allowing the reader to quickly and simplified find out what the article is about and what can be cited from it. Each of these parts should be devoted up to 2~bullet points.\vspace{3pt}\\
%\textbf{What are the main findings?}
% \begin{itemize}[labelsep=2.5mm,topsep=-3pt]
% \item First bullet.
% \item Second bullet.
% \end{itemize}\vspace{3pt}
%\textbf{What is the implication of the main finding?}
% \begin{itemize}[labelsep=2.5mm,topsep=-3pt]
% \item First bullet.
% \item Second bullet.
% \end{itemize}
%}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\setcounter{section}{-1} %% Remove this when starting to work on the template.
\section{How to Use this Template}

The template details the sections that can be used in a manuscript. Note that the order and names of article sections may differ from the requirements of the journal (e.g., the positioning of the Materials and Methods section). Please check the instructions on the authors' page of the journal to verify the correct order and names. For any questions, please contact the editorial office of the journal or support@mdpi.com. For LaTeX-related questions please contact latex@mdpi.com.%\endnote{This is an endnote.} % To use endnotes, please un-comment \printendnotes below (before References). Only journal Laws uses \footnote.

% The order of the section titles is different for some journals. Please refer to the "Instructions for Authors” on the journal homepage.

\section{Introduction}

The introduction should briefly place the study in a broad context and highlight why it is important. It should define the purpose of the work and its significance. The current state of the research field should be reviewed carefully and key publications cited. Please highlight controversial and diverging hypotheses when necessary. Finally, briefly mention the main aim of the work and highlight the principal conclusions. As far as possible, please keep the introduction comprehensible to scientists outside your particular field of research. Citing a journal paper \cite{ref-journal}. Now citing a book reference \cite{ref-book1,ref-book2} or other reference types \cite{ref-unpublish,ref-communication,ref-proceeding}. Please use the command \citep{ref-thesis,ref-url} for the following MDPI journals, which use author--date citation: Administrative Sciences, Arts, Econometrics, Economies, Genealogy, Humanities, IJFS, Journal of Intelligence, Journalism and Media, JRFM, Languages, Laws, Religions, Risks, Social Sciences, Literature.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Materials and Methods}

Materials and Methods should be described with sufficient details to allow others to replicate and build on published results. Please note that publication of your manuscript implicates that you must make all materials, data, computer code, and protocols associated with the publication available to readers. Please disclose at the submission stage any restrictions on the availability of materials or information. New methods and protocols should be described in detail while well-established methods can be briefly described and appropriately cited.

Research manuscripts reporting large datasets that are deposited in a publicly avail-able database should specify where the data have been deposited and provide the relevant accession numbers. If the accession numbers have not yet been obtained at the time of submission, please state that they will be provided during review. They must be provided prior to publication.

Interventionary studies involving animals or humans, and other studies require ethical approval must list the authority that provided approval and the corresponding ethical approval code.
\begin{quote}
This is an example of a quote.
\end{quote}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}

This section may be divided by subheadings. It should provide a concise and precise description of the experimental results, their interpretation as well as the experimental conclusions that can be drawn.
\subsection{Subsection}
\subsubsection{Subsubsection}

Bulleted lists look like this:
\begin{itemize}
\item	First bullet;
\item	Second bullet;
\item	Third bullet.
\end{itemize}

Numbered lists can be added as follows:
\begin{enumerate}
\item	First item; 
\item	Second item;
\item	Third item.
\end{enumerate}

The text continues here.

\subsection{Figures, Tables and Schemes}

All figures and tables should be cited in the main text as Figure~\ref{fig1}, Table~\ref{tab1}, etc.

\begin{figure}[H]
\includegraphics[width=10.5 cm]{Definitions/logo-mdpi}
\caption{This is a figure. Schemes follow the same formatting. If there are multiple panels, they should be listed as: (\textbf{a}) Description of what is contained in the first panel. (\textbf{b}) Description of what is contained in the second panel. Figures should be placed in the main text near to the first time they are cited. A caption on a single line should be centered.\label{fig1}}
\end{figure}   
\unskip

\begin{table}[H] 
\caption{This is a table caption. Tables should be placed in the main text near to the first time they are~cited.\label{tab1}}
\newcolumntype{C}{>{\centering\arraybackslash}X}
\begin{tabularx}{\textwidth}{CCC}
\toprule
\textbf{Title 1}	& \textbf{Title 2}	& \textbf{Title 3}\\
\midrule
Entry 1		& Data			& Data\\
Entry 2		& Data			& Data \textsuperscript{1}\\
\bottomrule
\end{tabularx}
\noindent{\footnotesize{\textsuperscript{1} Tables may have a footer.}}
\end{table}

The text continues here (Figure~\ref{fig2} and Table~\ref{tab2}).

% Example of a figure that spans the whole page width. The same concept works for tables, too.
\begin{figure}[H]
\begin{adjustwidth}{-\extralength}{0cm}
\centering
\includegraphics[width=15.5cm]{Definitions/logo-mdpi}
\end{adjustwidth}
\caption{This is a wide figure.\label{fig2}}
\end{figure}

\begin{table}[H]
\caption{This is a wide table.\label{tab2}}
	\begin{adjustwidth}{-\extralength}{0cm}
		\newcolumntype{C}{>{\centering\arraybackslash}X}
		\begin{tabularx}{\fulllength}{CCCC}
			\toprule
			\textbf{Title 1}	& \textbf{Title 2}	& \textbf{Title 3}     & \textbf{Title 4}\\
			\midrule
\multirow[m]{3}{*}{Entry 1 *}	& Data			& Data			& Data\\
			  	                   & Data			& Data			& Data\\
			             	      & Data			& Data			& Data\\
                   \midrule
\multirow[m]{3}{*}{Entry 2}    & Data			& Data			& Data\\
			  	                  & Data			& Data			& Data\\
			             	     & Data			& Data			& Data\\
                   \midrule
\multirow[m]{3}{*}{Entry 3}    & Data			& Data			& Data\\
			  	                 & Data			& Data			& Data\\
			             	    & Data			& Data			& Data\\
                  \midrule
\multirow[m]{3}{*}{Entry 4}   & Data			& Data			& Data\\
			  	                 & Data			& Data			& Data\\
			             	    & Data			& Data			& Data\\
			\bottomrule
		\end{tabularx}
	\end{adjustwidth}
	\noindent{\footnotesize{* Tables may have a footer.}}
\end{table}

%\begin{listing}[H]
%\caption{Title of the listing}
%\rule{\columnwidth}{1pt}
%\raggedright Text of the listing. In font size footnotesize, small, or normalsize. Preferred format: left aligned and single spaced. Preferred border format: top border line and bottom border line.
%\rule{\columnwidth}{1pt}
%\end{listing}

Text.

Text.

\subsection{Formatting of Mathematical Components}

This is the example 1 of equation:
\begin{linenomath}
\begin{equation}
a = 1,
\end{equation}
\end{linenomath}
the text following an equation need not be a new paragraph. Please punctuate equations as regular text.
%% If the documentclass option "submit" is chosen, please insert a blank line before and after any math environment (equation and eqnarray environments). This ensures correct linenumbering. The blank line should be removed when the documentclass option is changed to "accept" because the text following an equation should not be a new paragraph.

This is the example 2 of equation:
\begin{adjustwidth}{-\extralength}{0cm}
\begin{equation}
a = b + c + d + e + f + g + h + i + j + k + l + m + n + o + p + q + r + s + t + u + v + w + x + y + z
\end{equation}
\end{adjustwidth}

% Example of a page in landscape format (with table and table footnote).
%\startlandscape
%\begin{table}[H] %% Table in wide page
%\caption{This is a very wide table.\label{tab3}}
%	\begin{tabularx}{\textwidth}{CCCC}
%		\toprule
%		\textbf{Title 1}	& \textbf{Title 2}	& \textbf{Title 3}	& \textbf{Title 4}\\
%		\midrule
%		Entry 1		& Data			& Data			& This cell has some longer content that runs over two lines.\\
%		Entry 2		& Data			& Data			& Data\textsuperscript{1}\\
%		\bottomrule
%	\end{tabularx}
%	\begin{adjustwidth}{+\extralength}{0cm}
%		\noindent\footnotesize{\textsuperscript{1} This is a table footnote.}
%	\end{adjustwidth}
%\end{table}
%\finishlandscape

Please punctuate equations as regular text. Theorem-type environments (including propositions, lemmas, corollaries etc.) can be formatted as follows:
%% Example of a theorem:
\begin{Theorem}
Example text of a theorem.
\end{Theorem}

The text continues here. Proofs must be formatted as follows:

%% Example of a proof:
\begin{proof}[Proof of Theorem 1]
Text of the proof. Note that the phrase ``of Theorem 1'' is optional if it is clear which theorem is being referred to.
\end{proof}
The text continues here.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}

Authors should discuss the results and how they can be interpreted from the perspective of previous studies and of the working hypotheses. The findings and their implications should be discussed in the broadest context possible. Future research directions may also be highlighted.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions}

This section is not mandatory, but can be added to the manuscript if the discussion is unusually long or complex.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Patents}

This section is not mandatory, but may be added if there are patents resulting from the work reported in this manuscript.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\vspace{6pt}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% optional
%\supplementary{The following supporting information can be downloaded at:  \linksupplementary{s1}, Figure S1: title; Table S1: title; Video S1: title.}

% Only for journal Methods and Protocols:
% If you wish to submit a video article, please do so with any other supplementary material.
% \supplementary{The following supporting information can be downloaded at: \linksupplementary{s1}, Figure S1: title; Table S1: title; Video S1: title. A supporting video article is available at doi: link.}

% Only for journal Hardware:
% If you wish to submit a video article, please do so with any other supplementary material.
% \supplementary{The following supporting information can be downloaded at: \linksupplementary{s1}, Figure S1: title; Table S1: title; Video S1: title.\vspace{6pt}\\
%\begin{tabularx}{\textwidth}{lll}
%\toprule
%\textbf{Name} & \textbf{Type} & \textbf{Description} \\
%\midrule
%S1 & Python script (.py) & Script of python source code used in XX \\
%S2 & Text (.txt) & Script of modelling code used to make Figure X \\
%S3 & Text (.txt) & Raw data from experiment X \\
%S4 & Video (.mp4) & Video demonstrating the hardware in use \\
%... & ... & ... \\
%\bottomrule
%\end{tabularx}
%}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\authorcontributions{For research articles with several authors, a short paragraph specifying their individual contributions must be provided. The following statements should be used ``Conceptualization, X.X. and Y.Y.; methodology, X.X.; software, X.X.; validation, X.X., Y.Y. and Z.Z.; formal analysis, X.X.; investigation, X.X.; resources, X.X.; data curation, X.X.; writing---original draft preparation, X.X.; writing---review and editing, X.X.; visualization, X.X.; supervision, X.X.; project administration, X.X.; funding acquisition, Y.Y. All authors have read and agreed to the published version of the manuscript.'', please turn to the  \href{http://img.mdpi.org/data/contributor-role-instruction.pdf}{CRediT taxonomy} for the term explanation. Authorship must be limited to those who have contributed substantially to the work~reported.}

\funding{Please add: ``This research received no external funding'' or ``This research was funded by NAME OF FUNDER grant number XXX.'' and  and ``The APC was funded by XXX''. Check carefully that the details given are accurate and use the standard spelling of funding agency names at \url{https://search.crossref.org/funding}, any errors may affect your future funding.}

\institutionalreview{In this section, you should add the Institutional Review Board Statement and approval number, if relevant to your study. You might choose to exclude this statement if the study did not require ethical approval. Please note that the Editorial Office might ask you for further information. Please add “The study was conducted in accordance with the Declaration of Helsinki, and approved by the Institutional Review Board (or Ethics Committee) of NAME OF INSTITUTE (protocol code XXX and date of approval).” for studies involving humans. OR “The animal study protocol was approved by the Institutional Review Board (or Ethics Committee) of NAME OF INSTITUTE (protocol code XXX and date of approval).” for studies involving animals. OR “Ethical review and approval were waived for this study due to REASON (please provide a detailed justification).” OR “Not applicable” for studies not involving humans or animals.}

\informedconsent{Any research article describing a study involving humans should contain this statement. Please add ``Informed consent was obtained from all subjects involved in the study.'' OR ``Patient consent was waived due to REASON (please provide a detailed justification).'' OR ``Not applicable'' for studies not involving humans. You might also choose to exclude this statement if the study did not involve humans.

Written informed consent for publication must be obtained from participating patients who can be identified (including by the patients themselves). Please state ``Written informed consent has been obtained from the patient(s) to publish this paper'' if applicable.}

\dataavailability{We encourage all authors of articles published in MDPI journals to share their research data. In this section, please provide details regarding where data supporting reported results can be found, including links to publicly archived datasets analyzed or generated during the study. Where no new data were created, or where data is unavailable due to privacy or ethical restrictions, a statement is still required. Suggested Data Availability Statements are available in section ``MDPI Research Data Policies'' at \url{https://www.mdpi.com/ethics}.}

% Only for journal Nursing Reports
%\publicinvolvement{Please describe how the public (patients, consumers, carers) were involved in the research. Consider reporting against the GRIPP2 (Guidance for Reporting Involvement of Patients and the Public) checklist. If the public were not involved in any aspect of the research add: ``No public involvement in any aspect of this research''.}

% Only for journal Nursing Reports
%\guidelinesstandards{Please add a statement indicating which reporting guideline was used when drafting the report. For example, ``This manuscript was drafted against the XXX (the full name of reporting guidelines and citation) for XXX (type of research) research''. A complete list of reporting guidelines can be accessed via the equator network: \url{https://www.equator-network.org/}.}

\acknowledgments{In this section you can acknowledge any support given which is not covered by the author contribution or funding sections. This may include administrative and technical support, or donations in kind (e.g., materials used for experiments).}

\conflictsofinterest{Declare conflicts of interest or state ``The authors declare no conflicts of interest.'' Authors must identify and declare any personal circumstances or interest that may be perceived as inappropriately influencing the representation or interpretation of reported research results. Any role of the funders in the design of the study; in the collection, analyses or interpretation of data; in the writing of the manuscript; or in the decision to publish the results must be declared in this section. If there is no role, please state ``The funders had no role in the design of the study; in the collection, analyses, or interpretation of data; in the writing of the manuscript; or in the decision to publish the results''.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Optional

%% Only for journal Encyclopedia
%\entrylink{The Link to this entry published on the encyclopedia platform.}

\abbreviations{Abbreviations}{
The following abbreviations are used in this manuscript:\\

\noindent 
\begin{tabular}{@{}ll}
MDPI & Multidisciplinary Digital Publishing Institute\\
DOAJ & Directory of open access journals\\
TLA & Three letter acronym\\
LD & Linear dichroism
\end{tabular}
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Optional
\appendixtitles{no} % Leave argument "no" if all appendix headings stay EMPTY (then no dot is printed after "Appendix A"). If the appendix sections contain a heading then change the argument to "yes".
\appendixstart
\appendix
\section[\appendixname~\thesection]{}
\subsection[\appendixname~\thesubsection]{}
The appendix is an optional section that can contain details and data supplemental to the main text---for example, explanations of experimental details that would disrupt the flow of the main text but nonetheless remain crucial to understanding and reproducing the research shown; figures of replicates for experiments of which representative data are shown in the main text can be added here if brief, or as Supplementary Data. Mathematical proofs of results not central to the paper can be added as an appendix.

\begin{table}[H] 
\caption{This is a table caption.\label{tab5}}
\newcolumntype{C}{>{\centering\arraybackslash}X}
\begin{tabularx}{\textwidth}{CCC}
\toprule
\textbf{Title 1}	& \textbf{Title 2}	& \textbf{Title 3}\\
\midrule
Entry 1		& Data			& Data\\
Entry 2		& Data			& Data\\
\bottomrule
\end{tabularx}
\end{table}

\section[\appendixname~\thesection]{}
All appendix sections must be cited in the main text. In the appendices, Figures, Tables, etc. should be labeled, starting with ``A''---e.g., Figure A1, Figure A2, etc.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{adjustwidth}{-\extralength}{0cm}
%\printendnotes[custom] % Un-comment to print a list of endnotes

\reftitle{References}

% Please provide either the correct journal abbreviation (e.g. according to the “List of Title Word Abbreviations” http://www.issn.org/services/online-services/access-to-the-ltwa/) or the full name of the journal.
% Citations and References in Supplementary files are permitted provided that they also appear in the reference list here.

%=====================================
% References, variant A: external bibliography
%=====================================
%\bibliography{your_external_BibTeX_file}

%=====================================
% References, variant B: internal bibliography
%=====================================
\begin{thebibliography}{999}
% Reference 1
\bibitem[Author1(year)]{ref-journal}
Author~1, T. The title of the cited article. {\em Journal Abbreviation} {\bf 2008}, {\em 10}, 142--149.
% Reference 2
\bibitem[Author2(year)]{ref-book1}
Author~2, L. The title of the cited contribution. In {\em The Book Title}; Editor 1, F., Editor 2, A., Eds.; Publishing House: City, Country, 2007; pp. 32--58.
% Reference 3
\bibitem[Author3(year)]{ref-book2}
Author 1, A.; Author 2, B. \textit{Book Title}, 3rd ed.; Publisher: Publisher Location, Country, 2008; pp. 154--196.
% Reference 4
\bibitem[Author4(year)]{ref-unpublish}
Author 1, A.B.; Author 2, C. Title of Unpublished Work. \textit{Abbreviated Journal Name} year, \textit{phrase indicating stage of publication (submitted; accepted; in press)}.
% Reference 5
\bibitem[Author5(year)]{ref-communication}
Author 1, A.B. (University, City, State, Country); Author 2, C. (Institute, City, State, Country). Personal communication, 2012.
% Reference 6
\bibitem[Author6(year)]{ref-proceeding}
Author 1, A.B.; Author 2, C.D.; Author 3, E.F. Title of presentation. In Proceedings of the Name of the Conference, Location of Conference, Country, Date of Conference (Day Month Year); Abstract Number (optional), Pagination (optional).
% Reference 7
\bibitem[Author7(year)]{ref-thesis}
Author 1, A.B. Title of Thesis. Level of Thesis, Degree-Granting University, Location of University, Date of Completion.
% Reference 8
\bibitem[Author8(year)]{ref-url}
Title of Site. Available online: URL (accessed on Day Month Year).
\end{thebibliography}

% If authors have biography, please use the format below
%\section*{Short Biography of Authors}
%\bio
%{\raisebox{-0.35cm}{\includegraphics[width=3.5cm,height=5.3cm,clip,keepaspectratio]{Definitions/author1.pdf}}}
%{\textbf{Firstname Lastname} Biography of first author}
%
%\bio
%{\raisebox{-0.35cm}{\includegraphics[width=3.5cm,height=5.3cm,clip,keepaspectratio]{Definitions/author2.jpg}}}
%{\textbf{Firstname Lastname} Biography of second author}

% For the MDPI journals use author-date citation, please follow the formatting guidelines on http://www.mdpi.com/authors/references
% To cite two works by the same author: \citeauthor{ref-journal-1a} (\citeyear{ref-journal-1a}, \citeyear{ref-journal-1b}). This produces: Whittaker (1967, 1975)
% To cite two works by the same author with specific pages: \citeauthor{ref-journal-3a} (\citeyear{ref-journal-3a}, p. 328; \citeyear{ref-journal-3b}, p.475). This produces: Wong (1999, p. 328; 2000, p. 475)

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% for journal Sci
%\reviewreports{\\
%Reviewer 1 comments and authors’ response\\
%Reviewer 2 comments and authors’ response\\
%Reviewer 3 comments and authors’ response
%}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\PublishersNote{}
\end{adjustwidth}
\end{document}

### 2024-04-22 paper/paper.tex

@@
 
-Inland water bodies pose a unique challenge for analysis by remote sensing imagery due to their complex spectral characteristics and small-scale spatial variability; nevertheless, significant effort has gone into the development of techniques for estimating optically-active water quality parameters such as chlorophyll-a, blue-green algae, turbidity, and temperature (citation).
-Machine learning techniques have proven to be a powerful tool in this domain, facilitating the development of data-driven models which map observed reflectance to parameters of interest.
+Inland water bodies pose a unique challenge for analysis with remote sensing imagery due to their complex spectral characteristics and small-scale spatial variability; nevertheless, significant effort has driven the development of techniques for estimating optically-active water quality parameters such as chlorophyll-a, blue-green algae, turbidity, and temperature \cite{ritchie2003remote, bonansea2015using}.
+Machine learning has proven to be a powerful tool in this domain, with tree-based models, support vector machines, deep neural networks, and others demonstrating success \cite{thenkabail2018hyperspectral, sagan2020monitoring}
+
+facilitating the development of data-driven models mapping observed reflectance to parameters of interest.
+
 However, these methods rely on the availability of significant volumes of in-situ reference data needed to calibrate models and asses their uncertainty.
@@
 
+- Remote sensing used for oil spill analysis: extent and thickness mapping \cite{kokaly2013spectroscopic, leifer2012state}
+
+- Sun glitter remains a key challenge for sensing in the visible portion of the spectrum but multi and hyperspectral imagers have been used for oil spill identification and to identify their impacts on vegetation stress and mortality \cite{fingas2014review, khan2018modern}
+
+- Combining spectral indices such as the NDVI together with machine learning is a popular approach \cite{thenkabail2018hyperspectral, sagan2020monitoring, lu2021retrieval}
+
+- Ross et al.
+created a comprehensive dataset with over 600,000 water quality records matching optically active water quality parameters with associated satellite imagery from Landsat 5,7,8.
+To achieve this quantity of data, they needed records spanning 1984 to 2019 \cite{ross2019aquasat}
+- Aurin et al. took a similar route combining 30 years of remote sensing imagery with in-situ data from over 500 field compaigns for CDOM, organic carbon, etc... \cite{aurin2018remote}.
+
+- When ground truth data aren't available in sufficient quantity, classification into water quality categories is another approach (ground truth can be easier to obtain by expert analysis of scene) \cite{koponen2002lake}
+
 - supervised classification of water quality with XGBoost \cite{ghatkar2019classification}
@@
 Unsupervised classification
+- When no ground-truth data are available, unsupervised classification can still help partition imagery into groups or clusters.
+- Many data-driven, ML methods have been employed for the task including various matrix factorizations, k-nearest neighbors, fuzzy c-means, density estimation methods, etc.
+\cite{zhang2019hyperspectral}
+- Additionally unsupervised approaches can be used to perform nonlinear dimensionality reduction & pre-processing for supervised approaches.
+- Many unsupervised approaches can be challenging to interpret due to the lack of a relationship between the identified classes, e.g. is class 1 more similar to class 2 than class 10?
+- The Self Organizing Map is a popular technique which classifies data while enforcing topological relationships between the resulting classes such that classes close to each  other in the latent space are more similar than classes that are far apart.
+- SOM used for remote sensing imagery classification \cite{msom-remote-sensing}
+- SOM used for clustering and data compression of HSI cube-sat \cite{danielsen2021self}
+- SOM used for land-use and land-cover change analysis \cite{penfound2021analysis}
+- SOM has bee nused for identifying synoptic-scale patterns in wind and sea surface temperature data {som-satellite}
 
 endmember extraction, spectral unmixing
+- This is a related problem where the goal is to identify unique spectral signatures which combine (linearly or non-linearly) to produce the measured signal.
+- Having identified the set of endmembers, we then seek to determine their relative abundance in each pixel
+- Many statistical approaches for extracting endmembers \cite{berman2004ice}
+- Sparse PCA used for endmember extraction \cite{yousefi2016mineral}
+- Popular spectral unmixing approach is to treat "pure" endmembers as vertices of a simplex \cite{plaza2012endmember, nascimento2005vertex}
+- Convolutional Neural Networks with Autoencoder architectures are a popular ML approach which identify endmembers and perform non-linear unmixing \cite{palsson2020convolutional, non-negative-autoencoders, su2019daen, borsoi2019deep}
+- Self Organizing Map is another approach which has been used for endmember extraction together with neural networks for abundance mapping (unmixing) \cite{cantero2004analysis}
 
@@
 By flying close to the ground UAV-born hyperspectral imagers can produce centimeter-scale imagery while simulatenously limiting the need for complicated atmospheric corrections required for satellite platforms (citation).
+
+- applications include food quality & safety, medical diagnoses, precision agriculture, and forensic document examination \cite{khan2018modern}
+- Hyperspectral data were used to distinguish oils by type, e.g. crude, diesel, gasoline, and palm \cite{yang2020characterization}
+- Near-earth HSI addresses spatial, spectral, and temporal limitations of satellite and airborne platforms. UAV-based HSI enable fine-scale mapping \cite{banerjee2020uav}
+- Drones can be equipped with HSI and compute to enable rapid generation of spectral indices like the NDVI for applications such as precision agriculture \cite{horstrand2019uav}
+- UAV-based HSI can be georectified to centimeter-scales without need for ground control points by using on-board GPS and IMU \cite{arroyo2019implementation}
+- UAV-based HSI for turbidity estimation \cite{vogt2016near}
+- Polynomial regression models for chlorophyll-a, turbidity \cite{zhang2022selection}
 

### 2024-04-22 paper/paper.tex
updating abstract and intro

@@
 % Abstract (Do not insert blank lines, i.e. \\) 
-\abstract{FIXME}
+\abstract{Advancements in hyperspectral imaging technology have lead to a continual decrease in their size facilitating their inclusion as the payload of small, highly maneuverable unmanned aerial vehicles (UAVs) such as drones.
+Consequently, much work has gone into the development of UAV platforms and associated techniques
+- generative topographic mapping (GTM) as an efficient unsupervised method 
+- additionally the GTM can be used to perform nonlinear endmember extraction
+- We demonstrate the abilities of this approach using hyperspectral imagery collected at a pond in Montague, North Texas.
+- First, we apply the GTM to perform an unsupervised classification for all water pixels. 
+    - benefit: GTM classes maintain topological relationships in the latent space resulting in a smooth relationship between the obtained classes...
+    - ... revealing the small-scale spatial variability within the pond
+- Second, we utilize the GTM to identify reflectance spectra corresponding to specific features. 
+- Using these extracted features, we are able to map the extent of algae in the pond in addition to the time evolution of a rhodamine dye tracer. 
+- This approach demonstrates the value of unsupervised techniques coupled with UAV-born hyperspectral imaging for rapid water quality assessment.
+}
 
 % Keywords
-\keyword{Hyperspectral Imaging; Remote Sensing; Unsupervised Classification; Spectral Unmixing}
+\keyword{Hyperspectral Imaging; Remote Sensing; Unsupervised Classification; Endmember Extraction, Spectral Unmixing}
 
@@
 Inland water bodies pose a unique challenge for analysis with remote sensing imagery due to their complex spectral characteristics and small-scale spatial variability; nevertheless, significant effort has driven the development of techniques for estimating optically-active water quality parameters such as chlorophyll-a, blue-green algae, turbidity, and temperature \cite{ritchie2003remote, bonansea2015using}.
-Machine learning has proven to be a powerful tool in this domain, with tree-based models, support vector machines, deep neural networks, and others demonstrating success \cite{thenkabail2018hyperspectral, sagan2020monitoring}
+Machine learning has proven to be a powerful tool in this domain, with tree-based models, support vector machines, deep neural networks, and others providing data-driven approaches to map imagery to relevant quantities of interest \cite{thenkabail2018hyperspectral, ghatkar2019classification,sagan2020monitoring}.
+However, the training and evaluation of these models demands significant volumes of in-situ reference data.
+This process is both time consuming and expensive.
+For example, Aurin et al.
+developed a comprehensive dataset of combined in-situ measurements from over 500 field campaigns together with paired remote sensing imagery spanning over 30 years in order to develop accurate inventories for colored dissolved organic matter (CDOM), particulate organic carbon, and more \cite{aurin2018remote}.
+Similarly, Ross et al.
+combined over 600,000 water quality records for suspended sediment, dissolved organic carbon, chlorophyll-a, and Secchi-disk depth with paired imagery from Landsat.
 
-facilitating the development of data-driven models mapping observed reflectance to parameters of interest.
+Achieving this data volume 
+spanning 1984 to 2019.
 
-However, these methods rely on the availability of significant volumes of in-situ reference data needed to calibrate models and asses their uncertainty.
-For example Aurin et al...
-Similarly, Ross et al developed...
-The challenge posed by in-situ data collection together with long satellite revisit times limits the utility of remote sensing imagery for real-time applications.
-
+- Example: distinguishing species for harmful algal blooms (HABs) and distinguishing types of oil for oil spill monitoring
 - Remote sensing used for oil spill analysis: extent and thickness mapping \cite{kokaly2013spectroscopic, leifer2012state}
@@
 - Sun glitter remains a key challenge for sensing in the visible portion of the spectrum but multi and hyperspectral imagers have been used for oil spill identification and to identify their impacts on vegetation stress and mortality \cite{fingas2014review, khan2018modern}
-
-- Combining spectral indices such as the NDVI together with machine learning is a popular approach \cite{thenkabail2018hyperspectral, sagan2020monitoring, lu2021retrieval}
-
-- Ross et al.
-created a comprehensive dataset with over 600,000 water quality records matching optically active water quality parameters with associated satellite imagery from Landsat 5,7,8.
-To achieve this quantity of data, they needed records spanning 1984 to 2019 \cite{ross2019aquasat}
-- Aurin et al. took a similar route combining 30 years of remote sensing imagery with in-situ data from over 500 field compaigns for CDOM, organic carbon, etc... \cite{aurin2018remote}.
 

### 2024-04-22 paper/paper.tex
updating introduction some more...

@@
 
-Inland water bodies pose a unique challenge for analysis with remote sensing imagery due to their complex spectral characteristics and small-scale spatial variability; nevertheless, significant effort has driven the development of techniques for estimating optically-active water quality parameters such as chlorophyll-a, blue-green algae, turbidity, and temperature \cite{ritchie2003remote, bonansea2015using}.
-Machine learning has proven to be a powerful tool in this domain, with tree-based models, support vector machines, deep neural networks, and others providing data-driven approaches to map imagery to relevant quantities of interest \cite{thenkabail2018hyperspectral, ghatkar2019classification,sagan2020monitoring}.
-However, the training and evaluation of these models demands significant volumes of in-situ reference data.
-This process is both time consuming and expensive.
-For example, Aurin et al.
-developed a comprehensive dataset of combined in-situ measurements from over 500 field campaigns together with paired remote sensing imagery spanning over 30 years in order to develop accurate inventories for colored dissolved organic matter (CDOM), particulate organic carbon, and more \cite{aurin2018remote}.
-Similarly, Ross et al.
-combined over 600,000 water quality records for suspended sediment, dissolved organic carbon, chlorophyll-a, and Secchi-disk depth with paired imagery from Landsat.
+Inland water bodies pose a unique challenge for analysis with remote sensing imagery due to their complex spectral characteristics and small-scale spatial variability; nevertheless, significant effort has driven the development of techniques for estimating optically-active water quality parameters such as chlorophyll-a, blue-green algae, colored dissolved organic matter (CDOM), turbidity, and temperature \cite{koponen2002lake,ritchie2003remote, bonansea2015using}.
+Machine learning has proven to be a powerful tool in this domain, with tree-based models, support vector machines, deep neural networks, and others providing data-driven approaches to map imagery to parameters of interest \cite{thenkabail2018hyperspectral, ghatkar2019classification,sagan2020monitoring}.
+However, the training and evaluation of these models demands significant volumes of coincident, in situ data.
+Collecting these datasets is both time consuming and expensive often requiring observations spanning decades \cite{aurin2018remote, ross2019aquasat}.
 
-Achieving this data volume 
-spanning 1984 to 2019.
-
+- \cite{koponen2002lake}
 - Example: distinguishing species for harmful algal blooms (HABs) and distinguishing types of oil for oil spill monitoring
@@
 - Sun glitter remains a key challenge for sensing in the visible portion of the spectrum but multi and hyperspectral imagers have been used for oil spill identification and to identify their impacts on vegetation stress and mortality \cite{fingas2014review, khan2018modern}
-
-- When ground truth data aren't available in sufficient quantity, classification into water quality categories is another approach (ground truth can be easier to obtain by expert analysis of scene) \cite{koponen2002lake}
 
@@
 \begin{tabular}{@{}ll}
+CDOM & Colored Dissolved Organic Matter \\
 HAB & Harmful Algal Bloom \\ 

### 2024-04-22 paper/notes/outline.qmd
updated outline

@@
 ---
-title: "Paper Outline"
+title: "DRAFT: Introduction"
 format:
-    html:
-      self-contained: true
     pdf:
@@
 
-1. Characterization of Inland Water Bodies via Remote Sensing is Challenging
-   - complex spectral signatures
-   - limited spatial and temporal resolution (long satellite overpasses)
-   - nevertheless remote sensing has significant potential
-   - Harmful Algal Bloom (HAB) classification and mapping
-   - Oil Spill mapping
+1. Characterization of inland water quality with remote sensing is hard:
+   - Key limitations
+     - Spectral Resolution (visible and multi-spectral imaging)
+     - Spatial Resolution (large pixel sizes)
+     - Temporal Resolution (long satellite revisit times)
+     - Availability of in situ reference data
+   - When appropriate data are available, supervised ML approaches can map reflectance from remote sensing imagery to relevant water quality parameters
 
-2. Hyperspectral Imaging addresses spectral limitations of traditional remote sensing platforms
+2. Hyperpsectral imaging addresses spectral resolution
 
-3. Supervised methods
-   - Inversion of optically-active parameters
-   - Key limitation: availability of in-situ reference data
-     - Cite Aurin et al. and Ross et al. here
-   - When ground truth data are limited, classification is an alternative approach, e.g. "bad" to "good" water quality
+3. UAV addresses spatial and temporal resolution
 
-4. Unsupervised methods
-   - Unsupervised classification & ML Techniques
-   - Endmember Extraction and Spectral Unmixing
+4. Robot team addresses in situ data but what can we do if we don't already know which sensors to include?
 
-5. Robot Teams and HSI equipped Drones
+5. Unsupervised ML, Endmember extraction, and Spectral Unmixing
+   - k-NN, fuzzy c-means, matrix factorizations, etc.
+     - Are general but do not provide relationship between classes e.g. is class 1 more similar to class 2 or class 10?
+     - SOM addresses this by enforcing a topological relationship between clusters. Classes close to each other in the map are more similar than classes far apart
+   - Similarly, we may also be interested in the spectral characteristics of the learned *classes*, i.e. we want to think of the classes as *endmembers*.
+     - Traditional approaches rely on assumption of existence of "pure pixels" in the scene
+     - additionally, we may have to treat number of endmembers as a hyperparameter
 
-6. This paper
-   - Extend capabilities of robot team by showcasing capabilities for unsupervised classification
-   - Utilize GTM as Bayesian extension of SOM to enable unsupervised classification
-   - GTM also enables extraction of spectral features associated with each node
-   - Two case studies
-     1. GTM trained on water-only HSI data for lake
-        - Identify interesting regions for further examination by robot team
-        - Investigate spectral variability of water
-     2. GTM on dataset combining grass and water pixels with a simulated pollution event using Rhodamine tracer dye
-        - Use extracted GTM "endmembers" together with NS3 to map abundance
-        - Identify abundance of algae
-        - Map evolution of Rhodamine dye plume
+6. ...in this paper, we explore how the generative topographic mapping can be used to
+   - enable principled unsupervised classification of hyperspectral imagery by
+     - bayesian treatment (as compared to the self organizing map)
+     - topological ordering with smooth relationship between classes in latent space
+   - Train GTM on water pixels to obtain unsupervised classification mapping the detailed structure and small-scale variability of pond in Montague, North Texas
+     - Mapping can be further utilized for smart deployment of robot team (maybe save for discussion section?) i.e. requisition boat to bring reference sensors to "interesting" area
+   - Using combined dataset including land, water, and tracer dye, we demonstrate endmember extraction
+   - Using learned spectra together with NS3, we are able to evaluate extent of algae (application to harmful algal bloom mapping)
+   - Use learned spectra to map extent
 
@@
 1. Data Collection
-   - Describe collection at site in Montague, North Texas
+   - Describe collection site in Montague, North Texas
 2. Autonomous Robot Team
-   - Describe drone
+   - Describe drone setup
      - Alta X Quadcopter
-     - Resonon Pika XC2 HSI
-     - Intel Nuc compute
-   - describe georectification procedure
-3. Pre-Processing
+     - Resonon Pika SC2 HSI
+       - 462 wavelengths per pixel
+       - 1600 pixels per scan-line
+       - embedded GPS/IMU
+     - Intel NUC for on-board compute
+3. Pre-processing
 4. Generative Topographic Mapping
-  - Original SOM paper [@kohonen-som-1]
-  - Heskes reinterpretation of SOM algorithm as stochastic gradient descent on an energy suface i.e. justification with a cost function [@som-cost-function]
-  - GTM paper [@gtm-bishop-1]
-    - Describe algorithm
+   - Start off with Kohonen SOM paper. Key benefit:
+     - Topological relationship between classes i.e. similarity vs dissimilarity as a function of distance in mapping space (latent space)
+   - Limitations of SOM
+     - Not probabilistic
+     - No clear convergence criterion
+     - Heuristic based approach with no clear way to select hyperparameters like neighborhood function, number of classes, etc...
+     - Heskes did reinterpret algorithm as stochastic gradient descent on a cost function...
+   - GTM Algorithm
+     - Describe Algorithm
      - Describe parameters
-  - Bayesian Information Criterion and Akaike Information Criterion for Hyperparameter Optimization
-    - A full Bayesian treatment would involve marginalization over all possible parameter values but this is much more computational efficient.
-5. Abundance Mapping with the Normalized Spectral Similarity Score
-   - Describe different metrics for comparison of spectra
-     - Trade-off between metrics that distinguish changes in overall intensity and changes in  peak location, i.e. hue
+   - Bayesian information criterion and Akaike information criterion for hyperaprameter optimization
+       - As an alternative to performing full Bayesian treatment for marginalization over hyperparameters
+   - Abundance mapping with spectral similarity score
+     - paper comparing different distance functions for spectra
+       - curse of dimensionality can be pernicious here as we have > 462 wavelengths per pixel
+     - NS3 offers good trade off between spectral angle and euclidean-like distance (MSE)
+   - Describe use of thresholded NS3 for mapping abundance of GTM derived endmembers.
 
@@
 
-1. Two Case Studies
-2. GTM Fit on Full Dataset
-3. Hyperparameter Optimization with BIC
-4. Spectral Signature Identification
-5. Water Class Map (Water Only GTM)
-6. Algae Identification
-7. Dye Plume Identification
+1. GTM fit on Full Dataset
+2. Hyperparameter optimization
+3. Spectral signature identification
+4. Water Class Map (water-only GTM)
+5. Algae Identification
+7. Dye plume identification
 
@@
 
-1. GTM Applications in our Paper
+1. Overview of GTM applications in our paper
+   - Unsupervised classifciation of entire pond (no-dye)
+   - Endmember extraction of spectral signatures
 2. Comparison to existing approaches
-   - Unsupervised Classification
-   - Endmember Extraction & Spectral Unmixing
-   - Applications for either of these with Drones?
-3. Use of GTM and Unsupervised methods for Robot Team
-   - As a feature transformer/preprocessor for supervised methods (as opposed to, say, PCA or k-NN)
-   - identification of "interesting" regions for intelligent deployment of robotic boat
-     - don't continue collecting data in region that is similar
-     - provision boat to maximize data collection across distribution of GTM classes
+   - unsupervised classification methods (to general, no relationship between classes) 
+   - endmember extraction (mostly linear, assume existence of pure spectra)
+     - deep learning approaches to endmember extraction don't provide clear way to asses number of endmembers needed
+   - Applications of either of these with drones?
+     - Are there any papers using unsupervised methods or endmember extraction with drone based imagery (HSI or otherwise)?
+3. Use of GTM as component of Robot Team
+   - As a feature transformer / preprocessing step for supervised methods (as opposed to PCA, k-NN, etc...)
+   - identification of "interesting" regions for smart provisioning of robot boat
+     - Standard approach is to survey in a grid
+     - Can use classification map with real-time measurement to ensure optimal collection of data by collecting samples with from as many GTM classes as possible
 4. Extensions and future work
    - Incorporation of GTM extensions for efficient online-learning
-      - Currently GTM classes can quickly be applied (in real time) once they are trained to generate abundance maps e.g. chlorophyll, algae, etc...
-      - Key limitation of GTM is similar to SOM and k-NN: you need to compute distances between samples and latent nodes. This can scale poorly for very large datasets
-      - Bishop suggests a way to augment GTM training algorithm to enable online training of GTM.
-This could be deployed on-board the UAV to perform classification and identify spectral signatures in near-real time.
-      - Extensions of GTM [@gtm-bishop-2]
-    - Utilize the GTM with in-situ data collection to identify spectral signatures of specific algal species
+     - Comment on time to train standard GTM
+       - Key parameters are size of dataset (number of records) and number of RBF centers, $M$.
+     - Bishop suggest extension for on-line training of GTM using batches.
+     - All of this could be done in the field limiting the time and cost of surveying as well as optimizing the time-to-insights
+   - Utilize GTM with further in situ data collection to identify spectral signatures of specific algal species for HAB detection
+   - Same but for crude oil types...
 
 # Conclusions
-
-# NOTES AND CITATIONS
-
-1. Applications of remote sensing to water quality
-   - machine learning (XGBoost) used with paired remote sensing imagery and in-situ data to classify imagery into 5 categories including 3 for harmful algal blooms [@ghatkar2019classification]
-   - Spectral signatures of classes are complex and often overlapping [@thenkabail2018hyperspectral]
-   - Remote sensing used for oil spill analysis: extent and thickness mapping [@kokaly2013spectroscopic; @leifer2012state]
-   - Sun glitter remains a key challenge for sensing in the visible portion of the spectrum but multi and hyperspectral imagers have been used for oil spill identification and to identify their impacts on vegetation stress and mortality [@fingas2014review; @khan2018modern]
-
-2. Hyperspectral Imagery
-   - applications include food quality & safety, medical diagnoses, precision agriculture, and forensic document examination [@khan2018modern]
-   - Hyperspectral data were used to distinguish oils by type, e.g. crude, diesel, gasoline, and palm [@yang2020characterization]
-
-3. Supervised regression and classification for water quality
-   - common approach is inversion of optically-active water quality parameters such as chlorophyll-a, blue-green algae, turbidity, and temperature [@ritchie2003remote]
-   - Combining spectral indices such as the NDVI together with machine learning is a popular approach [@thenkabail2018hyperspectral; @sagan2020monitoring; @lu2021retrieval]
-   - Polynomial regression models for chlorophyll-a, turbidity [@zhang2022selection]
-   - Key limitation is collection of sufficient quantity of in-situ reference data
-   - Ross et al.
-created a comprehensive dataset with over 600,000 water quality records matching optically active water quality parameters with associated satellite imagery from Landsat 5,7,8.
-To achieve this quantity of data, they needed records spanning 1984 to 2019 [@ross2019aquasat]
-   - Aurin et al. took a similar route combining 30 years of remote sensing imagery with in-situ data from over 500 field compaigns for CDOM, organic carbon, etc... [@aurin2018remote].
-   - When ground truth data aren't available in sufficient quantity, classification into water quality categories is another approach (ground truth can be easier to obtain by expert analysis of scene) [@koponen2002lake]
-
-- When no ground-truth data are available, unsupervised classification can still help partition imagery into groups or clusters.
-   - Many data-driven, ML methods have been employed for the task including various matrix factorizations, k-nearest neighbors, fuzzy c-means, density estimation methods, etc.
-[@zhang2019hyperspectral]
-   - Additionally unsupervised approaches can be used to perform nonlinear dimensionality reduction & pre-processing for supervised approaches.
-   - SOM used for remote sensing imagery classification [@msom-remote-sensing]
-   - SOM used for clustering and data compression of HSI cube-sat [@danielsen2021self]
-   - SOM used for land-use and land-cover change analysis [@penfound2021analysis]
-
-4. Endmember Extraction and Spectral Unmixing
-   - This is a related problem where the goal is to identify unique spectral signatures which combine (linearly or non-linearly) to produce the measured signal.
-   - Having identified the set of endmembers, we then seek to determine their relative abundance in each pixel
-   - Many statistical approaches for extracting endmembers [@berman2004ice]
-   - Sparse PCA used for endmember extraction [@yousefi2016mineral]
-   - Popular spectral unmixing approach is to treat "pure" endmembers as vertices of a simplex [@plaza2012endmember; @nascimento2005vertex]
-   - Convolutional Neural Networks with Autoencoder architectures are a popular ML approach which identify endmembers and perform non-linear unmixing [@palsson2020convolutional; @non-negative-autoencoders; @su2019daen; @borsoi2019deep]
-   - Self Organizing Map is another approach which has been used for endmember extraction together with neural networks for abundance mapping (unmixing) [@cantero2004analysis]
-   - SOM has bee nused for identifying synoptic-scale patterns in wind and sea surface temperature data [@som-satellite]
-   - SOM used for HSI feature extraction [@som-hsi]
-   - drawback of autoencoder and other statistical approach is lack of a (topological) relationship between the classes/endmembers.
-Is endmember 1 closer to endmember 2 or endmember 10?
-The SOM addresses this.
-The GTM is even better...
-
-5. Abundance Mapping
-   - Many different spectral similarity functions exist for comparing spectra.
-Each have trade-offs between ability to distinguish differences in intensity versus differences in peak location (hue) [@deborah2015comprehensive]
-   - spectral angle mapper is popular similarity function used in endmember extraction. The SAM is nice because it is more sensitive to shape than scale [@jiang2020classification]
-   - spectral correlation mapper introduced to as a statistical alternative to SAM based on covariance instead of spectral angle [@de2000spectral]
-   - Normalized spectral similarity score developed by Nidamanuri et al. combine the mean-squared error with spectral angle for a happy medium [@nidamanuri2010normalized]
-
-6. Drone-based HSI
-   - Near-earth HSI addresses spatial, spectral, and temporal limitations of satellite and airborne platforms. UAV-based HSI enable fine-scale mapping [@banerjee2020uav]
-   - Drones can be equipped with HSI and compute to enable rapid generation of spectral indices like the NDVI for applications such as precision agriculture [@horstrand2019uav]
-   - UAV-based HSI can be georectified to centimeter-scales without need for ground control points by using on-board GPS and IMU [@arroyo2019implementation]
-   - UAV-based HSI for turbidity estimation [@vogt2016near]

### 2024-04-23 paper/outline.tex
updating intro & abstract

\documentclass{article}
\usepackage{graphicx} % Required for inserting images
\usepackage[margin=0.75in]{geometry}
\usepackage{outlines}

\begin{document}

\section*{Outline}
\begin{outline}[enumerate]
\1 Introduction
    \2 Characterization of inland water quality is hard
        \3 Key limitations:
            \4 Spectral resolution (visible and multi-spectral imaging)
            \4 Spatial Resolution (large pixel sizes)
            \4 Temporal Resolution (long satellite revisit times)
            \4 Availability of in situ reference data
        \3 When appropriate data are available, supervised ML approaches can map reflectance from remote sensing imagery to relevant water quality parameters
    \2 Hyperspectral Imaging addresses spectral resolution
    \2 UAV adresses spatial and temporal resolution
    \2 Robot team addresses in situ data... but what do we do if we don't already know which sensors to use? 
    \2 Unsupervised ML, Endmember extraction, and Spectral Unmixing
        \3 k-NN, fuzzy c-means, matrix factorizations, etc.
            \4 Are general but do not provide relationship between classes e.g. is class 1 more similar to class 2 or class 10?
        \3 SOM addresses this by enforcing a topological relationship between clusters. Classes close to each other in the map are more similar than classes far apart
        \3 Similarly, we may also be interested in the spectral characteristics of the learned *classes*, i.e. we want to think of the classes as *endmembers*.
            \4 Traditional approaches rely on assumption of existence of "pure pixels" in the scene
            \4 additionally, we may have to treat number of endmembers as a hyperparameter

    \2. ...in this paper, we explore how the generative topographic mapping can be used to
        \3 enable principled unsupervised classification of hyperspectral imagery by
            \4 Bayesian treatment (as compared to the self organizing map)
            \4 topological ordering with smooth relationship between classes in latent space
        \3 Train GTM on water pixels to obtain unsupervised classification mapping the detailed structure and small-scale variability of pond in Montague, North Texas
        \3 Mapping can be further utilized for smart deployment of robot team (maybe save for discussion section?) i.e. requisition boat to bring reference sensors to "interesting" area
        \3 Using combined dataset including land, water, and tracer dye, we demonstrate endmember extraction
        \3 Using learned spectra together with NS3, we are able to evaluate extent of algae (application to harmful algal bloom mapping)
        \3 Use learned spectra to map extent

\1 Materials and Methods
    \2 Data Collection
        \3 Describe collection site in Montague, North Texas
    \2 Autonomous Robot Team
        \3 Describe drone setup
            \4 Alta X Quadcopter
            \4 Resonon Pika SC2 HSI
                \5 462 wavelengths per pixel
               \5 1600 pixels per scan-line
               \5 embedded GPS/IMU
             \4 Intel NUC for on-board compute
    \2 Pre-processing
    \2 Generative Topographic Mapping
        \3 Start off with Kohonen SOM paper. Key benefit:
            \4 Topological relationship between classes i.e. similarity vs dissimilarity as a function of distance in mapping space (latent space)
        \3 Limitations of SOM
            \4 Not probabilistic
            \4 No clear convergence criterion
            \4 Heuristic based approach with no clear way to select hyperparameters like neighborhood function, number of classes, etc...
            \4 Heskes did reinterpret algorithm as stochastic gradient descent on a cost function...
        \3 GTM Algorithm
            \4 Describe Algorithm
            \4 Describe parameters
        \3 Bayesian information criterion and Akaike information criterion for hyperaprameter optimization
            \4 As an alternative to performing full Bayesian treatment for marginalization over hyperparameters
        \3 Abundance mapping with spectral similarity score
            \4 paper comparing different distance functions for spectra
                \5 curse of dimensionality can be pernicious here as we have > 462 wavelengths per pixel
            \4 NS3 offers good trade off between spectral angle and euclidean-like distance (MSE)
        \3 Describe use of thresholded NS3 for mapping abundance of GTM derived endmembers.

\1 Results
    \2  GTM fit on Full Dataset
    \2 Hyperparameter optimization
    \2 Spectral signature identification
    \2 Water Class Map (water-only GTM)
    \2 Algae Identification
    \2 Dye plume identification

\1 Discussion
    \2 Overview of GTM applications in our paper
        \3 Unsupervised classifciation of entire pond (no-dye)
        \3 Endmember extraction of spectral signatures
    \2 Comparison to existing approaches
        \3 unsupervised classification methods (to general, no relationship between classes) 
        \3 endmember extraction (mostly linear, assume existence of pure spectra)
        \3 deep learning approaches to endmember extraction don't provide clear way to asses number of endmembers needed
    \2 Applications of either of these with drones?
        \3 Are there any papers using unsupervised methods or endmember extraction with drone based imagery (HSI or otherwise)?
    \2. Use of GTM as component of Robot Team
        \3 As a feature transformer / preprocessing step for supervised methods (as opposed to PCA, k-NN, etc...)
        \3 identification of "interesting" regions for smart provisioning of robot boat
            \4 Standard approach is to survey in a grid
            \4 Can use classification map with real-time measurement to ensure optimal collection of data by collecting samples with from as many GTM classes as possible
    \2 Extensions and future work
        \3 Incorporation of GTM extensions for efficient online-learning
            \4 Comment on time to train standard GTM
                \5 Key parameters are size of dataset (number of records) and number of RBF centers, $M$.
            \4 Bishop suggest extension for on-line training of GTM using batches.
            \4 All of this could be done in the field limiting the time and cost of surveying as well as optimizing the time-to-insights
        \3 Utilize GTM with further in situ data collection to identify spectral signatures of specific algal species for HAB detection
        \3 Same but for crude oil types...

\1 Conclusions

\end{outline}

\bibliographystyle{plain}
\bibliography{paper/references.bib}

\end{document}

### 2024-04-23 paper/paper.tex
updating intro & abstract

@@
 % Abstract (Do not insert blank lines, i.e. \\) 
-\abstract{Advancements in hyperspectral imaging technology have lead to a continual decrease in their size facilitating their inclusion as the payload of small, highly maneuverable unmanned aerial vehicles (UAVs) such as drones.
-Consequently, much work has gone into the development of UAV platforms and associated techniques
+\abstract{
+Unmanned Aerial Vehicles (UAVs) equipped with hyperspectral imagers have emerged as a pivotal technology for the characterization of inland water bodies.
+The high spectral and spatial resolutions of these systems enable the retrieval of a plethora of optically-active water quality parameters via band ratio algorithms and machine learning methods.
+However, fitting and validating these models requires access to sufficient quantities of in situ reference data which are time-consuming and expensive to obtain.
+In this study, we demonstrate how the Generative Topographic Mapping (GTM), a Bayesian realisation of the Self Organizing Map (SOM), can be effectively used to enable unsupervised classification and nonlinear endmember extraction of UAV acquired hyperspectral imagery.
+
+to enable fully unsupervised classification UAV acquired hyperspectral imagery.
+
+The methods is fully Bayesian, providing an attractive method for the unsupervised classification of hyperspectral imagery.
+
 - generative topographic mapping (GTM) as an efficient unsupervised method 
@@
 
-Inland water bodies pose a unique challenge for analysis with remote sensing imagery due to their complex spectral characteristics and small-scale spatial variability; nevertheless, significant effort has driven the development of techniques for estimating optically-active water quality parameters such as chlorophyll-a, blue-green algae, colored dissolved organic matter (CDOM), turbidity, and temperature \cite{koponen2002lake,ritchie2003remote, bonansea2015using}.
-Machine learning has proven to be a powerful tool in this domain, with tree-based models, support vector machines, deep neural networks, and others providing data-driven approaches to map imagery to parameters of interest \cite{thenkabail2018hyperspectral, ghatkar2019classification,sagan2020monitoring}.
+Inland water bodies pose a unique challenge for analysis with remote sensing imagery due to their complex spectral characteristics and small-scale spatial variability.
+Nevertheless, significant effort has driven the development of techniques for estimating optically-active water quality parameters such as chlorophyll-a, blue-green algae, colored dissolved organic matter (CDOM), turbidity, and temperature from these sources\cite{koponen2002lake,ritchie2003remote, bonansea2015using}.
+Machine learning has proven to be a powerful tool in this domain, with tree-based models, support vector machines, deep neural networks, and others architectures providing data-driven approaches mapping imagery to parameters of interest \cite{thenkabail2018hyperspectral, ghatkar2019classification,sagan2020monitoring}.
 However, the training and evaluation of these models demands significant volumes of coincident, in situ data.
-Collecting these datasets is both time consuming and expensive often requiring observations spanning decades \cite{aurin2018remote, ross2019aquasat}.
+Collecting these datasets is time consuming and expensive, often requiring observations spanning decades \cite{aurin2018remote, ross2019aquasat}.
+
+To support and enhance real-time decision making, data-driven techniques...
+
+The characterization of inland water bodies by remote sensing imagery poses a distinct challenge due to their complex spectral characteristics and small-scale spatial variability \cite{ritchie2003remote}.
+Traditional multi-spectral imagers utilized by these sensing platforms have large pixel sizes, limited spectral bands, and long revisit times.
+- Nevertheless, ... 
+- Machine learning has proven... 
+- However, these techniques demand significant volumes of in situ data...
+- To address the
+
+Hyperspectral imaging
+- Hyperspectral imaging has emerged as a pivotal technology in this space.
+- With their increased spectral resolution, hyperspectral imagers (HSIs) capture the complex chemical, biological, environmental characteristics of reflectance spectra.
+(add references for specific applications of HSIs).
+- However HSIs alone do not address the limited spatial resolution of
+
+UAV
+- Advancements in hyperspectral imaging technology have lead to a continual decrease in their size facilitating their inclusion as the payload of small, highly maneuverable unmanned aerial vehicles (UAVs) such as drones.
+Consequently, much work has gone into the development of UAV platforms and associated techniques
+- Near-earth HSI addresses spatial, spectral, and temporal limitations of satellite and airborne platforms. UAV-based HSI enable fine-scale mapping \cite{banerjee2020uav}
+- Drones can be equipped with HSI and compute to enable rapid generation of spectral indices like the NDVI for applications such as precision agriculture \cite{horstrand2019uav}
+- UAV-based HSI can be georectified to centimeter-scales without need for ground control points by using on-board GPS and IMU \cite{arroyo2019implementation}
+- UAV-based HSI for turbidity estimation \cite{vogt2016near}
+- Polynomial regression models for chlorophyll-a, turbidity \cite{zhang2022selection}
+- For this to be usefull, we still need the in situ data.
+By combining a UAV born HSI with simultanous collection of ground truth data by a reference sensor equipped autonomous boat, we are able to rapidly train ML models to
+
+Unsupervised Methods
+- classic unsupervised ML
+- endmember extraction and spectral unmixing
+
+This paper...
+- our approach is to use the GTM (a fully Bayesian realization of the self organizing map developed by Biship et al.)
+- We demonstrate the capabilities using hyperspectral imagery collected at pond in Montague, North Texas. 
+- First we utilize the GTM to perform an unsupervised classification of the water-only pixels in order to reveal the centimeter-scale structures present in the pond
+- Next we showhow the GTM can be used to identify relevent spectral endmembers by 
+    1. using learned GTM class signatures to identity the abundance of algae (relevant for Harmful Algal Bloom detection)
+    2. use the GTM class signatures to map the evolution of a rhodamine tracer dye plume
 
@@
 
-- supervised classification of water quality with XGBoost \cite{ghatkar2019classification}
+Continued development of HSI technology has led to a dramatic reduction in their size enabling their utilize as the payload of small autonomous aerial vehicles (UAVs) (citation).
+By flying close to the ground UAV-born hyperspectral imagers can produce centimeter-scale imagery while simulatenously limiting the need for complicated atmospheric corrections required for satellite platforms (citation).
+
+- applications include food quality & safety, medical diagnoses, precision agriculture, and forensic document examination \cite{khan2018modern}
+- Hyperspectral data were used to distinguish oils by type, e.g. crude, diesel, gasoline, and palm \cite{yang2020characterization}
 
@@
 
-Hyperspectral imaging has emerged as a pivotal technology in this space.
-With their increased spectral resolution, hyperspectral imagers (HSIs) capture the complex chemical, biological, environmental characteristics of reflectance spectra.
-(add references for specific applications of HSIs).
-Continued development of HSI technology has led to a dramatic reduction in their size enabling their utilize as the payload of small autonomous aerial vehicles (UAVs) (citation).
-By flying close to the ground UAV-born hyperspectral imagers can produce centimeter-scale imagery while simulatenously limiting the need for complicated atmospheric corrections required for satellite platforms (citation).
-
-- applications include food quality & safety, medical diagnoses, precision agriculture, and forensic document examination \cite{khan2018modern}
-- Hyperspectral data were used to distinguish oils by type, e.g. crude, diesel, gasoline, and palm \cite{yang2020characterization}
-- Near-earth HSI addresses spatial, spectral, and temporal limitations of satellite and airborne platforms. UAV-based HSI enable fine-scale mapping \cite{banerjee2020uav}
-- Drones can be equipped with HSI and compute to enable rapid generation of spectral indices like the NDVI for applications such as precision agriculture \cite{horstrand2019uav}
-- UAV-based HSI can be georectified to centimeter-scales without need for ground control points by using on-board GPS and IMU \cite{arroyo2019implementation}
-- UAV-based HSI for turbidity estimation \cite{vogt2016near}
-- Polynomial regression models for chlorophyll-a, turbidity \cite{zhang2022selection}
-
 In our previous work, we introduced a robot team designed to rapidly characterize water composition in novel environments \cite{robot-team-1}. 
@@
 In this paper, we demonstrate how Generative Topographic Mapping (GTM), a Bayesian alternative to the Self Organizing Map, can be utilized to rapidly identify relevant spectral signatures within a scene.
+
 Using data collected at site in Montague, North Texas, we demonstrate the ability of this paradigm with two case studies.
@@
 Together with the normalized spectral similarity score, these spectral signatures are identified and their associated area in the scene is determined.
+
+In this study, we demonstrate how Generative Topographic Mapping (GTM), a Bayesian extension the Self Organizing Map, can be utilized to
 
@@
 \begin{tabular}{@{}ll}
-CDOM & Colored Dissolved Organic Matter \\
-HAB & Harmful Algal Bloom \\ 
-HSI & Hyperspectral Image \\ 
+UAV & Unmanned Aerial Vehicle \\
 SOM & Self Organizing Map \\ 
 GTM & Generative Topographic Mapping 
+HSI & Hyperspectral Image \\ 
+CDOM & Colored Dissolved Organic Matter \\
+HAB & Harmful Algal Bloom 
 \end{tabular}

### 2024-04-23 paper/paper.tex
Updates from Overleaf

@@
 In this study, we demonstrate how the Generative Topographic Mapping (GTM), a Bayesian realisation of the Self Organizing Map (SOM), can be effectively used to enable unsupervised classification and nonlinear endmember extraction of UAV acquired hyperspectral imagery.
-
-to enable fully unsupervised classification UAV acquired hyperspectral imagery.
-
-The methods is fully Bayesian, providing an attractive method for the unsupervised classification of hyperspectral imagery.
-
+Using imagery collected across a North Texas Pond, we demonstrate how th
+- to enable fully unsupervised classification UAV acquired hyperspectral imagery.
+-The methods is fully Bayesian, providing an attractive method for the unsupervised classification of hyperspectral imagery. 
 - generative topographic mapping (GTM) as an efficient unsupervised method 

### 2024-04-23 paper/draft-introduction.tex
separated introduction into draft, finished draft abstract (needs work), and added some points to the outline.

--- 2024-04-23 paper/paper.tex
@@
-%  LaTeX support: latex@mdpi.com
-%  For support, please attach all files needed for compiling as well as the log file, and specify your operating system, LaTeX version, and LaTeX editor.
+\documentclass{article}
+\usepackage{graphicx} % Required for inserting images
+\usepackage[margin=0.75in]{geometry}
 
-%=================================================================
-\documentclass[remotesensing,article,submit,pdftex,moreauthors]{Definitions/mdpi}
-
-%=================================================================
-% MDPI internal commands - do not modify
-\firstpage{1}
-\makeatletter
-\setcounter{page}{\@firstpage}
-\makeatother
-\pubvolume{1}
-\issuenum{1}
-\articlenumber{0}
-\pubyear{2024}
-\copyrightyear{2024}
-%\externaleditor{Academic Editor: Firstname Lastname}
-\datereceived{ }
-\daterevised{ } % Comment out if no revised date
-\dateaccepted{ }
-\datepublished{ }
-%\datecorrected{} % For corrected papers: "Corrected: XXX" date in the original paper.
-%\dateretracted{} % For corrected papers: "Retracted: XXX" date in the original paper.
-\hreflink{https://doi.org/} % If needed use \linebreak
-%\doinum{}
-%\pdfoutput=1 % Uncommented for upload to arXiv.org
-%\CorrStatement{yes}  % For updates
-
-%=================================================================
-% Add packages and commands here.
-The following packages are loaded in our class file: fontenc, inputenc, calc, indentfirst, fancyhdr, graphicx, epstopdf, lastpage, ifthen, float, amsmath, amssymb, lineno, setspace, enumitem, mathpazo, booktabs, titlesec, etoolbox, tabto, xcolor, colortbl, soul, multirow, microtype, tikz, totcount, changepage, attrib, upgreek, array, tabularx, pbox, ragged2e, tocloft, marginnote, marginfix, enotez, amsthm, natbib, hyperref, cleveref, scrextend, url, geometry, newfloat, caption, draftwatermark, seqsplit
-% cleveref: load \crefname definitions after \begin{document}
-
-%=================================================================
-% Please use the following mathematics environments: Theorem, Lemma, Corollary, Proposition, Characterization, Property, Problem, Example, ExamplesandDefinitions, Hypothesis, Remark, Definition, Notation, Assumption
-%% For proofs, please use the proof environment (the amsthm package is loaded by the MDPI class).
-
-%=================================================================
-% Full title of the paper (Capitalized)
-\Title{Robot Team III: Scotty Strikes Back}
-
-% MDPI internal command: Title for citation in the left column
-\TitleCitation{Title}
-
-% Author Orchid ID: enter ID or remove command
-\newcommand{\orcidauthorA}{0000-0002-5910-0183} % John
-\newcommand{\orcidauthorB}{0000-0003-2688-648X} % Lakitha
-\newcommand{\orcidauthorC}{0000-0002-9841-6703} % Shawhin
-\newcommand{\orcidauthorD}{0000-0003-2657-3416} % Prabuddha
-\newcommand{\orcidauthorE}{0000-0003-4265-9543} % David
-\newcommand{\orcidauthorF}{0000-0003-0667-2345} % Ash
-\newcommand{\orcidauthorG}{0000-0002-0126-218X} % Adam
-% \newcommand{\orcidauthorH}{} % Gokul
-
-% Authors, for the paper (add full first names)
-\Author{John Waczak \orcidA{}, Adam Aker \orcidG{}, Lakitha O.
-H.
-Wijeratne \orcidB{}, Shawhin Talebi \orcidC{}, Ashen Fernando \orcidF{}, Prabuddha M.
-H.
-Dewage \orcidD{}, Mazhar Iqbal, Matthew Lary, David Schaefer, Gokul Balagopal, and David J.
-Lary *\orcidE{}}
-
-%\longauthorlist{yes}
-
-% MDPI internal command: Authors, for metadata in PDF
-\AuthorNames{John Waczak, Adam Aker, Lakitha O. H. Wijeratne, Shawhin Talebi, Bharana Fernando, Prabuddha Hathurusinghe, Mazhar Iqbal, Matthew Lary, David Schaefer, Gokul Balagopal, and David J. Lary}
-
-% MDPI internal command: Authors, for citation in the left column
- \AuthorCitation{Waczak, J.; Aker, A.; Wijeratne, L.O.H.; Talebi, S.; Fernando, B.; Hathurusinghe, P.; Iqbal, M.; Lary, M.; Schaefer, Balagopal, G.; D.; Lary, D.J.}
-% If this is a Chicago style journal: Lastname, Firstname, Firstname Lastname, and Firstname Lastname.
-
-% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
-\address[1]{%
-Hanson Center for Space Sciences, University of Texas at Dallas, Richardson, TX 75080, USA;  john.waczak@utdallas.edu (J.W.); adam.aker@utdallas.edu (A.A.); lhw150030@utdallas.edu (L.O.H.W.); shawhintalebi@gmail.com (S.T.); ashen.fernando@utdallas.edu  (B.F.); pxh180012@utdallas.edu (P.M.H.D.); mazhar.iqbal@utdallas.edu (M.I.); MDL210001@utdallas.edu (M.L.); captdaveschaefer@gmail.com (D.S.); gokul.balagopal@utdallas.edu (G.B.);}
-
-% Contact information of the corresponding author
-\corres{Correspondence: david.lary@utdallas.edu} %; Tel.: (optional; include country code; if there are multiple corresponding authors, add author initials) +xx-xxxx-xxx-xxxx (F.L.)}
-
-% Abstract (Do not insert blank lines, i.e. \\) 
-\abstract{
-Unmanned Aerial Vehicles (UAVs) equipped with hyperspectral imagers have emerged as a pivotal technology for the characterization of inland water bodies.
-The high spectral and spatial resolutions of these systems enable the retrieval of a plethora of optically-active water quality parameters via band ratio algorithms and machine learning methods.
-However, fitting and validating these models requires access to sufficient quantities of in situ reference data which are time-consuming and expensive to obtain.
-In this study, we demonstrate how the Generative Topographic Mapping (GTM), a Bayesian realisation of the Self Organizing Map (SOM), can be effectively used to enable unsupervised classification and nonlinear endmember extraction of UAV acquired hyperspectral imagery.
-
-to enable fully unsupervised classification UAV acquired hyperspectral imagery.
-
-The methods is fully Bayesian, providing an attractive method for the unsupervised classification of hyperspectral imagery.
-
-- generative topographic mapping (GTM) as an efficient unsupervised method 
-- additionally the GTM can be used to perform nonlinear endmember extraction
-- We demonstrate the abilities of this approach using hyperspectral imagery collected at a pond in Montague, North Texas.
-- First, we apply the GTM to perform an unsupervised classification for all water pixels. 
-    - benefit: GTM classes maintain topological relationships in the latent space resulting in a smooth relationship between the obtained classes...
-    - ... revealing the small-scale spatial variability within the pond
-- Second, we utilize the GTM to identify reflectance spectra corresponding to specific features. 
-- Using these extracted features, we are able to map the extent of algae in the pond in addition to the time evolution of a rhodamine dye tracer. 
-- This approach demonstrates the value of unsupervised techniques coupled with UAV-born hyperspectral imaging for rapid water quality assessment.
-}
-
-% Keywords
-\keyword{Hyperspectral Imaging; Remote Sensing; Unsupervised Classification; Endmember Extraction, Spectral Unmixing}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \begin{document}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Introduction}
@@
 Continued development of HSI technology has led to a dramatic reduction in their size enabling their utilize as the payload of small autonomous aerial vehicles (UAVs) (citation).
-By flying close to the ground UAV-born hyperspectral imagers can produce centimeter-scale imagery while simulatenously limiting the need for complicated atmospheric corrections required for satellite platforms (citation).
+By flying close to the ground UAV-born hyperspectral imagers can produce centimeter-scale imagery while simultaneously limiting the need for complicated atmospheric corrections required for satellite platforms (citation).
 
@@
 - SOM used for land-use and land-cover change analysis \cite{penfound2021analysis}
-- SOM has bee nused for identifying synoptic-scale patterns in wind and sea surface temperature data {som-satellite}
+- SOM has been used for identifying synoptic-scale patterns in wind and sea surface temperature data {som-satellite}
 
@@
 
-By carefully coordinating the simultanous collection of ground-truth water quality data and hyperspectral imagery, this team can rapidly estimate concentrations of optically-active components including colored dissolved organic matter (CDOM), blue-green algae, crude oil, temperature as well
+By carefully coordinating the simultaneous collection of ground-truth water quality data and hyperspectral imagery, this team can rapidly estimate concentrations of optically-active components including colored dissolved organic matter (CDOM), blue-green algae, crude oil, temperature as well
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Materials and Methods}
+The method is fully Bayesian, providing an attractive method for the unsupervised classification of hyperspectral imagery.
 
-\begin{figure}[t]
-%\begin{adjustwidth}{-\extralength}{0cm}
-\centering
-%\includegraphics[width=15.5cm]{paper/figures/methods/sample-spectra.pdf}
-\includegraphics[width=\columnwidth]{paper/figures/methods/sample-spectra.pdf}
-%\end{adjustwidth}
-\caption{(a) Sample hyperspectral data cube.
-Spectra are plotted using their geographic position with the log10-reflectance colored along the z axis and a pseudo-color image on top.
-(b) Points taken from a sample hyperspectral data cube corresponding to algae, rhodamine dye, water, and dry grass.
-(c) The associated spectra for the exemplar points scaled so that the peak value of each spectrum is 1.
-(d) The location of the pond in Montague, Texas where data were collected for this study.\label{fig:sample-spectra}}
-\end{figure}
+- First, we apply the GTM to perform an unsupervised classification for all water pixels. 
+    - benefit: GTM classes maintain topological relationships in the latent space resulting in a smooth relationship between the obtained classes...
+    - ... revealing the small-scale spatial variability within the pond
+- Second, we utilize the GTM to identify reflectance spectra corresponding to specific features. 
+- Using these extracted features, we are able to map the extent of algae in the pond in addition to the time evolution of a rhodamine dye tracer.
 
-\subsection{Autonomous Robot Team}
-\subsection{Data Collection and Pre-Processing}
-
-- georetification \cite{muller2002program, baumker2001new, mostafa2000multi}
-
-- Julia programming language used \cite{bezanson2012julia}
-
-\subsection{Generative Topographic Mapping}
-
-- GTM paper oricinal \cite{gtm-bishop-1}
-- GTM developments paper \cite{gtm-biship-2}
-
-- implementation designed to work in MLJ ecosystem \cite{blaom2020mlj}
-
-\subsection{Image Segmentation with the Normalized Spectral Similarity Score}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Results}
-
-\subsection{GTM Fit}
-
-\begin{figure}[t]
-\centering
-\includegraphics[width=0.8\columnwidth]{paper/figures/results/square-ndvi.pdf}
-\caption{GTM Map: We plot the mean position of each data point in the latent space with points colored by the their associated NDVI values.
-The locations of the exemplar points for algae, rhodamine dye, water, and dry grass are shown in the latent space illustrating that the GTM has clearly distinguished land, near-land, and water pixels.\label{fig:}}
-\end{figure}
-
-\subsection{Hyperparamter Optimization with BIC}
-
-\begin{figure}[t]
-%\begin{adjustwidth}{-\extralength}{0cm}
-\centering
-\includegraphics[width=\textwidth]{paper/figures/results/bic.pdf}
-%\end{adjustwidth}
-\caption{Results of the hyperparameter search.
-Left: Variation of BIC with $m$ and $s$ for fixed $\alpha=0.1$.
-Right: Variation of the BIC with with $m$ and $\alpha$ for fixed $s=1.0$.
-The white star in each plot indicates the parameters with the lowest BIC.\label{fig:hp-results}}
-\end{figure}
-
-\subsection{Spectral Signatures}
-
-\begin{figure}[t]
-\begin{adjustwidth}{-\extralength}{0cm}
-\centering
-\includegraphics[width=15.5cm]{paper/figures/results/spectral-signatures.png}
-\end{adjustwidth}
-\caption{GTM Node Signatures corresponding to the exemplar spectra for the rhodamine plume (top left), algae (top right), dry grass (bottom left), and open water (bottom right).
-\label{fig:spectral-signautes}}
-\end{figure}
-
-\subsection{Water Class Map}
-
-\begin{figure}[t]
-\begin{adjustwidth}{-\extralength}{0cm}
-\centering
-\includegraphics[width=18.0cm]{paper/figures/results/gtm-water.png}
-\end{adjustwidth}
-\caption{Classification map for GTM trained solely on water pixels (no land and no rhodamine plume).
-\textbf{(a)} GTM applied to all water pixels colored by their associated position in the latent space.
-\textbf{(b)} Representation of the poitns from the training set in the latent space.
-The points are colored with $\xi_1$ corresponding to red and $\xi_2$ corresponding to blue.
-\textbf{(c)} The associated GTM spectral signatures, $\Psi_i$, corresponding to the four corners and center of the GTM latent space.
-\label{fig:gtm-water}}
-\end{figure}
-
-\subsection{Algae Identification}
-
-\begin{figure}[t]
-\begin{adjustwidth}{-\extralength}{0cm}
-\centering
-\includegraphics[width=15.5cm]{paper/figures/results/algae.png}
-\end{adjustwidth}
-\caption{\label{fig:algae-map}}
-\end{figure}
-
-\subsection{Rhodamine Dye Plume Identification}
-
-\begin{figure}[t]
-\begin{adjustwidth}{-\extralength}{0cm}
-\centering
-\includegraphics[width=15.5cm]{paper/figures/results/rhodamine.png}
-\end{adjustwidth}
-\caption{\label{fig:rhodamine-map}}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Discussion}
-
-\subsection{Summary of GTM applications in our paper}
-\subsection{Comparison to other existing approaches in literature}
-- Is there much done for classification in real time? 
-\subsection{Comparison to Spectral Unmixing}
-- Here we can re-interpret the GTM model as performing a type of Bayesian spectral unmixing and suggest how the method could be further extended to work in this domain.
-Critically, other traditional methods work by observing that pure endmembers form the vertices of a simplex.
-
-\subsection{Other applications of GTM}
-- Discuss use of GTM means/modes/responsibilities feature transformer for supervised models (e.g. better than PCA) 
-- Use of GTM classes to aid autonomous data collection by identification of "interesting" regons
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Conclusions}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\vspace{6pt}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% optional
-%\supplementary{The following supporting information can be downloaded at:  \linksupplementary{s1}, Figure S1: title; Table S1: title; Video S1: title.}
-
-% Only for journal Methods and Protocols:
-% If you wish to submit a video article, please do so with any other supplementary material.
-% \supplementary{The following supporting information can be downloaded at: \linksupplementary{s1}, Figure S1: title; Table S1: title; Video S1: title.
-A supporting video article is available at doi: link.}
-
-% Only for journal Hardware:
-% If you wish to submit a video article, please do so with any other supplementary material.
-% \supplementary{The following supporting information can be downloaded at: \linksupplementary{s1}, Figure S1: title; Table S1: title; Video S1: title.\vspace{6pt}\\
-%\begin{tabularx}{\textwidth}{lll}
-%\toprule
-%\textbf{Name} & \textbf{Type} & \textbf{Description} \\
-%\midrule
-%S1 & Python script (.py) & Script of python source code used in XX \\
-%S2 & Text (.txt) & Script of modelling code used to make Figure X \\
-%S3 & Text (.txt) & Raw data from experiment X \\
-%S4 & Video (.mp4) & Video demonstrating the hardware in use \\
-%... & ... & ... \\
-%\bottomrule
-%\end{tabularx}
-%}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\authorcontributions{FIXME}
-
-\funding{FIXME}
-
-\institutionalreview{Not applicable}
-
-\informedconsent{Not applicable}
-
-\dataavailability{We encourage all authors of articles published in MDPI journals to share their research data.
-In this section, please provide details regarding where data supporting reported results can be found, including links to publicly archived datasets analyzed or generated during the study.
-Where no new data were created, or where data is unavailable due to privacy or ethical restrictions, a statement is still required.
-Suggested Data Availability Statements are available in section ``MDPI Research Data Policies'' at \url{https://www.mdpi.com/ethics}.}
-
-\acknowledgments{In this section you can acknowledge any support given which is not covered by the author contribution or funding sections.
-This may include administrative and technical support, or donations in kind (e.g., materials used for experiments).}
-
-\conflictsofinterest{Declare conflicts of interest or state ``The authors declare no conflicts of interest.'' Authors must identify and declare any personal circumstances or interest that may be perceived as inappropriately influencing the representation or interpretation of reported research results.
-Any role of the funders in the design of the study; in the collection, analyses or interpretation of data; in the writing of the manuscript; or in the decision to publish the results must be declared in this section.
-If there is no role, please state ``The funders had no role in the design of the study; in the collection, analyses, or interpretation of data; in the writing of the manuscript; or in the decision to publish the results''.}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Optional
-
-%% Only for journal Encyclopedia
-%\entrylink{The Link to this entry published on the encyclopedia platform.}
-
-\abbreviations{Abbreviations}{
-The following abbreviations are used in this manuscript:\\
-
-\noindent 
-\begin{tabular}{@{}ll}
-UAV & Unmanned Aerial Vehicle \\
-SOM & Self Organizing Map \\ 
-GTM & Generative Topographic Mapping 
-HSI & Hyperspectral Image \\ 
-CDOM & Colored Dissolved Organic Matter \\
-HAB & Harmful Algal Bloom 
-\end{tabular}
-}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Optional
-\appendixtitles{yes} % Leave argument "no" if all appendix headings stay EMPTY (then no dot is printed after "Appendix A").
-If the appendix sections contain a heading then change the argument to "yes".
-\appendixstart
-\appendix
-\section[\appendixname~\thesection]{Hyperparameter Search Results}
-
-\begin{table}[H]
-  \caption{The top 25 models from the hyperparameter search. A variety of GTM were trained to explore the the impact of varying m, $\alpha$, and s.\label{tab:hp-search}}
-  \begin{adjustwidth}{-\extralength}{0cm}
-  \newcolumntype{C}{>{\centering\arraybackslash}X}
-  \begin{tabularx}{\fulllength}{CCCCCC}
-    \toprule
-    \textbf{m} & \textbf{$\alpha$} & \textbf{s} & \textbf{k} & \textbf{BIC} & \textbf{AIC} \\
-    \midrule
-    14 & 0.1 & 1.0 & 32 & \textbf{-1.918e8} & -1.926e8\\
-    13 & 0.01 & 1.0 & 32 & -1.917e8 & -1.923e8\\
-    16 & 0.01 & 1.5 & 32 & -1.917e8 & -1.926e8\\
-    14 & 10.0 & 1.0 & 32 & -1.917e8 & -1.924e8\\
-    16 & 0.001 & 1.5 & 32 & -1.917e8 & -1.926e8\\
-    13 & 1.0 & 1.0 & 32 & -1.917e8 & -1.923e8\\
-    13 & 10.0 & 1.0 & 32 & -1.917e8 & -1.923e8\\
-    14 & 0.001 & 1.5 & 32 & -1.916e8 & -1.924e8\\
-    13 & 0.1 & 1.0 & 32 & -1.916e8 & -1.923e8\\
-    14 & 0.01 & 1.0 & 32 & -1.916e8 & -1.924e8\\
-    15 & 0.01 & 1.5 & 32 & -1.916e8 & -1.925e8\\
-    14 & 0.01 & 1.5 & 32 & -1.916e8 & -1.923e8\\
-    15 & 1.0 & 1.0 & 32 & -1.916e8 & -1.924e8\\
-    18 & 0.01 & 1.5 & 32 & -1.916e8 & -1.928e8\\
-    12 & 0.01 & 1.0 & 32 & -1.916e8 & -1.921e8\\
-    15 & 0.01 & 0.5 & 32 & -1.915e8 & -1.924e8\\
-    17 & 1.0 & 1.0 & 32 & -1.915e8 & -1.926e8\\
-    16 & 0.1 & 1.0 & 32 & -1.915e8 & -1.925e8\\
-    18 & 0.001 & 1.5 & 32 & -1.915e8 & -1.928e8\\
-    13 & 0.001 & 1.0 & 32 & -1.915e8 & -1.922e8\\
-    12 & 1.0 & 1.0 & 32 & -1.915e8 & -1.921e8\\
-    17 & 0.001 & 1.5 & 32 & -1.915e8 & -1.926e8\\
-    15 & 0.001 & 1.5 & 32 & -1.915e8 & -1.923e8\\
-    15 & 10.0 & 1.0 & 32 & -1.915e8 & -1.923e8\\
-    12 & 0.1 & 1.5 & 32 & -1.915e8 & -1.92e8\\
-    \bottomrule
-  \end{tabularx}
-  \end{adjustwidth}
-\end{table}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{adjustwidth}{-\extralength}{0cm}
-%\printendnotes[custom] % Un-comment to print a list of endnotes
-
-\reftitle{References}
-
-%=====================================
-% References, variant A: external bibliography
-%=====================================
+\bibliographystyle{plain}
 \bibliography{paper/references.bib}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\PublishersNote{}
-\end{adjustwidth}
 \end{document}

### 2024-04-23 paper/paper.tex
separated introduction into draft, finished draft abstract (needs work), and added some points to the outline.

--- 2024-04-22 paper/paper.tex
@@
 % Abstract (Do not insert blank lines, i.e. \\) 
-\abstract{Advancements in hyperspectral imaging technology have lead to a continual decrease in their size facilitating their inclusion as the payload of small, highly maneuverable unmanned aerial vehicles (UAVs) such as drones.
-Consequently, much work has gone into the development of UAV platforms and associated techniques
-- generative topographic mapping (GTM) as an efficient unsupervised method 
-- additionally the GTM can be used to perform nonlinear endmember extraction
-- We demonstrate the abilities of this approach using hyperspectral imagery collected at a pond in Montague, North Texas.
-- First, we apply the GTM to perform an unsupervised classification for all water pixels. 
-    - benefit: GTM classes maintain topological relationships in the latent space resulting in a smooth relationship between the obtained classes...
-    - ... revealing the small-scale spatial variability within the pond
-- Second, we utilize the GTM to identify reflectance spectra corresponding to specific features. 
-- Using these extracted features, we are able to map the extent of algae in the pond in addition to the time evolution of a rhodamine dye tracer. 
-- This approach demonstrates the value of unsupervised techniques coupled with UAV-born hyperspectral imaging for rapid water quality assessment.
+\abstract{
+Unmanned Aerial Vehicles (UAVs) equipped with hyperspectral imagers have emerged as an essential technology for the characterization of inland water bodies.
+The high spectral and spatial resolutions of these systems enable the retrieval of a plethora of optically-active water quality parameters via band ratio algorithms and machine learning methods.
+However, fitting and validating these models requires access to sufficient quantities of in situ reference data which are time-consuming and expensive to obtain.
+In this study, we demonstrate how the Generative Topographic Mapping (GTM), a Bayesian realisation of the Self Organizing Map (SOM), enables highly-detailed unsupervised classification and nonlinear endmember extraction of UAV acquired hyperspectral imagery.
+Specifically, the topographic relationship between classes maintained by the GTM presents an attractive alternative to other unsupervised approaches which do not provide a notion of similarity between learned classes.
+Using data collected across a North Texas pond, we apply the GTM to create a detailed map segmenting the area into distinct regions of interest.
+Further, by interpreting the learned classes as distinct endmembers, the GTM can be used to rapidly identify and map unique water constituents.
+We demonstrate this capability by using the GTM to map algal abundance as well as the dispersion of a rhodamine dye tracer released into the water.
 }
@@
 \section{Introduction}
-
-Inland water bodies pose a unique challenge for analysis with remote sensing imagery due to their complex spectral characteristics and small-scale spatial variability; nevertheless, significant effort has driven the development of techniques for estimating optically-active water quality parameters such as chlorophyll-a, blue-green algae, colored dissolved organic matter (CDOM), turbidity, and temperature \cite{koponen2002lake,ritchie2003remote, bonansea2015using}.
-Machine learning has proven to be a powerful tool in this domain, with tree-based models, support vector machines, deep neural networks, and others providing data-driven approaches to map imagery to parameters of interest \cite{thenkabail2018hyperspectral, ghatkar2019classification,sagan2020monitoring}.
-However, the training and evaluation of these models demands significant volumes of coincident, in situ data.
-Collecting these datasets is both time consuming and expensive often requiring observations spanning decades \cite{aurin2018remote, ross2019aquasat}.
-
-- \cite{koponen2002lake}
-- Example: distinguishing species for harmful algal blooms (HABs) and distinguishing types of oil for oil spill monitoring
-- Remote sensing used for oil spill analysis: extent and thickness mapping \cite{kokaly2013spectroscopic, leifer2012state}
-
-- Sun glitter remains a key challenge for sensing in the visible portion of the spectrum but multi and hyperspectral imagers have been used for oil spill identification and to identify their impacts on vegetation stress and mortality \cite{fingas2014review, khan2018modern}
-
-- supervised classification of water quality with XGBoost \cite{ghatkar2019classification}
-
-Unsupervised classification
-- When no ground-truth data are available, unsupervised classification can still help partition imagery into groups or clusters.
-- Many data-driven, ML methods have been employed for the task including various matrix factorizations, k-nearest neighbors, fuzzy c-means, density estimation methods, etc.
-\cite{zhang2019hyperspectral}
-- Additionally unsupervised approaches can be used to perform nonlinear dimensionality reduction & pre-processing for supervised approaches.
-- Many unsupervised approaches can be challenging to interpret due to the lack of a relationship between the identified classes, e.g. is class 1 more similar to class 2 than class 10?
-- The Self Organizing Map is a popular technique which classifies data while enforcing topological relationships between the resulting classes such that classes close to each  other in the latent space are more similar than classes that are far apart.
-- SOM used for remote sensing imagery classification \cite{msom-remote-sensing}
-- SOM used for clustering and data compression of HSI cube-sat \cite{danielsen2021self}
-- SOM used for land-use and land-cover change analysis \cite{penfound2021analysis}
-- SOM has bee nused for identifying synoptic-scale patterns in wind and sea surface temperature data {som-satellite}
-
-endmember extraction, spectral unmixing
-- This is a related problem where the goal is to identify unique spectral signatures which combine (linearly or non-linearly) to produce the measured signal.
-- Having identified the set of endmembers, we then seek to determine their relative abundance in each pixel
-- Many statistical approaches for extracting endmembers \cite{berman2004ice}
-- Sparse PCA used for endmember extraction \cite{yousefi2016mineral}
-- Popular spectral unmixing approach is to treat "pure" endmembers as vertices of a simplex \cite{plaza2012endmember, nascimento2005vertex}
-- Convolutional Neural Networks with Autoencoder architectures are a popular ML approach which identify endmembers and perform non-linear unmixing \cite{palsson2020convolutional, non-negative-autoencoders, su2019daen, borsoi2019deep}
-- Self Organizing Map is another approach which has been used for endmember extraction together with neural networks for abundance mapping (unmixing) \cite{cantero2004analysis}
-
-Hyperspectral imaging has emerged as a pivotal technology in this space.
-With their increased spectral resolution, hyperspectral imagers (HSIs) capture the complex chemical, biological, environmental characteristics of reflectance spectra.
-(add references for specific applications of HSIs).
-Continued development of HSI technology has led to a dramatic reduction in their size enabling their utilize as the payload of small autonomous aerial vehicles (UAVs) (citation).
-By flying close to the ground UAV-born hyperspectral imagers can produce centimeter-scale imagery while simulatenously limiting the need for complicated atmospheric corrections required for satellite platforms (citation).
-
-- applications include food quality & safety, medical diagnoses, precision agriculture, and forensic document examination \cite{khan2018modern}
-- Hyperspectral data were used to distinguish oils by type, e.g. crude, diesel, gasoline, and palm \cite{yang2020characterization}
-- Near-earth HSI addresses spatial, spectral, and temporal limitations of satellite and airborne platforms. UAV-based HSI enable fine-scale mapping \cite{banerjee2020uav}
-- Drones can be equipped with HSI and compute to enable rapid generation of spectral indices like the NDVI for applications such as precision agriculture \cite{horstrand2019uav}
-- UAV-based HSI can be georectified to centimeter-scales without need for ground control points by using on-board GPS and IMU \cite{arroyo2019implementation}
-- UAV-based HSI for turbidity estimation \cite{vogt2016near}
-- Polynomial regression models for chlorophyll-a, turbidity \cite{zhang2022selection}
-
-In our previous work, we introduced a robot team designed to rapidly characterize water composition in novel environments \cite{robot-team-1}. 
-Joining UAV-borne HSI together with collection of in-situ reference data by an autonomous boat dramatically reduces the time required to generate .
-
-By carefully coordinating the simultanous collection of ground-truth water quality data and hyperspectral imagery, this team can rapidly estimate concentrations of optically-active components including colored dissolved organic matter (CDOM), blue-green algae, crude oil, temperature as well
-
-In this paper, we demonstrate how Generative Topographic Mapping (GTM), a Bayesian alternative to the Self Organizing Map, can be utilized to rapidly identify relevant spectral signatures within a scene.
-Using data collected at site in Montague, North Texas, we demonstrate the ability of this paradigm with two case studies.
-First, we demonstrate how the GTM can be used to provide a high-resolution unsupervised classification of a water body enabling the rapid identification of relevant regions which can then be used to direct further investigation by reference-sensor equipped boats.
-Second, we demonstrate the ability to perform endmember extraction and abundance mapping using imagery from a test-release of rhodamine dye.
-Together with the normalized spectral similarity score, these spectral signatures are identified and their associated area in the scene is determined.
 
@@
 \begin{tabular}{@{}ll}
+UAV & Unmanned Aerial Vehicle \\
+SOM & Self Organizing Map \\ 
+GTM & Generative Topographic Mapping  \\
+HSI & Hyperspectral Image \\ 
 CDOM & Colored Dissolved Organic Matter \\
-HAB & Harmful Algal Bloom \\ 
-HSI & Hyperspectral Image \\ 
-SOM & Self Organizing Map \\ 
-GTM & Generative Topographic Mapping 
+HAB & Harmful Algal Bloom 
 \end{tabular}

### 2024-04-24 paper/outline.tex
separated introduction into draft, finished draft abstract (needs work), and added some points to the outline.
updated outline

@@
 
+NOTE: We should move the description of the GTM algroithm to the top of the methods section.
+
 \section*{Outline}
@@
 \1 Introduction
-    \2 Characterization of inland water quality is hard
+    \2 Characterization of inland water quality via remote sensing is challenging for 4 key reasons:
         \3 Key limitations:
@@
 \1 Materials and Methods
-    \2 Data Collection
-        \3 Describe collection site in Montague, North Texas
+    \2 Overview paragraph about project, i.e.
+plan is to use the Generative Topographic Mapping to analyze drone-based hyperspectral imagery collected at a pond in Montague North Texas on 23-November 2021.
    \2 Autonomous Robot Team
@@
             \4 Resonon Pika SC2 HSI
-                \5 462 wavelengths per pixel
-               \5 1600 pixels per scan-line
-               \5 embedded GPS/IMU
+                \begin{itemize}
+                    \item 462 wavelengths per pixel
+                    \item 1600 pixels per scan-line
+                    \item embedded GPS/IMU
+               \end{itemize}
              \4 Intel NUC for on-board compute
+   \2 Data Collection
+        \3 Describe collection site in Montague, North Texas
+        \3 Collection was performed near solar noon to maximize incident light.
+At this time of year, the maximum solar elevation was $\approx 60^\circ$ so that sun-glint is not as much of an issue.
+(Cite sun-glint paper here).
     \2 Pre-processing
+        \3 Reflectance conversion of radiance imagery using on-board downwelling irradiance spectrometer
+        \3 Georectification of collected
+        \3 Wavelength selection to $\lambda \leq 900 \text{nm}$ as longer wavelengths are noisy
+        \3 Rescaling of spectra so that peak has $\tilde{R}  = 1.0$  to account for variability in incident lighting conditions due 
     \2 Generative Topographic Mapping
@@
             \4 paper comparing different distance functions for spectra
-                \5 curse of dimensionality can be pernicious here as we have > 462 wavelengths per pixel
+                \begin{itemize}
+                    \item curse of dimensionality can be pernicious here as we have > 462 wavelengths per pixel
+                \end{itemize}
             \4 NS3 offers good trade off between spectral angle and euclidean-like distance (MSE)
         \3 Describe use of thresholded NS3 for mapping abundance of GTM derived endmembers.
-
+    \2 Overview of two approaches
+        \3 GTM trained on water-only pixels for unsupervised classification of pond
+            \4 Goal: Reveal the small-scale spatial variability of pond water
+            \4 Discuss how unsupervised classification can help inform autnomous data collection by robot team (save for discussion?)
+        \3 GTM trained on combined dataset including water, land, and dye plume pixels
+            \4 Fit GTM
+            \4 Identify spectral signatures of key exemplar spectra (endmembers)
+            \4 Use endmembers to map algal abundance and time evolution of rhodamine dye plume
 \1 Results
@@
             \4 Comment on time to train standard GTM
-                \5 Key parameters are size of dataset (number of records) and number of RBF centers, $M$.
+                \begin{itemize}
+                    \item Key parameters are size of dataset (number of records) and number of RBF centers, $M$.
+                \end{itemize}
             \4 Bishop suggest extension for on-line training of GTM using batches.

### 2024-04-24 paper/paper.tex
updated with equations

@@
 %\end{adjustwidth}
-\caption{(a) Sample hyperspectral data cube.
+\caption{(\textbf{a}) Sample hyperspectral data cube.
 Spectra are plotted using their geographic position with the log10-reflectance colored along the z axis and a pseudo-color image on top.
-(b) Points taken from a sample hyperspectral data cube corresponding to algae, rhodamine dye, water, and dry grass.
-(c) The associated spectra for the exemplar points scaled so that the peak value of each spectrum is 1.
-(d) The location of the pond in Montague, Texas where data were collected for this study.\label{fig:sample-spectra}}
+(\textbf{b}) Points taken from a sample hyperspectral data cube corresponding to algae, rhodamine dye, water, and dry grass.
+(\textbf{c}) Reflectance spectra for the exemplar points scaled so the peak value of each spectrum is $1.0$.
+(\textbf{d}) The location of the pond in Montague, Texas where data were collected for this study.\label{fig:sample-spectra}}
 \end{figure}
 
-\subsection{Autonomous Robot Team}
-\subsection{Data Collection and Pre-Processing}
+at a pond in Montague, north Texas (shown in panel d of Fig~\ref{fig:sample-spectra})
+
+\subsection{UAV-Based Hyperspectral Imaging}
+A Freefly Alta-X autonomous quadcopter was used as a UAV platform for the robotic team.
+The Alta-X is specifically designed to carry cameras and has a payload of up to 35 lbs.
+We equipped the UAV with a Resonon Pika XC2 visible+near-infrared (VNIR) hyperspectral imager.
+For each image pixel, this camera samples 462 wavelengths ranging from 391 to 1011 nm.
+Additionally, the UAV includes an upward facing Ocean Optics UV-Vis-NIR spectrometer with a cosine corrector to capture the incident downwelling irradiance spectrum.
+Data collection by the hyperspectral imager is controlled by an attached Intel NUC small-form-factor computer.
+A second processing NUC is also included for onboard georectification and generation of data products.
+The collected hyperspectral images (HSIs) are stored locally on a solid state drive that is simultaneously mounted by the processing computer.
 
@@
 
+\subsection{Data Collection and Pre-Processing}
 - Julia programming language used \cite{bezanson2012julia}
@@
 
-- GTM paper oricinal \cite{gtm-bishop-1}
-- GTM developments paper \cite{gtm-biship-2}
+The Generative Topographic Mapping (GTM) is an unsupervised latent variable model developed by Bishop, Svens\'en, and Williams \cite{gtm-bishop-1}.
+The GTM takes inspiration from Kohonen's self-organizing map which enables an unsupervised classification of high dimensional data into a (typically) two-dimensional grid of nodes such that nodes near each near each other in the mapping space
 
+- mention
+
+\begin{equation}\label{eq:latent-prob}
+    p(\xi) = \frac{1}{K}\sum_k^K \delta(\xi - \xi_k)
+\end{equation}
+
+\begin{equation}\label{eq:data-prob}
+    p\left( \mathbf{x} \mid \xi, W, \beta \right)  = \mathcal{N}(\psi,\, \beta^{-1})
+\end{equation}
+
+\begin{equation}\label{eq:psi}
+    \psi = W\phi(\xi)
+\end{equation}
+
+Parameter prior (used for model weight regularization), i.e. assuming a Gaussian distribution with mean 0 for each weight where $\lVert \cdot \rVert_{F}$ is the Frobenius norm.
+\begin{equation}\label{eq:weight-prior}
+    p(W \mid \alpha) =  \left( \frac{\alpha}{2\pi} \right)^{MD/2}\exp\left(-\frac{\alpha}{2}\lVert W \rVert_{F}^2\right)
+\end{equation}
+\begin{equation}\label{eq:llh}
+    \mathcal{L}(W, \beta) = \sum_n^N \ln \left(\dfrac{1}{K}\sum_k^K p(\mathbf{x}_n \mid \xi_k, W, \beta) \right)
+\end{equation}
+
+Expectation Maximization Algorithm:
+
+Expectation step:
+\begin{equation}\label{eq:responsibility}
+    R_{kn} = p(\xi_k \mid \mathbf{x}_n, W, \beta) = \dfrac{p(\mathbf{x}_n \mid \xi_k, W, \beta)}{\sum\limits_{k'}^{K} p(\mathbf{x}_n \mid \xi_{k'}, W, \beta)}
+\end{equation}
+
+Maximization step:
+
+Definition of the diagonal matrix, $G$, with entries
+\begin{equation}
+    G_{kk}  = \sum\limits_n^N R_{kn}
+\end{equation}
+and all others $0$.
+
+\begin{equation}\label{eq:W-update}
+    \left(\Phi^T G \Phi + \dfrac{\alpha}{\beta}I \right)W_{\text{new}} = \Phi^T R X 
+\end{equation}
+
+\begin{equation}
+    \hat{\xi}_n = \sum_{k}^K R_{kn}\xi_k
+\end{equation}
+
+- GTM paper original \cite{gtm-bishop-1}
+- GTM developments paper \cite{gtm-bishop-2}
+- GTM code \cite{gtm-code}
 - implementation designed to work in MLJ ecosystem \cite{blaom2020mlj}

### 2024-04-28 paper/outline.tex
updated with equations
pushing changes to overleaf

@@
 
-NOTE: We should move the description of the GTM algroithm to the top of the methods section.
+NOTE: We should move the description of the GTM algorithm to the top of the methods section.
 
 \section*{Outline}
+We should recast the introduction to focus on unsupervised classification rather than endmember extraction...
+
+Characterization on inland waters via remote sensing is hard because (reasons).
+Hyperspectral imaging has emerged as a pivotal technology in this space due to the dramatic improvement in spectral resolution afforded by HSI platforms.
+Specifically, the combination with hyperspectral imaging together with autonomous collection by UAVs enables unparalleled spectral, spatial, and temporal resolution.
+Significant effort has developed the use of hyperspectral imagery for the inversion of optically-active water quality parameters.
+However, this approach requires access to sufficient quantities of in situ reference data to fit and evaluate each model.
+Traditionally, this process has required collocation of remote sensing imagery with data from field campaigns spanning decades (cite Aurin and Ross).
+Recently, we have demonstrated how combining autonomous collection of in situ data with UAV-borne hyperspectral imagery dramatically reduces the time required to collect these data.
+Both of these approaches require we already know what to look for.
+
 \begin{outline}[enumerate]
@@
     \2 Hyperspectral Imaging addresses spectral resolution
-    \2 UAV adresses spatial and temporal resolution
+    \2 UAV addresses spatial and temporal resolution
     \2 Robot team addresses in situ data... but what do we do if we don't already know which sensors to use? 

### 2024-04-28 paper/paper.tex
pushing changes to overleaf

--- 2024-04-23 paper/paper.tex
@@
 % Abstract (Do not insert blank lines, i.e. \\) 
-\abstract{
-Unmanned Aerial Vehicles (UAVs) equipped with hyperspectral imagers have emerged as an essential technology for the characterization of inland water bodies.
-The high spectral and spatial resolutions of these systems enable the retrieval of a plethora of optically-active water quality parameters via band ratio algorithms and machine learning methods.
-However, fitting and validating these models requires access to sufficient quantities of in situ reference data which are time-consuming and expensive to obtain.
-In this study, we demonstrate how the Generative Topographic Mapping (GTM), a Bayesian realisation of the Self Organizing Map (SOM), enables highly-detailed unsupervised classification and nonlinear endmember extraction of UAV acquired hyperspectral imagery.
-Specifically, the topographic relationship between classes maintained by the GTM presents an attractive alternative to other unsupervised approaches which do not provide a notion of similarity between learned classes.
-Using data collected across a North Texas pond, we apply the GTM to create a detailed map segmenting the area into distinct regions of interest.
-Further, by interpreting the learned classes as distinct endmembers, the GTM can be used to rapidly identify and map unique water constituents.
-We demonstrate this capability by using the GTM to map algal abundance as well as the dispersion of a rhodamine dye tracer released into the water.
-}
+\abstract{FIXME}
 
 % Keywords
-\keyword{Hyperspectral Imaging; Remote Sensing; Unsupervised Classification; Endmember Extraction, Spectral Unmixing}
+\keyword{Hyperspectral Imaging; Remote Sensing; Unsupervised Classification; Spectral Unmixing}
 
@@
 \section{Introduction}
+
+Inland water bodies pose a unique challenge for analysis by remote sensing imagery due to their complex spectral characteristics and small-scale spatial variability; nevertheless, significant effort has gone into the development of techniques for estimating optically-active water quality parameters such as chlorophyll-a, blue-green algae, turbidity, and temperature (citation).
+Machine learning techniques have proven to be a powerful tool in this domain, facilitating the development of data-driven models which map observed reflectance to parameters of interest.
+However, these methods rely on the availability of significant volumes of in-situ reference data needed to calibrate models and asses their uncertainty.
+For example Aurin et al...
+Similarly, Ross et al developed...
+The challenge posed by in-situ data collection together with long satellite revisit times limits the utility of remote sensing imagery for real-time applications.
+
+- supervised classification of water quality with XGBoost \cite{ghatkar2019classification}
+- Spectral signatures of classes are complex and often overlapping \cite{thenkabail2018hyperspectral}
+- Remote sensing used for oil spill analysis: extent and thickness mapping \cite{kokaly2013spectroscopic, leifer2012state}
+- Sun glitter remains a key challenge for sensing in the visible portion of the spectrum but multi and hyperspectral imagers have been used for oil spill identification and to identify their impacts on vegetation stress and mortality \cite{fingas2014review, khan2018modern}
+- common approach is inversion of optically-active water quality parameters such as chlorophyll-a, blue-green algae, turbidity, and temperature \cite{ritchie2003remote}
+- Combining spectral indices such as the NDVI together with machine learning is a popular approach \cite{thenkabail2018hyperspectral, sagan2020monitoring, lu2021retrieval}
+- Polynomial regression models for chlorophyll-a, turbidity \cite{zhang2022selection}
+- Ross et al.
+created a comprehensive dataset with over 600,000 water quality records matching optically active water quality parameters with associated satellite imagery from Landsat 5,7,8.
+To achieve this quantity of data, they needed records spanning 1984 to 2019 \cite{ross2019aquasat}
+- Aurin et al. took a similar route combining 30 years of remote sensing imagery with in-situ data from over 500 field compaigns for CDOM, organic carbon, etc... \cite{aurin2018remote}.
+- When ground truth data aren't available in sufficient quantity, classification into water quality categories is another approach (ground truth can be easier to obtain by expert analysis of scene) \cite{koponen2002lake}
+
+Unsupervised classification
+- When no ground-truth data are available, unsupervised classification can still help partition imagery into groups or clusters.
+- Many data-driven, ML methods have been employed for the task including various matrix factorizations, k-nearest neighbors, fuzzy c-means, density estimation methods, etc.
+\cite{zhang2019hyperspectral}
+- Additionally unsupervised approaches can be used to perform nonlinear dimensionality reduction & pre-processing for supervised approaches.
+- Many unsupervised approaches can be challenging to interpret due to the lack of a relationship between the identified classes, e.g. is class 1 more similar to class 2 than class 10?
+- The Self Organizing Map is a popular technique which classifies data while enforcing topological relationships between the resulting classes such that classes close to each  other in the latent space are more similar than classes that are far apart.
+- SOM used for remote sensing imagery classification \cite{msom-remote-sensing}
+- SOM used for clustering and data compression of HSI cube-sat \cite{danielsen2021self}
+- SOM used for land-use and land-cover change analysis \cite{penfound2021analysis}
+- SOM has bee nused for identifying synoptic-scale patterns in wind and sea surface temperature data {som-satellite}
+
+endmember extraction, spectral unmixing
+- This is a related problem where the goal is to identify unique spectral signatures which combine (linearly or non-linearly) to produce the measured signal.
+- Having identified the set of endmembers, we then seek to determine their relative abundance in each pixel
+- Many statistical approaches for extracting endmembers \cite{berman2004ice}
+- Sparse PCA used for endmember extraction \cite{yousefi2016mineral}
+- Popular spectral unmixing approach is to treat "pure" endmembers as vertices of a simplex \cite{plaza2012endmember, nascimento2005vertex}
+- Convolutional Neural Networks with Autoencoder architectures are a popular ML approach which identify endmembers and perform non-linear unmixing \cite{palsson2020convolutional, non-negative-autoencoders, su2019daen, borsoi2019deep}
+- Self Organizing Map is another approach which has been used for endmember extraction together with neural networks for abundance mapping (unmixing) \cite{cantero2004analysis}
+
+Hyperspectral imaging has emerged as a pivotal technology in this space.
+With their increased spectral resolution, hyperspectral imagers (HSIs) capture the complex chemical, biological, environmental characteristics of reflectance spectra.
+(add references for specific applications of HSIs).
+Continued development of HSI technology has led to a dramatic reduction in their size enabling their utilize as the payload of small autonomous aerial vehicles (UAVs) (citation).
+By flying close to the ground UAV-born hyperspectral imagers can produce centimeter-scale imagery while simulatenously limiting the need for complicated atmospheric corrections required for satellite platforms (citation).
+
+- applications include food quality & safety, medical diagnoses, precision agriculture, and forensic document examination \cite{khan2018modern}
+- Hyperspectral data were used to distinguish oils by type, e.g. crude, diesel, gasoline, and palm \cite{yang2020characterization}
+- Near-earth HSI addresses spatial, spectral, and temporal limitations of satellite and airborne platforms. UAV-based HSI enable fine-scale mapping \cite{banerjee2020uav}
+- Drones can be equipped with HSI and compute to enable rapid generation of spectral indices like the NDVI for applications such as precision agriculture \cite{horstrand2019uav}
+- UAV-based HSI can be georectified to centimeter-scales without need for ground control points by using on-board GPS and IMU \cite{arroyo2019implementation}
+- UAV-based HSI for turbidity estimation \cite{vogt2016near}
+
+In our previous work, we introduced a robot team designed to rapidly characterize water composition in novel environments \cite{robot-team-1}. 
+Joining UAV-borne HSI together with collection of in-situ reference data by an autonomous boat dramatically reduces the time required to generate .
+
+By carefully coordinating the simultanous collection of ground-truth water quality data and hyperspectral imagery, this team can rapidly estimate concentrations of optically-active components including colored dissolved organic matter (CDOM), blue-green algae, crude oil, temperature as well
+
+In this paper, we demonstrate how Generative Topographic Mapping (GTM), a Bayesian alternative to the Self Organizing Map, can be utilized to rapidly identify relevant spectral signatures within a scene.
+Using data collected at site in Montague, North Texas, we demonstrate the ability of this paradigm with two case studies.
+First, we demonstrate how the GTM can be used to provide a high-resolution unsupervised classification of a water body enabling the rapid identification of relevant regions which can then be used to direct further investigation by reference-sensor equipped boats.
+Second, we demonstrate the ability to perform endmember extraction and abundance mapping using imagery from a test-release of rhodamine dye.
+Together with the normalized spectral similarity score, these spectral signatures are identified and their associated area in the scene is determined.
 
@@
 \begin{tabular}{@{}ll}
-UAV & Unmanned Aerial Vehicle \\
+HAB & Harmful Algal Bloom \\ 
+HSI & Hyperspectral Image \\ 
 SOM & Self Organizing Map \\ 
-GTM & Generative Topographic Mapping  \\
-HSI & Hyperspectral Image \\ 
-CDOM & Colored Dissolved Organic Matter \\
-HAB & Harmful Algal Bloom 
+GTM & Generative Topographic Mapping 
 \end{tabular}

### 2024-04-30 3b-gtm-supervised.jl
added extra plots

@@ -198,6 +198,86 @@ save(joinpath(figures_path, "gtm-classes-selected.png"), fig)
 save(joinpath(figures_path, "gtm-classes-selected.pdf"), fig)
 
 
+
+
+# plot with in latent space with color by the Supervised boat values
+
+
+
+# CDOM
+fig = Figure();
+ax = CairoMakie.Axis(fig[1,1], xlabel="ξ₁", ylabel="ξ₂");
+mean_proj = DataFrame(MLJ.transform(mach, X))
+target_color = Vector(df_Y.CDOM)
+clims = (20.1, 21.6)
+cm = cgrad(:roma, rev=true)
+
+s = scatter!(ax, mean_proj.ξ₁, mean_proj.ξ₂, markersize=9, alpha=0.85, color=target_color, colormap=cm, colorrange=clims)
+cb = Colorbar(fig[1,2], label="CDOM", colorrange=clims, colormap=cm, lowclip = cm[1], highclip = cm[end])
+fig
+save(joinpath(figures_path, "CDOM.png"), fig)
+save(joinpath(figures_path, "CDOM.pdf"), fig)
+
+
+names(df_Y)
+
+# Temperature
+fig = Figure();
+ax = CairoMakie.Axis(fig[1,1], xlabel="ξ₁", ylabel="ξ₂");
+mean_proj = DataFrame(MLJ.transform(mach, X))
+target_color = Vector(df_Y.Temp3488)
+clims = (13.25, 13.95)
+cm = cgrad(:roma, rev=true)
+
+s = scatter!(ax, mean_proj.ξ₁, mean_proj.ξ₂, markersize=9, alpha=0.85, color=target_color, colormap=cm, colorrange=clims)
+cb = Colorbar(fig[1,2], label="Temperature", colorrange=clims, colormap=cm, lowclip = cm[1], highclip = cm[end])
+fig
+save(joinpath(figures_path, "temperature.png"), fig)
+save(joinpath(figures_path, "Temperature.pdf"), fig)
+
+
+# Na
+fig = Figure();
+ax = CairoMakie.Axis(fig[1,1], xlabel="ξ₁", ylabel="ξ₂");
+mean_proj = DataFrame(MLJ.transform(mach, X))
+target_color = Vector(df_Y.Na)
+clims = (200, 380)
+cm = cgrad(:roma, rev=true)
+
[…]

### 2024-04-30 7-extra-plots.jl

@@ -1,0 +1,36 @@
+using CairoMakie
+include("utils/makie-defaults.jl")
+
+set_theme!(mints_theme)
+update_theme!(
+    figure_padding=30,
+    Axis=(
+        xticklabelsize=20,
+        yticklabelsize=20,
+        xlabelsize=22,
+        ylabelsize=22,
+        titlesize=25,
+    ),
+    Colorbar=(
+        ticklabelsize=20,
+        labelsize=22
+    )
+)
+
+figures_path = joinpath("./figures", "extras")
+if !ispath(figures_path)
+    mkpath(figures_path)
+end
+
+
+k = 32
+m = 14
+s = 0.1
+
+ξ₁ = range(-1, stop=1.0, length=k)
+ξ₁ = range(-1, stop=1.0, length=k)
+
+
+
+# create figure with latent space grid and RBF points as well as
+# the "3d" projection of means in the "data" space

### 2024-04-30 paper/draft-introduction.tex

@@
 \section{Introduction}
+
+1. Remote sensing for water quality
+2. Unsupervised approaches for remote sensing
+3. UAV based Hyperspectral Imaging offers advantages for water composition analysis and characterization
+4. Approaches for "Unsupervised" Analysis of 
+    - Traditional Unsupervised Classification methods from ML
+    - Endmember Extraction and Spectral Unmixing (i.e. assumption of linear mixing and projection-based and statistical methods for endmember extraction + abundance mapping)
+5. Goal of this paper
+    - The generative topographic mapping is an attractive algorithm 
+        - Probabilistic interpretation
+        - Topographic relationship maintained between classes (not just orthogonality)
+        - clear fitting criterion with Maximum Likelihood Estimation / Maximum A-Posteriori Estimation
+    - We use the GTM to analyze hyperspectral imagery acquired by a drone over pond to 
+        - Present method for unsupervised characterization of small-scale variability of pond (i.e. identifying regions of interest which can be used to direct robot boat for further instructions) 
+        - Use GTM to extract endmember for algae and map the algal distribution near shore (relevant for Harmful Algal Bloom detection)
+        - Use GTM to extract endmember for rhodamine dye and use it to mape evolution of dye plume.
+This is a demonstration of source identification and mapping.
+Relevant for the identification of pollution events such as oil spills, industrial seepage, etc...
 
@@
 
+The primary goal of this study...
+
 \bibliographystyle{plain}

### 2024-04-30 paper/notes/outline.tex

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  letterpaper,
  DIV=11,
  numbers=noendperiod]{scrartcl}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\KOMAoption{captions}{tableheading}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Paper Outline},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Paper Outline}
\author{}
\date{}

\begin{document}
\maketitle

\section{Introduction}\label{introduction}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Characterization of Inland Water Bodies via Remote Sensing is
  Challenging

  \begin{itemize}
  \tightlist
  \item
    complex spectral signatures
  \item
    limited spatial and temporal resolution (long satellite overpasses)
  \item
    nevertheless remote sensing has significant potential
  \item
    Harmful Algal Bloom (HAB) classification and mapping
  \item
    Oil Spill mapping
  \end{itemize}
\item
  Hyperspectral Imaging addresses spectral limitations of traditional
  remote sensing platforms
\item
  Supervised methods

  \begin{itemize}
  \tightlist
  \item
    Inversion of optically-active parameters
  \item
    Key limitation: availability of in-situ reference data

    \begin{itemize}
    \tightlist
    \item
      Cite Aurin et al.~and Ross et al.~here
    \end{itemize}
  \item
    When ground truth data are limited, classification is an alternative
    approach, e.g.~``bad'' to ``good'' water quality
  \end{itemize}
\item
  Unsupervised methods

  \begin{itemize}
  \tightlist
  \item
    Unsupervised classification \& ML Techniques
  \item
    Endmember Extraction and Spectral Unmixing
  \end{itemize}
\item
  Robot Teams and HSI equipped Drones
\item
  This paper

  \begin{itemize}
  \tightlist
  \item
    Extend capabilities of robot team by showcasing capabilities for
    unsupervised classification
  \item
    Utilize GTM as Bayesian extension of SOM to enable unsupervised
    classification
  \item
    GTM also enables extraction of spectral features associated with
    each node
  \item
    Two case studies

    \begin{enumerate}
    \def\labelenumii{\arabic{enumii}.}
    \tightlist
    \item
      GTM trained on water-only HSI data for lake

      \begin{itemize}
      \tightlist
      \item
        Identify interesting regions for further examination by robot
        team
      \item
        Investigate spectral variability of water
      \end{itemize}
    \item
      GTM on dataset combining grass and water pixels with a simulated
      pollution event using Rhodamine tracer dye

      \begin{itemize}
      \tightlist
      \item
        Use extracted GTM ``endmembers'' together with NS3 to map
        abundance
      \item
        Identify abundance of algae
      \item
        Map evolution of Rhodamine dye plume
      \end{itemize}
    \end{enumerate}
  \end{itemize}
\end{enumerate}

\section{Materials and Methods}\label{materials-and-methods}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Data Collection

  \begin{itemize}
  \tightlist
  \item
    Describe collection at site in Montague, North Texas
  \end{itemize}
\item
  Autonomous Robot Team

  \begin{itemize}
  \tightlist
  \item
    Describe drone

    \begin{itemize}
    \tightlist
    \item
      Alta X Quadcopter
    \item
      Resonon Pika XC2 HSI
    \item
      Intel Nuc compute
    \end{itemize}
  \item
    describe georectification procedure
  \end{itemize}
\item
  Pre-Processing
\item
  Generative Topographic Mapping
\end{enumerate}

\begin{itemize}
\tightlist
\item
  Original SOM paper (Kohonen 1990)
\item
  Heskes reinterpretation of SOM algorithm as stochastic gradient
  descent on an energy suface i.e.~justification with a cost function
  (Heskes 1999)
\item
  GTM paper (Bishop, Svensén, and Williams 1998b)

  \begin{itemize}
  \tightlist
  \item
    Describe algorithm
  \item
    Describe parameters
  \end{itemize}
\item
  Bayesian Information Criterion and Akaike Information Criterion for
  Hyperparameter Optimization

  \begin{itemize}
  \tightlist
  \item
    A full Bayesian treatment would involve marginalization over all
    possible parameter values but this is much more computational
    efficient.
  \end{itemize}
\end{itemize}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{4}
\tightlist
\item
  Abundance Mapping with the Normalized Spectral Similarity Score

  \begin{itemize}
  \tightlist
  \item
    Describe different metrics for comparison of spectra

    \begin{itemize}
    \tightlist
    \item
      Trade-off between metrics that distinguish changes in overall
      intensity and changes in peak location, i.e.~hue
    \end{itemize}
  \end{itemize}
\end{enumerate}

\section{Results}\label{results}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Two Case Studies
\item
  GTM Fit on Full Dataset
\item
  Hyperparameter Optimization with BIC
\item
  Spectral Signature Identification
\item
  Water Class Map (Water Only GTM)
\item
  Algae Identification
\item
  Dye Plume Identification
\end{enumerate}

\section{Discussion}\label{discussion}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  GTM Applications in our Paper
\item
  Comparison to existing approaches

  \begin{itemize}
  \tightlist
  \item
    Unsupervised Classification
  \item
    Endmember Extraction \& Spectral Unmixing
  \item
    Applications for either of these with Drones?
  \end{itemize}
\item
  Use of GTM and Unsupervised methods for Robot Team

  \begin{itemize}
  \tightlist
  \item
    As a feature transformer/preprocessor for supervised methods (as
    opposed to, say, PCA or k-NN)
  \item
    identification of ``interesting'' regions for intelligent deployment
    of robotic boat

    \begin{itemize}
    \tightlist
    \item
      don't continue collecting data in region that is similar
    \item
      provision boat to maximize data collection across distribution of
      GTM classes
    \end{itemize}
  \end{itemize}
\item
  Extensions and future work

  \begin{itemize}
  \tightlist
  \item
    Incorporation of GTM extensions for efficient online-learning

    \begin{itemize}
    \tightlist
    \item
      Currently GTM classes can quickly be applied (in real time) once
      they are trained to generate abundance maps e.g.~chlorophyll,
      algae, etc\ldots{}
    \item
      Key limitation of GTM is similar to SOM and k-NN: you need to
      compute distances between samples and latent nodes. This can scale
      poorly for very large datasets
    \item
      Bishop suggests a way to augment GTM training algorithm to enable
      online training of GTM. This could be deployed on-board the UAV to
      perform classification and identify spectral signatures in
      near-real time.
    \item
      Extensions of GTM (Bishop, Svensén, and Williams 1998a)
    \end{itemize}
  \item
    Utilize the GTM with in-situ data collection to identify spectral
    signatures of specific algal species
  \end{itemize}
\end{enumerate}

\section{Conclusions}\label{conclusions}

\section{NOTES AND CITATIONS}\label{notes-and-citations}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Applications of remote sensing to water quality

  \begin{itemize}
  \tightlist
  \item
    machine learning (XGBoost) used with paired remote sensing imagery
    and in-situ data to classify imagery into 5 categories including 3
    for harmful algal blooms (Ghatkar, Singh, and Shanmugam 2019)
  \item
    Spectral signatures of classes are complex and often overlapping
    (Thenkabail, Lyon, and Huete 2018)
  \item
    Remote sensing used for oil spill analysis: extent and thickness
    mapping (Kokaly et al. 2013; Leifer et al. 2012)
  \item
    Sun glitter remains a key challenge for sensing in the visible
    portion of the spectrum but multi and hyperspectral imagers have
    been used for oil spill identification and to identify their impacts
    on vegetation stress and mortality (Fingas and Brown 2014; Khan et
    al. 2018)
  \end{itemize}
\item
  Hyperspectral Imagery

  \begin{itemize}
  \tightlist
  \item
    applications include food quality \& safety, medical diagnoses,
    precision agriculture, and forensic document examination (Khan et
    al. 2018)
  \item
    Hyperspectral data were used to distinguish oils by type,
    e.g.~crude, diesel, gasoline, and palm (Yang et al. 2020)
  \end{itemize}
\item
  Supervised regression and classification for water quality

  \begin{itemize}
  \tightlist
  \item
    common approach is inversion of optically-active water quality
    parameters such as chlorophyll-a, blue-green algae, turbidity, and
    temperature (Ritchie, Zimba, and Everitt 2003)
  \item
    Combining spectral indices such as the NDVI together with machine
    learning is a popular approach (Thenkabail, Lyon, and Huete 2018;
    Sagan et al. 2020; Lu et al. 2021)
  \item
    Polynomial regression models for chlorophyll-a, turbidity (D. Zhang,
    Zeng, and He 2022)
  \item
    Key limitation is collection of sufficient quantity of in-situ
    reference data
  \item
    Ross et al.~created a comprehensive dataset with over 600,000 water
    quality records matching optically active water quality parameters
    with associated satellite imagery from Landsat 5,7,8. To achieve
    this quantity of data, they needed records spanning 1984 to 2019
    (Ross et al. 2019)
  \item
    Aurin et al.~took a similar route combining 30 years of remote
    sensing imagery with in-situ data from over 500 field compaigns for
    CDOM, organic carbon, etc\ldots{} (Aurin, Mannino, and Lary 2018).
  \item
    When ground truth data aren't available in sufficient quantity,
    classification into water quality categories is another approach
    (ground truth can be easier to obtain by expert analysis of scene)
    (Koponen et al. 2002)
  \item
    When no ground-truth data are available, unsupervised classification
    can still help partition imagery into groups or clusters.
  \item
    Many data-driven, ML methods have been employed for the task
    including various matrix factorizations, k-nearest neighbors, fuzzy
    c-means, density estimation methods, etc. (L. Zhang et al. 2019)
  \item
    Additionally unsupervised approaches can be used to perform
    nonlinear dimensionality reduction \& pre-processing for supervised
    approaches.
  \item
    SOM used for remote sensing imagery classification (Wan and Fraser
    2000)
  \item
    SOM used for clustering and data compression of HSI cube-sat
    (Danielsen, Johansen, and Garrett 2021)
  \item
    SOM used for land-use and land-cover change analysis (Penfound and
    Vaz 2021)
  \end{itemize}
\item
  Endmember Extraction and Spectral Unmixing

  \begin{itemize}
  \tightlist
  \item
    This is a related problem where the goal is to identify unique
    spectral signatures which combine (linearly or non-linearly) to
    produce the measured signal.
  \item
    Having identified the set of endmembers, we then seek to determine
    their relative abundance in each pixel
  \item
    Many statistical approaches for extracting endmembers (Berman et al.
    2004)
  \item
    Sparse PCA used for endmember extraction (Yousefi et al. 2016)
  \item
    Popular spectral unmixing approach is to treat ``pure'' endmembers
    as vertices of a simplex {[}Plaza et al. (2012);
    nascimento2005vertex{]}
  \item
    Convolutional Neural Networks with Autoencoder architectures are a
    popular ML approach which identify endmembers and perform non-linear
    unmixing (Palsson, Ulfarsson, and Sveinsson 2020; Su et al. 2017,
    2019; Borsoi, Imbiriba, and Bermudez 2019)
  \item
    Self Organizing Map is another approach which has been used for
    endmember extraction together with neural networks for abundance
    mapping (unmixing) (Cantero et al. 2004)
  \item
    SOM has bee nused for identifying synoptic-scale patterns in wind
    and sea surface temperature data (Richardson, Risien, and
    Shillington 2003)
  \item
    SOM used for HSI feature extraction (Ceylan and Kaya 2021)
  \item
    drawback of autoencoder and other statistical approach is lack of a
    (topological) relationship between the classes/endmembers. Is
    endmember 1 closer to endmember 2 or endmember 10? The SOM addresses
    this. The GTM is even better\ldots{}
  \end{itemize}
\item
  Abundance Mapping

  \begin{itemize}
  \tightlist
  \item
    Many different spectral similarity functions exist for comparing
    spectra. Each have trade-offs between ability to distinguish
    differences in intensity versus differences in peak location (hue)
    (Deborah, Richard, and Hardeberg 2015)
  \item
    spectral angle mapper is popular similarity function used in
    endmember extraction. The SAM is nice because it is more sensitive
    to shape than scale (Jiang, Werff, and Meer 2020)
  \item
    spectral correlation mapper introduced to as a statistical
    alternative to SAM based on covariance instead of spectral angle (De
    Carvalho and Meneses 2000)
  \item
    Normalized spectral similarity score developed by Nidamanuri et
    al.~combine the mean-squared error with spectral angle for a happy
    medium (Nidamanuri and Zbell 2011)
  \end{itemize}
\item
  Drone-based HSI

  \begin{itemize}
  \tightlist
  \item
    Near-earth HSI addresses spatial, spectral, and temporal limitations
    of satellite and airborne platforms. UAV-based HSI enable fine-scale
    mapping (Banerjee, Raval, and Cullen 2020)
  \item
    Drones can be equipped with HSI and compute to enable rapid
    generation of spectral indices like the NDVI for applications such
    as precision agriculture (Horstrand et al. 2019)
  \item
    UAV-based HSI can be georectified to centimeter-scales without need
    for ground control points by using on-board GPS and IMU (Arroyo-Mora
    et al. 2019)
  \item
    UAV-based HSI for turbidity estimation (Vogt and Vogt 2016)
  \end{itemize}
\end{enumerate}

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-arroyo2019implementation}
Arroyo-Mora, J Pablo, Margaret Kalacska, Deep Inamdar, Raymond Soffer,
Oliver Lucanus, Janine Gorman, Tomas Naprstek, et al. 2019.
{``Implementation of a UAV--Hyperspectral Pushbroom Imager for
Ecological Monitoring.''} \emph{Drones} 3 (1): 12.

\bibitem[\citeproctext]{ref-aurin2018remote}
Aurin, Dirk, Antonio Mannino, and David J Lary. 2018. {``Remote Sensing
of CDOM, CDOM Spectral Slope, and Dissolved Organic Carbon in the Global
Ocean.''} \emph{Applied Sciences} 8 (12): 2687.

\bibitem[\citeproctext]{ref-banerjee2020uav}
Banerjee, Bikram Pratap, Simit Raval, and PJ Cullen. 2020.
{``UAV-Hyperspectral Imaging of Spectrally Complex Environments.''}
\emph{International Journal of Remote Sensing} 41 (11): 4136--59.

\bibitem[\citeproctext]{ref-berman2004ice}
Berman, Mark, Harri Kiiveri, Ryan Lagerstrom, Andreas Ernst, Rob Dunne,
and Jonathan F Huntington. 2004. {``ICE: A Statistical Approach to
Identifying Endmembers in Hyperspectral Images.''} \emph{IEEE
Transactions on Geoscience and Remote Sensing} 42 (10): 2085--95.

\bibitem[\citeproctext]{ref-gtm-bishop-2}
Bishop, Christopher M, Markus Svensén, and Christopher KI Williams.
1998a. {``Developments of the Generative Topographic Mapping.''}
\emph{Neurocomputing} 21 (1-3): 203--24.

\bibitem[\citeproctext]{ref-gtm-bishop-1}
---------. 1998b. {``GTM: The Generative Topographic Mapping.''}
\emph{Neural Computation} 10 (1): 215--34.

\bibitem[\citeproctext]{ref-borsoi2019deep}
Borsoi, Ricardo Augusto, Tales Imbiriba, and José Carlos Moreira
Bermudez. 2019. {``Deep Generative Endmember Modeling: An Application to
Unsupervised Spectral Unmixing.''} \emph{IEEE Transactions on
Computational Imaging} 6: 374--84.

\bibitem[\citeproctext]{ref-cantero2004analysis}
Cantero, MC, RM Perez, Pablo J Martinez, PL Aguilar, Javier Plaza, and
Antonio Plaza. 2004. {``Analysis of the Behavior of a Neural Network
Model in the Identification and Quantification of Hyperspectral
Signatures Applied to the Determination of Water Quality.''} In
\emph{Chemical and Biological Standoff Detection II}, 5584:174--85.
SPIE.

\bibitem[\citeproctext]{ref-som-hsi}
Ceylan, Oğuzhan, and Gülsen Taskin Kaya. 2021. {``Feature Selection
Using Self Organizing Map Oriented Evolutionary Approach.''} \emph{2021
IEEE International Geoscience and Remote Sensing Symposium IGARSS},
4003--6. \url{https://api.semanticscholar.org/CorpusID:238750026}.

\bibitem[\citeproctext]{ref-danielsen2021self}
Danielsen, Aksel S, Tor Arne Johansen, and Joseph L Garrett. 2021.
{``Self-Organizing Maps for Clustering Hyperspectral Images on-Board a
Cubesat.''} \emph{Remote Sensing} 13 (20): 4174.

\bibitem[\citeproctext]{ref-de2000spectral}
De Carvalho, O Abilio, and Paulo Roberto Meneses. 2000. {``Spectral
Correlation Mapper (SCM): An Improvement on the Spectral Angle Mapper
(SAM).''} In \emph{Summaries of the 9th JPL Airborne Earth Science
Workshop, JPL Publication 00-18}, 9:2. JPL publication Pasadena, CA,
USA.

\bibitem[\citeproctext]{ref-deborah2015comprehensive}
Deborah, Hilda, Noël Richard, and Jon Yngve Hardeberg. 2015. {``A
Comprehensive Evaluation of Spectral Distance Functions and Metrics for
Hyperspectral Image Processing.''} \emph{IEEE Journal of Selected Topics
in Applied Earth Observations and Remote Sensing} 8 (6): 3224--34.

\bibitem[\citeproctext]{ref-fingas2014review}
Fingas, Merv, and Carl Brown. 2014. {``Review of Oil Spill Remote
Sensing.''} \emph{Marine Pollution Bulletin} 83 (1): 9--23.

\bibitem[\citeproctext]{ref-ghatkar2019classification}
Ghatkar, Jayesh Ganpat, Rakesh Kumar Singh, and Palanisamy Shanmugam.
2019. {``Classification of Algal Bloom Species from Remote Sensing Data
Using an Extreme Gradient Boosted Decision Tree Model.''}
\emph{International Journal of Remote Sensing} 40 (24): 9412--38.

\bibitem[\citeproctext]{ref-som-cost-function}
Heskes, Tom. 1999. {``Energy Functions for Self-Organizing Maps.''} In
\emph{Kohonen Maps}, 303--15. Elsevier.

\bibitem[\citeproctext]{ref-horstrand2019uav}
Horstrand, Pablo, Raúl Guerra, Aythami Rodrı́guez, Marı́a Dı́az, Sebastián
López, and José Fco López. 2019. {``A UAV Platform Based on a
Hyperspectral Sensor for Image Capturing and on-Board Processing.''}
\emph{IEEE Access} 7: 66919--38.

\bibitem[\citeproctext]{ref-jiang2020classification}
Jiang, Tingxuan, Harald van der Werff, and Freek van der Meer. 2020.
{``Classification Endmember Selection with Multi-Temporal Hyperspectral
Data.''} \emph{Remote Sensing} 12 (10): 1575.

\bibitem[\citeproctext]{ref-khan2018modern}
Khan, Muhammad Jaleed, Hamid Saeed Khan, Adeel Yousaf, Khurram Khurshid,
and Asad Abbas. 2018. {``Modern Trends in Hyperspectral Image Analysis:
A Review.''} \emph{Ieee Access} 6: 14118--29.

\bibitem[\citeproctext]{ref-kohonen-som-1}
Kohonen, Teuvo. 1990. {``The Self-Organizing Map.''} \emph{Proceedings
of the IEEE} 78 (9): 1464--80.

\bibitem[\citeproctext]{ref-kokaly2013spectroscopic}
Kokaly, Raymond F, Brady R Couvillion, JoAnn M Holloway, Dar A Roberts,
Susan L Ustin, Seth H Peterson, Shruti Khanna, and Sarai C Piazza. 2013.
{``Spectroscopic Remote Sensing of the Distribution and Persistence of
Oil from the Deepwater Horizon Spill in Barataria Bay Marshes.''}
\emph{Remote Sensing of Environment} 129: 210--30.

\bibitem[\citeproctext]{ref-koponen2002lake}
Koponen, Sampsa, Jouni Pulliainen, Kari Kallio, and Martti Hallikainen.
2002. {``Lake Water Quality Classification with Airborne Hyperspectral
Spectrometer and Simulated MERIS Data.''} \emph{Remote Sensing of
Environment} 79 (1): 51--59.

\bibitem[\citeproctext]{ref-leifer2012state}
Leifer, Ira, William J Lehr, Debra Simecek-Beatty, Eliza Bradley, Roger
Clark, Philip Dennison, Yongxiang Hu, et al. 2012. {``State of the Art
Satellite and Airborne Marine Oil Spill Remote Sensing: Application to
the BP Deepwater Horizon Oil Spill.''} \emph{Remote Sensing of
Environment} 124: 185--209.

\bibitem[\citeproctext]{ref-lu2021retrieval}
Lu, Qikai, Wei Si, Lifei Wei, Zhongqiang Li, Zhihong Xia, Song Ye, and
Yu Xia. 2021. {``Retrieval of Water Quality from UAV-Borne Hyperspectral
Imagery: A Comparative Study of Machine Learning Algorithms.''}
\emph{Remote Sensing} 13 (19): 3928.

\bibitem[\citeproctext]{ref-nidamanuri2010normalized}
Nidamanuri, Rama Rao, and Bernd Zbell. 2011. {``Normalized Spectral
Similarity Score ( \({\hbox{NS}}^{3}\)) as an Efficient Spectral Library
Searching Method for Hyperspectral Image Classification.''} \emph{IEEE
Journal of Selected Topics in Applied Earth Observations and Remote
Sensing} 4: 226--40.
\url{https://api.semanticscholar.org/CorpusID:32567143}.

\bibitem[\citeproctext]{ref-palsson2020convolutional}
Palsson, Burkni, Magnus O Ulfarsson, and Johannes R Sveinsson. 2020.
{``Convolutional Autoencoder for Spectral--Spatial Hyperspectral
Unmixing.''} \emph{IEEE Transactions on Geoscience and Remote Sensing}
59 (1): 535--49.

\bibitem[\citeproctext]{ref-penfound2021analysis}
Penfound, Elissa, and Eric Vaz. 2021. {``Analysis of Wetland Landcover
Change in Great Lakes Urban Areas Using Self-Organizing Maps.''}
\emph{Remote Sensing} 13 (24): 4960.

\bibitem[\citeproctext]{ref-plaza2012endmember}
Plaza, Javier, Eligius MT Hendrix, Inmaculada Garcı́a, Gabriel Martı́n,
and Antonio Plaza. 2012. {``On Endmember Identification in Hyperspectral
Images Without Pure Pixels: A Comparison of Algorithms.''} \emph{Journal
of Mathematical Imaging and Vision} 42: 163--75.

\bibitem[\citeproctext]{ref-som-satellite}
Richardson, Anthony J, C Risien, and Frank Alan Shillington. 2003.
{``Using Self-Organizing Maps to Identify Patterns in Satellite
Imagery.''} \emph{Progress in Oceanography} 59 (2-3): 223--39.

\bibitem[\citeproctext]{ref-ritchie2003remote}
Ritchie, Jerry C, Paul V Zimba, and James H Everitt. 2003. {``Remote
Sensing Techniques to Assess Water Quality.''} \emph{Photogrammetric
Engineering \& Remote Sensing} 69 (6): 695--704.

\bibitem[\citeproctext]{ref-ross2019aquasat}
Ross, Matthew RV, Simon N Topp, Alison P Appling, Xiao Yang, Catherine
Kuhn, David Butman, Marc Simard, and Tamlin M Pavelsky. 2019.
{``AquaSat: A Data Set to Enable Remote Sensing of Water Quality for
Inland Waters.''} \emph{Water Resources Research} 55 (11): 10012--25.

\bibitem[\citeproctext]{ref-sagan2020monitoring}
Sagan, Vasit, Kyle T Peterson, Maitiniyazi Maimaitijiang, Paheding
Sidike, John Sloan, Benjamin A Greeling, Samar Maalouf, and Craig Adams.
2020. {``Monitoring Inland Water Quality Using Remote Sensing: Potential
and Limitations of Spectral Indices, Bio-Optical Simulations, Machine
Learning, and Cloud Computing.''} \emph{Earth-Science Reviews} 205:
103187.

\bibitem[\citeproctext]{ref-su2019daen}
Su, Yuanchao, Jun Li, Antonio Plaza, Andrea Marinoni, Paolo Gamba, and
Somdatta Chakravortty. 2019. {``DAEN: Deep Autoencoder Networks for
Hyperspectral Unmixing.''} \emph{IEEE Transactions on Geoscience and
Remote Sensing} 57 (7): 4309--21.

\bibitem[\citeproctext]{ref-non-negative-autoencoders}
Su, Yuanchao, Andrea Marinoni, Jun Li, Antonio Plaza, and Paolo Gamba.
2017. {``Nonnegative Sparse Autoencoder for Robust Endmember Extraction
from Remotely Sensed Hyperspectral Images.''} In \emph{2017 IEEE
International Geoscience and Remote Sensing Symposium (IGARSS)}, 205--8.
\url{https://doi.org/10.1109/IGARSS.2017.8126930}.

\bibitem[\citeproctext]{ref-thenkabail2018hyperspectral}
Thenkabail, Prasad S, John G Lyon, and Alfredo Huete. 2018.
\emph{Hyperspectral Indices and Image Classifications for Agriculture
and Vegetation}. CRC press.

\bibitem[\citeproctext]{ref-vogt2016near}
Vogt, Michael C, and Mark E Vogt. 2016. {``Near-Remote Sensing of Water
Turbidity Using Small Unmanned Aircraft Systems.''} \emph{Environmental
Practice} 18 (1): 18--31.

\bibitem[\citeproctext]{ref-msom-remote-sensing}
Wan, Weijian, and Donald Fraser. 2000. {``A Multiple Self-Organizing Map
Scheme for Remote Sensing Classification.''} In \emph{International
Workshop on Multiple Classifier Systems}, 300--309. Springer.

\bibitem[\citeproctext]{ref-yang2020characterization}
Yang, Junfang, Jianhua Wan, Yi Ma, Jie Zhang, and Yabin Hu. 2020.
{``Characterization Analysis and Identification of Common Marine Oil
Spill Types Using Hyperspectral Remote Sensing.''} \emph{International
Journal of Remote Sensing} 41 (18): 7163--85.

\bibitem[\citeproctext]{ref-yousefi2016mineral}
Yousefi, Bardia, Saeed Sojasi, Clemente Ibarra Castanedo, Georges
Beaudoin, François Huot, Xavier PV Maldague, Martin Chamberland, and
Erik Lalonde. 2016. {``Mineral Identification in Hyperspectral Imaging
Using Sparse-PCA.''} In \emph{Thermosense: Thermal Infrared Applications
XXXVIII}, 9861:312--22. SPIE.

\bibitem[\citeproctext]{ref-zhang2022selection}
Zhang, Dingyu, Siyu Zeng, and Weiqi He. 2022. {``Selection and
Quantification of Best Water Quality Indicators Using UAV-Mounted
Hyperspectral Data: A Case Focusing on a Local River Network in Suzhou
City, China.''} \emph{Sustainability} 14 (23): 16226.

\bibitem[\citeproctext]{ref-zhang2019hyperspectral}
Zhang, Lefei, Liangpei Zhang, Bo Du, Jane You, and Dacheng Tao. 2019.
{``Hyperspectral Image Unsupervised Classification by Robust Manifold
Matrix Factorization.''} \emph{Information Sciences} 485: 154--69.

\end{CSLReferences}

\end{document}

### 2024-04-30 paper/outline.tex
adding updates to introduction

--- 2024-04-24 paper/outline.tex
@@
 
-NOTE: We should move the description of the GTM algroithm to the top of the methods section.
+NOTE: We should move the description of the GTM algorithm to the top of the methods section.
 
 \section*{Outline}
+We should recast the introduction to focus on unsupervised classification rather than endmember extraction...
+
 \begin{outline}[enumerate]

### 2024-04-30 paper/paper.tex
adding updates to introduction

--- 2024-04-24 paper/paper.tex
@@
 
+1. Introductory paragraph
+
+Inland water bodies pose a unique challenge to characterization via remote sensing imagery due to their complex spectral characteristics and small-scale spatial variability (cite reviews here).
+Hyperspectral imaging has emerged as a pivotal technology in this space, where increased spectral resolution enables the inversion of optically-active water quality parameters such as colored dissolved organic matter (CDOM), chlorophyll-a, blue-green algae, turbidity, and temperature from captured reflectance spectra \cite{koponen2002lake,ritchie2003remote, bonansea2015using}.
+Machine learning has proven to be a powerful tool in this domain, with tree-based models, support vector machines, deep neural networks, and others architectures providing data-driven approaches to map observed reflectance spectra to parameters of interest \cite{thenkabail2018hyperspectral, ghatkar2019classification,sagan2020monitoring}.
+However, the training and evaluation of these models demands significant volumes of coincident, in situ data.
+Collecting these datasets is time consuming and expensive, often requiring observations spanning decades \cite{aurin2018remote, ross2019aquasat}.
+
+2. HSI equipped drones + in situ reference sensing --> still require that we know what to look for.
+
+due to the dramatic improvement in spectral resolution afforded by HSI platforms.
+Specifically, the combination with hyperspectral imaging together with autonomous collection by UAVs enables unparalleled spectral, spatial, and temporal resolution.
+Significant effort has developed the use of hyperspectral imagery for the inversion of optically-active water quality parameters.
+However, this approach requires access to sufficient quantities of in situ reference data to fit and evaluate each model.
+Traditionally, this process has required collocation of remote sensing imagery with data from field campaigns spanning decades (cite Aurin and Ross).
+Recently, we have demonstrated how combining autonomous collection of in situ data with UAV-borne hyperspectral imagery dramatically reduces the time required to collect these data.
+Both of these approaches require we already know what to look for.
+
+Nevertheless, significant effort has driven the development of techniques for estimating optically-active water quality parameters such as chlorophyll-a, blue-green algae, colored dissolved organic matter (CDOM), turbidity, and temperature from these sources.
+
+To take complete advantage we need unsupervised methods which can aid in the identification of sources (in order to alert our attention for further investigation) AND to identify relvant spectral signatures corresponding to quantities of interest.
+UAV based HSI address the spectral, spatial, and temporal resolution limitations of remote sensing platforms however, the collection of reference data remains the key limitation hindering the real time characterization of inland waters.
+In our previous work, we demonstrated how a coordinated approach combining UAV-borne HSI together with autonomous collection in situ data by an Unmanned Surface Vessel (USV) dramatically reduces the time needed to perform such analyses \cite{robot-team-1, robot-team-2}.
+Nevertheless, these approaches all presuppose that we already know what to look for- contaminants which we are not equipped to measure in situ can not be estimated in this paradigm.
+Therefore, the identification of unique sources purely from their reflectance signatures is of keen interest for two key reasons.
+First: we want to be able to direct in situ sensor payloads to areas of interest.
+Second, we desire to identify relevant spectral signatures which can readily be related to parameters of interest and used for source identification.
+
+3. Unsupervised methods in remote sensing and HSI
+- Describe classical ML methods such as k-means, fuzzy c-means, DBSCAN, etc in remote sensing and HSI. 
+    - Key limitation: how do you determine the correct number of classes?
+    - This can be very hard to do without a Bayesian interpretation for the model. e.g. you have to rely on heuristics like the Silhouette diagram.
+- Describe related problem of endmember extraction and spectral unmixing.
+Specifically there are geometric and statistical approaches to solving the problem with ML methods a popular choice for the identification of the relvant family of spectral endmembers.
+- To date, much of the focus on drone-based HSI platforms has been on supervised regression of parameters OR the determination of spectral indices.
+- An ideal unsupervised approach should accomplish three things
+    1. Segmentation of captured imagery into a set of distinct classes
+    2. Mapping of distinct spatial regions in collected imagery using the identified classes
+    3. Simultaneous identification of relevant spectral endmembers
+
+4. Generative topographic mapping 
+Describe the self organizing map (SOM) as an interesting approach due to the ability to 
+    - classify data 
+    - maintain topographic relationship between classes which can be used to produces a lower dimensional representation of data distribution (i.e. cluster identification)
+    - the weight vector for each class can be interpreted as an endmember in the original data space such that common classes can be interpreted as a family of spectral endmembers.
+Cite a few uses ofthe SOM for remote sensing and HSI analysis...
+However, the original SOM training algorithm relies on heuristics for controlling the updates during training and offers no clear probabilitistic interpreation.
+The GTM was developed to specifically address this, providing a fully Bayesian alternative to the SOM.
+
+5. This study
+
+In this study, we explore the use of the GTM to perform rapid unsupervised classification of hyperspectral imagery captured by
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \end{figure}
-
-at a pond in Montague, north Texas (shown in panel d of Fig~\ref{fig:sample-spectra})
-
-\subsection{UAV-Based Hyperspectral Imaging}
-A Freefly Alta-X autonomous quadcopter was used as a UAV platform for the robotic team.
-The Alta-X is specifically designed to carry cameras and has a payload of up to 35 lbs.
-We equipped the UAV with a Resonon Pika XC2 visible+near-infrared (VNIR) hyperspectral imager.
-For each image pixel, this camera samples 462 wavelengths ranging from 391 to 1011 nm.
-Additionally, the UAV includes an upward facing Ocean Optics UV-Vis-NIR spectrometer with a cosine corrector to capture the incident downwelling irradiance spectrum.
-Data collection by the hyperspectral imager is controlled by an attached Intel NUC small-form-factor computer.
-A second processing NUC is also included for onboard georectification and generation of data products.
-The collected hyperspectral images (HSIs) are stored locally on a solid state drive that is simultaneously mounted by the processing computer.
-
-- georetification \cite{muller2002program, baumker2001new, mostafa2000multi}
-
-\subsection{Data Collection and Pre-Processing}
-- Julia programming language used \cite{bezanson2012julia}
 
@@
 - implementation designed to work in MLJ ecosystem \cite{blaom2020mlj}
+
+at a pond in Montague, north Texas (shown in panel d of Fig~\ref{fig:sample-spectra})
+
+\subsection{UAV-Based Hyperspectral Imaging}
+A Freefly Alta-X autonomous quadcopter was used as a UAV platform for the robotic team.
+
+The Alta-X is specifically designed to carry cameras and has a payload of up to 35 lbs. We equipped the UAV with a Resonon Pika XC2 visible+near-infrared (VNIR) hyperspectral imager.
+
+For each image pixel, this camera samples 462 wavelengths ranging from 391 to 1011 nm.
+
+Additionally, the UAV includes an upward facing Ocean Optics UV-Vis-NIR spectrometer with a cosine corrector to capture the incident downwelling irradiance spectrum.
+
+Data collection by the hyperspectral imager is controlled by an attached Intel NUC small-form-factor computer.
+
+A second processing NUC is also included for onboard georectification and generation of data products.
+
+The collected hyperspectral images (HSIs) are stored locally on a solid state drive that is simultaneously mounted by the processing computer.
+
+- georetification \cite{muller2002program, baumker2001new, mostafa2000multi}
+
+\subsection{Data Collection and Pre-Processing}
+- Julia programming language used \cite{bezanson2012julia}
 

### 2024-04-30 paper/paper.tex
added update equation for  beta
updated intro some more...

@@
 
-1. Introductory paragraph
+%   1. Introductory paragraph
 
-Inland water bodies pose a unique challenge to characterization via remote sensing imagery due to their complex spectral characteristics and small-scale spatial variability (cite reviews here).
-Hyperspectral imaging has emerged as a pivotal technology in this space, where increased spectral resolution enables the inversion of optically-active water quality parameters such as colored dissolved organic matter (CDOM), chlorophyll-a, blue-green algae, turbidity, and temperature from captured reflectance spectra \cite{koponen2002lake,ritchie2003remote, bonansea2015using}.
-Machine learning has proven to be a powerful tool in this domain, with tree-based models, support vector machines, deep neural networks, and others architectures providing data-driven approaches to map observed reflectance spectra to parameters of interest \cite{thenkabail2018hyperspectral, ghatkar2019classification,sagan2020monitoring}.
-However, the training and evaluation of these models demands significant volumes of coincident, in situ data.
-Collecting these datasets is time consuming and expensive, often requiring observations spanning decades \cite{aurin2018remote, ross2019aquasat}.
+Inland water bodies pose a unique challenge to characterization via remote sensing imagery due to their complex spectral characteristics and small-scale spatial variability \cite{koponen2002lake,ritchie2003remote}.
+Hyperspectral imaging has emerged as a pivotal technology in this space, where the heightened spectral resolution enables the inversion of optically-active water quality parameters such as colored dissolved organic matter (CDOM), chlorophyll-a, blue-green algae, turbidity, and temperature from captured reflectance spectra \cite{brando2003satellite,moses2012estimation}.
 
-2. HSI equipped drones + in situ reference sensing --> still require that we know what to look for.
+Machine learning has proven to be a powerful tool in this domain, with tree-based models, support vector machines, deep neural networks, and others architectures providing data-driven approaches to map observed reflectance spectra to parameters of interest \cite{thenkabail2018hyperspectral,keller2018hyperspectral, ghatkar2019classification,sagan2020monitoring}.
+However, the calibration and evaluation of these models demands a significant volume of coincident, in situ data.
+Collecting paired datasets is time consuming and expensive, often relying on rare satellite overpasses spanning decades \cite{aurin2018remote, ross2019aquasat}.
+This requirement impedes the real-time assessment of water quality change, for example, during oil spills \cite{fingas2014review,fingas2017review}.
+
+% 2. HSI equipped drones + in situ reference sensing --> still require that we know what to look for.
 
@@
 
-3. Unsupervised methods in remote sensing and HSI
+% 3. Unsupervised methods in remote sensing and HSI
 - Describe classical ML methods such as k-means, fuzzy c-means, DBSCAN, etc in remote sensing and HSI. 
@@
 
-4. Generative topographic mapping 
+% 4. Generative topographic mapping 
 Describe the self organizing map (SOM) as an interesting approach due to the ability to 
@@
 
-5. This study
+% 5. This study
 
@@
     \left(\Phi^T G \Phi + \dfrac{\alpha}{\beta}I \right)W_{\text{new}} = \Phi^T R X 
+\end{equation}
+
+\begin{equation}
+    \frac{1}{\beta_{\text{new}}} = \frac{1}{ND} \sum\limits_{n}^{N} \sum\limits_{k}^{K} R_{kn} \left\lVert \psi_k - \mathbf{x}_n \right\rVert^2
 \end{equation}

### 2024-04-30 paper/paper.tex
moved things from the draft intro to the main document

@@
 Inland water bodies pose a unique challenge to characterization via remote sensing imagery due to their complex spectral characteristics and small-scale spatial variability \cite{koponen2002lake,ritchie2003remote}.
-Hyperspectral imaging has emerged as a pivotal technology in this space, where the heightened spectral resolution enables the inversion of optically-active water quality parameters such as colored dissolved organic matter (CDOM), chlorophyll-a, blue-green algae, turbidity, and temperature from captured reflectance spectra \cite{brando2003satellite,moses2012estimation}.
-
-Machine learning has proven to be a powerful tool in this domain, with tree-based models, support vector machines, deep neural networks, and others architectures providing data-driven approaches to map observed reflectance spectra to parameters of interest \cite{thenkabail2018hyperspectral,keller2018hyperspectral, ghatkar2019classification,sagan2020monitoring}.
+Hyperspectral imaging has emerged as a pivotal technology in this space, where the \textcolor{purple}{heightened} spectral resolution of hyperspectral images (HSIs) enables the inversion of optically-active water quality parameters such as colored dissolved organic matter, chlorophyll-a, blue-green algae, turbidity, and temperature from captured reflectance spectra \cite{brando2003satellite,moses2012estimation}.
+Machine learning techniques including tree-based models, support vector machines, deep neural networks, and other architectures have proven effective at mapping observed reflectance spectra to parameters of interest \cite{thenkabail2018hyperspectral,keller2018hyperspectral, ghatkar2019classification,sagan2020monitoring}.
 However, the calibration and evaluation of these models demands a significant volume of coincident, in situ data.
-Collecting paired datasets is time consuming and expensive, often relying on rare satellite overpasses spanning decades \cite{aurin2018remote, ross2019aquasat}.
+Collecting these paired datasets is time consuming and expensive, often relying on rare satellite overpasses spanning decades of collection \cite{aurin2018remote, ross2019aquasat}.
 This requirement impedes the real-time assessment of water quality change, for example, during oil spills \cite{fingas2014review,fingas2017review}.
@@
 
-due to the dramatic improvement in spectral resolution afforded by HSI platforms.
-Specifically, the combination with hyperspectral imaging together with autonomous collection by UAVs enables unparalleled spectral, spatial, and temporal resolution.
-Significant effort has developed the use of hyperspectral imagery for the inversion of optically-active water quality parameters.
-However, this approach requires access to sufficient quantities of in situ reference data to fit and evaluate each model.
-Traditionally, this process has required collocation of remote sensing imagery with data from field campaigns spanning decades (cite Aurin and Ross).
-Recently, we have demonstrated how combining autonomous collection of in situ data with UAV-borne hyperspectral imagery dramatically reduces the time required to collect these data.
-Both of these approaches require we already know what to look for.
+The continued development of hyperspectral imaging technology has resulted in significant size reductions enabling the deployment of hyperspectral imagers on unmanned aerial vehicles (UAVs) such as drones.
+UAVs are not only less expensive than their remote sensing counterparts but also limit the need for complicated atmospheric corrections and enable centimeter scale spatial resolution through low altitude flights \cite{adao2017hyperspectral, arroyo2019implementation, banerjee2020uav}.
 
-Nevertheless, significant effort has driven the development of techniques for estimating optically-active water quality parameters such as chlorophyll-a, blue-green algae, colored dissolved organic matter (CDOM), turbidity, and temperature from these sources.
+- UAV-based HSI can be georectified to centimeter-scales without need for ground control points by using on-board GPS and IMU \cite{arroyo2019implementation}
+- Drones can be equipped with HSI and compute to enable rapid generation of spectral indices like the NDVI for applications such as precision agriculture \cite{horstrand2019uav}
+
+UAV-based hyperspectral imagers have been used estimate chlorophyll-a, suspended solids, and  
+Many such systems have been utilized for the characterization of water quality and composition (cite examples).
+\cite{lu2021retrieval}
+\cite{zhang2022selection}
+
+The collection of requisite in situ data can be significantly accelerated by coordinating UAV flights with reference-sensor equipped autonomous boats \cite{robot-team-1, robot-team-2}. 
+However, this approach requires prior knowledge of expected contaminants in order to select relevant sensors
+
+- Hyperspectral data were used to distinguish oils by type, e.g. crude, diesel, gasoline, and palm \cite{yang2020characterization}
+
+Endmember extraction and unmixing approaches can leverage the ability of hyperspectral images to distinguish source types.
+Example: mineral identification from AVRIS data, algae classification for harmful algal bloom detection, and distinguishing types of oils based on slight difference in spectral shapes.
 
@@
 
+Unsupervised classification
+- When no ground-truth data are available, unsupervised classification can still help partition imagery into groups or clusters.
+- Many data-driven, ML methods have been employed for the task including various matrix factorizations, k-nearest neighbors, fuzzy c-means, density estimation methods, etc.
+\cite{zhang2019hyperspectral}
+- Additionally unsupervised approaches can be used to perform nonlinear dimensionality reduction & pre-processing for supervised approaches.
+- Many unsupervised approaches can be challenging to interpret due to the lack of a relationship between the identified classes, e.g. is class 1 more similar to class 2 than class 10?
+- The Self Organizing Map is a popular technique which classifies data while enforcing topological relationships between the resulting classes such that classes close to each  other in the latent space are more similar than classes that are far apart.
+- SOM used for remote sensing imagery classification \cite{msom-remote-sensing}
+- SOM used for clustering and data compression of HSI cube-sat \cite{danielsen2021self}
+- SOM used for land-use and land-cover change analysis \cite{penfound2021analysis}
+- SOM has been used for identifying synoptic-scale patterns in wind and sea surface temperature data {som-satellite}
+
+endmember extraction, spectral unmixing
+- This is a related problem where the goal is to identify unique spectral signatures which combine (linearly or non-linearly) to produce the measured signal.
+- Having identified the set of endmembers, we then seek to determine their relative abundance in each pixel
+- Many statistical approaches for extracting endmembers \cite{berman2004ice}
+- Sparse PCA used for endmember extraction \cite{yousefi2016mineral}
+- Popular spectral unmixing approach is to treat "pure" endmembers as vertices of a simplex \cite{plaza2012endmember, nascimento2005vertex}
+- Convolutional Neural Networks with Autoencoder architectures are a popular ML approach which identify endmembers and perform non-linear unmixing \cite{palsson2020convolutional, non-negative-autoencoders, su2019daen, borsoi2019deep}
+- Self Organizing Map is another approach which has been used for endmember extraction together with neural networks for abundance mapping (unmixing) \cite{cantero2004analysis}
+
 % 4. Generative topographic mapping 
@@
 In this study, we explore the use of the GTM to perform rapid unsupervised classification of hyperspectral imagery captured by
+
+    - The generative topographic mapping is an attractive algorithm 
+        - Probabilistic interpretation
+        - Topographic relationship maintained between classes (not just orthogonality)
+        - clear fitting criterion with Maximum Likelihood Estimation / Maximum A-Posteriori Estimation
+    - We use the GTM to analyze hyperspectral imagery acquired by a drone over pond to 
+        - Present method for unsupervised characterization of small-scale variability of pond (i.e. identifying regions of interest which can be used to direct robot boat for further instructions) 
+        - Use GTM to extract endmember for algae and map the algal distribution near shore (relevant for Harmful Algal Bloom detection)
+        - Use GTM to extract endmember for rhodamine dye and use it to mape evolution of dye plume.
+This is a demonstration of source identification and mapping.
+Relevant for the identification of pollution events such as oil spills, industrial seepage, etc...
+
+To support and enhance real-time decision making, data-driven techniques...
+
+Using data collected at site in Montague, North Texas, we demonstrate the ability of this paradigm with two case studies.
+First, we demonstrate how the GTM can be used to provide a high-resolution unsupervised classification of a water body enabling the rapid identification of relevant regions which can then be used to direct further investigation by reference-sensor equipped boats.
+Second, we demonstrate the ability to perform endmember extraction and abundance mapping using imagery from a test-release of rhodamine dye.
+Together with the normalized spectral similarity score, these spectral signatures are identified and their associated area in the scene is determined.
 

### 2024-05-06 paper/draft-introduction.tex
updating the outline

--- 2024-04-23 paper/draft-introduction.tex
@@
 \usepackage[margin=0.75in]{geometry}
+\usepackage{outlines}
 
@@
 
-Inland water bodies pose a unique challenge for analysis with remote sensing imagery due to their complex spectral characteristics and small-scale spatial variability.
-Nevertheless, significant effort has driven the development of techniques for estimating optically-active water quality parameters such as chlorophyll-a, blue-green algae, colored dissolved organic matter (CDOM), turbidity, and temperature from these sources\cite{koponen2002lake,ritchie2003remote, bonansea2015using}.
-Machine learning has proven to be a powerful tool in this domain, with tree-based models, support vector machines, deep neural networks, and others architectures providing data-driven approaches mapping imagery to parameters of interest \cite{thenkabail2018hyperspectral, ghatkar2019classification,sagan2020monitoring}.
-However, the training and evaluation of these models demands significant volumes of coincident, in situ data.
-Collecting these datasets is time consuming and expensive, often requiring observations spanning decades \cite{aurin2018remote, ross2019aquasat}.
+Inland water bodies pose a unique challenge to characterization via remote sensing imagery due to their complex spectral characteristics and small-scale spatial variability \cite{koponen2002lake,ritchie2003remote}.
+Hyperspectral imaging has emerged as a pivotal technology in this space, where the heightened spectral resolution of hyperspectral images (HSIs) enables the inversion of optically-active water quality parameters such as colored dissolved organic matter, chlorophyll-a, blue-green algae, turbidity, and temperature from captured reflectance spectra \cite{brando2003satellite,moses2012estimation}.
+Machine learning techniques including tree-based models, support vector machines, deep neural networks, and other architectures have proven effective at mapping observed reflectance spectra to parameters of interest \cite{thenkabail2018hyperspectral,keller2018hyperspectral, ghatkar2019classification,sagan2020monitoring}.
+However, the calibration and evaluation of these models demands a significant volume of coincident, in situ data.
+Collecting these paired datasets is time consuming and expensive, often relying on rare satellite overpasses spanning decades of collection \cite{aurin2018remote, ross2019aquasat}.
+This requirement impedes the real-time assessment of water quality change, for example, during oil spills \cite{fingas2014review,fingas2017review}.
 
-To support and enhance real-time decision making, data-driven techniques...
+The continued development of hyperspectral imaging technology has resulted in significant size reductions enabling the deployment of hyperspectral imagers on unmanned aerial vehicles (UAVs) such as drones.
+UAVs are not only less expensive than their remote sensing counterparts but also limit the need for complicated atmospheric corrections and enable centimeter scale spatial resolution through low altitude flights \cite{adao2017hyperspectral, arroyo2019implementation, banerjee2020uav}.
+UAV-based hyperspectral imaging has seen significant development for use in precision agriculture where the ability to rapidly asses spectral indices like the normalized difference vegetation index (NDVI) is a valuable tool for assessing crop health and irrigation efficiency \cite{horstrand2019uav}.
+Recently, UAV-based hyperspectral imaging has been utilized to asses a variety of optically active water quality parameters such as chlorophyll-a and suspended solids \cite{lu2021retrieval, zhang2022selection}.
+In our previous work, we introduced a coordinated robot team designed to enable near-real time characterization of water composition by synchronizing UAV-based HSI collection with in situ data collection by an autonomous boat \cite{robot-team-1, robot-team-2}.
+However, one key
 
-The characterization of inland water bodies by remote sensing imagery poses a distinct challenge due to their complex spectral characteristics and small-scale spatial variability \cite{ritchie2003remote}.
-Traditional multi-spectral imagers utilized by these sensing platforms have large pixel sizes, limited spectral bands, and long revisit times.
-- Nevertheless, ... 
-- Machine learning has proven... 
-- However, these techniques demand significant volumes of in situ data...
-- To address the
+rests with the availability reference sensors for each parameter of interest. That is, the ability to model water quality parameters depends on
 
-Hyperspectral imaging
-- Hyperspectral imaging has emerged as a pivotal technology in this space.
-- With their increased spectral resolution, hyperspectral imagers (HSIs) capture the complex chemical, biological, environmental characteristics of reflectance spectra.
-(add references for specific applications of HSIs).
-- However HSIs alone do not address the limited spatial resolution of
+In other words, which sensors to deploy depends on prior knowledge of expected contaminants-of-concern.
 
-UAV
-- Advancements in hyperspectral imaging technology have lead to a continual decrease in their size facilitating their inclusion as the payload of small, highly maneuverable unmanned aerial vehicles (UAVs) such as drones.
-Consequently, much work has gone into the development of UAV platforms and associated techniques
-- Near-earth HSI addresses spatial, spectral, and temporal limitations of satellite and airborne platforms. UAV-based HSI enable fine-scale mapping \cite{banerjee2020uav}
-- Drones can be equipped with HSI and compute to enable rapid generation of spectral indices like the NDVI for applications such as precision agriculture \cite{horstrand2019uav}
-- UAV-based HSI can be georectified to centimeter-scales without need for ground control points by using on-board GPS and IMU \cite{arroyo2019implementation}
-- UAV-based HSI for turbidity estimation \cite{vogt2016near}
-- Polynomial regression models for chlorophyll-a, turbidity \cite{zhang2022selection}
-- For this to be usefull, we still need the in situ data.
-By combining a UAV born HSI with simultanous collection of ground truth data by a reference sensor equipped autonomous boat, we are able to rapidly train ML models to
+% 3. Unsupervised methods in remote sensing and HSI
 
-Unsupervised Methods
-- classic unsupervised ML
-- endmember extraction and spectral unmixing
+To take complete advantage of acquired HSI, and in order to guide where to deploy reference sensors, the ability to
 
-This paper...
-- our approach is to use the GTM (a fully Bayesian realization of the self organizing map developed by Biship et al.)
-- We demonstrate the capabilities using hyperspectral imagery collected at pond in Montague, North Texas. 
-- First we utilize the GTM to perform an unsupervised classification of the water-only pixels in order to reveal the centimeter-scale structures present in the pond
-- Next we showhow the GTM can be used to identify relevent spectral endmembers by 
-    1. using learned GTM class signatures to identity the abundance of algae (relevant for Harmful Algal Bloom detection)
-    2. use the GTM class signatures to map the evolution of a rhodamine tracer dye plume
+we need unsupervised methods which can aid in the identification of sources (in order to alert our attention for further investigation) AND to identify relevant spectral signatures corresponding to quantities of interest.
 
-- \cite{koponen2002lake}
-- Example: distinguishing species for harmful algal blooms (HABs) and distinguishing types of oil for oil spill monitoring
-- Remote sensing used for oil spill analysis: extent and thickness mapping \cite{kokaly2013spectroscopic, leifer2012state}
+UAV based HSI address the spectral, spatial, and temporal resolution limitations of remote sensing platforms however, the collection of reference data remains the key limitation hindering the real time characterization of inland waters.
 
-- Sun glitter remains a key challenge for sensing in the visible portion of the spectrum but multi and hyperspectral imagers have been used for oil spill identification and to identify their impacts on vegetation stress and mortality \cite{fingas2014review, khan2018modern}
+In our previous work, we demonstrated how a coordinated approach combining UAV-borne HSI together with autonomous collection in situ data by an Unmanned Surface Vessel (USV) dramatically reduces the time needed to perform such analyses \cite{robot-team-1, robot-team-2}.
 
-Continued development of HSI technology has led to a dramatic reduction in their size enabling their utilize as the payload of small autonomous aerial vehicles (UAVs) (citation).
-By flying close to the ground UAV-born hyperspectral imagers can produce centimeter-scale imagery while simultaneously limiting the need for complicated atmospheric corrections required for satellite platforms (citation).
+Nevertheless, these approaches all presuppose that we already know what to look for- contaminants which we are not equipped to measure in situ can not be estimated in this paradigm.
 
-- applications include food quality & safety, medical diagnoses, precision agriculture, and forensic document examination \cite{khan2018modern}
+Therefore, the identification of unique sources purely from their reflectance signatures is of keen interest for two key reasons.
+First: we want to be able to direct in situ sensor payloads to areas of interest.
+Second, we desire to identify relevant spectral signatures which can readily be related to parameters of interest and used for source identification.
+
+Endmember extraction and unmixing approaches can leverage the ability of hyperspectral images to distinguish source types.
+Example: mineral identification from AVRIS data, algae classification for harmful algal bloom detection, and distinguishing types of oils based on slight difference in spectral shapes.
+
+- Describe classical ML methods such as k-means, fuzzy c-means, DBSCAN, etc in remote sensing and HSI. 
+    - Key limitation: how do you determine the correct number of classes?
+    - This can be very hard to do without a Bayesian interpretation for the model. e.g. you have to rely on heuristics like the Silhouette diagram.
+- Describe related problem of endmember extraction and spectral unmixing.
+Specifically there are geometric and statistical approaches to solving the problem with ML methods a popular choice for the identification of the relevant family of spectral endmembers.
+- To date, much of the focus on drone-based HSI platforms has been on supervised regression of parameters OR the determination of spectral indices.
+- An ideal unsupervised approach should accomplish three things
+    1. Segmentation of captured imagery into a set of distinct classes
+    2. Mapping of distinct spatial regions in collected imagery using the identified classes
+    3. Simultaneous identification of relevant spectral endmembers
+
 - Hyperspectral data were used to distinguish oils by type, e.g. crude, diesel, gasoline, and palm \cite{yang2020characterization}
+- add citation here about harmful algal blooms as examples
 
@@
 
-In our previous work, we introduced a robot team designed to rapidly characterize water composition in novel environments \cite{robot-team-1}. 
-Joining UAV-borne HSI together with collection of in-situ reference data by an autonomous boat dramatically reduces the time required to generate .
+% 4. Generative topographic mapping 
+Describe the self organizing map (SOM) as an interesting approach due to the ability to 
+    - classify data 
+    - maintain topographic relationship between classes which can be used to produces a lower dimensional representation of data distribution (i.e. cluster identification)
+    - the weight vector for each class can be interpreted as an endmember in the original data space such that common classes can be interpreted as a family of spectral endmembers.
+Cite a few uses ofthe SOM for remote sensing and HSI analysis...
+However, the original SOM training algorithm relies on heuristics for controlling the updates during training and offers no clear probabilitistic interpreation.
+The GTM was developed to specifically address this, providing a fully Bayesian alternative to the SOM.
 
-By carefully coordinating the simultaneous collection of ground-truth water quality data and hyperspectral imagery, this team can rapidly estimate concentrations of optically-active components including colored dissolved organic matter (CDOM), blue-green algae, crude oil, temperature as well
+% 5. This study
 
-In this paper, we demonstrate how Generative Topographic Mapping (GTM), a Bayesian alternative to the Self Organizing Map, can be utilized to rapidly identify relevant spectral signatures within a scene.
+In this study, we explore the use of the GTM to perform rapid unsupervised classification of hyperspectral imagery captured by
+
+    - The generative topographic mapping is an attractive algorithm 
+        - Probabilistic interpretation
+        - Topographic relationship maintained between classes (not just orthogonality)
+        - clear fitting criterion with Maximum Likelihood Estimation / Maximum A-Posteriori Estimation
+    - We use the GTM to analyze hyperspectral imagery acquired by a drone over pond to 
+        - Present method for unsupervised characterization of small-scale variability of pond (i.e. identifying regions of interest which can be used to direct robot boat for further instructions) 
+        - Use GTM to extract endmember for algae and map the algal distribution near shore (relevant for Harmful Algal Bloom detection)
+        - Use GTM to extract endmember for rhodamine dye and use it to mape evolution of dye plume.
+This is a demonstration of source identification and mapping.
+Relevant for the identification of pollution events such as oil spills, industrial seepage, etc...
+
+To support and enhance real-time decision making, data-driven techniques...
 
@@
 
-In this study, we demonstrate how Generative Topographic Mapping (GTM), a Bayesian extension the Self Organizing Map, can be utilized to
-
-The method is fully Bayesian, providing an attractive method for the unsupervised classification of hyperspectral imagery.
-
-- First, we apply the GTM to perform an unsupervised classification for all water pixels. 
-    - benefit: GTM classes maintain topological relationships in the latent space resulting in a smooth relationship between the obtained classes...
-    - ... revealing the small-scale spatial variability within the pond
-- Second, we utilize the GTM to identify reflectance spectra corresponding to specific features. 
-- Using these extracted features, we are able to map the extent of algae in the pond in addition to the time evolution of a rhodamine dye tracer.
-
 \bibliographystyle{plain}

### 2024-05-06 paper/outline.tex
updating the outline

--- 2024-04-24 paper/outline.tex
@@
 
-NOTE: We should move the description of the GTM algroithm to the top of the methods section.
-
-\section*{Outline}
+\section*{Introduction}
 \begin{outline}[enumerate]
-\1 Introduction
-    \2 Characterization of inland water quality via remote sensing is challenging for 4 key reasons:
-        \3 Key limitations:
-            \4 Spectral resolution (visible and multi-spectral imaging)
-            \4 Spatial Resolution (large pixel sizes)
-            \4 Temporal Resolution (long satellite revisit times)
-            \4 Availability of in situ reference data
-        \3 When appropriate data are available, supervised ML approaches can map reflectance from remote sensing imagery to relevant water quality parameters
-    \2 Hyperspectral Imaging addresses spectral resolution
-    \2 UAV adresses spatial and temporal resolution
-    \2 Robot team addresses in situ data... but what do we do if we don't already know which sensors to use? 
-    \2 Unsupervised ML, Endmember extraction, and Spectral Unmixing
-        \3 k-NN, fuzzy c-means, matrix factorizations, etc.
-            \4 Are general but do not provide relationship between classes e.g. is class 1 more similar to class 2 or class 10?
-        \3 SOM addresses this by enforcing a topological relationship between clusters. Classes close to each other in the map are more similar than classes far apart
-        \3 Similarly, we may also be interested in the spectral characteristics of the learned *classes*, i.e. we want to think of the classes as *endmembers*.
-            \4 Traditional approaches rely on assumption of existence of "pure pixels" in the scene
-            \4 additionally, we may have to treat number of endmembers as a hyperparameter
-
-    \2. ...in this paper, we explore how the generative topographic mapping can be used to
-        \3 enable principled unsupervised classification of hyperspectral imagery by
-            \4 Bayesian treatment (as compared to the self organizing map)
-            \4 topological ordering with smooth relationship between classes in latent space
-        \3 Train GTM on water pixels to obtain unsupervised classification mapping the detailed structure and small-scale variability of pond in Montague, North Texas
-        \3 Mapping can be further utilized for smart deployment of robot team (maybe save for discussion section?) i.e. requisition boat to bring reference sensors to "interesting" area
-        \3 Using combined dataset including land, water, and tracer dye, we demonstrate endmember extraction
-        \3 Using learned spectra together with NS3, we are able to evaluate extent of algae (application to harmful algal bloom mapping)
-        \3 Use learned spectra to map extent
-
-\1 Materials and Methods
-    \2 Overview paragraph about project, i.e.
-plan is to use the Generative Topographic Mapping to analyze drone-based hyperspectral imagery collected at a pond in Montague North Texas on 23-November 2021.
-   \2 Autonomous Robot Team
-        \3 Describe drone setup
-            \4 Alta X Quadcopter
-            \4 Resonon Pika SC2 HSI
-                \begin{itemize}
-                    \item 462 wavelengths per pixel
-                    \item 1600 pixels per scan-line
-                    \item embedded GPS/IMU
-               \end{itemize}
-             \4 Intel NUC for on-board compute
-   \2 Data Collection
-        \3 Describe collection site in Montague, North Texas
-        \3 Collection was performed near solar noon to maximize incident light.
-At this time of year, the maximum solar elevation was $\approx 60^\circ$ so that sun-glint is not as much of an issue.
-(Cite sun-glint paper here).
-    \2 Pre-processing
-        \3 Reflectance conversion of radiance imagery using on-board downwelling irradiance spectrometer
-        \3 Georectification of collected
-        \3 Wavelength selection to $\lambda \leq 900 \text{nm}$ as longer wavelengths are noisy
-        \3 Rescaling of spectra so that peak has $\tilde{R}  = 1.0$  to account for variability in incident lighting conditions due 
-    \2 Generative Topographic Mapping
-        \3 Start off with Kohonen SOM paper. Key benefit:
-            \4 Topological relationship between classes i.e. similarity vs dissimilarity as a function of distance in mapping space (latent space)
-        \3 Limitations of SOM
-            \4 Not probabilistic
-            \4 No clear convergence criterion
-            \4 Heuristic based approach with no clear way to select hyperparameters like neighborhood function, number of classes, etc...
-            \4 Heskes did reinterpret algorithm as stochastic gradient descent on a cost function...
-        \3 GTM Algorithm
-            \4 Describe Algorithm
-            \4 Describe parameters
-        \3 Bayesian information criterion and Akaike information criterion for hyperaprameter optimization
-            \4 As an alternative to performing full Bayesian treatment for marginalization over hyperparameters
-        \3 Abundance mapping with spectral similarity score
-            \4 paper comparing different distance functions for spectra
-                \begin{itemize}
-                    \item curse of dimensionality can be pernicious here as we have > 462 wavelengths per pixel
-                \end{itemize}
-            \4 NS3 offers good trade off between spectral angle and euclidean-like distance (MSE)
-        \3 Describe use of thresholded NS3 for mapping abundance of GTM derived endmembers.
-    \2 Overview of two approaches
-        \3 GTM trained on water-only pixels for unsupervised classification of pond
-            \4 Goal: Reveal the small-scale spatial variability of pond water
-            \4 Discuss how unsupervised classification can help inform autnomous data collection by robot team (save for discussion?)
-        \3 GTM trained on combined dataset including water, land, and dye plume pixels
-            \4 Fit GTM
-            \4 Identify spectral signatures of key exemplar spectra (endmembers)
-            \4 Use endmembers to map algal abundance and time evolution of rhodamine dye plume
-\1 Results
-    \2  GTM fit on Full Dataset
-    \2 Hyperparameter optimization
-    \2 Spectral signature identification
-    \2 Water Class Map (water-only GTM)
-    \2 Algae Identification
-    \2 Dye plume identification
-
-\1 Discussion
-    \2 Overview of GTM applications in our paper
-        \3 Unsupervised classifciation of entire pond (no-dye)
-        \3 Endmember extraction of spectral signatures
-    \2 Comparison to existing approaches
-        \3 unsupervised classification methods (to general, no relationship between classes) 
-        \3 endmember extraction (mostly linear, assume existence of pure spectra)
-        \3 deep learning approaches to endmember extraction don't provide clear way to asses number of endmembers needed
-    \2 Applications of either of these with drones?
-        \3 Are there any papers using unsupervised methods or endmember extraction with drone based imagery (HSI or otherwise)?
-    \2. Use of GTM as component of Robot Team
-        \3 As a feature transformer / preprocessing step for supervised methods (as opposed to PCA, k-NN, etc...)
-        \3 identification of "interesting" regions for smart provisioning of robot boat
-            \4 Standard approach is to survey in a grid
-            \4 Can use classification map with real-time measurement to ensure optimal collection of data by collecting samples with from as many GTM classes as possible
-    \2 Extensions and future work
-        \3 Incorporation of GTM extensions for efficient online-learning
-            \4 Comment on time to train standard GTM
-                \begin{itemize}
-                    \item Key parameters are size of dataset (number of records) and number of RBF centers, $M$.
-                \end{itemize}
-            \4 Bishop suggest extension for on-line training of GTM using batches.
-            \4 All of this could be done in the field limiting the time and cost of surveying as well as optimizing the time-to-insights
-        \3 Utilize GTM with further in situ data collection to identify spectral signatures of specific algal species for HAB detection
-        \3 Same but for crude oil types...
-
-\1 Conclusions
+\1 Challenges for water quality characterization
+    \2 Traditional remote sensing suffers from limited spatial, spectral, and temporal resolution
+        \3 Papers: 
+            \4 Broad bands of traditional remote sensing multi-spectral imagers struggle to discern water quality parameters like chlorophyll-a in highly turbid waters due to domintation of spectral signal by sediments.
+\cite{ritchie2003remote}
+            \4 Small, irregular shape of Finnish lakes makes analysis using medium resolution satellite data challenging due to reflectance from shore and near-shore vegetation.
+"If even a small protion of a pixel is covered by land the retrieval of water quality variables may not be possible" \cite{koponen2002lake}
+    \2 This can be addressed by combining two technologies: 
+        \3 Hyperspectral Imaging (improved spectral resolution)
+        \3 Unmanned Aerial Vehicles (improves spatial and temporal resolution)
+        \3 Studies using UAV + HSI for water quality: "UAV-based HSI has emerged as a promising tool..."
+            \4 UAV-based HSI addresses resolution, signal-to-noise, and cloud cover issues that plague traditional satellite and high-altitude platforms\cite{banerjee2020uav}
+            \4 UAV-based HSI for ecological monitoring \cite{arroyo2019implementation}
+    \2 Remaining challenge: how can we extract meaningful insights from complicated spectral data?
+        \3 "Ability for measuring hundreds of bands raises complexity when considering the sheer quantity of acquired data, whose usefulness depends on both calibration and corrective tasks occruing in pre- and post-flight stages" \cite{adao2017hyperspectral}
+\1 Current approaches for characterization
+    \2 Band ratios and ML for water quality parameter inversion
+        \3 Papers:
+            \4 UAV-based HSI with embedding processing capabiltieis enables rapid generation of spectral indices like the NDVI \cite{horstrand2019uav}
+            \4 UAV-based multispectral imaging used to map turbidity (Secchi Disk Depth) \cite{vogt2016near}
+            \4 Polynomial combinations of HSI bands from UAV-mounted HSI imagery used to map turbidity, total nitrogen, total phosphorus, ammonia nitrogen, permanganate index, and chemical oxygen demand \cite{zhang2022selection}
+            \4 ML methods used to map UAV-based HSI to chlorophyll-a and suspended solids. Tree-based methods worked best (Catboost) \cite{lu2021retrieval}
+            \4 Estimation of CDOM, Chlorophyll-a, turbidity, diatoms, and green-algae from HSI using Machine Learning with hyperspectral sensor and two in-situ sensor probes \cite{keller2018hyperspectral}
+            \4 ICA and SVM used for HSI classification (not water) \cite{liang2010hyperspectral}
+            \4 SVA and LDA used for HSI classification (not water) \cite{zhangSVM2014}
+    \2 Key limitation: need for in situ reference data
+        \3 For remote sensing applications, collection of datasets relies on serendipitous satellite overpasses.
+Acquiring sufficient volumes of data needed to calibrate and evaluate inversion models can require decades of observations \cite{aurin2018remote,ross2019aquasat}
+        \3 Can be addressed by extending the UAV approach to coordinated robot teams including autonomous in situ sensing \cite{robot-team-1, robot-team-2}
+    \2 Remaining limitation: insights are limited by available reference sensors 
+        \3 Sensor require careful calibration, for example using specific standards which may not be relevant for local water
+    \2 Need: unsupervised methods to extract meaningful information from full HSI data
+        \3 Useful for deciding where/how to collect reference data
+        \3 Useful for time-critical applications where identification of anomolous sources is important e.g. oil spills, industrial waste, etc... 
+\1 Unsupervised ML Approaches
+    \2 Dimensionality reduction
+        \3 PCA is a popular choice but is a linear model
+            \4 Mineral endmembers identified from HSI using sparse PCA \cite{yousefi2016mineral}
+        \3 t-SNE can be computationally complex 
+    \2 Unsupervised Classification
+        \3 k-NN, fuzzy c-means, DBSCAN, etc... matrix factorizations
+        \3 Can be helpful for identifying clusters, but don't aid in visualization by themselves
+            \4 Robust Manifold Matrix Factorization used for dimensionality reduction \textit{and} clustering \cite{zhang2019hyperspectral}
+    \2 Endmember extraction and Spectral Unmixing
+        \3 Geometric and statistical methods
+        \3 Can be useful for identifying relevant endmembers but may rely on presence of pure pixels
+            \4 This assumption may not be valid for water leafing reflectance spectra due to interactions of suspended solids, dissolved components, etc...
+        \3 Once endmembers are identified, abundance maps can be produced using spectral angle mapper (e.g. cosine distance) or unmixing models
+            \4 However, this does not itself to visualization
+    \2 Need: a (principled) method which simultaneously enables dimensionality reduction \textit{and} classification/endmember extraction.
+\1 Self Organizing Map and Generative Topographic Mapping
+    \2 Self Organizing map developed by Kohonen based on principle that \textit{neurons that fire together, wire together}.
+        \3 Construct an unsupervised classification into $N=n\times n$ classes
+        \3 Classes are arranged in a regular grid of typically two dimensions with different possible topologies:
+            \4 square vs hexagonal lattice (number of neighbors)
+            \4 rectangular, cylindrical, toroidal topologies based on boundary conditions
+        \3 The map is trained to enforce similarity between neighboring nodes, that is, classes near each other in the SOM grid are more similar than nodes that are far apart. 
+        \3 In the context of HSI and water quality, the SOM grid can be used to perform dimensionality reduction via the location of each datapoint in the SOM grid 
+        \3 SOM nodes can \textit{also} be interpreted as endmembers/classes
+            \4 Add citations here, e.g. David's dust sources paper and other SOM related HSI studies
+            \4 SOM used for endmember extraction with HSI for water quality \cite{cantero2004analysis}
+    \2 Drawbacks of SOM
+        \3 Training algorithm is based on heuristic approach
+        \3 Does not provide a clear probabilistic interpretation or a principled approach for hyperparameter choice
+    \2 The Generative Topographic Mapping addresses these issues
+        \3 Fully Bayesian re-interpretation of the SOM
+        \3 Enables both dimensionality reduction \textit{and} classification like the SOM
+        \3 Uses of the GTM in HSI:
+\1 Paper overview
+    \2 Development of the GTM as an efficient approach for dimensionality reduction and unsupervised classification of UAV-acquired HSI
+    \2 Apply the GTM to data collected at a pond in Montague, North Texas
+    \2 Train a GTM on water-only pixels to provide detailed segmentation water spectra
+    \2 Use GTM on combined dataset for demonstration of endmember extraction and source mapping
+        \3 Algal abundance (i.e. application to algal bloom)
+            \4 Comment on potential application for algal species discrimination
+        \3 Rhodamine Dye Tracer simulating dispersion of pollution source
+            \4 Application to oil spill identification
 
 \end{outline}
+
+\section*{Materials and Methods}
+\section*{Results}
+\section*{Discussion}
+\section*{Conclusions}
 

### 2024-05-06 paper/paper.tex
updating the outline

--- 2024-04-30 paper/paper.tex
@@
 
-1. Introductory paragraph
-
-Inland water bodies pose a unique challenge to characterization via remote sensing imagery due to their complex spectral characteristics and small-scale spatial variability (cite reviews here).
-Hyperspectral imaging has emerged as a pivotal technology in this space, where increased spectral resolution enables the inversion of optically-active water quality parameters such as colored dissolved organic matter (CDOM), chlorophyll-a, blue-green algae, turbidity, and temperature from captured reflectance spectra \cite{koponen2002lake,ritchie2003remote, bonansea2015using}.
-Machine learning has proven to be a powerful tool in this domain, with tree-based models, support vector machines, deep neural networks, and others architectures providing data-driven approaches to map observed reflectance spectra to parameters of interest \cite{thenkabail2018hyperspectral, ghatkar2019classification,sagan2020monitoring}.
-However, the training and evaluation of these models demands significant volumes of coincident, in situ data.
-Collecting these datasets is time consuming and expensive, often requiring observations spanning decades \cite{aurin2018remote, ross2019aquasat}.
-
-2. HSI equipped drones + in situ reference sensing --> still require that we know what to look for.
-
-due to the dramatic improvement in spectral resolution afforded by HSI platforms.
-Specifically, the combination with hyperspectral imaging together with autonomous collection by UAVs enables unparalleled spectral, spatial, and temporal resolution.
-Significant effort has developed the use of hyperspectral imagery for the inversion of optically-active water quality parameters.
-However, this approach requires access to sufficient quantities of in situ reference data to fit and evaluate each model.
-Traditionally, this process has required collocation of remote sensing imagery with data from field campaigns spanning decades (cite Aurin and Ross).
-Recently, we have demonstrated how combining autonomous collection of in situ data with UAV-borne hyperspectral imagery dramatically reduces the time required to collect these data.
-Both of these approaches require we already know what to look for.
-
-Nevertheless, significant effort has driven the development of techniques for estimating optically-active water quality parameters such as chlorophyll-a, blue-green algae, colored dissolved organic matter (CDOM), turbidity, and temperature from these sources.
-
-To take complete advantage we need unsupervised methods which can aid in the identification of sources (in order to alert our attention for further investigation) AND to identify relvant spectral signatures corresponding to quantities of interest.
-UAV based HSI address the spectral, spatial, and temporal resolution limitations of remote sensing platforms however, the collection of reference data remains the key limitation hindering the real time characterization of inland waters.
-In our previous work, we demonstrated how a coordinated approach combining UAV-borne HSI together with autonomous collection in situ data by an Unmanned Surface Vessel (USV) dramatically reduces the time needed to perform such analyses \cite{robot-team-1, robot-team-2}.
-Nevertheless, these approaches all presuppose that we already know what to look for- contaminants which we are not equipped to measure in situ can not be estimated in this paradigm.
-Therefore, the identification of unique sources purely from their reflectance signatures is of keen interest for two key reasons.
-First: we want to be able to direct in situ sensor payloads to areas of interest.
-Second, we desire to identify relevant spectral signatures which can readily be related to parameters of interest and used for source identification.
-
-3. Unsupervised methods in remote sensing and HSI
-- Describe classical ML methods such as k-means, fuzzy c-means, DBSCAN, etc in remote sensing and HSI. 
-    - Key limitation: how do you determine the correct number of classes?
-    - This can be very hard to do without a Bayesian interpretation for the model. e.g. you have to rely on heuristics like the Silhouette diagram.
-- Describe related problem of endmember extraction and spectral unmixing.
-Specifically there are geometric and statistical approaches to solving the problem with ML methods a popular choice for the identification of the relvant family of spectral endmembers.
-- To date, much of the focus on drone-based HSI platforms has been on supervised regression of parameters OR the determination of spectral indices.
-- An ideal unsupervised approach should accomplish three things
-    1. Segmentation of captured imagery into a set of distinct classes
-    2. Mapping of distinct spatial regions in collected imagery using the identified classes
-    3. Simultaneous identification of relevant spectral endmembers
-
-4. Generative topographic mapping 
-Describe the self organizing map (SOM) as an interesting approach due to the ability to 
-    - classify data 
-    - maintain topographic relationship between classes which can be used to produces a lower dimensional representation of data distribution (i.e. cluster identification)
-    - the weight vector for each class can be interpreted as an endmember in the original data space such that common classes can be interpreted as a family of spectral endmembers.
-Cite a few uses ofthe SOM for remote sensing and HSI analysis...
-However, the original SOM training algorithm relies on heuristics for controlling the updates during training and offers no clear probabilitistic interpreation.
-The GTM was developed to specifically address this, providing a fully Bayesian alternative to the SOM.
-
-5. This study
-
-In this study, we explore the use of the GTM to perform rapid unsupervised classification of hyperspectral imagery captured by
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
+In this study, we explore the use of Generative Topographic Mapping as a tool for the unsupervised classification and nonlinear endmember extraction of hyperspectral imagery.
+To this end, HSIs were collected using a USV mounted hyperspectral imager at a pond in Montague, North Texas on on 23 November 2020 (shown in panel (d) of Fig~\ref{fig:sample-spectra}).
+In this section we shall first provide a detailed description of the GTM algorithm with subsequent subsections reserved to describe the specific data collection, processing, and analysis techniques employed in the study.
+
+%%% --- Generative 0Topographic Mapping ------------------------------------
 \subsection{Generative Topographic Mapping}
+
+The Generative Topographic Mapping (GTM) was developed by Bishop, Svens\'en, and Williams as a Bayesian realization of Kohonen's Self-organizing Map (SOM) \cite{gtm-bishop-1}. Specifically, the
 
@@
 The GTM takes inspiration from Kohonen's self-organizing map which enables an unsupervised classification of high dimensional data into a (typically) two-dimensional grid of nodes such that nodes near each near each other in the mapping space
+
+As outlined in the introduction, an effective unsupervised classification algorithm for hyperspectral imagery should accomplish the following three things: 
+\begin{enumerate}
+    \item Segmentation of captured imagery into a set of distinct classes
+    \item Mapping of distinct spatial regions in collected imagery using the identified classes
+    \item Simultaneous identification of relevant spectral endmembers
+\end{enumerate}
 
@@
 - Use of GTM classes to aid autonomous data collection by identification of "interesting" regons
+
+- Drones can be equipped with HSI and compute to enable rapid generation of spectral indices like the NDVI for applications such as precision agriculture \cite{horstrand2019uav}
 

### 2024-05-06 paper/outline.tex
updated outline

@@
 \usepackage{outlines}
+\usepackage{amsmath}
 
@@
     \2 Unsupervised Classification
-        \3 k-NN, fuzzy c-means, DBSCAN, etc... matrix factorizations
+        \3 k-NN, fuzzy c-means, DBSCAN, matrix factorizations, Autoencoders
         \3 Can be helpful for identifying clusters, but don't aid in visualization by themselves
             \4 Robust Manifold Matrix Factorization used for dimensionality reduction \textit{and} clustering \cite{zhang2019hyperspectral}
+            \4 Use a convolutional autoencoder with softmax to perfrom endmember extraction and unmixing \cite{palsson2020convolutional}
+            \4 Combine stacked autoencoders and a variational autoencoder for unmixing \cite{su2019daen}
+            \4 Develop a VAE model for unmixing of HSI \cite{borsoi2019deep}
     \2 Endmember extraction and Spectral Unmixing
         \3 Geometric and statistical methods
+            \4 statistical approach \cite{berman2004ice}
+            \4 vertex component analysis: endmembers are vertices of simplex. Note these methods still generally require us to know the number of endmembers in advance \cite{nascimento2005vertex}
+            \4 Non-negative matrix factorization is a popular approach using matrix decomposition into non-negative endmember and abundance matrices.
+constrained NMF generally requires extensive parameter tuning and generalized NMF takes a long time to compute and rely on assumption of linear mixing \cite{Feng2022HyperspectralUB}
         \3 Can be useful for identifying relevant endmembers but may rely on presence of pure pixels
             \4 This assumption may not be valid for water leafing reflectance spectra due to interactions of suspended solids, dissolved components, etc...
-        \3 Once endmembers are identified, abundance maps can be produced using spectral angle mapper (e.g. cosine distance) or unmixing models
+        \3 Once endmembers are identified, abundance maps can be produced using spectral angle mapper (e.g. cosine distance) or unmixing assuming a mixing model
             \4 However, this does not itself to visualization
-    \2 Need: a (principled) method which simultaneously enables dimensionality reduction \textit{and} classification/endmember extraction.
+            \4 Common assumption is linear mixing however strong nonlinear mixing effects are present in many situations \cite{heylen2014review}
+        \3 The increased spatial resolution affored by UAV platforms suggests UAV-based HSI platforms are ideal for endmember extraction as likelihood of pure endmembers in UAV-based HSI is signifcantly higher than in satelliteet imagery
+            \4 Use unmixing of UAV-hsi to discover endmembers for satellite hsi \cite{alvarez2020can}
+            \4 Use unmixing of UAV-hsi as endmember library for satellite HSI \cite{gu2023intrinsic}
+    \2 Goal: a (principled) method which simultaneously enables dimensionality reduction \textit{and} classification/endmember extraction.
 \1 Self Organizing Map and Generative Topographic Mapping
-    \2 Self Organizing map developed by Kohonen based on principle that \textit{neurons that fire together, wire together}.
+    \2 Self Organizing map developed by Kohonen based on principle that \textit{neurons that fire together, wire together}. \cite{kohonen-som-1}
         \3 Construct an unsupervised classification into $N=n\times n$ classes
@@
             \4 SOM used for endmember extraction with HSI for water quality \cite{cantero2004analysis}
+            \4 SOM classes used to classify synoptic scale wind and surface temperature data from satellite observations \cite{som-satellite}
+            \4 SOM for feature selection and dimensionality reduction of HSI \cite{som-hsi}
+            \4 SOM for classification of remote sensing imagery \cite{msom-remote-sensing}
+            \4 SOM for land cover change detection from remote sensing imagery \cite{penfound2021analysis}
+            \4 Use SOM for on-board compression of HSI from CubeSat \cite{danielsen2021self}
     \2 Drawbacks of SOM
@@
     \2 The Generative Topographic Mapping addresses these issues
-        \3 Fully Bayesian re-interpretation of the SOM
+        \3 Fully Bayesian re-interpretation of the SOM \cite{gtm-bishop-1, gtm-bishop-2}
         \3 Enables both dimensionality reduction \textit{and} classification like the SOM
@@
 \section*{Materials and Methods}
+
+\begin{outline}[enumerate]
+\1 Overview Paragraph
+\1 Description of GTM Algorithm and Implementation
+    \2 Algorithm  Description
+    \2 Hyperparameter selection and model comparison with the Bayesian Information Criterion rather than full Bayesian treatment (e.g. marginalization over model hyperparameters)
+    \2 Implementation 
+        \3 julia programming language \cite{bezanson2012julia}
+        \3
+\1 Description of UAV + HSI System
+    \2 Description of Drone
+        \3 Alta X Quadcopter
+        \3 Resonon Pika SC2 Hyperspectral Imager
+            \4 462 Wavelengths per pixel
+            \4 1600 Pixels per scanline
+            \4 Embedded GPS/IMU
+        \3 Intel NUC for real-time processing of collected imagery
+    \2 Reflectance Conversion 
+        \3 citation from last paper
+    \2 Georectification
+        \3 Georectification papers \cite{muller2002program, baumker2001new, mostafa2000multi}
+        \3 MLJ package \cite{blaom2020mlj}
+        \3 GTM package implementation \cite{GenerativeTopographicMapping.jl}
+\1 Data Collection
+    \2 Data collected at pond in Montague, North Texas
+        \3 Data collected 23 November 2020
+        \3 Collection performed near solar noon cooresponding to a maximum solar elevation of $\approx 60^\circ$ to maximize incident light. At this angle, sunglint was not a signficant problem
+        \3 First flight with no dye
+        \3 Two subsequent flights after Rhodamine dye release immediately after and 15 minutes after
+    \2 Data preprocessing
+        \3 Wavelengths limited to $\lambda \leq 900$ nm as longer wavelengths were noisy
+        \3 All pixels scaled to a maximum value of $1.0$ to account for incident light variability
+\1 GTM Study
+    \2 GTM trained on water-only pixels    
+        \3 Use resulting GTM to produce detailed map of segmented pixel spectra
+    \2 GTM trained on combined dataset
+        \3 Hyperparamer optimization using BIC
+        \3 Endember identification using exemplar pixel spectra for water, grass, rhodamine, and algae
+        \3 Mapping endmembers using NS3 
+            \4 spectral distance functions evaluated. There is a trade-off between ability to distinguish difference is hue (wavelength shifts) and intensity (peak size)\cite{deborah2015comprehensive}
+            \4 spectral angle mapper is a popular technique for comparison of candidate spectra using the angle as defined by the standard euclidean dot-product
+            \4 spectral correlation mapper uses correlation instead as an improvement \cite{de2000spectral}
+            \4 Normalized Spectral Similarity Score (NS3) is suggested by \cite{nidamanuri2010normalized} to provide a compromise between the Euclidean-like RMSE and the spectral angle with 
+                \begin{align}
+                    \text{RMSE}(R_1, R_2) &= \sqrt{\frac{1}{N-1}\sum_\lambda \left(R_{1}(\lambda) - R_2(\lambda) \right)^2} \\
+                    \cos\theta &= \frac{\langle R_1 , R_2 \rangle}{\lVert R_1\rVert \lVert R_2 \rVert} \\
+                    \text{NS3}(R_1, R_2) &= \sqrt{RMSE(R_1, R_2)^2 + (1-\cos\theta)^2}
+                \end{align}
+            \4 Algae using NS3
+            \4 Mapping Rhodmaine tracer evolution using NS3
+\end{outline}
+
 \section*{Results}
+\begin{outline}[enumerate]
+\1 GTM fit on water-only pixels
+\1 GTM fit on combined dataset
+    \2 Hyperparameter optimization
+    \2 Map in Latent Space
+    \2 Endmember extraction using exemplar spectra 
+    \2 Algae Mapping
+    \2 Rhodamine Mapping
+\end{outline}
+
 \section*{Discussion}
+
+\begin{outline}[enumerate]
+\1 Overview of GTM applications introduced in the paper
+    \2 Dimensionality reduction
+        \3 Latent space representation
+        \3 GTM classes plotted on background map
+    \2 Nonlinear endmember extraction
+        \3 Algae 
+        \3 Rhodamine
+\1 Comparison to existing approaches
+    \2 Unsupervised classification
+        \3 Most do not enforce a relationship between classes .e.g. if we use k-NN, is class 1 "closer" to class 2 than to class 10? 
+    \2 Endmember extraction 
+        \3 most approaches assume linear relationship between classes, e.g. NMF, etc... 
+        \3 GTM nodes are non-linearly mapped from latent space to data space via the map $\psi$
+    \2 Deep learning approaches
+        \3 Time intensive and require you to decide on number of endmembers in network architecture
+        \3 VAE do enable probabilistic approach which is nice...
+\1 Applications of either of these with drones 
+    \2 Increased interest in using drone-based HSI to build endmember libraries due to finer spatial resolution potentially limiting the degree of mixing within each pixel
+    \2 Few papers yet directly doing unmixing of drone-based HSI
+\1 Use of GTM as part of a coordinated robot team for water quality and composition
+    \2 As a non-linear feature transformer / preprocessing step for supervised methods (as opposed to PCA or k-NN) 
+    \2 Classification maps can be used to identify "interesting regions" for intelligent provisioning of robot boat.
+        \3 Without any prior knowledge, the standard approach would be to send out robotic boat in a grid to collect data
+        \3 GTM class map outlines spatial variability within pond and be used to desing smart sampling strategy. That is, construct a route so as to maximize sampled class variability 
+            \4 Since the ultimate goal in this modality is to collect training data for drone, we don't want to spend time collecting "useless" data in regions that are spatially uniform
+\1 Extensions and future work
+    \2 Incorportation of extensions to GTM algorithim from \cite{gtm-bishop-2} for efficient online training of GTM.
+        \3 If optimized, this could be performed on the drone in near real time.
+        \3 Class maps are significantly smaller than full HSI and can be streamed to ground station quickly
+    \2 Use  GTM framework and additional datacollection to explore potential for 
+        \3 discrimination of Algal speices from reflectance spectra for Harmful Algal Bloom (HAB) identification
+            \4 Algae species classified from remote snesing imagery using tree based models\cite{ghatkar2019classification}
+        \3 Discrimination of industrial pollution sources, e.g. for oilspill detection and mapping
+            \4 Oil types can be distinguished from hyperspectral imagery using SVM into categories such as crude oil, fuel, diesel, gasoline, and palm oil \cite{yang2020characterization}
+\end{outline}
+
 \section*{Conclusions}

### 2024-05-07 paper/paper.tex
updated intro with second paragraph

@@
 \section{Introduction}
+
+% 1. Traditional Strategies for Remote Sensing
+
+Inland water bodies pose a unique challenge to characterization by remote sensing imagery due to their complex spectral characteristics and small scale spatial variability.
+The broad bands of multi-spectral imagers combined with the irregular shape of lakes and rivers results in pixels with highly mixed signals that are easily dominated by reflectance from shore and near-shore vegetation sources \cite{koponen2002lake, ritchie2003remote}.
+Recently, the combination of hyperspectral imaging with unmanned aerial vehicles (UAVs) such as drones has emerged as a powerful approach to simultaneously address the spectral, spatial, and temporal limitations of traditional high-altitude and satellite-based collection \cite{adao2017hyperspectral,arroyo2019implementation}.
+UAVs are significantly less expensive to deploy than their remote sensing counterparts, and low altitude flights enable centimeter-scale sampling while limiting the need for complicated atmospheric corrections \cite{banerjee2020uav}.
+However, the significant increase in data volume generated by these systems presents a new challenge, namely, how to efficiently extract water quality parameters of interest from intricate pixel spectra.
+
+% 2. Current Approaches
+
+Significant research efforts have focused on the deveħopment of techniques and algorithms to invert water quality parameters from UAV-captured hyperspectral images (HSIs).
+On-board compute installed alongside hyperspectral imagers can enable the rapid evaluation of spectral indices from HSI band ratios \cite{horstrand2019uav}.
+These band ratios and polynomial combinations of bands have been used to successfully invert optically-active water quality parameters such as turbidity directly from UAV acquired imagery \cite{vogt2016near, zhang2022selection}.
+Supervised machine learning techniques such as tree-based models, support vector machines, and neural networks have also been used to estimate a plethora of parameters such as colored dissolved organic matter, chlorophyll a, blue-green algae, and suspended sediment concentrations \cite{keller2018hyperspectral, lu2021retrieval}.
+The calibration and evaluation of these data-driven models demand a significant volume of coincident, in situ data.
+This can be addressed by coordinating UAV flights with reference data collection by autonomous robotic boats \cite{robot-team-1, robot-team-2}.
+Nevertheless, this approach relies on prior knowledge of expected sources in order to select appropriate reference instruments for model validation.
+The presence of unanticipated contaminants cannot be directly identified in this paradigm.
+
+% 3. Unsupervised ML
+To complement these capabilities...
+
+Therefore, unsupervised techniques to segment HSI and identify unique spectral signatures are needed for near-real time applications where detailed knowledge of contaminants may not be available
+
+% 4. SOM and GTM
+
+% 5. This paper...
 

### 2024-05-07 paper/paper.tex
updated introduction further...

@@
 % 3. Unsupervised ML
-To complement these capabilities...
 
-Therefore, unsupervised techniques to segment HSI and identify unique spectral signatures are needed for near-real time applications where detailed knowledge of contaminants may not be available
+To take full advantage of the information contained in HSI for near-real time water quality analysis, unsupervised methods that segment HSI pixels and can extract relevant spectral endmembers are needed.
+
+A variety of technqiues have
+
+Data-driven techniques such as principal component analysis (PCA), independent component analysis (ICA), and k-nearest neighbors (kNN) have been widely employed for HSI analysis \cite{yousefi2016mineral,} (other citations).
+While useful, these methods
+
+Geometric methods such as vertex component analysis (VCA) offer another approach based on the fact that endmember spectra form vertices of a simplex \cite{}. However, these approaches
+
+Deep learning strategies such as autoencoders...
+
+Principal component analysis (PCA) has been widely used for this task as it enables both dimensionality reduction and feature extraction.
+
+Principal component analysis (PCA) is a popular choice for both dimenionsality reduction and feature extraction.
+
+Principal component analysis (PCA) is a common method utilized for both dimensionality reduction and feature extraction
+
+Dimensionality Reduction
+Unsupervised Classification
+Endmember Extraction and Spectral Unmixing
+
+Goal: a principled unsupervised method that can both segment HSI imagery and aid in the extraction of relevant spectral endmembers
 

### 2024-05-07 paper/paper.tex
Added sample paragraph 3

--- 2024-05-07 paper/paper.tex
@@
 % 3. Unsupervised ML
-To complement these capabilities...
 
-Therefore, unsupervised techniques to segment HSI and identify unique spectral signatures are needed for near-real time applications where detailed knowledge of contaminants may not be available
+% To take full advantage of the information contained in HSI for near-real time water quality analysis, unsupervised methods that segment HSI pixels and can extract relevant spectral endmembers are needed.
+
+% A variety of technqiues have
+
+% Data-driven techniques such as principal component analysis (PCA), independent component analysis (ICA), and k-nearest neighbors (kNN) have been widely employed for HSI analysis \cite{yousefi2016mineral,} (other citations).
+While useful, these methods
+
+% Geometric methods such as vertex component analysis (VCA) offer another approach based on the fact that endmember spectra form vertices of a simplex \cite{}. However, these approaches
+
+% Deep learning strategies such as autoencoders...
+
+% Principal component analysis (PCA) has been widely used for this task as it enables both dimensionality reduction and feature extraction.
+
+% Principal component analysis (PCA) is a popular choice for both dimenionsality reduction and feature extraction.
+
+% Principal component analysis (PCA) is a common method utilized for both dimensionality reduction and feature extraction
+
+Unsupervised machine learning (ML) approaches have been widely explored for hyperspectral image (HSI) analysis, including dimensionality reduction, unsupervised classification, and endmember extraction and spectral unmixing.
+Dimensionality reduction techniques, such as principal component analysis (PCA) and t-Distributed Stochastic Neighbor Embedding (t-SNE), aim to represent the high-dimensional HSI data in a lower-dimensional space while preserving relevant information.
+However, PCA is a linear model, and t-SNE can be computationally complex.
+Unsupervised classification methods, including k-nearest neighbors (k-NN), fuzzy c-means, DBSCAN, matrix factorizations, and autoencoders, can identify clusters in the data, but do not directly aid in visualization.
+Robust Manifold Matrix Factorization has been used for both dimensionality reduction and clustering, while convolutional autoencoders with softmax have been employed for endmember extraction and unmixing.
+Additionally, stacked autoencoders and variational autoencoders (VAEs) have been combined for unmixing tasks.
+Endmember extraction and spectral unmixing techniques, such as geometric and statistical methods, vertex component analysis, and non-negative matrix factorization (NMF), have been widely explored.
+While these methods can identify relevant endmembers, they often rely on the presence of pure pixels or require prior knowledge of the number of endmembers, which may not be valid in certain scenarios.
+Once endmembers are identified, abundance maps can be produced using techniques like spectral angle mapper or unmixing assuming a mixing model, but these do not directly facilitate visualization.
+The increased spatial resolution afforded by unmanned aerial vehicle (UAV) platforms suggests that UAV-based HSI is ideal for endmember extraction, as the likelihood of pure endmembers is significantly higher than in satellite imagery.
+Consequently, unmixing of UAV-HSI has been used to discover endmembers for satellite HSI or as an endmember library.
+The goal is to develop a principled method that simultaneously enables dimensionality reduction and classification/endmember extraction for HSI analysis.
 
 % 4. SOM and GTM
+
+Among unsupervised methods, Kohonen's self-organizing map provides an attractive compromise enabling both dimensionality reduction and classification...
 

### 2024-05-08 paper/draft-introduction.tex
updated intro with second paragraph
updated introduction

--- 2024-04-23 paper/draft-introduction.tex
@@
 
-Inland water bodies pose a unique challenge for analysis with remote sensing imagery due to their complex spectral characteristics and small-scale spatial variability.
-Nevertheless, significant effort has driven the development of techniques for estimating optically-active water quality parameters such as chlorophyll-a, blue-green algae, colored dissolved organic matter (CDOM), turbidity, and temperature from these sources\cite{koponen2002lake,ritchie2003remote, bonansea2015using}.
-Machine learning has proven to be a powerful tool in this domain, with tree-based models, support vector machines, deep neural networks, and others architectures providing data-driven approaches mapping imagery to parameters of interest \cite{thenkabail2018hyperspectral, ghatkar2019classification,sagan2020monitoring}.
-However, the training and evaluation of these models demands significant volumes of coincident, in situ data.
-Collecting these datasets is time consuming and expensive, often requiring observations spanning decades \cite{aurin2018remote, ross2019aquasat}.
+Inland water bodies present a unique challenge to characterization by remote sensing imagery due to their complex spectral characteristics and small scale spatial variability.
+The broad bands of multi-spectral imagers coupled with the irregular shape of lakes and rivers results in pixels with highly mixed signals that are easily dominated by reflectance from shore and near-shore vegetation sources \cite{koponen2002lake, ritchie2003remote}.
+Recently, the combination of hyperspectral imaging with unmanned aerial vehicles (UAV) such as drones has emerged as a powerful approach to simultaneously address the spectral, spatial, and temporal limitations of traditional high-altitude and satellite-based collection \cite{adao2017hyperspectral,arroyo2019implementation}.
+UAV are significantly less expensive to deploy than their remote sensing counterparts, and low altitude flights enable centimeter-scale sampling while limiting the need for complicated atmospheric corrections \cite{banerjee2020uav}.
+However, the significant increase in data volume generated by these systems presents a new challenge, namely, how to efficiently extract water quality parameters of interest from intricate pixel spectra.
 
-To support and enhance real-time decision making, data-driven techniques...
+Extensive research efforts have been dedicated to the development of techniques and algorithms to retrieve water quality parameters from UAV-captured hyperspectral images (HSI).
+On-board compute installed alongside hyperspectral imagers can enable the rapid evaluation of spectral indices from HSI band ratios \cite{horstrand2019uav}.
+These band ratios and polynomial combinations of bands have been used to successfully invert optically-active water quality parameters such as turbidity directly from UAV acquired imagery \cite{vogt2016near, zhang2022selection}.
+Supervised machine learning techniques such as tree-based models, support vector machines, and neural networks have also been used to estimate a plethora of parameters such as colored dissolved organic matter, chlorophyll a, blue-green algae, and suspended sediment concentrations \cite{keller2018hyperspectral, lu2021retrieval}.
+The calibration and evaluation of these data-driven models demand a significant volume of coincident, in situ data.
+This can be addressed by coordinating UAV flights with reference data collection using autonomous robotic boats \cite{robot-team-1, robot-team-2}.
+Nevertheless, this approach relies on prior knowledge of expected sources in order to select appropriate reference instruments for model validation.
+The presence of unanticipated contaminants cannot be directly identified in this paradigm.
 
-The characterization of inland water bodies by remote sensing imagery poses a distinct challenge due to their complex spectral characteristics and small-scale spatial variability \cite{ritchie2003remote}.
-Traditional multi-spectral imagers utilized by these sensing platforms have large pixel sizes, limited spectral bands, and long revisit times.
-- Nevertheless, ... 
-- Machine learning has proven... 
-- However, these techniques demand significant volumes of in situ data...
-- To address the
+% Unsupervised methods
 
-Hyperspectral imaging
-- Hyperspectral imaging has emerged as a pivotal technology in this space.
-- With their increased spectral resolution, hyperspectral imagers (HSIs) capture the complex chemical, biological, environmental characteristics of reflectance spectra.
-(add references for specific applications of HSIs).
-- However HSIs alone do not address the limited spatial resolution of
+To address this limitation, unsupervised methods that segment HSI pixels and can identify unique spectral signatures are needed.
 
-UAV
-- Advancements in hyperspectral imaging technology have lead to a continual decrease in their size facilitating their inclusion as the payload of small, highly maneuverable unmanned aerial vehicles (UAVs) such as drones.
-Consequently, much work has gone into the development of UAV platforms and associated techniques
-- Near-earth HSI addresses spatial, spectral, and temporal limitations of satellite and airborne platforms. UAV-based HSI enable fine-scale mapping \cite{banerjee2020uav}
-- Drones can be equipped with HSI and compute to enable rapid generation of spectral indices like the NDVI for applications such as precision agriculture \cite{horstrand2019uav}
-- UAV-based HSI can be georectified to centimeter-scales without need for ground control points by using on-board GPS and IMU \cite{arroyo2019implementation}
-- UAV-based HSI for turbidity estimation \cite{vogt2016near}
-- Polynomial regression models for chlorophyll-a, turbidity \cite{zhang2022selection}
-- For this to be usefull, we still need the in situ data.
-By combining a UAV born HSI with simultanous collection of ground truth data by a reference sensor equipped autonomous boat, we are able to rapidly train ML models to
+These methods generally fall into two categories: dimensionality reduction and endmember extraction.
+For example, principle component analysis (PCA) and t-distributed stochastic neighbor embedding (tSNE) make it possible to reduce high dimensional HSI into two or three dimensions where they can be visualized and compared \cite{}.
 
-Unsupervised Methods
-- classic unsupervised ML
-- endmember extraction and spectral unmixing
+In remote sensing, where reference data is sparse, many approaches have been explored and generally fall into two categories: dimensionality reduction, and unsupervised classification and endmember extraction.
 
-This paper...
-- our approach is to use the GTM (a fully Bayesian realization of the self organizing map developed by Biship et al.)
-- We demonstrate the capabilities using hyperspectral imagery collected at pond in Montague, North Texas. 
-- First we utilize the GTM to perform an unsupervised classification of the water-only pixels in order to reveal the centimeter-scale structures present in the pond
-- Next we showhow the GTM can be used to identify relevent spectral endmembers by 
-    1. using learned GTM class signatures to identity the abundance of algae (relevant for Harmful Algal Bloom detection)
-    2. use the GTM class signatures to map the evolution of a rhodamine tracer dye plume
+Algorithms such as principal component analysis (PCA) and t-distributed stochastic neighbor embedding (t-SNE) enable the visualization of HSI by reducing their dimensionality \cite{yousefi2016mineral}.
 
-- \cite{koponen2002lake}
-- Example: distinguishing species for harmful algal blooms (HABs) and distinguishing types of oil for oil spill monitoring
-- Remote sensing used for oil spill analysis: extent and thickness mapping \cite{kokaly2013spectroscopic, leifer2012state}
+including dimensionality reduction techniques like principal component analysis (PCA) and clustering algorithms such as k-nearest neighbors \cite{yousefi2016mineral,cariou2015unsupervised}.
+Non-negative matrix factorization is another popular approach which seeks to simultaneously extract endmembers with their associated abundances via a constrained matrix decomposition \cite{Feng2022HyperspectralUB}.
+Similarly, the advent of deep learning has led to a profusion of neural network based models utilizing an autoencoder architecture to enable nonlinear endmember extraction \cite{su2019daen,borsoi2019deep,palsson2020convolutional}.
+Other methods such as vertex component analysis (VCA) are tailored specifically for hyperspectral data based on the observation that endmember spectra form the vertices of a simplex \cite{nascimento2005vertex}.
+Ultimately, an ideal approach for near-real time analysis of water quality from HSI should enable both visualization via dimensionality reduction and identification of relevant spectral endmembers.
 
-- Sun glitter remains a key challenge for sensing in the visible portion of the spectrum but multi and hyperspectral imagers have been used for oil spill identification and to identify their impacts on vegetation stress and mortality \cite{fingas2014review, khan2018modern}
+% 4. SOM
 
-Continued development of HSI technology has led to a dramatic reduction in their size enabling their utilize as the payload of small autonomous aerial vehicles (UAVs) (citation).
-By flying close to the ground UAV-born hyperspectral imagers can produce centimeter-scale imagery while simultaneously limiting the need for complicated atmospheric corrections required for satellite platforms (citation).
+- SOM is an attractive approach as it enables both dimensionality reduction and unsupervised classification. 
+- Furthermore, the SOM maintains a topological relationship between learned classes such that classes near each other in the SOM grid are more similar than classes far apart
+- SOM can be used to nonlinearly map data into a 2-dimensional space
+- Weight vectors corresponding to each SOM node can be interpreted as spectral endmembers.
+The training algorithm updates weights based on euclidean distance between data record and each node weight vector
+- Add citations and examples of use in remote sensing / HSI
+-
 
-- applications include food quality & safety, medical diagnoses, precision agriculture, and forensic document examination \cite{khan2018modern}
-- Hyperspectral data were used to distinguish oils by type, e.g. crude, diesel, gasoline, and palm \cite{yang2020characterization}
+Among unsupervised methods, Kohonen's self-organizing map provides an attractive compromise enabling both dimensionality reduction and classification...
 
-Unsupervised classification
-- When no ground-truth data are available, unsupervised classification can still help partition imagery into groups or clusters.
-- Many data-driven, ML methods have been employed for the task including various matrix factorizations, k-nearest neighbors, fuzzy c-means, density estimation methods, etc.
-\cite{zhang2019hyperspectral}
-- Additionally unsupervised approaches can be used to perform nonlinear dimensionality reduction & pre-processing for supervised approaches.
-- Many unsupervised approaches can be challenging to interpret due to the lack of a relationship between the identified classes, e.g. is class 1 more similar to class 2 than class 10?
-- The Self Organizing Map is a popular technique which classifies data while enforcing topological relationships between the resulting classes such that classes close to each  other in the latent space are more similar than classes that are far apart.
-- SOM used for remote sensing imagery classification \cite{msom-remote-sensing}
-- SOM used for clustering and data compression of HSI cube-sat \cite{danielsen2021self}
-- SOM used for land-use and land-cover change analysis \cite{penfound2021analysis}
-- SOM has been used for identifying synoptic-scale patterns in wind and sea surface temperature data {som-satellite}
+% 5. GTM
 
-endmember extraction, spectral unmixing
-- This is a related problem where the goal is to identify unique spectral signatures which combine (linearly or non-linearly) to produce the measured signal.
-- Having identified the set of endmembers, we then seek to determine their relative abundance in each pixel
-- Many statistical approaches for extracting endmembers \cite{berman2004ice}
-- Sparse PCA used for endmember extraction \cite{yousefi2016mineral}
-- Popular spectral unmixing approach is to treat "pure" endmembers as vertices of a simplex \cite{plaza2012endmember, nascimento2005vertex}
-- Convolutional Neural Networks with Autoencoder architectures are a popular ML approach which identify endmembers and perform non-linear unmixing \cite{palsson2020convolutional, non-negative-autoencoders, su2019daen, borsoi2019deep}
-- Self Organizing Map is another approach which has been used for endmember extraction together with neural networks for abundance mapping (unmixing) \cite{cantero2004analysis}
-
-In our previous work, we introduced a robot team designed to rapidly characterize water composition in novel environments \cite{robot-team-1}. 
-Joining UAV-borne HSI together with collection of in-situ reference data by an autonomous boat dramatically reduces the time required to generate .
-
-By carefully coordinating the simultaneous collection of ground-truth water quality data and hyperspectral imagery, this team can rapidly estimate concentrations of optically-active components including colored dissolved organic matter (CDOM), blue-green algae, crude oil, temperature as well
-
-In this paper, we demonstrate how Generative Topographic Mapping (GTM), a Bayesian alternative to the Self Organizing Map, can be utilized to rapidly identify relevant spectral signatures within a scene.
+% 6. This paper...
 

### 2024-05-08 paper/paper.tex
updated introduction

--- 2024-05-07 paper/paper.tex
@@
 \section{Introduction}
-
-% 1. Traditional Strategies for Remote Sensing
-
-Inland water bodies pose a unique challenge to characterization by remote sensing imagery due to their complex spectral characteristics and small scale spatial variability.
-The broad bands of multi-spectral imagers combined with the irregular shape of lakes and rivers results in pixels with highly mixed signals that are easily dominated by reflectance from shore and near-shore vegetation sources \cite{koponen2002lake, ritchie2003remote}.
-Recently, the combination of hyperspectral imaging with unmanned aerial vehicles (UAVs) such as drones has emerged as a powerful approach to simultaneously address the spectral, spatial, and temporal limitations of traditional high-altitude and satellite-based collection \cite{adao2017hyperspectral,arroyo2019implementation}.
-UAVs are significantly less expensive to deploy than their remote sensing counterparts, and low altitude flights enable centimeter-scale sampling while limiting the need for complicated atmospheric corrections \cite{banerjee2020uav}.
-However, the significant increase in data volume generated by these systems presents a new challenge, namely, how to efficiently extract water quality parameters of interest from intricate pixel spectra.
-
-% 2. Current Approaches
-
-Significant research efforts have focused on the deveħopment of techniques and algorithms to invert water quality parameters from UAV-captured hyperspectral images (HSIs).
-On-board compute installed alongside hyperspectral imagers can enable the rapid evaluation of spectral indices from HSI band ratios \cite{horstrand2019uav}.
-These band ratios and polynomial combinations of bands have been used to successfully invert optically-active water quality parameters such as turbidity directly from UAV acquired imagery \cite{vogt2016near, zhang2022selection}.
-Supervised machine learning techniques such as tree-based models, support vector machines, and neural networks have also been used to estimate a plethora of parameters such as colored dissolved organic matter, chlorophyll a, blue-green algae, and suspended sediment concentrations \cite{keller2018hyperspectral, lu2021retrieval}.
-The calibration and evaluation of these data-driven models demand a significant volume of coincident, in situ data.
-This can be addressed by coordinating UAV flights with reference data collection by autonomous robotic boats \cite{robot-team-1, robot-team-2}.
-Nevertheless, this approach relies on prior knowledge of expected sources in order to select appropriate reference instruments for model validation.
-The presence of unanticipated contaminants cannot be directly identified in this paradigm.
-
-% 3. Unsupervised ML
-To complement these capabilities...
-
-Therefore, unsupervised techniques to segment HSI and identify unique spectral signatures are needed for near-real time applications where detailed knowledge of contaminants may not be available
-
-% 4. SOM and GTM
-
-% 5. This paper...
 

### 2024-05-13 paper/draft-introduction.tex
updating draft intro

@@
 
-Extensive research efforts have been dedicated to the development of techniques and algorithms to retrieve water quality parameters from UAV-captured hyperspectral images (HSI).
+Significant research efforts have focused on the development of techniques and algorithms to retrieve water quality parameters from UAV-captured hyperspectral images (HSI).
 On-board compute installed alongside hyperspectral imagers can enable the rapid evaluation of spectral indices from HSI band ratios \cite{horstrand2019uav}.
@@
 Nevertheless, this approach relies on prior knowledge of expected sources in order to select appropriate reference instruments for model validation.
-The presence of unanticipated contaminants cannot be directly identified in this paradigm.
+The presence of unanticipated contaminants cannot be directly identified in this sensing paradigm.
 
-% Unsupervised methods
+Extending the capabilities of UAV-based hyperspectral imaging to enable water quality monitoring in real-time scenarios where contaminant sources may not be known in advance requires two additional capabilities: dimensionality reduction techniques to permit the visual comparison of HSI, and endmember extraction techniques which can identify spectral signatures corresponding to unique sources within the imaging scene.
+In remote sensing where reference data are typically sparse, many such techniques have been developed.
+Principal component analysis and t-distributed stochastic neighbor embedding are popular dimensionality reduction approaches commonly used to reduce HSI to two or three dimensions for visualization \cite{tyo2003principal,zhang2015hyperspectral}.
+For endmember extraction there are a variety of approaches including geometric methods like vertex component analysis, statistical methods like k-nearest neighbors and non-negative matrix factorization, and deep learning methods based on autoencoder architectures \cite{heylen2014review,nascimento2005vertex, Feng2022HyperspectralUB, cariou2015unsupervised, su2019daen, borsoi2019deep, palsson2020convolutional}.
+Ultimately, an ideal approach should enable both visualization and identification of relevant spectral endmembers.
 
-To address this limitation, unsupervised methods that segment HSI pixels and can identify unique spectral signatures are needed.
+Among the numerous unsupervised approaches, Kohonen's self-organizing map (SOM) is an attractive option which 
+simultaneously accomplishes both tasks.
 
-These methods generally fall into two categories: dimensionality reduction and endmember extraction.
-For example, principle component analysis (PCA) and t-distributed stochastic neighbor embedding (tSNE) make it possible to reduce high dimensional HSI into two or three dimensions where they can be visualized and compared \cite{}.
+provides a low dimensional representation of data into
 
-In remote sensing, where reference data is sparse, many approaches have been explored and generally fall into two categories: dimensionality reduction, and unsupervised classification and endmember extraction.
+SOM have been used
 
-Algorithms such as principal component analysis (PCA) and t-distributed stochastic neighbor embedding (t-SNE) enable the visualization of HSI by reducing their dimensionality \cite{yousefi2016mineral}.
+However, the SOM suffers from key limitations
 
-including dimensionality reduction techniques like principal component analysis (PCA) and clustering algorithms such as k-nearest neighbors \cite{yousefi2016mineral,cariou2015unsupervised}.
-Non-negative matrix factorization is another popular approach which seeks to simultaneously extract endmembers with their associated abundances via a constrained matrix decomposition \cite{Feng2022HyperspectralUB}.
-Similarly, the advent of deep learning has led to a profusion of neural network based models utilizing an autoencoder architecture to enable nonlinear endmember extraction \cite{su2019daen,borsoi2019deep,palsson2020convolutional}.
-Other methods such as vertex component analysis (VCA) are tailored specifically for hyperspectral data based on the observation that endmember spectra form the vertices of a simplex \cite{nascimento2005vertex}.
-Ultimately, an ideal approach for near-real time analysis of water quality from HSI should enable both visualization via dimensionality reduction and identification of relevant spectral endmembers.
+\cite{cantero2004analysis} -- use SOM for endmember extraction
+\cite{som-satellite} -- SOM For synoptic scale wind and sea surface temperature patterns
+\cite{som-hsi} -- SOM for HSI feature extraction
+\cite{danielsen2021self} -- SOM for compression of HSI from cubesat
+
+To address these shortcomings, Bishop, Svensen, and () developted the Generative Topographic mapping (GTM) \cite{gtm-bishop-1}. The GTM is a fully Bayesian realization of the SOM which
+
+In this paper, we
 
@@
 
-- SOM is an attractive approach as it enables both dimensionality reduction and unsupervised classification. 
-- Furthermore, the SOM maintains a topological relationship between learned classes such that classes near each other in the SOM grid are more similar than classes far apart
-- SOM can be used to nonlinearly map data into a 2-dimensional space
-- Weight vectors corresponding to each SOM node can be interpreted as spectral endmembers.
+% - SOM is an attractive approach as it enables both dimensionality reduction and unsupervised classification. 
+% - Furthermore, the SOM maintains a topological relationship between learned classes such that classes near each other in the SOM grid are more similar than classes far apart
+% - SOM can be used to nonlinearly map data into a 2-dimensional space
+% - Weight vectors corresponding to each SOM node can be interpreted as spectral endmembers.
 The training algorithm updates weights based on euclidean distance between data record and each node weight vector
-- Add citations and examples of use in remote sensing / HSI
--
+% - Add citations and examples of use in remote sensing / HSI
+% -
 
-Among unsupervised methods, Kohonen's self-organizing map provides an attractive compromise enabling both dimensionality reduction and classification...
+% Among unsupervised methods, Kohonen's self-organizing map provides an attractive compromise enabling both dimensionality reduction and classification...
 
@@
 
-Using data collected at site in Montague, North Texas, we demonstrate the ability of this paradigm with two case studies.
+% Using data collected at site in Montague, North Texas, we demonstrate the ability of this paradigm with two case studies.
 First, we demonstrate how the GTM can be used to provide a high-resolution unsupervised classification of a water body enabling the rapid identification of relevant regions which can then be used to direct further investigation by reference-sensor equipped boats.

### 2024-05-13 paper/draft-introduction.tex
updated introduction

--- 2024-05-08 paper/draft-introduction.tex
@@
 Nevertheless, this approach relies on prior knowledge of expected sources in order to select appropriate reference instruments for model validation.
-The presence of unanticipated contaminants cannot be directly identified in this paradigm.
+The presence of unanticipated contaminants cannot be directly identified in this sensing paradigm.
 
-% Unsupervised methods
+Extending the capabilities of UAV-based hyperspectral imaging to enable water quality monitoring in real-time scenarios where contaminant sources may not be known in advance requires two additional capabilities: dimensionality reduction techniques to permit the visual comparison of HSI, and endmember extraction techniques which can identify spectral signatures corresponding to unique sources within the imaging scene.
+In remote sensing where reference data are typically sparse, many approaches have been explored.
+For example, principal component analysis (PCA) and t-distributed stochastic neighbor embedding (tSNE) are dimensionality reduction methods commonly used to reduce HSI to two or three dimensions for visualization \cite{tyo2003principal,zhang2015hyperspectral}.
+Similarly for endmember extraction there are a variety of established approaches including geometric methods like vertex component analysis, statistical methods like k-nearest neighbors and non-negative matrix factorization (NMF), and deep learning methods based on autoencoder architectures \cite{heylen2014review,nascimento2005vertex, Feng2022HyperspectralUB, cariou2015unsupervised, su2019daen, borsoi2019deep, palsson2020convolutional}.
+Methods based on linear mappings like PCA and NMF are often too restrictive for HSI where the assumption of linear mixing is easily broken.
+However, the increased complexity of nonlinear methods like tSNE and autoencoders often lead to significant increases in computation time.
+An ideal approach should enable both visualization and nonlinear identification of relevant spectral endmembers.
 
-To address this limitation, unsupervised methods that segment HSI pixels and can identify unique spectral signatures are needed.
+% neurons that wire together, fire together
+The self-organizing map (SOM) developed by Teuvo Kohonen is an unsupervised machine learning method which nonlinearly maps high dimensional data to the nodes of a two-dimensional grid \cite{kohonen-som-1}.
+By preserving the topological relationship between nodes during training, the SOM ensures that similar spectral signatures are mapped together such that related HSI pixels naturally cluster together.
+This presents an attractive compromise by enabling the simultaneous visualization of HSI data and endmember extraction via the weight vector associated with each SOM node \cite{cantero2004analysis, duran2007time,som-hsi}.
+When reference data are available, the SOM can be utilized to provide semi-supervised labeling of HSI spectra \cite{riese2019supervised}.
+Furthermore, Danielsen et al.
+demonstrated that the dimensionality reduction offered by the SOM can even be used for on-board data compression of HSIs acquired by a CubeSat \cite{danielsen2021self}.
+Despite these clear capabilities, the SOM relies on a heuristic training algorithm with hyperparemeters that can be challenging to tune and offers no direct probabilistic interpretation.
+To address these limitations, Bishop et al.
+developed the generative topographic mapping (GTM), a Bayesian latent-variable model inspired by the SOM \cite{gtm-bishop-1}.
+The GTM has been utilized in a variety of domains including drug design and chemical data visualization but has yet to see adoption for the analysis of hyperspectral imagery \cite{kireeva2012generative, gaspar2015chemical, horvath2019generative}.
 
-These methods generally fall into two categories: dimensionality reduction and endmember extraction.
-For example, principle component analysis (PCA) and t-distributed stochastic neighbor embedding (tSNE) make it possible to reduce high dimensional HSI into two or three dimensions where they can be visualized and compared \cite{}.
-
-In remote sensing, where reference data is sparse, many approaches have been explored and generally fall into two categories: dimensionality reduction, and unsupervised classification and endmember extraction.
-
-Algorithms such as principal component analysis (PCA) and t-distributed stochastic neighbor embedding (t-SNE) enable the visualization of HSI by reducing their dimensionality \cite{yousefi2016mineral}.
-
-including dimensionality reduction techniques like principal component analysis (PCA) and clustering algorithms such as k-nearest neighbors \cite{yousefi2016mineral,cariou2015unsupervised}.
-Non-negative matrix factorization is another popular approach which seeks to simultaneously extract endmembers with their associated abundances via a constrained matrix decomposition \cite{Feng2022HyperspectralUB}.
-Similarly, the advent of deep learning has led to a profusion of neural network based models utilizing an autoencoder architecture to enable nonlinear endmember extraction \cite{su2019daen,borsoi2019deep,palsson2020convolutional}.
-Other methods such as vertex component analysis (VCA) are tailored specifically for hyperspectral data based on the observation that endmember spectra form the vertices of a simplex \cite{nascimento2005vertex}.
-Ultimately, an ideal approach for near-real time analysis of water quality from HSI should enable both visualization via dimensionality reduction and identification of relevant spectral endmembers.
-
-% 4. SOM
-
-- SOM is an attractive approach as it enables both dimensionality reduction and unsupervised classification. 
-- Furthermore, the SOM maintains a topological relationship between learned classes such that classes near each other in the SOM grid are more similar than classes far apart
-- SOM can be used to nonlinearly map data into a 2-dimensional space
-- Weight vectors corresponding to each SOM node can be interpreted as spectral endmembers.
-The training algorithm updates weights based on euclidean distance between data record and each node weight vector
-- Add citations and examples of use in remote sensing / HSI
--
-
-Among unsupervised methods, Kohonen's self-organizing map provides an attractive compromise enabling both dimensionality reduction and classification...
-
-% 5. GTM
-
-% 6. This paper...
+In this paper, we explore the application of the GTM to UAV-acquired HSI for the characterization water quality. Specifically, we use the GTM to
 

### 2024-05-14 paper/paper.tex
updated introduction
Updated methods with description of GTM algorithm

@@
 % Full title of the paper (Capitalized)
-\Title{Robot Team III: Scotty Strikes Back}
+% \Title{Robot Team III: Scotty Strikes Back}
+\Title{Visualization and Unsupervised Classification of Water Composition with UAV-based Hyperspectral Imaging and Generative Topographic Mapping}
 
@@
 
+Inland water bodies present a unique challenge to characterization by remote sensing imagery due to their complex spectral characteristics and small scale spatial variability.
+The broad bands of multi-spectral imagers coupled with the irregular shape of lakes and rivers results in pixels with highly mixed signals that are easily dominated by reflectance from shore and near-shore vegetation sources \cite{koponen2002lake, ritchie2003remote}.
+Recently, the combination of hyperspectral imaging with unmanned aerial vehicles (UAV) such as drones has emerged as a powerful approach to simultaneously address the spectral, spatial, and temporal limitations of traditional high-altitude and satellite-based collection \cite{adao2017hyperspectral,arroyo2019implementation}.
+UAV are significantly less expensive to deploy than their remote sensing counterparts, and low altitude flights enable centimeter-scale sampling while limiting the need for complicated atmospheric corrections \cite{banerjee2020uav}.
+However, the significant increase in data volume generated by these systems presents a new challenge, namely, how to efficiently extract water quality parameters of interest from intricate pixel spectra.
+
+Significant research efforts have focused on the development of techniques and algorithms to retrieve water quality parameters from UAV-captured hyperspectral images (HSI).
+On-board compute installed alongside hyperspectral imagers can enable the rapid evaluation of spectral indices from HSI band ratios \cite{horstrand2019uav}.
+These band ratios and polynomial combinations of bands have been used to successfully invert optically-active water quality parameters such as turbidity directly from UAV acquired imagery \cite{vogt2016near, zhang2022selection}.
+Supervised machine learning techniques such as tree-based models, support vector machines, and neural networks have also been used to estimate a plethora of parameters such as colored dissolved organic matter, chlorophyll a, blue-green algae, and suspended sediment concentrations \cite{keller2018hyperspectral, lu2021retrieval}.
+The calibration and evaluation of these data-driven models demand a significant volume of coincident, in situ data.
+This can be addressed by coordinating UAV flights with reference data collection using autonomous robotic boats \cite{robot-team-1, robot-team-2}.
+Nevertheless, this approach relies on prior knowledge of expected sources in order to select appropriate reference instruments for model validation.
+The presence of unanticipated contaminants cannot be directly identified in this sensing paradigm.
+
+Extending the capabilities of UAV-based hyperspectral imaging to enable water quality monitoring in real-time scenarios where contaminant sources may not be known in advance requires two additional capabilities: dimensionality reduction techniques to permit the visual comparison of HSI, and endmember extraction techniques which can identify spectral signatures corresponding to unique sources within the imaging scene.
+In remote sensing where reference data are typically sparse, many approaches have been explored.
+For example, principal component analysis (PCA) and t-distributed stochastic neighbor embedding (tSNE) are dimensionality reduction methods commonly used to reduce HSI to two or three dimensions for visualization \cite{tyo2003principal,zhang2015hyperspectral}.
+Similarly for endmember extraction there are a variety of established approaches including geometric methods like vertex component analysis, statistical methods like k-nearest neighbors and non-negative matrix factorization (NMF), and deep learning methods based on autoencoder architectures \cite{heylen2014review,nascimento2005vertex, Feng2022HyperspectralUB, cariou2015unsupervised, su2019daen, borsoi2019deep, palsson2020convolutional}.
+Methods based on linear mappings like PCA and NMF are often too restrictive for HSI where the assumption of linear mixing is easily broken.
+However, the increased complexity of nonlinear methods like tSNE and autoencoders often lead to significant increases in computation time.
+An ideal approach should enable both visualization and nonlinear identification of relevant spectral endmembers.
+
+% neurons that wire together, fire together
+The self-organizing map (SOM) developed by Teuvo Kohonen is an unsupervised machine learning method which non-linearly maps high dimensional data to the nodes of a two-dimensional grid \cite{kohonen-som-1}.
+By preserving the topological relationship between nodes during training, the SOM ensures that similar spectral signatures are mapped together such that related HSI pixels naturally cluster together.
+This presents an attractive compromise by enabling the simultaneous visualization of HSI data and endmember extraction via the weight vector associated with each SOM node \cite{cantero2004analysis, duran2007time,som-hsi}.
+When reference data are available, the SOM can be utilized to provide semi-supervised labeling of HSI spectra \cite{riese2019supervised}.
+Furthermore, Danielsen et al.
+demonstrated that the dimensionality reduction offered by the SOM can even be used for on-board data compression of HSIs acquired by a CubeSat \cite{danielsen2021self}.
+Despite these clear capabilities, the SOM relies on a heuristic training algorithm with hyperparameters that can be challenging to tune and offers no direct probabilistic interpretation.
+To address these limitations, Bishop et al.
+developed the generative topographic mapping (GTM), a Bayesian latent-variable model inspired by the SOM \cite{gtm-bishop-1}.
+The GTM has been utilized in a variety of domains including drug design and chemical data visualization but has yet to see adoption for the analysis of hyperspectral imagery \cite{kireeva2012generative, gaspar2015chemical, horvath2019generative}.
+
+In this paper, we explore the application of the GTM to UAV-acquired HSI for the characterization water quality.
+Using data collected at pond in Montague, North Texas, we first train a GTM using water-only pixels to produce a low-dimensional representation of the collected HSI.
+We use this mapping to explore the highly detailed small-scale variability within the pond and discuss how this can be used to guide reference data collection.
+Next, we demonstrate how the GTM can be utilized to identify relevant spectral endmembers from a combined dataset including land pixels, algae, water, and a simulated contaminant plume using a rhodamine tracer dye.
+Once identified, these endmembers can be used to rapidly map the abundance of spectral features within the pond.
+We demonstrate this capability by mapping the abundance of algae near the shore and to track the evolution of a rhodamine dye plume.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-In this study, we explore the use of Generative Topographic Mapping as a tool for the unsupervised classification and nonlinear endmember extraction of hyperspectral imagery.
-To this end, HSIs were collected using a USV mounted hyperspectral imager at a pond in Montague, North Texas on on 23 November 2020 (shown in panel (d) of Fig~\ref{fig:sample-spectra}).
-In this section we shall first provide a detailed description of the GTM algorithm with subsequent subsections reserved to describe the specific data collection, processing, and analysis techniques employed in the study.
+In this study, we explore the use of the GTM as a tool for dimensionality reduction and non-linear endmember extraction of hyperspectral imagery.
+To this end, a dataset of HSI were collected at a pond in Montague, North Texas (shown in panel d of Fig~\ref{fig:sample-spectra}) on 23 November 2020 using a UAV mounted hyperspectral imager configured as described in \cite{robot-team-1, robot-team-2}.
+In this section we first provide a detailed overview of the GTM algorithm.
+Next we describe the UAV platform used for HSI collection and the various steps used to processes captured HSI.
+Finally, we describe the GTM training procedure, hyperparameter optimization, and subsequent visualization methods used to analyze trained GTM models in this study.
 
@@
 
-The Generative Topographic Mapping (GTM) was developed by Bishop, Svens\'en, and Williams as a Bayesian realization of Kohonen's Self-organizing Map (SOM) \cite{gtm-bishop-1}. Specifically, the
+The GTM is a probabilistic latent variable model inspired by the SOM for visualizing and clustering high-dimensional data.
+Like the SOM, the GTM assumes vectors $\mathbf{x}$ in the high-dimensional data space (here representing reflectance spectra) are constrained to a low-dimensional embedded manifold.
+The SOM describes this manifold using a regular grid of nodes, each having an associated weight vector defining the mapping from the manifold to the data space.
+The position of each data record $\mathbf{x}$ in the manifold is assigned as the position of the node whose weight vector has the minimum distance to $\mathbf{x}$.
+An iterative training procedure updates the weight vectors of each node to fit the manifold to the data such that nodes near each other in the SOM grid correspond to similar records (in the Euclidean sense) in the data space.
 
-The Generative Topographic Mapping (GTM) is an unsupervised latent variable model developed by Bishop, Svens\'en, and Williams \cite{gtm-bishop-1}.
-The GTM takes inspiration from Kohonen's self-organizing map which enables an unsupervised classification of high dimensional data into a (typically) two-dimensional grid of nodes such that nodes near each near each other in the mapping space
-
-As outlined in the introduction, an effective unsupervised classification algorithm for hyperspectral imagery should accomplish the following three things: 
-\begin{enumerate}
-    \item Segmentation of captured imagery into a set of distinct classes
-    \item Mapping of distinct spatial regions in collected imagery using the identified classes
-    \item Simultaneous identification of relevant spectral endmembers
-\end{enumerate}
-
-- mention
-
+The GTM mimics the grid of the SOM by assuming data are generated from latent variables $\xi$ which are constrained to the $K$-many nodes of a regular grid.
+This assumption corresponds to establishing a prior distribution on the latent space of the form
 \begin{equation}\label{eq:latent-prob}
@@
 \end{equation}
+where $\delta(\cdot)$ is the Dirac delta function.
 
-\begin{equation}\label{eq:data-prob}
-    p\left( \mathbf{x} \mid \xi, W, \beta \right)  = \mathcal{N}(\psi,\, \beta^{-1})
-\end{equation}
-
-\begin{equation}\label{eq:psi}
-    \psi = W\phi(\xi)
-\end{equation}
-
-Parameter prior (used for model weight regularization), i.e. assuming a Gaussian distribution with mean 0 for each weight where $\lVert \cdot \rVert_{F}$ is the Frobenius norm.
-\begin{equation}\label{eq:weight-prior}
-    p(W \mid \alpha) =  \left( \frac{\alpha}{2\pi} \right)^{MD/2}\exp\left(-\frac{\alpha}{2}\lVert W \rVert_{F}^2\right)
-\end{equation}
+Points $\xi$ in this latent space are mapped to the embedded manifold by a non-linear function $\psi$ parameterized by weights $W$.
+However, since real data is rarely noise-free, this embedded manifold will not be perfectly thin.
+To account for this the points $\xi$ are described in the data space by a radially symmetric Gaussian distribution, $\mathcal{N}(\psi(\xi),\, \beta^{-1})$, with mean $\psi(\xi)$ and variance $\beta^{-1}$.
+For a dataset consisting of $N$-many i.i.d.
+records, this choice yields a log-likelihood function of the form
 \begin{equation}\label{eq:llh}
@@
 \end{equation}
+which can be maximized to fit $W$ and $\beta^{-1}$.
 
-Expectation Maximization Algorithm:
-
-Expectation step:
-\begin{equation}\label{eq:responsibility}
-    R_{kn} = p(\xi_k \mid \mathbf{x}_n, W, \beta) = \dfrac{p(\mathbf{x}_n \mid \xi_k, W, \beta)}{\sum\limits_{k'}^{K} p(\mathbf{x}_n \mid \xi_{k'}, W, \beta)}
+The function $\psi$ is typically chosen to be given by the sum of $M$-many radial basis functions (RBF) evenly distributed in the latent space so that $\psi(\xi) = W\phi(\xi)$ with centers $\mu_m$ and width $\sigma$ so that
+\begin{equation}
+    \phi_m(\xi) = \exp\left(-\dfrac{\lVert \xi - \mu_m \rVert^2}{2\sigma^2}\right).
+\end{equation}
+Together the number of RBFs, $M$, and $\sigma$ are hyperparameters for the model which govern the smoothness of the resulting manifold in the data space.
+Additionally, sparsity of $W$ can be controlled by an additional hyperparameter $\alpha$ corresponding to a prior distribution over the weights given by
+\begin{equation}\label{eq:weight-prior}
+    p(W \mid \alpha) =  \left( \frac{\alpha}{2\pi} \right)^{MD/2}\exp\left(-\frac{\alpha}{2}\left\lVert W \right\rVert_{F}^2\right).
 \end{equation}
 
-Maximization step:
+The use of RBFs for $\psi$ enables maximizing $\mathcal{L}(W,\beta)$ via an Expectation-Maximization routine.
+During each step of the fitting process, the responsibility of the $k$th node for the $n$th data record is computed as
+\begin{equation}\label{eq:responsibility}
+    R_{kn} = p(\xi_k \mid \mathbf{x}_n, W, \beta) = \dfrac{p(\mathbf{x}_n \mid \xi_k, W, \beta)}{\sum\limits_{k'}^{K} p(\mathbf{x}_n \mid \xi_{k'}, W, \beta)}.
+\end{equation}
+Together these responsibilities form a matrix with entries $R_{kn}$ which are kept fixed during the maximization step.
+The maximization step is performed by updating the weights $W$ and variance $\beta^{-1}$ according to
+\begin{align}\label{eq:m-step}
+    W_{\text{new}} &= \left(\Phi^T G \Phi + \dfrac{\alpha}{\beta}I \right)^{-1} \Phi^T R X  \\
+    \frac{1}{\beta_{\text{new}}} &= \frac{1}{ND} \sum\limits_{n}^{N} \sum\limits_{k}^{K} R_{kn} \left\lVert \psi_k - \mathbf{x}_n \right\rVert^2
+\end{align}
+where $\Phi_{km} = \phi_m(\xi_k)$, $X$ is the data matrix formed by concatenating the records $\mathbf{x}_n$, and $G$ is a diagonal matrix with $G_{kk} = \sum\limits_n^N R_{kn}$.
 
-Definition of the diagonal matrix, $G$, with entries
-\begin{equation}
-    G_{kk}  = \sum\limits_n^N R_{kn}
-\end{equation}
-and all others $0$.
-
-\begin{equation}\label{eq:W-update}
-    \left(\Phi^T G \Phi + \dfrac{\alpha}{\beta}I \right)W_{\text{new}} = \Phi^T R X 
-\end{equation}
-
-\begin{equation}
-    \frac{1}{\beta_{\text{new}}} = \frac{1}{ND} \sum\limits_{n}^{N} \sum\limits_{k}^{K} R_{kn} \left\lVert \psi_k - \mathbf{x}_n \right\rVert^2
-\end{equation}
-
-\begin{equation}
-    \hat{\xi}_n = \sum_{k}^K R_{kn}\xi_k
-\end{equation}
+For this study, a freely available implementation of the GTM was developed in the Julia programming language \cite{gtm-code}. Julia is a dy
 
@@
 \subsection{Image Segmentation with the Normalized Spectral Similarity Score}
+
+\begin{equation}
+    \hat{\xi}_n = \sum_{k}^K R_{kn}\xi_k
+\end{equation}
 
@@
 - Drones can be equipped with HSI and compute to enable rapid generation of spectral indices like the NDVI for applications such as precision agriculture \cite{horstrand2019uav}
+
+- Mention how Bishop suggests extensions to GTM for on-line learning which could be used to accelerate training for large HSI datasets
 

### 2024-05-15 7-extra-plots.jl
added plots for GTM diagram

@@ -23,14 +23,81 @@ if !ispath(figures_path)
 end
 
 
-k = 32
-m = 14
-s = 0.1
+k = 15
+m = 5
+
+
 
 ξ₁ = range(-1, stop=1.0, length=k)
-ξ₁ = range(-1, stop=1.0, length=k)
+ξ₂ = range(-1, stop=1.0, length=k)
+
+μ₁ = range(-1, stop=1.0, length=m)
+μ₂ = range(-1, stop=1.0, length=m)
 
 
 
-# create figure with latent space grid and RBF points as well as
-# the "3d" projection of means in the "data" space
+s = 0.01
+function rbf_val(x,y)
+    out = 0.0
+    for i ∈ 1:m, j ∈ 1:m
+        out += exp(-((x-μ₁[i])^2 + (y - μ₂[j])^2)/(2*s))
+    end
+    out
+end
+
+cm = cgrad([CairoMakie.RGBA(0,0,0,0), mints_colors[2]]) #CairoMakie.RGBA(1,0,0, 1)])
+
+
+
+mints_colors[2]
+
+fig =Figure();
+ax = Axis(
+    fig[1,1],
+    xlabel="ξ₁", ylabel="ξ₂",
+    aspect=DataAspect(),
+    xticklabelsvisible=false, xticksvisible=false, xminorticksvisible=false, xgridvisible=false, xminorgridvisible=false, xlabelsize=30,
+    yticklabelsvisible=false, yticksvisible=false, yminorticksvisible=false, ygridvisible=false, yminorgridvisible=false, ylabelsize=30,
+);
+
+hidespines!(ax)
+
+pts_x = [(Point(ξ₁[i], -1), Point(ξ₁[i], 1)) for i ∈ 1:k]
+pts_y = [(Point(-1, ξ₂[i]), Point(1, ξ₂[i])) for i ∈ 1:k]
+pts_joined = [Point(ξ₁[i], ξ₂[j]) for i ∈ 1:k for j ∈ 1:k]
+
+pts_rbf = [Point(μ₁[i], μ₂[j]) for i ∈ 1:m for j ∈ 1:m]
+
+linesegments!(ax, pts_x, color=:lightgray, linewidth=2)
+linesegments!(ax, pts_y, color=:lightgray, linewidth=2)
+heatmap!(ax, -1:0.005:1, -1:0.005:1, rbf_val, colormap=cm)
[…]

### 2024-05-15 7-extra-plots.jl
updated figures

@@ -88,7 +88,7 @@ zs = [fz(xs[i], ys[j]) for i∈1:k, j∈1:k]
 
 pts = [Point3f(xs[i], ys[j], zs[i,j]) for i in 1:k for j ∈ 1:k]
 
-fig = Figure();
+fig = Figure(size=(800,800));
 ax = Axis3(fig[1,1]);
 hidedecorations!(ax)
 hidespines!(ax)
@@ -98,6 +98,6 @@ hidespines!(ax)
 wireframe!(ax, xs, ys, zs, color=:lightgray);
 scatter!(ax, pts, color=mints_colors[3], markersize=10)
 
-save(joinpath(figures_path, "gtm-projected.svg"), fig)
+save(joinpath(figures_path, "gtm-projected.png"), fig)
 
 fig

### 2024-05-15 paper/paper.tex
updated methods section

@@
 
-%%% --- Generative 0Topographic Mapping ------------------------------------
+%%% --- Generative Topographic Mapping ------------------------------------
 \subsection{Generative Topographic Mapping}
@@
 The GTM is a probabilistic latent variable model inspired by the SOM for visualizing and clustering high-dimensional data.
-Like the SOM, the GTM assumes vectors $\mathbf{x}$ in the high-dimensional data space (here representing reflectance spectra) are constrained to a low-dimensional embedded manifold.
+Like the SOM, the GTM assumes vectors $\mathbf{x}$ in the $d$-dimensional data space (here representing reflectance spectra) are constrained to a low-dimensional embedded manifold.
 The SOM describes this manifold using a regular grid of nodes, each having an associated weight vector defining the mapping from the manifold to the data space.
-The position of each data record $\mathbf{x}$ in the manifold is assigned as the position of the node whose weight vector has the minimum distance to $\mathbf{x}$.
-An iterative training procedure updates the weight vectors of each node to fit the manifold to the data such that nodes near each other in the SOM grid correspond to similar records (in the Euclidean sense) in the data space.
+The position of each data record $\mathbf{x}$ in the manifold is assigned to the position of the node whose weight vector has the minimum Euclidean distance to $\mathbf{x}$.
+An iterative training procedure updates the weight vectors of each node to fit the manifold to the data such that nodes near each other in the SOM grid correspond to similar records in the data space.
 
@@
 
-Points $\xi$ in this latent space are mapped to the embedded manifold by a non-linear function $\psi$ parameterized by weights $W$.
+Points $\xi$ in this latent space are mapped to the embedded manifold by a non-linear function $\psi$ parameterized by weights $W$ as illustrated in Fig~\ref{fig:gtm-diagram}.
 However, since real data is rarely noise-free, this embedded manifold will not be perfectly thin.
-To account for this the points $\xi$ are described in the data space by a radially symmetric Gaussian distribution, $\mathcal{N}(\psi(\xi),\, \beta^{-1})$, with mean $\psi(\xi)$ and variance $\beta^{-1}$.
-For a dataset consisting of $N$-many i.i.d.
+To account for this, the points $\xi$ are described in the data space by a radially symmetric Gaussian distribution, $\mathcal{N}(\psi(\xi),\, \beta^{-1})$, with mean $\psi(\xi)$ and variance $\beta^{-1}$.
+For a dataset $\left\{ \mathbf{x}_n\right\}\limits_{n=1}^N$ consisting of $N$-many i.i.d.
 records, this choice yields a log-likelihood function of the form
@@
 \end{equation}
-which can be maximized to fit $W$ and $\beta^{-1}$.
+which can be maximized to yield optimal $W$ and $\beta^{-1}$.
 
-The function $\psi$ is typically chosen to be given by the sum of $M$-many radial basis functions (RBF) evenly distributed in the latent space so that $\psi(\xi) = W\phi(\xi)$ with centers $\mu_m$ and width $\sigma$ so that
+\begin{figure}[t]
+%\begin{adjustwidth}{-\extralength}{0cm}
+\centering
+%\includegraphics[width=\columnwidth]{paper/figures/methods/gtm-diagram.pdf}
+\includegraphics[width=\columnwidth]{paper/figures/methods/gtm-diagram.png}
+%\end{adjustwidth}
+\caption{Illustration of the GTM algorithm.
+On the left is a regular grid of $K$ many points (green) in the latent space represented by their coordinates $\xi_1$ and $\xi_2$.
+In red are the $M$-many RBF which define the mapping $\psi$ from the latent space to the data space.
+Points in the latent space are mapped non-linearly to the data space yielding an embedded manifold in $\mathcal{R}^d$, here illustrated in three dimensions.\label{fig:gtm-diagram}}
+\end{figure}
+
+The function $\psi$ is typically chosen to be given by the sum of $M$-many radial basis functions (RBF) evenly distributed in the latent space such that $\psi(\xi) = W\phi(\xi)$ with centers $\mu_m$ and width $\sigma$ so that
 \begin{equation}
@@
 Together the number of RBFs, $M$, and $\sigma$ are hyperparameters for the model which govern the smoothness of the resulting manifold in the data space.
-Additionally, sparsity of $W$ can be controlled by an additional hyperparameter $\alpha$ corresponding to a prior distribution over the weights given by
+Additionally, sparsity of $W$ can be enforced by introducing an additional hyperparameter $\alpha$ corresponding to a prior distribution over the weights given by
 \begin{equation}\label{eq:weight-prior}
@@
 Together these responsibilities form a matrix with entries $R_{kn}$ which are kept fixed during the maximization step.
-The maximization step is performed by updating the weights $W$ and variance $\beta^{-1}$ according to
+The maximization step is then performed by updating the weights $W$ and variance $\beta^{-1}$ according to
 \begin{align}\label{eq:m-step}
@@
 where $\Phi_{km} = \phi_m(\xi_k)$, $X$ is the data matrix formed by concatenating the records $\mathbf{x}_n$, and $G$ is a diagonal matrix with $G_{kk} = \sum\limits_n^N R_{kn}$.
+This process is repeated until the log-likelihood converges to a predetermined tolerance level.
 
-For this study, a freely available implementation of the GTM was developed in the Julia programming language \cite{gtm-code}. Julia is a dy
+Slices $R_{[:,n]}$ of the matrix $R$ define the responsibility of each latent node $\xi_k$ for the $n$th data record $\mathbf{x}_n$.
+Therefore the final responsibility matrix computed at the end of the GTM training routine can be used to represent each record in the latent space by the mean:
+\begin{equation}
+    \hat{\xi}_n = \sum_{k}^K R_{kn}\xi_k.
+\end{equation}
 
-- GTM paper original \cite{gtm-bishop-1}
-- GTM developments paper \cite{gtm-bishop-2}
-- GTM code \cite{gtm-code}
-- implementation designed to work in MLJ ecosystem \cite{blaom2020mlj}
-
-at a pond in Montague, north Texas (shown in panel d of Fig~\ref{fig:sample-spectra})
+A freely available implementation of the GTM algorithm was developed for this study and is accessible at \cite{gtm-code}.
+The code is written in the Julia programming language and complies with the Machine Learning in Julia (MLJ) common interface \cite{bezanson2012julia,blaom2020mlj}.
 
 \subsection{UAV-Based Hyperspectral Imaging}
-A Freefly Alta-X autonomous quadcopter was used as a UAV platform for the robotic team.
 
-The Alta-X is specifically designed to carry cameras and has a payload of up to 35 lbs. We equipped the UAV with a Resonon Pika XC2 visible+near-infrared (VNIR) hyperspectral imager.
+A Freefly Alta-X autonomous quadcopter was used as the UAV platform in this study.
+This UAV was equipped with a Resonon Pika XC2 visible+near-infrared (VNIR) hyperspectral imager to capture HSI with $462$ wavelengths per pixel ranging from $391$ to $1011$ nm.
+This imager is in a pushbroom configuration so that HSI are captured one scan-line at a time resulting in data cubes consisting of $1000$ scan-lines with $1600$ pixels each.
+Additionally, the camera includes an embedded GPS/INS unit to enable georectification of collected HSI.
+An upward facing Ocean Optics UV-Vis NIR spectrometer with a cosine corrector was also included to provide measurements of the incident solar irradiance spectrum.
+Data collection and processing was controlled by an attached Intel NUC small-form-factor computer.
 
-For each image pixel, this camera samples 462 wavelengths ranging from 391 to 1011 nm.
+To account for the variability of incident light, raw HSI are converted to units of reflectance using the downwelling irradiance spectrum simultaneously captured with each HSI.
+With the hyperspectral imager oriented to nadir, the reflectance is given by
+\begin{equation}\label{eq:reflectance}
+    R(\lambda) = \pi L(\lambda)/E_d(\lambda)
+\end{equation}
+where $L$ is the spectral radiance, $E_d$ is the downwelling irradiance, and a factor of $\pi$ steradians results from assuming the water surface is Lambertian (diffuse) \cite{ruddick2019review}.
+HSI collection was performed near solar noon to maximize the amount of incident sunlight illuminating the water.
+For the site in North Texas, this corresponded to an average solar zenith angle of $54.9^\circ$ for 23 November 2020 resulting in HSI with negligible sunglint effects.
 
-Additionally, the UAV includes an upward facing Ocean Optics UV-Vis-NIR spectrometer with a cosine corrector to capture the incident downwelling irradiance spectrum.
-
-Data collection by the hyperspectral imager is controlled by an attached Intel NUC small-form-factor computer.
-
-A second processing NUC is also included for onboard georectification and generation of data products.
-
-The collected hyperspectral images (HSIs) are stored locally on a solid state drive that is simultaneously mounted by the processing computer.
-
-- georetification \cite{muller2002program, baumker2001new, mostafa2000multi}
+After conversion to reflectance, each HSI must also be georectified to assign geographic coordinates to each pixel.
+The UAV flights were performed at an approximate altitude of $50$ m above the water so that the imaging surface is essentially flat.
+Consequently, HSI were rapidly georectified to a $10$ cm resolution using position and orientation data from the embedded GPS/INS as outlined in \cite{muller2002program, baumker2001new, mostafa2000multi}.
+An example hyperspectral data cube is illustrated in panel (a) of Fig~\ref{fig:spectral-signatures} where the log10-reflectance values are plotted along the z-axis for each pixel in the scene and a psuedo-color image at the top of the data cube illustrates how the water would appear to the human eye from the perspective of the UAV.
 
 \subsection{Data Collection and Pre-Processing}
-- Julia programming language used \cite{bezanson2012julia}
 
 \subsection{Image Segmentation with the Normalized Spectral Similarity Score}
-
-\begin{equation}
-    \hat{\xi}_n = \sum_{k}^K R_{kn}\xi_k
-\end{equation}
 
@@
 \caption{GTM Node Signatures corresponding to the exemplar spectra for the rhodamine plume (top left), algae (top right), dry grass (bottom left), and open water (bottom right).
-\label{fig:spectral-signautes}}
+\label{fig:spectral-signatures}}
 \end{figure}

### 2024-05-15 paper/paper.tex
updated methods section

@@
 Next we describe the UAV platform used for HSI collection and the various steps used to processes captured HSI.
-Finally, we describe the GTM training procedure, hyperparameter optimization, and subsequent visualization methods used to analyze trained GTM models in this study.
+Finally, we describe the two case studies presented in this paper for the utilization of the GTM as a HSI visualization tool and as an endmember extraction technique.
 
@@
 After conversion to reflectance, each HSI must also be georectified to assign geographic coordinates to each pixel.
-The UAV flights were performed at an approximate altitude of $50$ m above the water so that the imaging surface is essentially flat.
+The UAV flights were performed at an approximate altitude of $50$ m above the water so that the imaging surface can be considered to be flat.
 Consequently, HSI were rapidly georectified to a $10$ cm resolution using position and orientation data from the embedded GPS/INS as outlined in \cite{muller2002program, baumker2001new, mostafa2000multi}.
@@
 
-\subsection{Data Collection and Pre-Processing}
+As a final processing step before training each GTM, we limit the wavelengths of each HSI to $\lambda \leq 900$ nm as wavelengths above $900$ nm showed significant noise.
+Additionally, each spectrum was re-scaled to a peak value of $1.0$ to account for incident light variability between HSI.
 
-\subsection{Image Segmentation with the Normalized Spectral Similarity Score}
+\subsection{GTM Case Studies}
+
+To explore the ability of the GTM to segment HSI pixels and aid in source identification we first consider a dataset of water-only spectra consisting of $>36,000$ records identified from collected HSI by a normalized difference water index (NDWI) greater than $0.25$ where the NDWI is defined as
+\begin{equation}
+    \text{NDWI} = \dfrac{R(550) - R(860)}{R(550) + R(860)}.
+\end{equation}
+We then apply the trained GTM to all water-only pixels to visualize the distribution learned by the GTM and examine spectra associated with a subset of nodes in the latent space in order to assess the small-scale spatial variability within the pond.
+
+Next, we consider a dataset of combined HSI pixels including both water and land.
+To simulate the dispersion of a potential contaminant source, a Rhodamine tracer dye was released into the western portion of the pond and two additional UAV flights were used to collect HSI capturing the evolution of the resulting plume.
+From these HSI a collection of $>145,000$ were sampled for model training.
+Exemplar spectra for water, grass, algae, and rhodamine dye were identified from a sample HSI as shown in panels (b) and (c) of Fig~\ref{fig:sample-spectra}.
+Using these spectra, we explore the trained GTM and use it to extract spectral signatures corresponding to these endmember categories.
+Values for the hyperparameters $m$, $s$, and $\alpha$, are determined by training multiple GTM models and selecting values which minimize the Bayesian Information Criterion (BIC) given by
+\begin{equation}
+    \text{BIC} = 2P\ln(N) - 2\mathcal{L}
+\end{equation}
+where $P$ is the total number of model parameters, $N$ is the number of records in the dataset, and $\mathcal{L}$ is the log-likelihood defined in Eq~\ref{eq:llh}.
+
+The normalized spectral similarity score (NS3) introduced by Nidamanuri and Zbell combines the root-mean-square (RMS) difference together with the spectral angle to provide a spectral distance function\cite{nidamanuri2010normalized}.
+For two spectra $R_1(\lambda)$ and $R_2(\lambda)$ it is defined by
+\begin{equation}
+    \text{NS3}(R_1, R_2) &= \sqrt{RMS(R_1, R_2)^2 + (1-\cos\theta)^2}
+\end{equation}
+where
+\begin{align}
+    \text{RMS}(R_1, R_2) &= \sqrt{\frac{1}{N-1}\sum_\lambda \left(R_{1}(\lambda) - R_2(\lambda) \right)^2} \\
+    \cos\theta &= \frac{\langle R_1 , R_2 \rangle}{\lVert R_1\rVert \lVert R_2 \rVert}.
+\end{align}
+We use the NS3 together with identified spectral endmembers to map the abundance of algae near the shore as well as the evolution of the Rhodamine dye plume.
 

### 2024-05-16 paper/paper.tex
updated results section

@@
 %%% --- Generative Topographic Mapping ------------------------------------
-\subsection{Generative Topographic Mapping}
+\subsection{Generative Topographic Mapping}\label{sec:gtm-overview}
 
@@
 \end{equation}
-Together the number of RBFs, $M$, and $\sigma$ are hyperparameters for the model which govern the smoothness of the resulting manifold in the data space.
+The width $\sigma$ is taken as the distance between neighboring RBF centers adjusted by a scale factor labelled $s$.
+to Together the number of RBFs, $M$, and $s$ are hyperparameters for the model which govern the smoothness of the resulting manifold in the data space.
 Additionally, sparsity of $W$ can be enforced by introducing an additional hyperparameter $\alpha$ corresponding to a prior distribution over the weights given by
@@
 Slices $R_{[:,n]}$ of the matrix $R$ define the responsibility of each latent node $\xi_k$ for the $n$th data record $\mathbf{x}_n$.
-Therefore the final responsibility matrix computed at the end of the GTM training routine can be used to represent each record in the latent space by the mean:
+Therefore the final responsibility matrix after GTM training can be used to represent each record in the latent space by the mean:
 \begin{equation}
@@
 
-\subsection{GTM Case Studies}
+\subsection{GTM Case Studies}\label{sec:case-studies}
 
@@
 
-\subsection{GTM Fit}
-
-\begin{figure}[t]
-\centering
-\includegraphics[width=0.8\columnwidth]{paper/figures/results/square-ndvi.pdf}
-\caption{GTM Map: We plot the mean position of each data point in the latent space with points colored by the their associated NDVI values.
-The locations of the exemplar points for algae, rhodamine dye, water, and dry grass are shown in the latent space illustrating that the GTM has clearly distinguished land, near-land, and water pixels.\label{fig:}}
-\end{figure}
-
-\subsection{Hyperparamter Optimization with BIC}
-
-\begin{figure}[t]
-%\begin{adjustwidth}{-\extralength}{0cm}
-\centering
-\includegraphics[width=\textwidth]{paper/figures/results/bic.pdf}
-%\end{adjustwidth}
-\caption{Results of the hyperparameter search.
-Left: Variation of BIC with $m$ and $s$ for fixed $\alpha=0.1$.
-Right: Variation of the BIC with with $m$ and $\alpha$ for fixed $s=1.0$.
-The white star in each plot indicates the parameters with the lowest BIC.\label{fig:hp-results}}
-\end{figure}
-
-\subsection{Spectral Signatures}
-
-\begin{figure}[t]
-\begin{adjustwidth}{-\extralength}{0cm}
-\centering
-\includegraphics[width=15.5cm]{paper/figures/results/spectral-signatures.png}
-\end{adjustwidth}
-\caption{GTM Node Signatures corresponding to the exemplar spectra for the rhodamine plume (top left), algae (top right), dry grass (bottom left), and open water (bottom right).
-\label{fig:spectral-signatures}}
-\end{figure}
-
-\subsection{Water Class Map}
+\subsection{Water-only Pixel Segmentation}
 
@@
 \textbf{(a)} GTM applied to all water pixels colored by their associated position in the latent space.
-\textbf{(b)} Representation of the poitns from the training set in the latent space.
-The points are colored with $\xi_1$ corresponding to red and $\xi_2$ corresponding to blue.
-\textbf{(c)} The associated GTM spectral signatures, $\Psi_i$, corresponding to the four corners and center of the GTM latent space.
+\textbf{(b)} Representation of the points from the training set in the latent space.
+Color is assigned to each point by mapping $\xi_1$ to red and $\xi_2$ to blue.
+\textbf{(c)} GTM spectral signatures, $\psi(\xi_k)$, corresponding to nodes from the four corners and center of the GTM latent space.
 \label{fig:gtm-water}}
@@
 
-\subsection{Algae Identification}
+A GTM with $K=32\times 32$ nodes was trained on the dataset of water-only pixels described in Sec~\ref{sec:case-studies} in order to explore the distribution of reflectance spectra captured by the HSI.
+The resulting GTM is visualized in Fig~\ref{fig:gtm-water}.
+First, we note that the GTM has utilized most of the embedding space as illustrated panel (b) where the position corresponding to the mean responsibility of each data point $\hat{\xi}_n$ has been plotted in the latent space.
+For this water-only gtm, the spectra appear to be largely clustered toward the left edge, bottom, and right edge of the embedding space.
+Spectral signatures corresponding to GTM nodes from the four corners and center of the latent space are shown in panel (c) illustrating the spectral variability represented across the latent space.
+
+To visualize the distribution of HSI learned by the GTM, we can associate a color with each dimension of the latent space.
+In Fig~\ref{fig:gtm-water} we have used the red channel to represent $\xi_1$ and the blue channel to represent $\xi_2$.
+Applying the trained GTM to compute mean node responsibilities for all water-pixels in collected HSI allows us to illustrate the spatial distribution of the spectra on a map as shown in panel (a).
+Here we observe a clear distinction between water near the shore and water in the middle of the pond.
+Additionally, the eastern alcove of the pond is significantly more blue than the rest of the water reflecting the limited flow through this region.
+The close proximity of highly dissimilar GTM classes illustrated by sharp color gradients in the map captures the small-scale spatial variability typical of inland water bodies like this pond.
+
+\subsection{Endmember Extraction}
+
+\begin{figure}[t]
+\centering
+\includegraphics[width=0.8\columnwidth]{paper/figures/results/square-ndvi.pdf}
+%\caption{GTM Map: We plot the mean position of each data point in the latent space with points colored by the their associated NDVI values.
+The locations of the exemplar points for algae, rhodamine dye, water, and dry grass are shown in the latent space illustrating that the GTM has clearly distinguished land, near-land, and water pixels.\label{fig:gtm-ndvi}}
+\caption{GTM latent space visualization: Each sample spectrum from the dataset of combined HSI covering shore, water, and the rhodamine dye release are plotted in the GTM latent space at the location of the mean node responsibility, $\hat{\xi}_n$.
+Points are colored according to the NDVI computed from the original reflectance spectra.
+The locations of exemplar spectra for algae, rhodamine, water, and grass in the latent space are included as color-filled circles.
+Spectra from the shore and water are clearly separated to different regions of the latent space.\label{fig:gtm-ndvi}}
+\end{figure}
+
+A second GTM was trained on the combined dataset of HSI covering shore, water, and the rhodamine tracer dye release.
+The resulting class map is shown in Fig~\ref{fig:ndvi} where each sample has been plotted at the position of the mean node responsibility and colored by the normalized difference vegetation index (NDVI) computed from the original reflectance spectrum.
+The NDVI is a spectral index sensitive to variations in vegetation health with negative values corresponding to water, small positive values corresponding to sparse vegetation, and large values near $1$ corresponding to dense vegetation \cite{thenkabail2018hyperspectral}.
+From this, we see that the GTM can clearly separate spectra by vegetation content with high values concentrated to the left of the embedding space and negative, water-based pixels concentrated to the right.
+Additionally, the position of exemplar spectra for algae, rhodamine, water, and grass are included as color-filled circles further illustrating the classification obtained by the GTM.
+
+%\begin{figure}[t]
+\begin{figure}[h]
+\centering
+\includegraphics[width=\textwidth]{paper/figures/results/bic.pdf}
+%\end{adjustwidth}
+\caption{Results of the hyperparameter search.
+Left: Variation of BIC with $m$ and $s$ for fixed $\alpha=0.1$.
+Right: Variation of the BIC with with $m$ and $\alpha$ for fixed $s=1.0$.
+The white star in each plot indicates the parameters with the lowest BIC across the entire parameter search.\label{fig:hp-results}}
+\end{figure}
+
+As outlined in Sec~\ref{sec:gtm-overview}, there are three hyperparameters that need to be chosen to fit a GTM model: the number of RBF centers $M=m^2$ with $m$ the number along each axis, the scale factor $s$ which controls the RBF overlap, and the regularization parameter $\alpha$.
+To choose appropriate values we performed a grid search for $m$ values between $2$ and $20$, $s$ values between $0.1$ and $3.0$, and $\alpha$ values between $0.001$ and $10.0$.
+The best values were determined as those which minimized the BIC, with $m=14$, $s=1.0$, and $\alpha=0.1$, respectively.
+Heatmaps comparing the BIC for different hyperparameter values are provided in Fig~\ref{fig:hp-results} and a table of the top 25 models is given in Appendix A.
+Since the matrix of RBF activations $\Phi$ need only be computed once at the GTM initialization step, the number of latent nodes given by $K=k^2$ can be chosen to be large enough to ensure a smooth mapping.
+We found that a value of $k=32$ provided a sufficient number of GTM nodes without significantly impacting training time.
+
+\begin{figure}[t]
+\begin{adjustwidth}{-\extralength}{0cm}
+\centering
+\includegraphics[width=15.5cm]{paper/figures/results/spectral-signatures.png}
+\end{adjustwidth}
+\caption{Spectral signatures $\psi(\xi_k)$ corresponding to GTM nodes with non-zero responsibility for exemplar spectra corresponding to the rhodamine dye plume (top left), algae (top right), dry grass (bottom left), and open water (bottom right).
+A pseudo-color image is inset into each plot wit the location of the exemplar spectrum marked with a white circle.
+\label{fig:spectral-signatures}}
+\end{figure}
+
+Using the trained GTM we can compute spectral signatures for each node in the latent space via the non-linear mapping $\psi$.
+The responsibilities $R_{kn}$ therefore correspond to the contributions of the $kth$ node $\xi_k$ to the $n$th sample spectrum in the dataset.
+Endmembers are identified by nodes with non-zero responsibility values for each of the selected exemplar spectra and are plotted in Fig~\ref{fig:spectral-signatures}.
+We note that these extracted endmembers are able to accurately capture the shape of exemplar spectrum including thin reflectance peaks.
+The endmembers also manage to smoothly interpolate through noisy wavelengths as shown by the water and algae endmembers for $\lambda > 800$ nm.
+
+\subsection{Abundance Mapping with NS3}
 
@@
 \end{adjustwidth}
-\caption{\label{fig:algae-map}}
+\caption{Using the spectral endmember assigned to algae from the trained GTM together with the NS3, we are able to estimate the algal abundance near the shore.
+(\textbf{left}) NS3 values of pixels below a threshold of 0.4275 corresponding to a total area of $587.3$ $m^2$.
+(\textbf{right}) A picture of the pond near the shore showing the algae.\label{fig:algae-map}}
 \end{figure}
 
-\subsection{Rhodamine Dye Plume Identification}
+Spectral endmembers extracted from the GTM can be used to map the abundance of water constituents in HSI by selecting HSI pixels with a NS3 value below a given threshold.
+In Fig~\ref{fig:algae-map} we demonstrate this by using the extracted algae endmember spectrum to map algal abundance near the western shore of the pond.
+Identifying algae pixels as those with an NS3 value below 0.4725, we estimate the area subsumed by algae in Fig~\ref{fig:algae-map} to be $587.3$ $m^2$.
+Similarly, we are able to track the evolution of the rhodamine dye release into the western portion of the pond by computing the NS3 for HSI collected across multiple flights.
+In the left panel of Fig~\ref{fig:rhodamine-map}, we see that the dye plume initially encompasses an area of roughly $330.7$ $m^2$.
+A second flight performed 15 minutes later reveals the dye plume to have increased to an area of $1495.6$ $m^2$.
+The increase in NS3 values between flights reflects the dilution of dye as it diffused into the water.
 
@@
 \end{adjustwidth}
-\caption{\label{fig:rhodamine-map}}
+\caption{Using the spectral endmember assigned to Rhodamine from the trained GTM together with the NS3, we are able to map the evolution of the dye plume across two UAV flights.
+(\textbf{left}) The initial dye plume corresponding to a total area of $330.7$ $m^2$.
+(\textbf{right}) The same dye plume imaged approximately 15 minutes later now subsumes a total area of $1495.6$ $m^2$.\label{fig:rhodamine-map}}
 \end{figure}
@@
 \section{Discussion}
+
+- Use of UAV-based HSI for water quality assessment is becoming increasingly popular
+- Most studies have focused on supervised techniques mapping captured spectra to specific parameters of interest 
+- This approach relies on reference data which are expensive to collect and require prior knowledge of expected sources in order to select appropriate sensors 
+- Spectral unmixing provides an HSI-only alternative however this approach relies on libraries of known endmember spectra
+- Most studies exploring endmember extraction from UAV-based HSI are focused on extracting pure endmembers which can then be utilized by satellite based HSI
+
+% smart collection of reference data
+- The map of the distribution of GTM classes throughout the pond in Fig~\ref{fig:gtm-water} reveals the significant spatial variability typical of inland water bodies.
+Knowledge of this distribution is highly relevant for the collection of in situ reference data where taking samples as little as a meter apart could clearly lead to significant differences in measurements.
+- As a consequence, one useful application for the GTM is for intelligent reference data collection.
+- The ability of any supervised model to infer water quality parameters from captured HSI relies on a sufficient volume of reference measurements which reflect the true distribution in the water
+- In our previous work, we demonstrated how autonomous in situ data collection leads to significantly greater data volumes, which in turn, improves water parameter inversion models. 
+- Using the distribution obtained by the GTM, an autonomous boat can be routed to maximially sample the observed distribution
+- Cite papers where this has been done in other contexts...
+e.g.
+guided navigation of autonomous boat based on some cost function, for example, like the expected improvement functions used for Bayesian optimization strategies.
+- Such a strategy would likely reduce the sampling time and improve data quality by sampling a broader portion of the distribution. 
+- Since the GTM captures the distribution of observed spectra, we are also not constrained to the distribution of a single parameter as measured by the boat
+- That is, we previously observed the distributions for parameters like CDOM, Chlorophyll, Temperature, Turbidity, Blue-green algae, etc.
+were highly dissimilar so that guiding the boat using a single reference sensor would not accurately reflect the comprehensive distribution of water composition.
 
@@
 \begin{table}[H]
-  \caption{The top 25 models from the hyperparameter search. A variety of GTM were trained to explore the the impact of varying m, $\alpha$, and s.\label{tab:hp-search}}
+  \caption{The top 25 models from the hyperparameter search.
+A variety of GTM were trained to explore the the impact of varying m, $\alpha$, and s.
+The Bayesian Information Criterion (BIC) and Akaike Information Criterion (AIC) are given in the final two columns which can be used for hyperparameter selection.
+\label{tab:hp-search}}
   \begin{adjustwidth}{-\extralength}{0cm}
@@
     \midrule
-    14 & 0.1 & 1.0 & 32 & \textbf{-1.918e8} & -1.926e8\\
+    \textbf{14} & \textbf{0.1} & \textbf{1.0} & \textbf{32} & \textbf{-1.918e8} & \textbf{-1.926e8}\\
     13 & 0.01 & 1.0 & 32 & -1.917e8 & -1.923e8\\

### 2024-05-17 paper/paper.tex

@@
 An upward facing Ocean Optics UV-Vis NIR spectrometer with a cosine corrector was also included to provide measurements of the incident solar irradiance spectrum.
+The configuration of the UAV with the attached hyperspectral imager is shown in Fig~\ref{fig:drone}.
 Data collection and processing was controlled by an attached Intel NUC small-form-factor computer.
@@
 For the site in North Texas, this corresponded to an average solar zenith angle of $54.9^\circ$ for 23 November 2020 resulting in HSI with negligible sunglint effects.
+
+\begin{figure}[t]
+\begin{adjustwidth}{-\extralength}{0cm}
+\centering
+\includegraphics[width=15.5cm]{paper/figures/methods/annotated-drone.pdf}
+\end{adjustwidth}
+\caption{The UAV platform: (\textbf{a}) The Resonon Pika XC2 hyperspectral imager.
+(\text{b}) The Freefly Alta-X with the attached hyperspectral imager, processing computer, and downwelling irradiance spectrometer.
+\label{fig:drone}}
+\end{figure}
 
@@
 \begin{figure}[h]
+\begin{adjustwidth}{-\extralength}{0cm}
 \centering
-\includegraphics[width=\textwidth]{paper/figures/results/bic.pdf}
-%\end{adjustwidth}
+\includegraphics[width=17.0cm]{paper/figures/results/bic.pdf}
+\end{adjustwidth}
 \caption{Results of the hyperparameter search.
-Left: Variation of BIC with $m$ and $s$ for fixed $\alpha=0.1$.
-Right: Variation of the BIC with with $m$ and $\alpha$ for fixed $s=1.0$.
+(\textbf{left}) Variation of BIC with $m$ and $s$ for fixed $\alpha=0.1$.
+(\textbf{right}) Variation of the BIC with with $m$ and $\alpha$ for fixed $s=1.0$.
 The white star in each plot indicates the parameters with the lowest BIC across the entire parameter search.\label{fig:hp-results}}
@@
 \centering
-\includegraphics[width=15.5cm]{paper/figures/results/spectral-signatures.png}
+\includegraphics[width=17.0cm]{paper/figures/results/spectral-signatures.png}
 \end{adjustwidth}
@@
 \centering
-\includegraphics[width=15.5cm]{paper/figures/results/algae.png}
+\includegraphics[width=17.0cm]{paper/figures/results/algae.png}
 \end{adjustwidth}
@@
 \centering
-\includegraphics[width=15.5cm]{paper/figures/results/rhodamine.png}
+\includegraphics[width=17.0cm]{paper/figures/results/rhodamine.png}
 \end{adjustwidth}
@@
 UAV & Unmanned Aerial Vehicle \\
+GTM & Generative Topographic Mapping  \\
 SOM & Self Organizing Map \\ 
-GTM & Generative Topographic Mapping  \\
 HSI & Hyperspectral Image \\ 
-CDOM & Colored Dissolved Organic Matter \\
-HAB & Harmful Algal Bloom 
+PCA & Principal Component Analysis \\ 
+tSNE & t-Distributed Stochastic Neighbor Embedding \\
+MLJ & Machine Learning in Julia \\ 
+VNIR & Visible + Near-Infrared \\
+NDWI & Normalized Difference Water Index \\ 
+NS3 & Normalized Spectral Similarity Score
 \end{tabular}

### 2024-05-18 paper/paper.tex
updated with first draft

@@
 % \Title{Robot Team III: Scotty Strikes Back}
-\Title{Visualization and Unsupervised Classification of Water Composition with UAV-based Hyperspectral Imaging and Generative Topographic Mapping}
+\Title{Unsupervised Characterization of Water Composition with UAV-based Hyperspectral Imaging and Generative Topographic Mapping}
 
@@
 However, fitting and validating these models requires access to sufficient quantities of in situ reference data which are time-consuming and expensive to obtain.
-In this study, we demonstrate how the Generative Topographic Mapping (GTM), a Bayesian realisation of the Self Organizing Map (SOM), enables highly-detailed unsupervised classification and nonlinear endmember extraction of UAV acquired hyperspectral imagery.
-Specifically, the topographic relationship between classes maintained by the GTM presents an attractive alternative to other unsupervised approaches which do not provide a notion of similarity between learned classes.
-Using data collected across a North Texas pond, we apply the GTM to create a detailed map segmenting the area into distinct regions of interest.
-Further, by interpreting the learned classes as distinct endmembers, the GTM can be used to rapidly identify and map unique water constituents.
-We demonstrate this capability by using the GTM to map algal abundance as well as the dispersion of a rhodamine dye tracer released into the water.
+In this study, we demonstrate how the Generative Topographic Mapping (GTM), a Bayesian realisation of the Self-organizing Map, can be used to visualize high-dimensional hyperspectral imagery and extract spectral signatures corresponding to unique endmembers present in the water.
+Using data collected across a North Texas pond, we first apply the GTM to visualize the distribution of captured reflectance spectra revealing small-scale spatial variability of water composition.
+Next, we demonstrate how the nodes of the fitted GTM can be interpreted as unique spectral endmembers.
+Using extracted endmembers together with the normalized spectral similarity score, we are able to efficiently map the abundance of near shore algae as well as the evolution of a rhodamine tracer dye used to simulate water contamination by a localized source.
+%This study highlights the value of combining UAV-based hyperspectral imaging with unsupervised methods for...
 }
@@
 % Keywords
-\keyword{Hyperspectral Imaging; Remote Sensing; Unsupervised Classification; Endmember Extraction, Spectral Unmixing}
+\keyword{Hyperspectral Imaging; Remote Sensing; Unsupervised Classification; Endmember Extraction; Generative Topographic Mapping}
 
@@
 Once identified, these endmembers can be used to rapidly map the abundance of spectral features within the pond.
-We demonstrate this capability by mapping the abundance of algae near the shore and to track the evolution of a rhodamine dye plume.
+We demonstrate this capability by using GTM to map the abundance of algae near the shore as well as the dispersion of a rhodamine dye plume.
 
@@
 \end{equation}
-The width $\sigma$ is taken as the distance between neighboring RBF centers adjusted by a scale factor labelled $s$.
+The width $\sigma$ is taken to be the distance between neighboring RBF centers multiplied by a scale factor, $s$.
 to Together the number of RBFs, $M$, and $s$ are hyperparameters for the model which govern the smoothness of the resulting manifold in the data space.
@@
 \caption{Spectral signatures $\psi(\xi_k)$ corresponding to GTM nodes with non-zero responsibility for exemplar spectra corresponding to the rhodamine dye plume (top left), algae (top right), dry grass (bottom left), and open water (bottom right).
-A pseudo-color image is inset into each plot wit the location of the exemplar spectrum marked with a white circle.
+A pseudo-color image is inset into each plot with the location of the exemplar spectrum marked with a white circle.
 \label{fig:spectral-signatures}}
@@
 
-\begin{figure}[t]
+\newpage
+
+\begin{figure}
 \begin{adjustwidth}{-\extralength}{0cm}
 \centering
-\includegraphics[width=17.0cm]{paper/figures/results/rhodamine.png}
+\includegraphics[width=14.5cm]{paper/figures/results/rhodamine-vertical.png}
 \end{adjustwidth}
@@
 (\textbf{left}) The initial dye plume corresponding to a total area of $330.7$ $m^2$.
-(\textbf{right}) The same dye plume imaged approximately 15 minutes later now subsumes a total area of $1495.6$ $m^2$.\label{fig:rhodamine-map}}
+(\textbf{right}) The same dye plume imaged approximately 15 minutes later extends to a total area of $1495.6$ $m^2$.\label{fig:rhodamine-map}}
 \end{figure}
+
+\newpage
 
@@
 
-- Use of UAV-based HSI for water quality assessment is becoming increasingly popular
-- Most studies have focused on supervised techniques mapping captured spectra to specific parameters of interest 
-- This approach relies on reference data which are expensive to collect and require prior knowledge of expected sources in order to select appropriate sensors 
-- Spectral unmixing provides an HSI-only alternative however this approach relies on libraries of known endmember spectra
-- Most studies exploring endmember extraction from UAV-based HSI are focused on extracting pure endmembers which can then be utilized by satellite based HSI
+% Overview of previous approaches
 
-% smart collection of reference data
-- The map of the distribution of GTM classes throughout the pond in Fig~\ref{fig:gtm-water} reveals the significant spatial variability typical of inland water bodies.
-Knowledge of this distribution is highly relevant for the collection of in situ reference data where taking samples as little as a meter apart could clearly lead to significant differences in measurements.
-- As a consequence, one useful application for the GTM is for intelligent reference data collection.
-- The ability of any supervised model to infer water quality parameters from captured HSI relies on a sufficient volume of reference measurements which reflect the true distribution in the water
-- In our previous work, we demonstrated how autonomous in situ data collection leads to significantly greater data volumes, which in turn, improves water parameter inversion models. 
-- Using the distribution obtained by the GTM, an autonomous boat can be routed to maximially sample the observed distribution
-- Cite papers where this has been done in other contexts...
-e.g.
-guided navigation of autonomous boat based on some cost function, for example, like the expected improvement functions used for Bayesian optimization strategies.
-- Such a strategy would likely reduce the sampling time and improve data quality by sampling a broader portion of the distribution. 
-- Since the GTM captures the distribution of observed spectra, we are also not constrained to the distribution of a single parameter as measured by the boat
-- That is, we previously observed the distributions for parameters like CDOM, Chlorophyll, Temperature, Turbidity, Blue-green algae, etc.
-were highly dissimilar so that guiding the boat using a single reference sensor would not accurately reflect the comprehensive distribution of water composition.
+The application of UAV-based hyperspectral imaging (HSI) for water quality assessment is gaining significant traction due to its ability to efficiently capture detailed spectral data at high spatial resolutions.
+Most studies using UAVs have employed supervised techniques that map the captured spectra to specific water quality parameters of interest.
+For example Lu et al.
+evaluated a variety of machine learning methods for the extraction of chlorophyll-a and suspended sediment concentrations from HSI using data from UAV flights over 33 sampling locations \cite{lu2021retrieval}.
+However, this approach relies on the collection of paired in situ data which can be challenging to obtain in sufficient quantities to facilitate model training.
+Additionally, supervised methods require prior knowledge of expected sources in order to select and calibrate suitable reference instruments.
+Using models purpose-built for specific water quality parameters like chlorophyll a or turbidity discards potential information contained in HSI which can reveal unanticipated sources.
+Therefore, unsupervised methods which aid in the visualization of HSI and enable the identification of spectral endmembers within the imaging scene are needed to complement these supervised approaches.
 
-\subsection{Summary of GTM applications in our paper}
-\subsection{Comparison to other existing approaches in literature}
-- Is there much done for classification in real time? 
-\subsection{Comparison to Spectral Unmixing}
-- Here we can re-interpret the GTM model as performing a type of Bayesian spectral unmixing and suggest how the method could be further extended to work in this domain.
-Critically, other traditional methods work by observing that pure endmembers form the vertices of a simplex.
+Recently some researchers have begun to explore endmember extraction using UAV-based imaging where the increased spatial resolution provided by UAV platforms is hypothesised to yield more pure pixels than their satellite counterparts.
+For instance, Alvarez et al.
+used multi-spectral UAV imagery to extract endmembers which were then applied to unmix remote sensing data for plant abundance estimation across a broad region of France \cite{alvarez2020can}.
+Similarly, Gu et al.
+explored UAV-to-satellite hyperspectral unmixing by applying endmembers extracted from UAV-based HSI to satellite observations \cite{gu2023intrinsic}.
+These studies underscore the potential of combining endmember extraction techniques with UAV-based HSI which has yet to see widespread adoption for water quality analysis.
 
-\subsection{Other applications of GTM}
-- Discuss use of GTM means/modes/responsibilities feature transformer for supervised models (e.g. better than PCA) 
-- Use of GTM classes to aid autonomous data collection by identification of "interesting" regons
+In this study, we explored the GTM as a unsupervised approach to simultaneously enable visualization of HSI and perform endmember extraction of spectral signatures.
+First, we showed how the representation of data in the latent space of the GTM given by the mean responsibility can be used to visualize the small-scale structures within inland water bodies as evidenced in Fig~\ref{fig:gtm-water}.
+In particular, we note that the sharp gradients in colors found when visualizing the spatial distribution of GTM nodes across the pond reflects significant variability in water composition at the sub-meter scale.
+This has important consequences for water quality assessment where the location of in situ data collection will have a strong impact on the ability of any model to predict water quality parameters from HSI.
+Visualizing the distribution of spectra from collected HSI is therefore highly relevant to guide in situ data collection as shifting the sampling location by as little as a meter can lead to significant differences in measured values.
 
-- Drones can be equipped with HSI and compute to enable rapid generation of spectral indices like the NDVI for applications such as precision agriculture \cite{horstrand2019uav}
+Based on these observations, one clear application of the GTM for real-time water quality assessment is for intelligently guided reference data collection.
+In our previous work, we showed that coordinating UAV-based hyperspectral imaging with in situ data collection by an autonomous boat can dramatically improve the inversion of water quality parameters from HSI pixels \cite{robot-team-2}.
+However, the spatial distributions of parameters such as chlorophyll-a, blue-green algae, and temperature are often highly dissimilar posing a challenge for optimal route planning.
+Since the GTM estimates the full distribution of reflectance spectra and not a single water quality parameter, one could construct a prize collecting travelling salesman problem (PC-TSP) which seeks to find the optimal route maximizing the area explored in the GTM latent space \cite{balas2007prize}.
+Similar approaches have been used to guide data collection with autonomous vehicles to optimize data quality subject to resource constraints \cite{suryan2020learning}.
 
-- Mention how Bishop suggests extensions to GTM for on-line learning which could be used to accelerate training for large HSI datasets
+The second application of the GTM presented in this study is for the extraction of endmembers corresponding to unique sources observed in the HSI.
+Here the GTM is an attractive choice as it does not rely on the assumption of linear mixing and the presence of pure pixels which are easily broken by the effects of multiple scattering in realistic scenes \cite{nonlinearity-in-hsi}.
+Additionally, because the GTM is a probabilistic model, the values of model hyperparameters can be selected objectively using information criteria like the BIC.
+This is a clear advantage over similar methods like the SOM on which the GTM is based.
+Endmembers identified for exemplar spectra corresponding to rhodamine, algae, grass, and water demonstrate that the method can successfully identify spectral signatures corresponding to diverse sources.
+If a set of labeled spectra for known sources are available, their representation in the GTM latent space could be used to perform as semi-supervised classification similar to the SOM approach developed by Riese et al.
+\cite{riese2019supervised}.
+
+% something about abundance mapping
+Once endmembers are identified, estimating their abundance using the NS3 provides a quick method to map the distribution of sources in water.
+We note that the spatial distribution of algae mapped in Fig~\ref{fig:algae-map} realistically reflects clustering of algae near the shore.
+Additionally, the ability for the UAV to quickly survey the same area in rapid succession is highly advantageous for tracking the diffusion of point sources as demonstrated in Fig~\ref{fig:rhodamine-map} for the rhodmaine dye release.
+
+The primary limitation of the GTM is that the number of nodes increases exponentially with the dimension of the latent space.
+However when constrained to two dimensions in order to enable visualization of the resulting map, the number of latent nodes has a negligible impact compared to the size of the dataset on which it is trained.
+Additionally, the GTM considers individual HSI pixels and does not exploit spatial structure like other methods such as convolutional autoencoders and non-negative matrix factorization using super-pixels \cite{palsson2020convolutional,Feng2022HyperspectralUB}.
+Extensions to the GTM have been proposed to enable batch training for large datasets as well as manifold-aligned noise models which replace the fixed variance parameter $\beta^{-1}$ with a full covariance matrix \cite{gtm-bishop-2}.
+
+Finally, we note that the representation obtained by the GTM can be used for non-linear feature extraction to improve supervised models.
+Traditionally, PCA is used as a common preprocessing technique to reduce high-dimensional HSI by keeping the first $r$-many principal components.
+For example, Uddin et al.
+report improved classification of HSI by using PCA to extract features for a support vector machine \cite{uddin2021pca}.
+The GTM can similarly be used to provide a sparse representation of the input data via the latent node responsibilities $R_{kn}$ obtained for each record.
 
@@
 \section{Conclusions}
+
+In this study we present the GTM as an useful unsupervised method for the visualization of UAV-based hyperspectral imagery and associated extraction of spectral endmembers.
+Using data collected at a North Texas pond, we demonstrate how the latent space of the GTM can be used to visualize the distribution of observed reflectance spectra revealing the small-scale spatial variability of water composition.
+Spectral signatures extracted from GTM nodes are used to successfully map the abundance of algae near the shore and to track the evolution of a rhodamine tracer dye plume.
+These examples illustrate the power of combining unsupervised learning with UAV-based HSI collection for the characterization of water composition.
 

### 2024-05-20 paper/paper.tex
updated w/ Adam's suggestions

@@
 Recently, the combination of hyperspectral imaging with unmanned aerial vehicles (UAV) such as drones has emerged as a powerful approach to simultaneously address the spectral, spatial, and temporal limitations of traditional high-altitude and satellite-based collection \cite{adao2017hyperspectral,arroyo2019implementation}.
-UAV are significantly less expensive to deploy than their remote sensing counterparts, and low altitude flights enable centimeter-scale sampling while limiting the need for complicated atmospheric corrections \cite{banerjee2020uav}.
+UAVs are significantly less expensive to deploy than their remote sensing counterparts, and low altitude flights enable centimeter-scale sampling while limiting the need for complicated atmospheric corrections \cite{banerjee2020uav}.
 However, the significant increase in data volume generated by these systems presents a new challenge, namely, how to efficiently extract water quality parameters of interest from intricate pixel spectra.
@@
 In this paper, we explore the application of the GTM to UAV-acquired HSI for the characterization water quality.
-Using data collected at pond in Montague, North Texas, we first train a GTM using water-only pixels to produce a low-dimensional representation of the collected HSI.
+Using data collected at a pond in Montague, North Texas, we first train a GTM using water-only pixels to produce a low-dimensional representation of the collected HSI.
 We use this mapping to explore the highly detailed small-scale variability within the pond and discuss how this can be used to guide reference data collection.
@@
 Once identified, these endmembers can be used to rapidly map the abundance of spectral features within the pond.
-We demonstrate this capability by using GTM to map the abundance of algae near the shore as well as the dispersion of a rhodamine dye plume.
+We demonstrate this capability by using a GTM to map the abundance of algae near the shore as well as the dispersion of a rhodamine dye plume.
 
@@
 The width $\sigma$ is taken to be the distance between neighboring RBF centers multiplied by a scale factor, $s$.
-to Together the number of RBFs, $M$, and $s$ are hyperparameters for the model which govern the smoothness of the resulting manifold in the data space.
+Together the number of RBFs, $M$, and $s$ are hyperparameters for the model which govern the smoothness of the resulting manifold in the data space.
 Additionally, sparsity of $W$ can be enforced by introducing an additional hyperparameter $\alpha$ corresponding to a prior distribution over the weights given by
@@
 To simulate the dispersion of a potential contaminant source, a Rhodamine tracer dye was released into the western portion of the pond and two additional UAV flights were used to collect HSI capturing the evolution of the resulting plume.
-From these HSI a collection of $>145,000$ were sampled for model training.
+From these HSI a collection of $>145,000$ pixels were sampled for model training.
 Exemplar spectra for water, grass, algae, and rhodamine dye were identified from a sample HSI as shown in panels (b) and (c) of Fig~\ref{fig:sample-spectra}.
@@
 First, we note that the GTM has utilized most of the embedding space as illustrated panel (b) where the position corresponding to the mean responsibility of each data point $\hat{\xi}_n$ has been plotted in the latent space.
-For this water-only gtm, the spectra appear to be largely clustered toward the left edge, bottom, and right edge of the embedding space.
+For this water-only GTM, the spectra appear to be largely clustered toward the left edge, bottom, and right edge of the embedding space.
 Spectral signatures corresponding to GTM nodes from the four corners and center of the latent space are shown in panel (c) illustrating the spectral variability represented across the latent space.
@@
 A second GTM was trained on the combined dataset of HSI covering shore, water, and the rhodamine tracer dye release.
-The resulting class map is shown in Fig~\ref{fig:ndvi} where each sample has been plotted at the position of the mean node responsibility and colored by the normalized difference vegetation index (NDVI) computed from the original reflectance spectrum.
+The resulting class map is shown in Fig~\ref{fig:gtm-ndvi} where each sample has been plotted at the position of the mean node responsibility and colored by the normalized difference vegetation index (NDVI) computed from the original reflectance spectrum.
 The NDVI is a spectral index sensitive to variations in vegetation health with negative values corresponding to water, small positive values corresponding to sparse vegetation, and large values near $1$ corresponding to dense vegetation \cite{thenkabail2018hyperspectral}.
@@
 \caption{Using the spectral endmember assigned to algae from the trained GTM together with the NS3, we are able to estimate the algal abundance near the shore.
-(\textbf{left}) NS3 values of pixels below a threshold of 0.4275 corresponding to a total area of $587.3$ $m^2$.
-(\textbf{right}) A picture of the pond near the shore showing the algae.\label{fig:algae-map}}
+(\textbf{Top}) NS3 values of pixels below a threshold of 0.4275 corresponding to a total area of $587.3$ $m^2$.
+(\textbf{Bottom}) A picture of the pond near the shore showing the algae.\label{fig:algae-map}}
 \end{figure}
@@
 The second application of the GTM presented in this study is for the extraction of endmembers corresponding to unique sources observed in the HSI.
-Here the GTM is an attractive choice as it does not rely on the assumption of linear mixing and the presence of pure pixels which are easily broken by the effects of multiple scattering in realistic scenes \cite{nonlinearity-in-hsi}.
+Here the GTM is an attractive choice as it does not rely on the assumption of linear mixing and the presence of pure pixels which are easily broken by the effects of multiple scattering in realistic scenarios \cite{nonlinearity-in-hsi}.
 Additionally, because the GTM is a probabilistic model, the values of model hyperparameters can be selected objectively using information criteria like the BIC.

### 2024-05-20 paper/paper.tex
Updates from Overleaf

@@
 
-Inland water bodies present a unique challenge to characterization by remote sensing imagery due to their complex spectral characteristics and small scale spatial variability.
-The broad bands of multi-spectral imagers coupled with the irregular shape of lakes and rivers results in pixels with highly mixed signals that are easily dominated by reflectance from shore and near-shore vegetation sources \cite{koponen2002lake, ritchie2003remote}.
-Recently, the combination of hyperspectral imaging with unmanned aerial vehicles (UAV) such as drones has emerged as a powerful approach to simultaneously address the spectral, spatial, and temporal limitations of traditional high-altitude and satellite-based collection \cite{adao2017hyperspectral,arroyo2019implementation}.
-UAVs are significantly less expensive to deploy than their remote sensing counterparts, and low altitude flights enable centimeter-scale sampling while limiting the need for complicated atmospheric corrections \cite{banerjee2020uav}.
-However, the significant increase in data volume generated by these systems presents a new challenge, namely, how to efficiently extract water quality parameters of interest from intricate pixel spectra.
+Inland water bodies present a unique challenge to characterization by remote sensing imagery due to their complex spectral characteristics and small-scale spatial variability.
+The broad bands of multispectral imagers coupled with the irregular shape of lakes and rivers result in pixels with highly mixed signals that are easily dominated by reflectance from shore and nearshore vegetation sources \cite{koponen2002lake, ritchie2003remote}.
+Recently, the combination of hyperspectral imaging with unmanned aerial vehicles (UAVs), such as drones, has emerged as a powerful approach to simultaneously address the spectral, spatial and temporal limitations of traditional high-altitude and satellite-based collection \cite{adao2017hyperspectral,arroyo2019implementation}.
+UAVs are significantly less expensive to deploy than their satellite or aircraft based remote sensing counterparts, and low-altitude flights enable centimeter-scale sampling while limiting the need for complicated atmospheric corrections \cite{banerjee2020uav}.
+However, the significant increase in the data volume generated by these systems presents a new challenge, namely, how to efficiently extract water quality parameters of interest from intricate pixel spectra.
 
-Significant research efforts have focused on the development of techniques and algorithms to retrieve water quality parameters from UAV-captured hyperspectral images (HSI).
+Significant research efforts have focused on the development of techniques and algorithms to retrieve water quality parameters, e.g.
+from UAV-captured hyperspectral images (HSI).
 On-board compute installed alongside hyperspectral imagers can enable the rapid evaluation of spectral indices from HSI band ratios \cite{horstrand2019uav}.
-These band ratios and polynomial combinations of bands have been used to successfully invert optically-active water quality parameters such as turbidity directly from UAV acquired imagery \cite{vogt2016near, zhang2022selection}.
-Supervised machine learning techniques such as tree-based models, support vector machines, and neural networks have also been used to estimate a plethora of parameters such as colored dissolved organic matter, chlorophyll a, blue-green algae, and suspended sediment concentrations \cite{keller2018hyperspectral, lu2021retrieval}.
-The calibration and evaluation of these data-driven models demand a significant volume of coincident, in situ data.
+These band ratios and polynomial combinations of bands have been used to successfully invert optically active water quality parameters such as turbidity directly from UAV acquired imagery \cite{vogt2016near, zhang2022selection}.
+Supervised machine learning techniques such as tree-based models, support vector machines, and neural networks have also been used to estimate a wide range of parameters such as colored dissolved organic matter, chlorophyll A, blue-green algae, and suspended sediment concentrations \cite{keller2018hyperspectral, lu2021retrieval}.
+The calibration and evaluation of these data-driven models require a significant volume of coincident in situ data.
 This can be addressed by coordinating UAV flights with reference data collection using autonomous robotic boats \cite{robot-team-1, robot-team-2}.
-Nevertheless, this approach relies on prior knowledge of expected sources in order to select appropriate reference instruments for model validation.
+However, this approach relies on prior knowledge of expected sources in order to select appropriate reference instruments for model validation.
 The presence of unanticipated contaminants cannot be directly identified in this sensing paradigm.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\authorcontributions{FIXME}
 
-\funding{FIXME}
+\authorcontributions{Methodology, J.W.
+and D.J.L; conceptualization, D.J.L.; software, J.W.; field deployment and preparation, J.W., A.A., L.O.H.W., S.T., B.F., P.M.H.D., M.I., M.L., D.S., and D.J.L.; validation, J.W.; formal analysis, J.W.; investigation J.W.; resources, D.J.L.; data curation, J.W., A.A., L.O.H.W., and D.J.L; writing---original draft preparation, J.W.; writing---review and editing, J.W.
+and D.J.L.; visualization, J.W.; supervision, D.J.L.; project administration, D.J.L.; funding acquisition, D.J.L.
+All authors have read and agreed to the published version of the manuscript.
+}
+
+\funding{This research was funded by the following grants: the Texas National Security Network Excellence Fund award for Environmental Sensing Security Sentinels; the SOFWERX award for Machine Learning for Robotic Teams %Please ensure meaning has been retained.
+	and NSF Award OAC-2115094; support from the University of Texas at Dallas Office of Sponsored Programs, Dean of Natural Sciences and Mathematics, and Chair of the Physics Department is gratefully acknowledged; TRECIS CC* Cyberteam (NSF \#2019135); NSF OAC-2115094 Award; and EPA P3 grant number 84057001-0.
+}
 
@@
 
-\acknowledgments{In this section you can acknowledge any support given which is not covered by the author contribution or funding sections.
-This may include administrative and technical support, or donations in kind (e.g., materials used for experiments).}
+\acknowledgments{Don MacLaughlin, Scotty MacLaughlin, and the city of Plano, TX, are gratefully acknowledged for allowing us to deploy the autonomous robot team on their property.
+Christopher Simmons is gratefully acknowledged for his computational support.
+We thank  Antonio Mannino for his advice with regard to selecting the robotic boat's sensing suite.
+Annette Rogers is gratefully acknowledged for supporting the arrangement of insurance coverage.
+Steven Lyles is gratefully acknowledged for supporting the arrangement of a secure place for the robot team.
+The authors acknowledge the OIT-Cyberinfrastructure Research Computing group at the University of Texas at Dallas and the TRECIS CC* Cyberteam (NSF \#2019135) for providing HPC resources that contributed to this research; the authors also acknowledge their receipt of the %Please ensure meaning has been retained.
+	NSF OAC-2115094 Award and EPA P3 grant number 84057001-0.}
 
-\conflictsofinterest{Declare conflicts of interest or state ``The authors declare no conflicts of interest.'' Authors must identify and declare any personal circumstances or interest that may be perceived as inappropriately influencing the representation or interpretation of reported research results.
-Any role of the funders in the design of the study; in the collection, analyses or interpretation of data; in the writing of the manuscript; or in the decision to publish the results must be declared in this section.
-If there is no role, please state ``The funders had no role in the design of the study; in the collection, analyses, or interpretation of data; in the writing of the manuscript; or in the decision to publish the results''.}
+\conflictsofinterest{The authors declare no conflicts of interest.}
 

### 2024-06-18 paper/paper.tex
changed Bayesian -> probabilistic

@@
 However, fitting and validating these models requires access to sufficient quantities of in situ reference data which are time-consuming and expensive to obtain.
-In this study, we demonstrate how the Generative Topographic Mapping (GTM), a Bayesian realisation of the Self-organizing Map, can be used to visualize high-dimensional hyperspectral imagery and extract spectral signatures corresponding to unique endmembers present in the water.
+In this study, we demonstrate how the Generative Topographic Mapping (GTM), a probabilistic realization of the Self-organizing Map, can be used to visualize high-dimensional hyperspectral imagery and extract spectral signatures corresponding to unique endmembers present in the water.
 Using data collected across a North Texas pond, we first apply the GTM to visualize the distribution of captured reflectance spectra revealing small-scale spatial variability of water composition.
@@
 To address these limitations, Bishop et al.
-developed the generative topographic mapping (GTM), a Bayesian latent-variable model inspired by the SOM \cite{gtm-bishop-1}.
+developed the generative topographic mapping (GTM), a probabilistic latent-variable model inspired by the SOM \cite{gtm-bishop-1}.
 The GTM has been utilized in a variety of domains including drug design and chemical data visualization but has yet to see adoption for the analysis of hyperspectral imagery \cite{kireeva2012generative, gaspar2015chemical, horvath2019generative}.
@@
 To account for this, the points $\xi$ are described in the data space by a radially symmetric Gaussian distribution, $\mathcal{N}(\psi(\xi),\, \beta^{-1})$, with mean $\psi(\xi)$ and variance $\beta^{-1}$.
-For a dataset $\left\{ \mathbf{x}_n\right\}\limits_{n=1}^N$ consisting of $N$-many i.i.d.
-records, this choice yields a log-likelihood function of the form
+For a dataset $\left\{ \mathbf{x}_n\right\}\limits_{n=1}^N$ consisting of $N$-many records, this choice yields a log-likelihood function of the form
 \begin{equation}\label{eq:llh}
@@
 \begin{equation}
-    \text{BIC} = 2P\ln(N) - 2\mathcal{L}
+    \text{BIC} = P\ln(N) - 2\mathcal{L}
 \end{equation}
@@
 The resulting GTM is visualized in Fig~\ref{fig:gtm-water}.
-First, we note that the GTM has utilized most of the embedding space as illustrated panel (b) where the position corresponding to the mean responsibility of each data point $\hat{\xi}_n$ has been plotted in the latent space.
-For this water-only GTM, the spectra appear to be largely clustered toward the left edge, bottom, and right edge of the embedding space.
+First, we note that the GTM has utilized most of the latent space as illustrated panel (b) where the position corresponding to the mean responsibility of each data point $\hat{\xi}_n$ has been plotted.
+For this water-only GTM, the spectra appear to be largely clustered toward the left edge, bottom, and right edge of the latent space.
 Spectral signatures corresponding to GTM nodes from the four corners and center of the latent space are shown in panel (c) illustrating the spectral variability represented across the latent space.
@@
 The NDVI is a spectral index sensitive to variations in vegetation health with negative values corresponding to water, small positive values corresponding to sparse vegetation, and large values near $1$ corresponding to dense vegetation \cite{thenkabail2018hyperspectral}.
-From this, we see that the GTM can clearly separate spectra by vegetation content with high values concentrated to the left of the embedding space and negative, water-based pixels concentrated to the right.
+From this, we see that the GTM can clearly separate spectra by vegetation content with high values concentrated to the left of the latent space and negative, water-based pixels concentrated to the right.
 Additionally, the position of exemplar spectra for algae, rhodamine, water, and grass are included as color-filled circles further illustrating the classification obtained by the GTM.
@@
 \authorcontributions{Methodology, J.W.
-and D.J.L; conceptualization, D.J.L.; software, J.W.; field deployment and preparation, J.W., A.A., L.O.H.W., S.T., B.F., P.M.H.D., M.I., M.L., D.S., and D.J.L.; validation, J.W.; formal analysis, J.W.; investigation J.W.; resources, D.J.L.; data curation, J.W., A.A., L.O.H.W., and D.J.L; writing---original draft preparation, J.W.; writing---review and editing, J.W.
+and D.J.L; conceptualization, D.J.L.; software, J.W.; field deployment and preparation, J.W., A.A., L.O.H.W., S.T., B.F., P.M.H.D., M.I., M.L., D.S., G.B., and D.J.L.; validation, J.W.; formal analysis, J.W.; investigation J.W.; resources, D.J.L.; data curation, J.W., A.A., L.O.H.W., and D.J.L; writing---original draft preparation, J.W.; writing---review and editing, J.W.
 and D.J.L.; visualization, J.W.; supervision, D.J.L.; project administration, D.J.L.; funding acquisition, D.J.L.
@@
 
-\dataavailability{We encourage all authors of articles published in MDPI journals to share their research data.
-In this section, please provide details regarding where data supporting reported results can be found, including links to publicly archived datasets analyzed or generated during the study.
-Where no new data were created, or where data is unavailable due to privacy or ethical restrictions, a statement is still required.
-Suggested Data Availability Statements are available in section ``MDPI Research Data Policies'' at \url{https://www.mdpi.com/ethics}.}
+\dataavailability{The original data presented in the study are made openly available via the Open Storage Network at \url{https://ncsa.osn.xsede.org/ees230012-bucket01/RobotTeam/unsupervised/gtm-data/}.
+The open-source implementation of the GTM used in this study is available at \url{https://github.com/john-waczak/GenerativeTopographicMapping.jl}.}
 
@@
 Steven Lyles is gratefully acknowledged for supporting the arrangement of a secure place for the robot team.
-The authors acknowledge the OIT-Cyberinfrastructure Research Computing group at the University of Texas at Dallas and the TRECIS CC* Cyberteam (NSF \#2019135) for providing HPC resources that contributed to this research; the authors also acknowledge their receipt of the %Please ensure meaning has been retained.
-	NSF OAC-2115094 Award and EPA P3 grant number 84057001-0.}
+The authors acknowledge the OIT-Cyberinfrastructure Research Computing group at the University of Texas at Dallas and the TRECIS CC* Cyberteam (NSF \#2019135) for providing HPC resources that contributed to this research; the authors also acknowledge their receipt of the NSF OAC-2115094 Award and EPA P3 grant number 84057001-0.}
 

### 2024-06-18 1-datacube-exemplar-viz.jl
updated figures to change grass -> ground

@@ -218,7 +218,7 @@ important_coords = Dict(
         "R" => Data[1:462, 285, 100] ./ maximum(Data[1:idx_900, 285, 100]),
         "λs" => λs,
     ),
-    "grass" => Dict(
+    "ground" => Dict(
         "x" => xs[400],
         "y" => xs[80],
         "R" => Data[1:462, 400, 80] ./ maximum(Data[1:idx_900, 400, 80]),
@@ -264,7 +264,7 @@ s_p = scatter!(ax, [80], [235], marker=:circle, color=red, markersize=10, stroke
 s_w = scatter!(ax, [400], [320], marker=:circle, color=blue, markersize=10, strokewidth=stroke_width, strokecolor=darkblue)
 s_g = scatter!(ax, [400], [80], marker=:circle, color=brown, markersize=10, strokewidth=stroke_width, strokecolor=darkbrown)
 
-fig[1,1] = Legend(fig, [s_a, s_p, s_w, s_g], ["Algae", "Rhodamine", "Water", "Grass"], framevisible=false, orientation=:horizontal, padding=(0,0,0,0), labelsize=13, height=-5)
+fig[1,1] = Legend(fig, [s_a, s_p, s_w, s_g], ["Algae", "Rhodamine", "Water", "Ground"], framevisible=false, orientation=:horizontal, padding=(0,0,0,0), labelsize=13, height=-5)
 
 fig
 
@@ -279,10 +279,10 @@ ax = CairoMakie.Axis(fig[2,1], xlabel="λ (nm)", ylabel="Scaled Reflectance",);
 la = lines!(ax, important_coords["algae"]["λs"][1:idx_900], important_coords["algae"]["R"][1:idx_900], linewidth=2, label="Algae", color=green)
 lp = lines!(ax, important_coords["plume"]["λs"][1:idx_900], important_coords["plume"]["R"][1:idx_900], linewidth=2, label="Rhodamine", color=red)
 lw = lines!(ax, important_coords["water"]["λs"][1:idx_900], important_coords["water"]["R"][1:idx_900], linewidth=2, label="Water", color=blue)
-lg = lines!(ax, important_coords["grass"]["λs"][1:idx_900], important_coords["grass"]["R"][1:idx_900], linewidth=2, label="Grass", color=tan)
[…]

### 2024-06-18 3-gtm-robotteam.jl

@@ -97,7 +97,7 @@ M = gtm_mdl.M                          # RBF centers
 Ψ = rpt[:W] * rpt[:Φ]'                 # Projected Node Means
 llhs = rpt[:llhs]
 
-# compute responsabilities and projections
+# compute responsibilities and projections
 Rs = predict_responsibility(mach, X)
 mean_proj = DataFrame(MLJ.transform(mach, X))
 mean_proj_2 = DataFrame(MLJ.transform(mach, X2))
@@ -122,7 +122,7 @@ end
 sample_coords["plume"]["img_idxs"] = CartesianIndex(80, 235)
 sample_coords["water"]["img_idxs"] = CartesianIndex(400, 320)
 sample_coords["algae"]["img_idxs"] = CartesianIndex(285, 100)
-sample_coords["grass"]["img_idxs"] = CartesianIndex(400, 80)
+sample_coords["ground"]["img_idxs"] = CartesianIndex(400, 80)
 
 
 
@@ -277,8 +277,8 @@ save(joinpath(figures_path, "square-water-spec.png"), fig)
 
 
 # GRASS
-idx_grass = sample_coords["grass"]["idx_data"]
-idx_sort = sortperm(Rs[idx_grass, :], rev=true)
+idx_ground = sample_coords["ground"]["idx_data"]
+idx_sort = sortperm(Rs[idx_ground, :], rev=true)
 
 fig = Figure();
 ax = CairoMakie.Axis(fig[2,1], xlabel="λ (nm)", ylabel="Scaled Reflectance");
@@ -292,10 +292,10 @@ hidedecorations!(ax_inset)
 hidespines!(ax_inset)
 
 ls = []
-l_1 = lines!(ax, wavelengths[1:idx_900], Vector(X[idx_grass, 1:idx_900]), linewidth=2)
+l_1 = lines!(ax, wavelengths[1:idx_900], Vector(X[idx_ground, 1:idx_900]), linewidth=2)
 push!(ls, l_1)
 for i ∈ 1:3
-    if Rs[idx_grass, idx_sort[i]] > 0
+    if Rs[idx_ground, idx_sort[i]] > 0
         l_2 = lines!(ax, wavelengths[1:idx_900], Ψ[1:idx_900, idx_sort[i]], linewidth=2, color=(mints_colors[2], 1/i))
     end
     if i == 1
@@ -303,16 +303,16 @@ for i ∈ 1:3
     end
 end
 
[…]

### 2024-06-18 utils/viz.jl

@@ -4,7 +4,6 @@ using PythonCall
 using Images
 using HDF5
 using ProgressMeter
-
 export get_background_satmap
 
 

### 2024-06-19 1-datacube-exemplar-viz.jl
updated figures

@@ -259,6 +259,12 @@ img = heatmap!(ax, rgb_image);
 
 stroke_width = 1.5
 
+
+scatter!(ax, [285], [100], marker=:circle, color=:white , markersize=13, strokewidth=stroke_width, strokecolor=:white, )
+scatter!(ax, [80], [235], marker=:circle, color=:white , markersize=13, strokewidth=stroke_width, strokecolor=:white)
+scatter!(ax, [400], [320], marker=:circle, color=:white , markersize=13, strokewidth=stroke_width, strokecolor=:white)
+scatter!(ax, [400], [80], marker=:circle, color=:white, markersize=13, strokewidth=stroke_width, strokecolor=:white)
+
 s_a = scatter!(ax, [285], [100], marker=:circle, color=green, markersize=10, strokewidth=stroke_width, strokecolor=darkgreen, )
 s_p = scatter!(ax, [80], [235], marker=:circle, color=red, markersize=10, strokewidth=stroke_width, strokecolor=darkred)
 s_w = scatter!(ax, [400], [320], marker=:circle, color=blue, markersize=10, strokewidth=stroke_width, strokecolor=darkblue)

### 2024-06-19 6-make-maps.jl

@@ -51,14 +51,13 @@ e= -97.712413
 
 satmap = get_background_satmap(w,e,s,n)
 
+lon_min, lon_max = (-97.7168, -97.7125)
+lat_min, lat_max = (33.70075, 33.7035)
 
 CairoMakie.activate!()
 fig = Figure(size=(800,600));
 ax = CairoMakie.Axis(fig[1,1]);
 bg = heatmap!(ax, (satmap.w - lon_min)..(satmap.e - lon_min), (satmap.s - lat_min)..(satmap.n - lat_min), satmap.img);
-
-lon_min, lon_max = (-97.7168, -97.7125)
-lat_min, lat_max = (33.70075, 33.7035)
 
 # add 30 meter scale bar
 lines!(ax, [λ_scale_l - lon_min, λ_scale_r - lon_min], [ϕ_scale - lat_min, ϕ_scale - lat_min], color=:white, linewidth=5)
@@ -286,6 +285,45 @@ R, X, Y, Lon, Lat, ij_inbounds, IsInbounds = get_data_for_heatmap(f_list_dye[1])
 rmse_plume = zeros(size(R,2), size(R,3));
 ns3_plume = zeros(size(R,2), size(R,3));
 cos_plume = zeros(size(R,2), size(R,3));
+
+
+
+# compute distribution of ns3 for this datacube
+# using Statistics
+
+# R, X, Y, Lon, Lat, ij_inbounds, IsInbounds = get_data_for_heatmap(f_list_dye[1]);
+# R1, X1, Y1, Lon1, Lat1, ij_inbounds1, IsInbounds1 = get_data_for_heatmap(f_list_dye[2]);
+# R2, X2, Y2, Lon2, Lat2, ij_inbounds2, IsInbounds2 = get_data_for_heatmap(f_list_dye[3]);
+
+
+
+# idx_notnan = findall(.!(isnan.(R[1,:,:])))
+# R[:, idx_notnan]
+# ns3_plume_vals_1 = [NS3(R[:, idx], R_plume) for idx in idx_notnan]
+
+# idx_notnan = findall(.!(isnan.(R1[1,:,:])))
+# R1[:, idx_notnan]
+# ns3_plume_vals_2 = [NS3(R1[:, idx], R_plume) for idx in idx_notnan]
+
+# idx_notnan = findall(.!(isnan.(R2[1,:,:])))
[…]

### 2024-06-19 paper/paper.tex
Updates from Overleaf

@@
 \abstract{
-Unmanned Aerial Vehicles (UAVs) equipped with hyperspectral imagers have emerged as an essential technology for the characterization of inland water bodies.
+Unmanned Aerial Vehicles equipped with hyperspectral imagers have emerged as an essential technology for the characterization of inland water bodies.
 The high spectral and spatial resolutions of these systems enable the retrieval of a plethora of optically-active water quality parameters via band ratio algorithms and machine learning methods.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Introduction}
+\section{Introduction}\label{introduction}
 
@@
 Once identified, these endmembers can be used to rapidly map the abundance of spectral features within the pond.
-We demonstrate this capability by using a GTM to map the abundance of algae near the shore as well as the dispersion of a rhodamine dye plume.
+We demonstrate this capability by using a GTM to map the abundance of filamentous blue-green algae near the shore as well as the dispersion of a rhodamine dye plume.
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Materials and Methods}
+\section{Materials and Methods}\label{materials-and-methods}
 
@@
 Spectra are plotted using their geographic position with the log10-reflectance colored along the z axis and a pseudo-color image on top.
-(\textbf{b}) Points taken from a sample hyperspectral data cube corresponding to algae, rhodamine dye, water, and dry grass.
+(\textbf{b}) Points taken from a sample hyperspectral data cube corresponding to algae, rhodamine dye, water, and ground (dirt and dry grass).
 (\textbf{c}) Reflectance spectra for the exemplar points scaled so the peak value of each spectrum is $1.0$.
@@
 
-In this study, we explore the use of the GTM as a tool for dimensionality reduction and non-linear endmember extraction of hyperspectral imagery.
+In this study, we explore the use of the GTM as a tool for dimensionality reduction and unsupervised endmember extraction of hyperspectral imagery.
 To this end, a dataset of HSI were collected at a pond in Montague, North Texas (shown in panel d of Fig~\ref{fig:sample-spectra}) on 23 November 2020 using a UAV mounted hyperspectral imager configured as described in \cite{robot-team-1, robot-team-2}.
@@
 Next we describe the UAV platform used for HSI collection and the various steps used to processes captured HSI.
-Finally, we describe the two case studies presented in this paper for the utilization of the GTM as a HSI visualization tool and as an endmember extraction technique.
+\textcolor{red}{Finally, we describe the two case studies presented in this paper for the utilization of the GTM as a HSI visualization tool and as an endmember extraction technique.}
 
@@
 Consequently, HSI were rapidly georectified to a $10$ cm resolution using position and orientation data from the embedded GPS/INS as outlined in \cite{muller2002program, baumker2001new, mostafa2000multi}.
-An example hyperspectral data cube is illustrated in panel (a) of Fig~\ref{fig:spectral-signatures} where the log10-reflectance values are plotted along the z-axis for each pixel in the scene and a psuedo-color image at the top of the data cube illustrates how the water would appear to the human eye from the perspective of the UAV.
+An example hyperspectral data cube is illustrated in panel (a) of Fig~\ref{fig:sample-spectra} where the log10-reflectance values are plotted along the z-axis for each pixel in the scene and a psuedo-color image at the top of the data cube illustrates how the water would appear to the human eye from the perspective of the UAV.
 
@@
 \begin{equation}
-    \text{NDWI} = \dfrac{R(550) - R(860)}{R(550) + R(860)}.
+    \text{NDWI} = \dfrac{R(550) - R(860)}{R(550) + R(860)}
 \end{equation}
+where $\lambda=550$ and $\lambda=860$ nm were chosen to represent the green and NIR bands as described in \cite{ndwi}.
 We then apply the trained GTM to all water-only pixels to visualize the distribution learned by the GTM and examine spectra associated with a subset of nodes in the latent space in order to assess the small-scale spatial variability within the pond.
@@
 From these HSI a collection of $>145,000$ pixels were sampled for model training.
-Exemplar spectra for water, grass, algae, and rhodamine dye were identified from a sample HSI as shown in panels (b) and (c) of Fig~\ref{fig:sample-spectra}.
+Exemplar spectra from water, ground, algae, and rhodamine dye were identified from a sample HSI as shown in panels (b) and (c) of Fig~\ref{fig:sample-spectra}.
 Using these spectra, we explore the trained GTM and use it to extract spectral signatures corresponding to these endmember categories.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Results}
+\section{Results}\label{results}
 
@@
 
-A GTM with $K=32\times 32$ nodes was trained on the dataset of water-only pixels described in Sec~\ref{sec:case-studies} in order to explore the distribution of reflectance spectra captured by the HSI.
+A GTM with $K=32\times 32$ nodes was trained on the dataset of water-only pixels described in Section ~\ref{sec:case-studies} in order to explore the distribution of reflectance spectra captured by the HSI.
 The resulting GTM is visualized in Fig~\ref{fig:gtm-water}.
@@
 \includegraphics[width=0.8\columnwidth]{paper/figures/results/square-ndvi.pdf}
-%\caption{GTM Map: We plot the mean position of each data point in the latent space with points colored by the their associated NDVI values.
-The locations of the exemplar points for algae, rhodamine dye, water, and dry grass are shown in the latent space illustrating that the GTM has clearly distinguished land, near-land, and water pixels.\label{fig:gtm-ndvi}}
 \caption{GTM latent space visualization: Each sample spectrum from the dataset of combined HSI covering shore, water, and the rhodamine dye release are plotted in the GTM latent space at the location of the mean node responsibility, $\hat{\xi}_n$.
 Points are colored according to the NDVI computed from the original reflectance spectra.
-The locations of exemplar spectra for algae, rhodamine, water, and grass in the latent space are included as color-filled circles.
+The locations of exemplar spectra for algae, rhodamine, water, and ground points in the latent space are included as color-filled circles.
 Spectra from the shore and water are clearly separated to different regions of the latent space.\label{fig:gtm-ndvi}}
@@
 From this, we see that the GTM can clearly separate spectra by vegetation content with high values concentrated to the left of the latent space and negative, water-based pixels concentrated to the right.
-Additionally, the position of exemplar spectra for algae, rhodamine, water, and grass are included as color-filled circles further illustrating the classification obtained by the GTM.
+Additionally, the position of exemplar spectra for algae, rhodamine, water, and ground points are included as color-filled circles further illustrating the classification obtained by the GTM.
 
@@
 \end{adjustwidth}
-\caption{Spectral signatures $\psi(\xi_k)$ corresponding to GTM nodes with non-zero responsibility for exemplar spectra corresponding to the rhodamine dye plume (top left), algae (top right), dry grass (bottom left), and open water (bottom right).
+\caption{Spectral signatures $\psi(\xi_k)$ corresponding to GTM nodes with non-zero responsibility for exemplar spectra corresponding to the rhodamine dye plume (top left), algae (top right), ground (bottom left), and open water (bottom right).
 A pseudo-color image is inset into each plot with the location of the exemplar spectrum marked with a white circle.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Discussion}
+\section{Discussion}\label{discussion}
 
@@
 
-The second application of the GTM presented in this study is for the extraction of endmembers corresponding to unique sources observed in the HSI.
+The second application of the GTM presented in this study is for the unsupervised extraction of endmembers corresponding to unique sources observed in the HSI.
 Here the GTM is an attractive choice as it does not rely on the assumption of linear mixing and the presence of pure pixels which are easily broken by the effects of multiple scattering in realistic scenarios \cite{nonlinearity-in-hsi}.
@@
 This is a clear advantage over similar methods like the SOM on which the GTM is based.
-Endmembers identified for exemplar spectra corresponding to rhodamine, algae, grass, and water demonstrate that the method can successfully identify spectral signatures corresponding to diverse sources.
+Endmembers identified for exemplar spectra corresponding to rhodamine, algae, ground, and water pixels demonstrate that the method can successfully identify spectral signatures corresponding to diverse sources.
 If a set of labeled spectra for known sources are available, their representation in the GTM latent space could be used to perform as semi-supervised classification similar to the SOM approach developed by Riese et al.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Conclusions}
+\section{Conclusions}\label{conclusions}
 

### 2024-06-20 paper/paper.tex
Updated text with edits for Reviewer #2

@@
 from UAV-captured hyperspectral images (HSI).
-On-board compute installed alongside hyperspectral imagers can enable the rapid evaluation of spectral indices from HSI band ratios \cite{horstrand2019uav}.
+On-board computers installed alongside hyperspectral imagers can enable the rapid evaluation of spectral indices from HSI band ratios \cite{horstrand2019uav}.
 These band ratios and polynomial combinations of bands have been used to successfully invert optically active water quality parameters such as turbidity directly from UAV acquired imagery \cite{vogt2016near, zhang2022selection}.
@@
 In this study, we explore the use of the GTM as a tool for dimensionality reduction and unsupervised endmember extraction of hyperspectral imagery.
-To this end, a dataset of HSI were collected at a pond in Montague, North Texas (shown in panel d of Fig~\ref{fig:sample-spectra}) on 23 November 2020 using a UAV mounted hyperspectral imager configured as described in \cite{robot-team-1, robot-team-2}.
+To this end, a dataset of HSI was collected at a pond in Montague, North Texas (shown in panel d of Fig~\ref{fig:sample-spectra}) on 23 November 2020 using a UAV mounted hyperspectral imager configured as described in \cite{robot-team-1, robot-team-2}.
 In this section we first provide a detailed overview of the GTM algorithm.
-Next we describe the UAV platform used for HSI collection and the various steps used to processes captured HSI.
+Next we describe the UAV platform used for HSI collection and the various steps used to process captured HSI.
 \textcolor{red}{Finally, we describe the two case studies presented in this paper for the utilization of the GTM as a HSI visualization tool and as an endmember extraction technique.}
@@
 
-Points $\xi$ in this latent space are mapped to the embedded manifold by a non-linear function $\psi$ parameterized by weights $W$ as illustrated in Fig~\ref{fig:gtm-diagram}.
+Points $\xi$ in this latent space are mapped to the embedded manifold by a non-linear function $\psi$.
 However, since real data is rarely noise-free, this embedded manifold will not be perfectly thin.
-To account for this, the points $\xi$ are described in the data space by a radially symmetric Gaussian distribution, $\mathcal{N}(\psi(\xi),\, \beta^{-1})$, with mean $\psi(\xi)$ and variance $\beta^{-1}$.
-For a dataset $\left\{ \mathbf{x}_n\right\}\limits_{n=1}^N$ consisting of $N$-many records, this choice yields a log-likelihood function of the form
+To account for this, the points $\xi$ are described in the data space by a radially symmetric Gaussian distribution, $\mathcal{N}(\psi(\xi),\, \beta^{-1})$, with mean $\psi(\xi)$ and precision $\beta$.
+For a dataset $\left\{ \mathbf{x}_n\right\}\limits_{n=1}^N$ consisting of $N$-many records, this choice yields a log likelihood function of the form
 \begin{equation}\label{eq:llh}
@@
 \end{equation}
-which can be maximized to yield optimal $W$ and $\beta^{-1}$.
+which can be maximized to obtain optimal $W$ and $\beta^{-1}$.
 
-\begin{figure}[t]
+The nonlinear mapping $\psi$ is typically chosen to be given by the sum of $M$-many radial basis functions (RBF) evenly distributed in the latent space such that $\psi(\xi) = W\phi(\xi)$ with centers $\mu_m$ and width $\sigma$ so that
+\begin{equation}
+    \phi_m(\xi) = \exp\left(-\dfrac{\lVert \xi - \mu_m \rVert^2}{2\sigma^2}\right).
+\end{equation}
+The width $\sigma$ is taken to be the distance between neighboring RBF centers multiplied by a scale factor, $s$.
+Together the number of RBFs, $M$, and $s$ are hyperparameters for the model which govern the smoothness of the resulting manifold in the data space.
+A visual representation of the GTM is shown in Fig~\ref{fig:gtm-diagram}.
+Additionally, sparsity of $W$ can be enforced by introducing an additional hyperparameter $\alpha$ corresponding to a prior distribution over the weights given by
+\begin{equation}\label{eq:weight-prior}
+    p(W \mid \alpha) =  \left( \frac{\alpha}{2\pi} \right)^{MD/2}\exp\left(-\frac{\alpha}{2}\left\lVert W \right\rVert_{F}^2\right).
+\end{equation}
+
+\begin{figure}[h!]
 %\begin{adjustwidth}{-\extralength}{0cm}
@@
 On the left is a regular grid of $K$ many points (green) in the latent space represented by their coordinates $\xi_1$ and $\xi_2$.
-In red are the $M$-many RBF which define the mapping $\psi$ from the latent space to the data space.
-Points in the latent space are mapped non-linearly to the data space yielding an embedded manifold in $\mathcal{R}^d$, here illustrated in three dimensions.\label{fig:gtm-diagram}}
+In red are the $M$-many radial basis functions which define the mapping $\psi$ from the latent space to the data space.
+Points in the latent space are mapped non-linearly to the data space yielding an embedded manifold in $\mathcal{R}^d$, here shown in three dimensions.\label{fig:gtm-diagram}}
 \end{figure}
 
-The function $\psi$ is typically chosen to be given by the sum of $M$-many radial basis functions (RBF) evenly distributed in the latent space such that $\psi(\xi) = W\phi(\xi)$ with centers $\mu_m$ and width $\sigma$ so that
-\begin{equation}
-    \phi_m(\xi) = \exp\left(-\dfrac{\lVert \xi - \mu_m \rVert^2}{2\sigma^2}\right).
-\end{equation}
-The width $\sigma$ is taken to be the distance between neighboring RBF centers multiplied by a scale factor, $s$.
-Together the number of RBFs, $M$, and $s$ are hyperparameters for the model which govern the smoothness of the resulting manifold in the data space.
-Additionally, sparsity of $W$ can be enforced by introducing an additional hyperparameter $\alpha$ corresponding to a prior distribution over the weights given by
-\begin{equation}\label{eq:weight-prior}
-    p(W \mid \alpha) =  \left( \frac{\alpha}{2\pi} \right)^{MD/2}\exp\left(-\frac{\alpha}{2}\left\lVert W \right\rVert_{F}^2\right).
-\end{equation}
-
-The use of RBFs for $\psi$ enables maximizing $\mathcal{L}(W,\beta)$ via an Expectation-Maximization routine.
-During each step of the fitting process, the responsibility of the $k$th node for the $n$th data record is computed as
+The occurrence of the sum within the natural logarithm in Eq~\ref{eq:llh} prevents an analytic solution for the maximum of $\mathcal{L}$.
+However, the form of $\psi$ enables an Expectation-Maximization routine which optimizes $\mathcal{L}$ by iteratively increasing a lower bound given by the expected complete-data log likelihood \cite{em-algorithm}.
+During the expectation step of the fitting process the posterior probability $p(\xi_k \mid \mathbf{x}_n, W, \beta)$, i.e.
+the \textit{responsibility} of the $k$th node for explaining the $n$th record, is computed as
 \begin{equation}\label{eq:responsibility}
@@
 where $\Phi_{km} = \phi_m(\xi_k)$, $X$ is the data matrix formed by concatenating the records $\mathbf{x}_n$, and $G$ is a diagonal matrix with $G_{kk} = \sum\limits_n^N R_{kn}$.
-This process is repeated until the log-likelihood converges to a predetermined tolerance level.
+This process is repeated until the log likelihood converges to a predetermined tolerance level.
 
@@
 \begin{equation}\label{eq:reflectance}
-    R(\lambda) = \pi L(\lambda)/E_d(\lambda)
+    \rho(\lambda) = \pi L(\lambda)/E_d(\lambda)
 \end{equation}
-where $L$ is the spectral radiance, $E_d$ is the downwelling irradiance, and a factor of $\pi$ steradians results from assuming the water surface is Lambertian (diffuse) \cite{ruddick2019review}.
+where $L$ is the spectral radiance, $E_d$ is the downwelling irradiance, and a factor of $\pi$ steradians results from assuming Lambertian (diffuse) upwelling radiance \cite{ruddick2019review}.
 HSI collection was performed near solar noon to maximize the amount of incident sunlight illuminating the water.
@@
 \begin{equation}
-    \text{NDWI} = \dfrac{R(550) - R(860)}{R(550) + R(860)}
+    \text{NDWI} = \dfrac{\rho(550) - \rho(860)}{\rho(550) + \rho(860)}
 \end{equation}
@@
 \end{equation}
-where $P$ is the total number of model parameters, $N$ is the number of records in the dataset, and $\mathcal{L}$ is the log-likelihood defined in Eq~\ref{eq:llh}.
+where $P$ is the total number of model parameters, $N$ is the number of records in the dataset, and $\mathcal{L}$ is the log likelihood defined in Eq~\ref{eq:llh}.
 
 The normalized spectral similarity score (NS3) introduced by Nidamanuri and Zbell combines the root-mean-square (RMS) difference together with the spectral angle to provide a spectral distance function\cite{nidamanuri2010normalized}.
-For two spectra $R_1(\lambda)$ and $R_2(\lambda)$ it is defined by
+For two spectra $\rho_1(\lambda)$ and $\rho_2(\lambda)$ it is defined by
 \begin{equation}
-    \text{NS3}(R_1, R_2) &= \sqrt{RMS(R_1, R_2)^2 + (1-\cos\theta)^2}
+    \text{NS3}(\rho_1, \rho_2) &= \sqrt{RMS(\rho_1, \rho_2)^2 + (1-\cos\theta)^2}
 \end{equation}
@@
 \begin{align}
-    \text{RMS}(R_1, R_2) &= \sqrt{\frac{1}{N-1}\sum_\lambda \left(R_{1}(\lambda) - R_2(\lambda) \right)^2} \\
-    \cos\theta &= \frac{\langle R_1 , R_2 \rangle}{\lVert R_1\rVert \lVert R_2 \rVert}.
+    \text{RMS}(\rho_1, \rho_2) &= \sqrt{\frac{1}{N-1}\sum_\lambda \left(\rho_{1}(\lambda) - \rho_2(\lambda) \right)^2} \\
+    \cos\theta &= \frac{\langle \rho_1 , \rho_2 \rangle}{\lVert \rho_1\rVert \lVert \rho_2 \rVert}.
 \end{align}
@@
 
-A second GTM was trained on the combined dataset of HSI covering shore, water, and the rhodamine tracer dye release.
-The resulting class map is shown in Fig~\ref{fig:gtm-ndvi} where each sample has been plotted at the position of the mean node responsibility and colored by the normalized difference vegetation index (NDVI) computed from the original reflectance spectrum.
+A second GTM was trained on the combined dataset of HSI with pixels covering shore, water, and the rhodamine tracer dye release.
+The resulting \textcolor{red}{latent space distribution} is shown in Fig~\ref{fig:gtm-ndvi} where each sample has been plotted at the position of the mean node responsibility and colored by the normalized difference vegetation index (NDVI) computed from the original reflectance spectrum.
 The NDVI is a spectral index sensitive to variations in vegetation health with negative values corresponding to water, small positive values corresponding to sparse vegetation, and large values near $1$ corresponding to dense vegetation \cite{thenkabail2018hyperspectral}.
 From this, we see that the GTM can clearly separate spectra by vegetation content with high values concentrated to the left of the latent space and negative, water-based pixels concentrated to the right.
-Additionally, the position of exemplar spectra for algae, rhodamine, water, and ground points are included as color-filled circles further illustrating the classification obtained by the GTM.
+Additionally, the position of exemplar spectra for algae, rhodamine, water, and ground points are included as color-filled circles further illustrating the separation of distinct sources obtained by the GTM.
+\textcolor{red}{We note that a portion of the near-shore algae was observed to float at the surface of the water near the shore.
+Consequently, the position of the exemplar spectrum in Fig~\ref{fig:gtm-ndvi} used for algae occurs at a non-negative NDVI.
+The accumulation of records with negative NDVI values in close proximity (in the latent space) to the selected algae spectrum reflects the presence of algae within the water.}
 
@@
 
-Using the trained GTM we can compute spectral signatures for each node in the latent space via the non-linear mapping $\psi$.
+\textcolor{red}{Examining the latent space distribution learned by the GTM provides an unsupervised method to extract endmember spectra corresponding to unique sources within the imaging scene.
+To demonstrate this clearly, we consider the exemplar spectra visually identified in Fig~\ref{fig:sample-spectra}.} Using the trained GTM we can compute spectral signatures for each node in the latent space via the non-linear mapping $\psi$.
 The responsibilities $R_{kn}$ therefore correspond to the contributions of the $kth$ node $\xi_k$ to the $n$th sample spectrum in the dataset.
-Endmembers are identified by nodes with non-zero responsibility values for each of the selected exemplar spectra and are plotted in Fig~\ref{fig:spectral-signatures}.
-We note that these extracted endmembers are able to accurately capture the shape of exemplar spectrum including thin reflectance peaks.
-The endmembers also manage to smoothly interpolate through noisy wavelengths as shown by the water and algae endmembers for $\lambda > 800$ nm.
+\textcolor{red}{Endmembers are identified by nodes with non-zero responsibility values for each of the selected exemplar spectra and are plotted in Fig~\ref{fig:spectral-signatures}.} We note that these extracted endmembers are able to accurately capture the shape of each exemplar spectrum including thin reflectance peaks.
+The GTM has also managed to smoothly interpolate through noisy wavelengths as shown by the water and algae endmembers for $\lambda > 800$ nm.
 
@@
 
-Spectral endmembers extracted from the GTM can be used to map the abundance of water constituents in HSI by selecting HSI pixels with a NS3 value below a given threshold.
-In Fig~\ref{fig:algae-map} we demonstrate this by using the extracted algae endmember spectrum to map algal abundance near the western shore of the pond.
-Identifying algae pixels as those with an NS3 value below 0.4725, we estimate the area subsumed by algae in Fig~\ref{fig:algae-map} to be $587.3$ $m^2$.
+Spectral endmembers extracted from the GTM can be used to map the abundance of water constituents in HSI by identifying pixels with a NS3 value below a given threshold.
+In Fig~\ref{fig:algae-map} we demonstrate this by using the extracted algae endmember spectrum to map the abundance of filamentous blue-green algae near the western shore of the pond.
+\textcolor{red}{The distribution of NS3 values was multi-modal with the majority of values having an NS3 $>0.4$ corresponding to dissimilar spectra.
+Consequently, a cutoff threshold of $0.4725$ was chosen to distinguish algae values corresponding to the $20$th-quantile of the NS3 distribution.} The estimated area subsumed by algae is shown in Fig~\ref{fig:algae-map} and calculated to be $587.3$ $m^2$.
 Similarly, we are able to track the evolution of the rhodamine dye release into the western portion of the pond by computing the NS3 for HSI collected across multiple flights.
-In the left panel of Fig~\ref{fig:rhodamine-map}, we see that the dye plume initially encompasses an area of roughly $330.7$ $m^2$.
+\textcolor{red}{Here the same cutoff threshold was used corresponding to the $33$rd-quantile of the NS3 distribution.} In the left panel of Fig~\ref{fig:rhodamine-map}, we see that the dye plume initially encompasses an area of roughly $330.7$ $m^2$.
 A second flight performed 15 minutes later reveals the dye plume to have increased to an area of $1495.6$ $m^2$.

### 2024-06-24 paper/paper.tex

@@
 We use this mapping to explore the highly detailed small-scale variability within the pond and discuss how this can be used to guide reference data collection.
-Next, we demonstrate how the GTM can be utilized to identify relevant spectral endmembers from a combined dataset including land pixels, algae, water, and a simulated contaminant plume using a rhodamine tracer dye.
+Next, we demonstrate how the GTM can be utilized to identify relevant spectral endmembers from a combined dataset including ground pixels, near-shore filamentous blue-green algae, water, and a simulated contaminant plume using a rhodamine tracer dye.
 Once identified, these endmembers can be used to rapidly map the abundance of spectral features within the pond.
@@
 In this study, we explore the use of the GTM as a tool for dimensionality reduction and unsupervised endmember extraction of hyperspectral imagery.
-To this end, a dataset of HSI was collected at a pond in Montague, North Texas (shown in panel d of Fig~\ref{fig:sample-spectra}) on 23 November 2020 using a UAV mounted hyperspectral imager configured as described in \cite{robot-team-1, robot-team-2}.
-In this section we first provide a detailed overview of the GTM algorithm.
-Next we describe the UAV platform used for HSI collection and the various steps used to process captured HSI.
-\textcolor{red}{Finally, we describe the two case studies presented in this paper for the utilization of the GTM as a HSI visualization tool and as an endmember extraction technique.}
+To this end, a dataset of HSI was collected at a pond in Montague, North Texas on 23 November 2020 using a UAV mounted hyperspectral imager configured as described in \cite{robot-team-1, robot-team-2}.
+The pond spans an area $<0.1$ $\text{km}^2$ and has a maximum depth of $3$ m.
+Across multiple flights, many data cubes like the example shown in panel a of Fig~\ref{fig:sample-spectra} were collected Additionally, rhodamine dye was released and subsequently imaged to simulate a localized contaminant source.
+Using the HSI shown in Fig~\ref{fig:sample-spectra}, a set of exemplar spectra were visually identified corresponding to near-shore filamentous blue-green algae, water, rhodamine dye, and the ground (a combination of soil and dry grass).
+The location of these samples is shown in panel b and their corresponding reflectance spectra are visualized in panel c.
+In the remainder of this section we first provide a detailed overview of the GTM algorithm.
+Next we describe the UAV platform used for HSI collection and the various steps employed to process captured HSI.
+Finally, we describe the two case studies presented in this paper for the utilization of the GTM as a HSI visualization tool and as an endmember extraction technique.
 
@@
 
-Points $\xi$ in this latent space are mapped to the embedded manifold by a non-linear function $\psi$.
+Points $\xi$ in this latent space are mapped to the embedded manifold by a non-linear function $\psi$ parameterized by weights $W$.
 However, since real data is rarely noise-free, this embedded manifold will not be perfectly thin.
@@
 Therefore the final responsibility matrix after GTM training can be used to represent each record in the latent space by the mean:
-\begin{equation}
+\begin{equation}\label{eq:mean-resp}
     \hat{\xi}_n = \sum_{k}^K R_{kn}\xi_k.
@@
 HSI collection was performed near solar noon to maximize the amount of incident sunlight illuminating the water.
-For the site in North Texas, this corresponded to an average solar zenith angle of $54.9^\circ$ for 23 November 2020 resulting in HSI with negligible sunglint effects.
+For the site in North Texas, this corresponded to an average solar zenith angle of $54.9^\circ$ for 23 November 2020.
+During flights, the hyperspectral imager was oriented to nadir resulting in HSI with negligible sunglint effects.
 
@@
 \caption{The UAV platform: (\textbf{a}) The Resonon Pika XC2 hyperspectral imager.
-(\text{b}) The Freefly Alta-X with the attached hyperspectral imager, processing computer, and downwelling irradiance spectrometer.
+(\textbf{b}) The Freefly Alta-X with the attached hyperspectral imager, processing computer, and downwelling irradiance spectrometer.
 \label{fig:drone}}
@@
 Next, we consider a dataset of combined HSI pixels including both water and land.
-To simulate the dispersion of a potential contaminant source, a Rhodamine tracer dye was released into the western portion of the pond and two additional UAV flights were used to collect HSI capturing the evolution of the resulting plume.
+To simulate the dispersion of a potential contaminant source, a rhodamine tracer dye was released into the western portion of the pond and two additional UAV flights were used to collect HSI capturing the evolution of the resulting plume.
 From these HSI a collection of $>145,000$ pixels were sampled for model training.
-Exemplar spectra from water, ground, algae, and rhodamine dye were identified from a sample HSI as shown in panels (b) and (c) of Fig~\ref{fig:sample-spectra}.
-Using these spectra, we explore the trained GTM and use it to extract spectral signatures corresponding to these endmember categories.
+To demonstrate the ability to extract relevant spectral endmembers from the trained GTM, we utilize the exemplar spectra corresponding to open water, ground, near-shore algae, and rhodamine dye visually identified in Fig~\ref{fig:sample-spectra}.
+The responsibility matrix obtained by the trained GTM provides a probability distribution over the latent space nodes for every data record.
+We identify relevant endmembers for each reference spectrum via the set of nodes with non-zero responsibility.
+We then utilize the mapping $\psi$ to obtain the spectral representation of these nodes.
+Since the GTM is unsupervised, these extracted endmembers do not correspond to individual records, but rather spectral features learned by the GTM.
+
 Values for the hyperparameters $m$, $s$, and $\alpha$, are determined by training multiple GTM models and selecting values which minimize the Bayesian Information Criterion (BIC) given by 
@@
 \end{align}
-We use the NS3 together with identified spectral endmembers to map the abundance of algae near the shore as well as the evolution of the Rhodamine dye plume.
+
+We use the NS3 together with extracted GTM endmembers to map the abundance of filamentous blue-green algae near the shore as well as the evolution of the rhodamine dye plume.
+To do this, the GTM node with maximum responsibility is selected from the endmembers identified for the algae and rhodamine reference spectra.
+The distribution of NS3 values between each endmember and all HSI pixels is then used to select a suitable cutoff threshold.
+The area subsumed by HSI pixels with an NS3 below this threshold is then used to infer the spatial extent of algae and rhodamine in the pond.
+
+% add something about thresholding here...
 
@@
 \end{adjustwidth}
-\caption{Classification map for GTM trained solely on water pixels (no land and no rhodamine plume).
-\textbf{(a)} GTM applied to all water pixels colored by their associated position in the latent space.
-\textbf{(b)} Representation of the points from the training set in the latent space.
-Color is assigned to each point by mapping $\xi_1$ to red and $\xi_2$ to blue.
-\textbf{(c)} GTM spectral signatures, $\psi(\xi_k)$, corresponding to nodes from the four corners and center of the GTM latent space.
-\label{fig:gtm-water}}
+\caption{Visualization of the GTM trained solely on water pixels (no land and no rhodamine plume).
+Each data point is colored by computing the mean latent space position, $\hat{\xi}_n$ according to Eq~\ref{eq:mean-resp} with red corresponding to the $\xi_1$ coordinate and blue corresponding to the $\xi_2$ coordinate.
+\textbf{(a)} Spatial distribution learned by the GTM.
+\textbf{(b)} The distribution of data in the latent space.
+Nodes corresponding to the four corners and center of the latent space are identified with cyan markers.
+\textbf{(c)} reflectance spectra corresponding to the selected GTM nodes computed via the nonlinear mapping $\psi(\xi_k$).\label{fig:gtm-water}}
 \end{figure}
@@
 Here we observe a clear distinction between water near the shore and water in the middle of the pond.
-Additionally, the eastern alcove of the pond is significantly more blue than the rest of the water reflecting the limited flow through this region.
-The close proximity of highly dissimilar GTM classes illustrated by sharp color gradients in the map captures the small-scale spatial variability typical of inland water bodies like this pond.
+Additionally, the eastern alcove of the pond is significantly more blue than the rest of the water.
+We note that flow through this region is restricted by the small entrance near the cyan square shown in panel a.
+However, despite its similar depth and surface characteristics, the GTM has distinguished this region from the rest of the water.
+
+The close proximity of highly dissimilar GTM classes illustrated by sharp color gradients in the map reflects the small-scale spatial variability typical of inland water bodies like this pond.
+For example, water near the cyan diamond in panel a of Fig~\ref{fig:gtm-water} includes both blue and red pixels in close proximity.
+As shown in panel b, these colors correspond to the top-left and bottom-right corners of the GTM latent space.
+As the GTM model maps neighboring nodes to similar spectra, this region includes maximally dissimilar pixels.
+Furthermore, this point is away from the shore where the depth is approximately uniform.
+The precision parameter $\beta$ learned by the GTM enables the model to account for spectral variability so that effects from localized disturbances such as surface ripples are constrained to small displacements in the latent space.
+With these considerations in mind, contrast in this region of the pond therefore reflects real variability of water composition, surface sediments, and vegetation.
 
@@
 The locations of exemplar spectra for algae, rhodamine, water, and ground points in the latent space are included as color-filled circles.
-Spectra from the shore and water are clearly separated to different regions of the latent space.\label{fig:gtm-ndvi}}
+Spectra from the shore and open water are clearly separated to different regions of the latent space.\label{fig:gtm-ndvi}}
 \end{figure}
@@
 \end{adjustwidth}
-\caption{Spectral signatures $\psi(\xi_k)$ corresponding to GTM nodes with non-zero responsibility for exemplar spectra corresponding to the rhodamine dye plume (top left), algae (top right), ground (bottom left), and open water (bottom right).
+\caption{Spectral signatures $\psi(\xi_k)$ corresponding to GTM nodes with non-zero responsibility for the rhodamine dye plume (top left), near-shore algae (top right), ground (bottom left), and open water (bottom right).
 A pseudo-color image is inset into each plot with the location of the exemplar spectrum marked with a white circle.
@@
 \caption{Using the spectral endmember assigned to algae from the trained GTM together with the NS3, we are able to estimate the algal abundance near the shore.
-(\textbf{Top}) NS3 values of pixels below a threshold of 0.4275 corresponding to a total area of $587.3$ $m^2$.
-(\textbf{Bottom}) A picture of the pond near the shore showing the algae.\label{fig:algae-map}}
+(\textbf{Left}) NS3 values of pixels below a threshold of 0.4275 corresponding to a total area of $587.3$ $m^2$.
+(\textbf{Right}) A picture of the pond near the shore showing the algae.\label{fig:algae-map}}
 \end{figure}
@@
 In Fig~\ref{fig:algae-map} we demonstrate this by using the extracted algae endmember spectrum to map the abundance of filamentous blue-green algae near the western shore of the pond.
-\textcolor{red}{The distribution of NS3 values was multi-modal with the majority of values having an NS3 $>0.4$ corresponding to dissimilar spectra.
-Consequently, a cutoff threshold of $0.4725$ was chosen to distinguish algae values corresponding to the $20$th-quantile of the NS3 distribution.} The estimated area subsumed by algae is shown in Fig~\ref{fig:algae-map} and calculated to be $587.3$ $m^2$.
+\textcolor{red}{The distribution of NS3 values was multi-modal with a majority of pixels above $0.4$ belonging to dissimilar spectra.
+Consequently, a cutoff threshold of $0.4275$ was chosen to distinguish algae values corresponding to the $20$th-quantile of the NS3 distribution.} The estimated area subsumed by algae is shown in Fig~\ref{fig:algae-map} and calculated to be $587.3$ $m^2$.
 Similarly, we are able to track the evolution of the rhodamine dye release into the western portion of the pond by computing the NS3 for HSI collected across multiple flights.
@@
 First, we showed how the representation of data in the latent space of the GTM given by the mean responsibility can be used to visualize the small-scale structures within inland water bodies as evidenced in Fig~\ref{fig:gtm-water}.
-In particular, we note that the sharp gradients in colors found when visualizing the spatial distribution of GTM nodes across the pond reflects significant variability in water composition at the sub-meter scale.
+In particular, we note that the sharp gradients in colors found when visualizing the spatial distribution of GTM nodes across the pond reflects significant spectral variability at the sub-meter scale.
 This has important consequences for water quality assessment where the location of in situ data collection will have a strong impact on the ability of any model to predict water quality parameters from HSI.
@@
 Endmembers identified for exemplar spectra corresponding to rhodamine, algae, ground, and water pixels demonstrate that the method can successfully identify spectral signatures corresponding to diverse sources.
-If a set of labeled spectra for known sources are available, their representation in the GTM latent space could be used to perform as semi-supervised classification similar to the SOM approach developed by Riese et al.
+If a set of labeled spectra for known sources are available, their representation in the GTM latent space could be used to perform a semi-supervised classification similar to the SOM approach developed by Riese et al.
 \cite{riese2019supervised}.
+For example, in clear waters where variation in sediments and vegetation at the pond floor will contribute to observed spectral variability, the distribution learned by the GTM can be paired with in-situ data to classify the surface type.
 
@@
 Additionally, the GTM considers individual HSI pixels and does not exploit spatial structure like other methods such as convolutional autoencoders and non-negative matrix factorization using super-pixels \cite{palsson2020convolutional,Feng2022HyperspectralUB}.
-Extensions to the GTM have been proposed to enable batch training for large datasets as well as manifold-aligned noise models which replace the fixed variance parameter $\beta^{-1}$ with a full covariance matrix \cite{gtm-bishop-2}.
+Extensions to the GTM have been proposed to enable batch training for large datasets as well as manifold-aligned noise models which replace the fixed precision parameter $\beta$ with a full covariance matrix \cite{gtm-bishop-2}.
 
@@
 
-In this study we present the GTM as an useful unsupervised method for the visualization of UAV-based hyperspectral imagery and associated extraction of spectral endmembers.
+In this study we present the GTM as a useful unsupervised method for the visualization of UAV-based hyperspectral imagery and associated extraction of spectral endmembers.
 Using data collected at a North Texas pond, we demonstrate how the latent space of the GTM can be used to visualize the distribution of observed reflectance spectra revealing the small-scale spatial variability of water composition.

### 2024-06-24 paper/paper.tex
Updated to address reviewer comments for first round of review

@@
 
-Significant research efforts have focused on the development of techniques and algorithms to retrieve water quality parameters, e.g.
-from UAV-captured hyperspectral images (HSI).
-On-board computers installed alongside hyperspectral imagers can enable the rapid evaluation of spectral indices from HSI band ratios \cite{horstrand2019uav}.
+Significant research efforts have focused on the development of techniques and algorithms to retrieve water quality parameters from UAV-captured hyperspectral images (HSI).
+On-board computers installed alongside hyperspectral imagers enable the rapid evaluation of spectral indices from HSI band ratios \cite{horstrand2019uav}.
 These band ratios and polynomial combinations of bands have been used to successfully invert optically active water quality parameters such as turbidity directly from UAV acquired imagery \cite{vogt2016near, zhang2022selection}.
 Supervised machine learning techniques such as tree-based models, support vector machines, and neural networks have also been used to estimate a wide range of parameters such as colored dissolved organic matter, chlorophyll A, blue-green algae, and suspended sediment concentrations \cite{keller2018hyperspectral, lu2021retrieval}.
-The calibration and evaluation of these data-driven models require a significant volume of coincident in situ data.
-This can be addressed by coordinating UAV flights with reference data collection using autonomous robotic boats \cite{robot-team-1, robot-team-2}.
+The calibration and evaluation of these data-driven models requires a significant volume of coincident in situ data.
+This can be streamlined by coordinating UAV flights with reference data collection using autonomous robotic boats \cite{robot-team-1, robot-team-2}.
+New proximal sensors for water quality parameters are continually being developed \cite{proximal-sensing-1, proximal-sensing-2}.
 However, this approach relies on prior knowledge of expected sources in order to select appropriate reference instruments for model validation.
@@
 Data collection and processing was controlled by an attached Intel NUC small-form-factor computer.
+A detailed description of the UAV platform can be found in \cite{robot-team-1}.
 
@@
 A second GTM was trained on the combined dataset of HSI with pixels covering shore, water, and the rhodamine tracer dye release.
-The resulting \textcolor{red}{latent space distribution} is shown in Fig~\ref{fig:gtm-ndvi} where each sample has been plotted at the position of the mean node responsibility and colored by the normalized difference vegetation index (NDVI) computed from the original reflectance spectrum.
+The resulting latent space distribution is shown in Fig~\ref{fig:gtm-ndvi} where each sample has been plotted at the position of the mean node responsibility and colored by the normalized difference vegetation index (NDVI) computed from the original reflectance spectrum.
 The NDVI is a spectral index sensitive to variations in vegetation health with negative values corresponding to water, small positive values corresponding to sparse vegetation, and large values near $1$ corresponding to dense vegetation \cite{thenkabail2018hyperspectral}.
@@
 Additionally, the position of exemplar spectra for algae, rhodamine, water, and ground points are included as color-filled circles further illustrating the separation of distinct sources obtained by the GTM.
-\textcolor{red}{We note that a portion of the near-shore algae was observed to float at the surface of the water near the shore.
+We note that a portion of the near-shore algae was observed to float at the surface of the water near the shore.
 Consequently, the position of the exemplar spectrum in Fig~\ref{fig:gtm-ndvi} used for algae occurs at a non-negative NDVI.
-The accumulation of records with negative NDVI values in close proximity (in the latent space) to the selected algae spectrum reflects the presence of algae within the water.}
+The accumulation of records with negative NDVI values in close proximity (in the latent space) to the selected algae spectrum reflects the presence of algae within the water.
 
@@
 
-\textcolor{red}{Examining the latent space distribution learned by the GTM provides an unsupervised method to extract endmember spectra corresponding to unique sources within the imaging scene.
-To demonstrate this clearly, we consider the exemplar spectra visually identified in Fig~\ref{fig:sample-spectra}.} Using the trained GTM we can compute spectral signatures for each node in the latent space via the non-linear mapping $\psi$.
+Examining the latent space distribution learned by the GTM provides an unsupervised method to extract endmember spectra corresponding to unique sources within the imaging scene.
+To demonstrate this clearly, we consider the exemplar spectra visually identified in Fig~\ref{fig:sample-spectra}.
+Using the trained GTM we can compute spectral signatures for each node in the latent space via the non-linear mapping $\psi$.
 The responsibilities $R_{kn}$ therefore correspond to the contributions of the $kth$ node $\xi_k$ to the $n$th sample spectrum in the dataset.
-\textcolor{red}{Endmembers are identified by nodes with non-zero responsibility values for each of the selected exemplar spectra and are plotted in Fig~\ref{fig:spectral-signatures}.} We note that these extracted endmembers are able to accurately capture the shape of each exemplar spectrum including thin reflectance peaks.
+Endmembers are identified by nodes with non-zero responsibility values for each of the selected exemplar spectra and are plotted in Fig~\ref{fig:spectral-signatures}.
+We note that these extracted endmembers are able to accurately capture the shape of each exemplar spectrum including thin reflectance peaks.
 The GTM has also managed to smoothly interpolate through noisy wavelengths as shown by the water and algae endmembers for $\lambda > 800$ nm.
@@
 In Fig~\ref{fig:algae-map} we demonstrate this by using the extracted algae endmember spectrum to map the abundance of filamentous blue-green algae near the western shore of the pond.
-\textcolor{red}{The distribution of NS3 values was multi-modal with a majority of pixels above $0.4$ belonging to dissimilar spectra.
-Consequently, a cutoff threshold of $0.4275$ was chosen to distinguish algae values corresponding to the $20$th-quantile of the NS3 distribution.} The estimated area subsumed by algae is shown in Fig~\ref{fig:algae-map} and calculated to be $587.3$ $m^2$.
+The distribution of NS3 values was multi-modal with a majority of pixels above $0.4$ belonging to dissimilar spectra.
+Consequently, a cutoff threshold of $0.4275$ was chosen to distinguish algae values corresponding to the $20$th-quantile of the NS3 distribution.
+The estimated area subsumed by algae is shown in Fig~\ref{fig:algae-map} and calculated to be $587.3$ $m^2$.
 Similarly, we are able to track the evolution of the rhodamine dye release into the western portion of the pond by computing the NS3 for HSI collected across multiple flights.
-\textcolor{red}{Here the same cutoff threshold was used corresponding to the $33$rd-quantile of the NS3 distribution.} In the left panel of Fig~\ref{fig:rhodamine-map}, we see that the dye plume initially encompasses an area of roughly $330.7$ $m^2$.
+Here the same cutoff threshold was used corresponding to the $33$rd-quantile of the NS3 distribution.
+In the left panel of Fig~\ref{fig:rhodamine-map}, we see that the dye plume initially encompasses an area of roughly $330.7$ $m^2$.
 A second flight performed 15 minutes later reveals the dye plume to have increased to an area of $1495.6$ $m^2$.
@@
 The second application of the GTM presented in this study is for the unsupervised extraction of endmembers corresponding to unique sources observed in the HSI.
-Here the GTM is an attractive choice as it does not rely on the assumption of linear mixing and the presence of pure pixels which are easily broken by the effects of multiple scattering in realistic scenarios \cite{nonlinearity-in-hsi}.
+Here the GTM is an attractive choice as it does not rely on the assumptions of linear mixing and the presence of pure pixels which are easily broken by the effects of multiple scattering in realistic scenarios \cite{nonlinearity-in-hsi}.
 Additionally, because the GTM is a probabilistic model, the values of model hyperparameters can be selected objectively using information criteria like the BIC.
@@
 Spectral signatures extracted from GTM nodes are used to successfully map the abundance of algae near the shore and to track the evolution of a rhodamine tracer dye plume.
-These examples illustrate the power of combining unsupervised learning with UAV-based HSI collection for the characterization of water composition.
+These examples illustrate the power of combining unsupervised learning with UAV-based hyperspectral imaging for the characterization of water composition.
+Future work will further develop the GTM as a tool to guide in-situ data collection and enable contaminant localization for real-time applications.
 
